### 2019-01-30 microbiome_online.Rmd
Imported Scripts

---
title: "R Notebook"
output: html_notebook
---

This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

Combining 2 dat asets

ps_merge <- merge_phyloseq(pseq, atlas1006)

 get_variable(batmicrobiome,""LATITUDE"" )

 get_variable(batmicrobiome,"LONGITUDE" )

sample_data(batmicrobiome)

sample_names(batmicrobiome)

batmicrobiome[LONGITUDE] <- sample(50:1000)

sample_data(atlas1006)$human = factor( get_variable(atlas1006) %in% c("Feces", "Mock", "Skin", "Tongue") )

sample_data(atlas1006)$human = sample(50:1000)

sample_data(atlas1006)$human = rep(3,1151) %%% THIS WORKS !!!!

sample_data(atlas1006)$LATITUDE = rep(17,1151)
sample_data(atlas1006)$LONGITUDE = rep(-77,1151) 
get_variable(atlas1006,"LATITUDE" )
map_phyloseq(atlas1006)

map_phyloseq(batmicrobiome)

rep(3,12)

devtools::install_github("UrbanInstitute/urbnmapr")

atlas1006

sample_variables(atlas1006)

sample_variables(batmicrobiome)[56]  THIS WORKS

batmicrobiome = sample_data(data.frame(
  LONGITUDE = sample(50:1000)))

    sampledata = sample_data(data.frame(
  Location = sample(LETTERS[1:4], size=nsamples(physeq), replace=TRUE),
  Depth = sample(50:1000, size=nsamples(physeq), replace=TRUE),
  row.names=sample_names(physeq),
  stringsAsFactors=FALSE

  sampledata = sample_data(data.frame(
  Location = sample(LETTERS[1:4], size=nsamples(physeq), replace=TRUE),
  Depth = sample(50:1000, size=nsamples(physeq), replace=TRUE),
  row.names=sample_names(physeq),
  stringsAsFactors=FALSE
))

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.

```{r}
	setwd("/omega/bashscripts")

	## Get data
	library(foreign)
	library(stargazer)
	library(psych)
	library(plyr)
	library(xtable)
	library(car)
	#library(dplyr)
	library(scales)
	library(reshape) 
	library(gmodels)
	library(nnet)
	library(tidyr)
	library(sjmisc)
	library(RColorBrewer)
  library(htmltools)
	library(broom)

library(hrbrthemes)
library(gcookbook)
library(tidyverse)

library(phyloseq)
library(microbiome)
library(knitr)
library(ggpubr)

library(phylogeo)

data(atlas1006)   
# Rename the example data (which is a phyloseq object)
pseq <- atlas1006

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

```{r}
tab <- global(pseq, index = "all")
stargazer(tab, type = "text")

```

```{r}
####### Alpha diversity #######
ps1 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
tab <- diversities(ps1, index = "all")
stargazer(tab, type = "text",summary = NULL)
kable(head(tab))
```

```{r}
#  Prepare data for vizualisation
ps1.meta <- meta(ps1)
kable(head(ps1.meta))

stargazer(ps1.meta, type = "text",summary = NULL)
# Add the diversity table to metadata
ps1.meta$Shannon <- tab$shannon 
ps1.meta$InverseSimpson <- tab$inverse_simpson

# Compare differences in Shannon index between bmi group of the study subjects

# create a list of pairwise comparisons
bmi <- levels(ps1.meta$bmi_group) # get the variables

# make a pairwise list that we want to compare.
bmi.pairs <- combn(seq_along(bmi), 2, simplify = FALSE, FUN = function(i)bmi[i])

print(bmi.pairs)
```

```{r}
# CREATING VIOLIN PLOT
# the number in the palettes must equal the number of groups in bmi.pairs
p1 <- ggviolin(ps1.meta, x = "bmi_group", y = "Shannon",
 add = "boxplot", fill = "bmi_group", palette = c("#a6cee3", "#b2df8a", "#fdbf6f","#a6cee3", "#b2df8a", "#fdbf6f")) # 
print(p1)
```

```{r}

# Pairwise comparision using non-parametric test (Wilcoxon test).
p1 <- p1 + stat_compare_means(comparisons = bmi.pairs) 
print(p1)
```

```{r}
####### RICHNESS #######

tab <- richness(pseq)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

```

```{r}
###### DOMINANCE #####

# The dominance index refers to the abundance of the most abundant species. Various dominance indices are available (see the function help for a list of options).

# Absolute abundances for the single most abundant taxa in each sample
tab <- dominance(pseq, index = "all")
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

```

```{r}
######### RARENESS and low abundance ######

tab <- rarity(pseq, index = "all")
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)
```

```{r}
#### COVERAGE

# The coverage index gives the number of groups needed to have a given proportion of the ecosystem occupied (by default 0.5 ie 50%).

tab <- coverage(pseq, threshold = 0.5)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)
```

```{r}
#### ABUNDANCE

# The core_abundance function refers to the relative proportion of the core species. Non-core abundance provides the complement (1-x; see noncore_abundance).

tab <- core_abundance(pseq, detection = .1/100, prevalence = 50/100)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

```

```{r}
##### GINI INDEX

#Gini index is a common measure for inequality in economical income. The inverse gini index (1/x) can also be used as a community diversity measure.

tab <- inequality(pseq)
kable(head(tab))
plot(tab)
```

```{r}
###### EVENNESS

# Various evenness measures are also available.

tab <- evenness(pseq, "all")
kable(head(tab))
plot(tab)
```

MICROBIOME COMPOSITION

```{r}
# Just use prevalent taxa to speed up examples
# (not absolute counts used in this example)

# Pick sample subset
library(phyloseq)

library(hrbrthemes)
library(gcookbook)
library(tidyverse)
library(dplyr)
library(microbiome)

######## THIS ALTERED THE CORE DATA ###############
#pseq <- core(atlas1006, detection = 10^3, prevalence = 95/100)
#pseq2 <- subset_samples(pseq, group == "DI" & nationality == "AFR" & timepoint.within.group == 1)

# Try another theme
# from https://github.com/hrbrmstr/hrbrthemes

# Limit the analysis on core taxa and specific sample group
p <- plot_composition(pseq,
              taxonomic.level = "OTU",
                      sample.sort = "nationality",
                      x.label = "nationality") +
     guides(fill = guide_legend(ncol = 1)) +
     scale_y_percent() +
     labs(x = "Samples", y = "Relative abundance (%)",
                                   title = "Relative abundance data",
                                   subtitle = "Subtitle",
                                   caption = "Caption text.") + 
     theme_ipsum(grid="Y")
print(p)

# Averaged by group
p <- plot_composition(pseq,
                      average_by = "bmi_group", transform = "compositional")
print(p)

```

```{r}
tmp <- plot_composition(pseq, plot.type = "heatmap", transform = "compositional", 
            sample.sort = "neatmap", otu.sort = "neatmap", mar = c(6, 13, 1, 1))

print(tmp)
```

PLOTTING TAXA Prevlence

```{r}
data(atlas1006)

# Use sample and taxa subset to speed up example
p0 <- subset_samples(atlas1006, DNA_extraction_method == "r")

# Define detection and prevalence thresholds to filter out rare taxa
p0 <- core(p0, detection = 10, prevalence = 0)

# For the available taxonomic levels
plot_taxa_prevalence(p0, "Phylum", detection = 10)
```

PLOTTING RELAVANCE ABUNDANCE

```{r}

library(phyloseq)
ps2 <- data(atlas1006)
# for example purpose we will remove samples with less than 10000
ps2 = prune_samples(sample_sums(ps1)>=2000, ps1)

# To speed up the example we will use only those OTUs that are detected 100 times and present in 50% of the samples.

pseq2 <- microbiome::core(ps2, detection = 100, prevalence = .5)

# Improve the plotting MODIFIED 
tax_table(pseq2)[tax_table(pseq2)[,"Phylum"]== "Clostridium", "Phylum" ] <- "Clostridium Phylum"

# We will also remove the "f__" patterns for cleaner labels
tax_table(pseq2)[,colnames(tax_table(pseq2))] <- gsub(tax_table(pseq2)[,colnames(tax_table(pseq2))],pattern="[a-z]__",replacement="")

# merge at family level.

pseq.fam <- aggregate_taxa(pseq2, "Phylum")

p.fam <- plot_composition(pseq.fam, sample.sort = NULL, otu.sort = NULL,
  x.label = "ibd_subtype", plot.type = "barplot", verbose = FALSE)

print(p.fam)

```

PLOTTING RELTIVE ABUNDANCE

```{r}

### REQUIREDS THE DATA FROM ABOVE 
pseq.famrel <- transform(pseq.fam, "compositional")

p.famrel <- plot_composition(pseq.famrel, sample.sort = NULL, otu.sort = NULL,
  x.label = "ibd_subtype", plot.type = "barplot", verbose = FALSE)

print(p.famrel)

# further improvements can be done as follows

p.famrel <- plot_composition(pseq.famrel, 
                             sample.sort = NULL, 
                             otu.sort = NULL, 
                             x.label = "ibd_subtype", 
                             plot.type = "barplot", 
                             verbose = FALSE) + 
  guides(fill = guide_legend(ncol = 1)) + 
  scale_y_percent() + 
  labs(x = "Samples", 
       y = "Relative abundance (%)",
       title = "Relative abundance data", 
       subtitle = "Subtitle",
       caption = "Caption text.") + 
     theme_ipsum(grid="Y")

print(p.famrel)
```

HEAT MAP

```{r}
pseq.famlog <- transform(pseq.fam, "log10")

p.famrel.heatmap <- plot_composition(pseq.famlog, 
                             sample.sort = NULL, 
                             otu.sort = NULL, 
                             x.label = "ibd_subtype", 
                             plot.type = "heatmap", 
                             verbose = FALSE)

print(p.famrel.heatmap)
```

Plot taxa prevalence

```{r}

library(microbiome)
data(peerj32)

# Rename the data
pseq <- peerj32$phyloseq

# Calculate compositional version of the data
# (relative abundances)
pseq.rel <- microbiome::transform(pseq, "compositional")

core.taxa.standard <- core_members(pseq.rel, detection = 0, prevalence = 50/100)

```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

```{r}
library(ggplot2)
library(gridExtra)
library(phylogeo)

data(mountainsoil)

```

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

### 2019-01-30 Installs_RPackages.R
Imported Scripts

@@ -1,0 +1,10 @@
+# restore_packages.R
+#
+# installs each package from the stored list of packages
+
+
+# alter the directory. Remember microbiome is a GITHUB respository 
+# THe file io defined in Stre_list_RPackages.R 
+load("/omega/bashscripts/microbiome/Required_coreRpackages.rda")
+
+for (count in 1:length(installedpackages)) install.packages(installedpackages[count])

### 2019-01-30 Stores_list_RPackages.R

@@ -1,0 +1,12 @@
+# store_packages.R
+#
+# stores a list of the currently installed packages
+# Update as required
+
+tmp = installed.packages()
+
+# Run AFTER you have execugted the script
+installedpackages = as.vector(tmp[is.na(tmp[,"Priority"]), 1])
+
+# alter the directory. Remember microbiome is a GITHUB respository 
+save(installedpackages, file="/omega/bashscripts/microbiome/Required_coreRpackages.rda")

### 2019-01-30 push2github_local.sh
Imported Scripts

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+
+# This pushes the content of the repository directory back onto github
+pwd
+git add --all
+git commit -m "Imported Scripts"
+git push origin master
+

### 2019-01-30 push2github_local.sh
updated

@@ -2,8 +2,12 @@
 
 
 # This pushes the content of the repository directory back onto github
+# must be ran from the repository directory using the command   ./push2github_local.sh
+# If you alter this remember to make the file executable with   chmod a+x push2github_local.sh
+
+
 pwd
 git add --all
-git commit -m "Imported Scripts"
+git commit -m "updated"
 git push origin master
 

### 2019-01-30 MRD_microbiome_working.Rmd
updated

---
title: "MRD_microbiome_working"
author: "Reis-Dehabadi, M"
date: "1/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}

library(microbiome)
library(knitr)
library(ggpubr)
library(dplyr)

# Data from http://dx.doi.org/10.1038/ncomms7342
data(atlas1006)   
# Rename the example data (which is a phyloseq object)
pseq <- atlas1006

data(dietswap)  
pseqds <- dietswap
print(dietswap)

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

# Richness
tab2 <- richness(pseqds)
kable(head(tab2))

```

ou can also embed plots, for example:

```{r , echo=FALSE}

# Richness
tab2 <- richness(pseqds)
kable(head(tab2))

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-01-30 microbiome_online_MRD_working.Rmd
updated

--- 2019-01-30 microbiome_online.Rmd
@@
 ```{r}
-	setwd("/omega/bashscripts")
+	setwd("/Users/rezahaji-dehabadi/scripts/workspace")
 
@@
 library(tidyverse)
-
+library(gdtools)
 library(phyloseq)

### 2019-02-06 MRD_microbiome_working.Rmd
updated

@@
 
-```{r cars}
+```{r data set-up}
 
+setwd("/Users/rezahaji-dehabadi/scripts/workspace")
+
+	## Get data
+	library(foreign)
+	library(stargazer)
+	library(psych)
+	library(plyr)
+	library(xtable)
+	library(car)
+	#library(dplyr)
+	library(scales)
+	library(reshape) 
+	library(gmodels)
+	library(nnet)
+	library(tidyr)
+	library(sjmisc)
+	library(RColorBrewer)
+  library(htmltools)
+	library(broom)
+
+library(hrbrthemes)
+library(gcookbook)
+library(tidyverse)
+library(gdtools)
+library(phyloseq)
 library(microbiome)
@@
 library(ggpubr)
+
+library(phylogeo)
 library(dplyr)
@@
 data(atlas1006)
+
 # Rename the example data (which is a phyloseq object)
@@
 
-data(dietswap)  
-pseqds <- dietswap
-print(dietswap)
+```
+
+Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.
+
+```{r summary}
+
+# Summarise contents of phyloseq
+summarize_phyloseq(pseq)
 
@@
 
-## Including Plots
+```{r alpha diversity}
 
-You can also embed plots, for example:
+# Alpha diversity
+ps1 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
+tab <- diversities(ps1, index = "all")
+stargazer(tab, type = "text",summary = NULL)
+kable(head(tab))
+plot(tab)
 
-```{r pressure, echo=FALSE}
+```
+
+```{r richness}
 
 # Richness
-tab2 <- richness(pseqds)
-kable(head(tab2))
+tab <- richness(pseq)
+kable(head(tab))
+stargazer(tab, type = "text",summary = NULL)
+plot(tab)
 
@@
 
-```{r , echo=FALSE}
+```{r abundance}
 
-# Richness
-tab2 <- richness(pseqds)
-kable(head(tab2))
+# Abundance table
+
+# Absolute abundances for the single most abundant taxa in each sample
+tab <- dominance(pseq, index = "all")
+kable(head(tab))
+stargazer(tab, type = "text",summary = NULL)
+plot(tab)
+
+# Abunance histogram
+
+# Visualize population densities for specific taxa
+plot_density(pseq, "Dialister") + ggtitle("Absolute abundance")
+
+# Same with log10 compositional abundances
+x <- microbiome::transform(pseq, "compositional")
+tax <- "Dialister"
+plot_density(x, tax, log10 = TRUE) +
+  ggtitle("Relative abundance") +
+  xlab("Relative abundance (%)")
+
+```
+```{r dominance}
+
+dominance(pseq, index = "all", rank = 1, relative = TRUE, aggregate = TRUE)
 
@@
 Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+```{r rarity and low abundance}
+
+# Rarity and low abundance
+tab <- rarity(pseq, index = "all")
+kable(head(tab))
+stargazer(tab, type = "text",summary = NULL)
+plot(tab)
+
+```
+```{r core abundance}
+
+# Core abundance
+# relative proportion of the core species
+tab <- core_abundance(pseq, detection = 0.1/100, prevalence = 50/100)
+kable(head(tab))
+stargazer(tab, type = "text",summary = NULL)
+plot(tab)
+
+```
+
+```{r gini index}
+
+# Gini index
+# measure for relative inequality in economical income
+# measures community diversity
+# Gini index is between 0-1
+# increasing gini index implies increasing inequality
+tab <- inequality(pseq)
+kable(head(tab))
+stargazer(tab, type = "text",summary = NULL)
+plot(tab)
+
+```
+```{r evenness}
+
+# Evenness
+tab <- evenness(pseq, "all", zeroes = TRUE)
+kable(head(tab))
+stargazer(tab, type = "text",summary = NULL)
+plot(tab)
+
+```
+```{r prepare data for visualisation}
+
+# Prepare data for vizualisation
+
+# Get metadata
+ps1.meta <- meta(ps1)
+kable(head(ps1.meta))
+stargazer(ps1.meta, type = "text",summary = NULL)
+
+# Add the diversity table to metadata
+ps1.meta$Shannon <- tab$shannon 
+ps1.meta$InverseSimpson <- tab$inverse_simpson
+
+# Compare differences in Shannon index between bmi group of the study subjects
+
+# Create a list of pairwise comaprisons
+bmi <- levels(ps1.meta$bmi_group) # get the variables
+
+# Make a pairwise list that we want to compare.
+bmi.pairs <- combn(seq_along(bmi), 2, simplify = FALSE, FUN = function(i)bmi[i])
+
+print(bmi.pairs)
+
+```
+
+```{r violin plot}
+
+# Violin plot
+p1 <- ggviolin(ps1.meta, x = "bmi_group", y = "Shannon",
+ add = "boxplot", fill = "bmi_group", palette = c("#a6cee3", "#b2df8a", "#fdbf6f","#a6cee3", "#b2df8a", "#fdbf6f"))
+
+print(p1)
+
+```
+
+```{r add mean comparison p-values to a ggplot}
+
+# Add mean comparison p-values to a ggplot
+p1 <- p1 + stat_compare_means(comparisons = bmi.pairs) 
+print(p1)
+
+```
+# https://rpkgs.datanovia.com/ggpubr/index.html
+# Package for elegant data visualization in R
+
+```{r coverage}
+
+# Coverage
+
+# The coverage index gives the number of groups needed to have a given proportion of the ecosystem occupied (by default 0.5 ie 50%).
+
+tab <- coverage(pseq, threshold = 0.5)
+kable(head(tab))
+stargazer(tab, type = "text",summary = NULL)
+plot(tab)
+
+```
+
+MICROBIOME COMPOSITION
+
+```{r composition}
+
+# Just use prevalent taxa to speed up examples
+# (not absolute counts used in this example)
+
+# Pick sample subset
+library(phyloseq)
+
+library(hrbrthemes)
+library(gcookbook)
+library(tidyverse)
+library(dplyr)
+library(microbiome)
+
+######## THIS ALTERED THE CORE DATA ###############
+#pseq <- core(atlas1006, detection = 10^3, prevalence = 95/100)
+#pseq2 <- subset_samples(pseq, group == "DI" & nationality == "AFR" & timepoint.within.group == 1)
+
+# Try another theme
+# from https://github.com/hrbrmstr/hrbrthemes
+
+# Limit the analysis on core taxa and specific sample group
+p <- plot_composition(pseq,
+              taxonomic.level = "OTU",
+                      sample.sort = "nationality",
+                      x.label = "nationality") +
+     guides(fill = guide_legend(ncol = 1)) +
+     scale_y_percent() +
+     labs(x = "Samples", y = "Relative abundance (%)",
+                                   title = "Relative abundance data",
+                                   subtitle = "Subtitle",
+                                   caption = "Caption text.") + 
+     theme_ipsum(grid="Y")
+print(p)
+
+# Averaged by group
+p <- plot_composition(pseq,
+                      average_by = "bmi_group", transform = "compositional")
+print(p)
+
+```
+```{r composition heatmap}
+
+# Composition heatmap
+tmp <- plot_composition(pseq, plot.type = "heatmap", transform = "compositional", 
+            sample.sort = "neatmap", otu.sort = "neatmap", mar = c(6, 13, 1, 1))
+
+print(tmp)
+
+```
+
+```{r plot taxa prevalence}
+
+# Plot taxa prevalence
+# overview of OTU prevalences alongwith their taxonomic affiliations
+# will aid in checking if you filter OTUs based on prevalence, then what taxonomic affliations will be lost
+
+data(atlas1006)
+
+# Use sample and taxa subset to speed up example
+p0 <- subset_samples(atlas1006, DNA_extraction_method == "r")
+
+# Define detection and prevalence thresholds to filter out rare taxa
+p0 <- core(p0, detection = 10, prevalence = 0)
+
+# For the available taxonomic levels
+plot_taxa_prevalence(p0, "Phylum", detection = 10)
+
+```
+
+```{r}
+
+```
+
+CORE MICROBIOME
+```{r load example data for core microbiome}
+
+# Load data
+library(microbiome)
+data(peerj32)
+
+# Rename the data
+pseq <- peerj32$phyloseq
+
+# Calculate compositional version of the data
+# (relative abundances)
+pseq.rel <- microbiome::transform(pseq, "compositional")
+
+```
+
+```{r prevalence of taxonomic groups}
+
+# Prevalence of taxonomi groups
+
+# Relative population frequencies; at 1% compositional abundance threshold
+head(prevalence(pseq.rel, detection = 1/100, sort = TRUE))
+
+# Absolute population frequencies (sample count)
+head(prevalence(pseq.rel, detection = 1/100, sort = TRUE, count = TRUE))
+
+```
+
+```{r core analysis}
+
+# Core microbiota analysis
+
+# Names of core taxa
+core.taxa.standard <- core_members(pseq.rel, detection = 0, prevalence = 50/100)
+
+# Full phyloseq object of core data
+pseq.core <- core(pseq.rel, detection = 0, prevalence = .5)
+
+# Retrieving the associated taxa names from the phyloseq object
+core.taxa <- taxa(pseq.core)
+
+# Total core abundance in each sample (sum of abundances of the core members)
+core.abundance <- sample_sums(core(pseq.rel, detection = .01, prevalence = .95))
+
+```
+
+CORE VISUALIZATION
+```{r core line plots}
+
+# Core line plots
+
+# Determine core microbiota across various abundance/prevalence thresholds with the blanket analysis
+
+# With compositional (relative) abundances
+det <- c(0, 0.1, 0.5, 2, 5, 20)/100
+prevalences <- seq(.05, 1, .05)
+plot_core(pseq.rel, prevalences = prevalences, detections = det, plot.type = "lineplot") + xlab("Relative Abundance (%)")
+
+```
+
+```{r core heat maps}
+
+# Core heat maps
+
+# Core with compositionals:
+prevalences <- seq(.05, 1, .05)
+detections <- 10^seq(log10(1e-3), log10(.2), length = 10)
+
+# Also define gray color palette
+gray <- gray(seq(0,1,length=5))
+p <- plot_core(pseq.rel, plot.type = "heatmap", colours = gray,
+    prevalences = prevalences, detections = detections) +
+    xlab("Detection Threshold (Relative Abundance (%))")
+print(p)
+
+# Core with absolute counts and horizontal view:
+# and minimum population prevalence (given as percentage)
+detections <- 10^seq(log10(1), log10(max(abundances(pseq))/10), length = 10)
+
+library(RColorBrewer)
+p <- plot_core(pseq, plot.type = "heatmap", 
+             prevalences = prevalences,
+             detections = detections,
+         colours = rev(brewer.pal(5, "Spectral")),
+         min.prevalence = .2, horizontal = TRUE)
+print(p)
+
+```
+
+As it can be seen, we see only OTu IDs and this may not be useful to interpret the data. We need to repreoccess this figure to include taxonomic information. We can do this as follows:
+
+library(RColorBrewer)
+library(knitr)
+# Core with absolute counts and vertical view:
+# and minimum population prevalence (given as percentage)
+detections <- 10^seq(log10(1), log10(max(abundances(pseq.2))/10), length = 10)
+
+healthycore <- plot_core(pseq.2, plot.type = "heatmap", 
+             prevalences = prevalences,
+             detections = detections,
+         colours = rev(brewer.pal(5, "Spectral")),
+         min.prevalence = .90, horizontal = F)
+# get the data used for plotting 
+df <- healthycore$data
+
+# get the list of OTUs
+list <- df$Taxa
+
+# check the OTU ids
+# print(list)
+
+# get the taxonomy data
+tax <- tax_table(pseq.2)
+tax <- as.data.frame(tax)
+
+# add the OTus to last column
+tax$OTU <- rownames(tax)
+
+# select taxonomy of only 
+# those OTUs that are used in the plot
+tax2 <- dplyr::filter(tax, rownames(tax) %in% list)
+
+# head(tax2)
+
+# We will merege all the column into one except the Doamin as all is bacteria in this case
+tax.unit <- tidyr::unite(tax2, Taxa_level,c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species", "OTU"), sep = "_;", remove = TRUE)
+
+tax.unit$Taxa_level <- gsub(pattern="[a-z]__",replacement="", tax.unit$Taxa_level)
+
+# add this new information into the plot data df
+
+df$Taxa <- tax.unit$Taxa_level
+
+# you can see now we have the taxonomic information
+knitr::kable(head(df))
+
+# replace the data in the plot object
+healthycore$data <- df
+
+plot(healthycore + theme(axis.text.y = element_text(face="italic")))
+
+```{r}
+
+```
+
+MICROBIOME STABILITY ANALYSIS
+
+Get example data - HITChip Atlas of 130 genus-like taxa across 1006 healthy western adults. A subset of 76 subjects have also short time series available for temporal stability analysis:
+```{r microbiome stability analysis}
+
+# Load the example data
+set.seed(134)
+library(microbiome)
+library(dplyr)
+data(atlas1006)
+
+# Rename the example data
+pseq <- atlas1006
+
+# Focus on specific subset
+pseq <- pseq %>% subset_samples(DNA_extraction_method == "r")
+
+# Use relative abundances
+pseq <- microbiome::transform(pseq, "compositional")
+
+# Keep only the prevalent taxa to speed up examples
+pseq <- core(pseq, detection = .1/100, prevalence = 99/100)
+
+# For cross-sectional analysis, use only the baseline time point:
+pseq0 <- baseline(pseq)
+
+```
+
+It has been reported that certain microbial groups exhibit bi-stable abundance distributions with distinct peaks at low and high abundances, and an instable intermediate abundance range.
+Instability at the intermediate abundance range is hence one indicator of bi-stability.
+Lahti et al.
+2014 used straightforward correlation analysis to quantify how the distance from the intermediate abundance region (50% quantile) is associated with the observed shifts between consecutive time points.
+```{r intermediate stability quantification}
+
+# Intermediate stability quantification
+intermediate.stability <- intermediate_stability(pseq, output = "scores")
+
+```
+
+BIMODALITY ANALYSIS
+
+Bimodality of the abundance distribution provides an indirect indicator of bistability, although other explanations such as sampling biases etc.
+should be controlled.
+Multiple bimodality scores are available.
+```{r}
+
+# Load the example data
+library(microbiome)
+library(dplyr)
+data(atlas1006)
+
+# Rename the example data
+pseq <- atlas1006
+
+# Focus on specific DNA extraction method
+pseq <- pseq %>% subset_samples(DNA_extraction_method == "r")
+
+# Keep prevalent taxa (HITChip signal >3 in >20 percent of the samples)
+pseq <- core(pseq, detection = 10^3, prevalence = .2)
+
+# Use relative abundances
+pseq <- microbiome::transform(pseq, "compositional")
+
+# For cross-sectional analysis, include
+# only the zero time point:
+pseq0 <- subset_samples(pseq, time == 0)
+
+# Bimodality is better estimated from log10 abundances
+pseq0.log10 <- microbiome::transform(pseq0, "log10")
+bimodality <- bimodality(pseq0.log10, method = "potential_analysis", bs.iter = 20)
+
+```
+
+```{r bimodality visualization}
+
+# Visualize population densities for unimodal and bimodal groups
+
+# Pick the most and least bimodal taxa as examples
+unimodal  <- names(sort(bimodality))[[1]]
+bimodal  <- rev(names(sort(bimodality)))[[1]]
+
+# Visualize population frequencies
+library(ggplot2)
+theme_set(theme_bw(20))
+p1 <- plot_density(pseq, variable = unimodal, log10 = TRUE) 
+p2 <- plot_density(pseq, variable = bimodal,  log10 = TRUE) 
+library(gridExtra)
+library(ggplot2)
+grid.arrange(p1, p2, nrow = 1)
+
+```
+
+```{r tipping point detection}
+
+# Tipping point detection
+
+# Log10 abundance for a selected taxonomic group
+tax <- bimodal
+
+# Detect tipping points detection at log10 abundances 
+x <- log10(abundances(pseq)[tax,])
+
+# Bootstrapped potential analysis to identify potential minima
+set.seed(3432)
+potential.minima <- potential_analysis(log10(abundances(pseq)[tax,]), bs.iter = 50)$minima
+# Same with earlywarnings package (without bootstrap ie. less robust)
+# library(earlywarnings)
+# res <- livpotential_ews(x)$min.points
+
+# Identify the potential minimum location as a tipping point candidate
+# and cast the tipping back to the original (non-log) space:
+tipping.point <- 10^potential.minima
+
+# Identify potential minima in cross-section population data as tipping point candidates.
+
+# Log10 abundance for a selected taxonomic group
+# Pick the most bimodal taxa as an example
+tax  <- names(which.max(bimodality.score))
+
+# Detect tipping points detection at log10 abundances 
+x <- abundances(microbiome::transform(pseq, "log10"))[tax,]
+
+# Bootstrapped potential analysis to identify potential minima
+# in practice, use more bootstrap iterations
+potential.minima <- potential_analysis(x, bs.iter = 10)$minima
+
+# Same with earlywarnings package (without bootstrap ie. less robust)
+# library(earlywarnings)
+# res <- livpotential_ews(x)$min.points
+
+# Identify the potential minimum location as a tipping point candidate
+# and cast the tipping back to the original (non-log) space:
+tipping.point <- 10^potential.minima
+
+print(tipping.point)
+
+```
+
+Pick subset of the HITChip Atlas data set and plot the subject abundance variation lineplot (Variation tip plot) and Bimodality hotplot for a given taxon as in Lahti et al.
+2014.
+The Dialister has bimodal population distribution and reduced temporal stability within subjects at intermediate abundances.
+```{r variation lineplot and bimodality hotplot}
+
+# Variation lineplot and bimodality hotplot
+
+# We set the tipping point manually in this example
+
+# Bimodality hotplot:
+# Consider a unique sample from each subject: the baseline time point 
+p <- hotplot(pseq0, tax, tipping.point = 0.005)
+print(p)
+
+# Visualize bimodality
+pv <- plot_tipping(pseq, tax, tipping.point = 0.005)
+print(pv)
+
+```
+
+Comparing bimodality and intermediate stability
+The analysis suggests that bimodal population distribution across individuals is often associated with instable intermediate abundances within individuals.
+The specific bi-stable groups in the upper left corner were suggested to constitute bistable tipping elements of the human intestinal microbiota in Lahti et al.
+Nat.
+Comm.
+5:4344, 2014:
+```{r comparing bimodality and intermediate stability}
+
+# Comparing bimodality and intermediate stability
+
+taxa <- taxa(pseq0)
+df <- data.frame(group = taxa,
+                 intermediate.stability = intermediate.stability[taxa],
+         bimodality = bimodality.score[taxa])
+
+theme_set(theme_bw(20))
+p <- ggplot(df,
+       aes(x = intermediate.stability, y = bimodality, label = group)) +
+       geom_text() +
+       geom_point() 
+print(p)
+
+```
+
+```{r}
+
+```

### 2019-02-06 Installs_RPackages.R

@@ -5,6 +5,6 @@
 
 # alter the directory. Remember microbiome is a GITHUB respository 
 # THe file io defined in Stre_list_RPackages.R 
-load("/omega/bashscripts/microbiome/Required_coreRpackages.rda")
+load("/Users/rezahaji-dehabadi/scripts/microbiome/Required_coreRpackages.rda")
 
 for (count in 1:length(installedpackages)) install.packages(installedpackages[count])

### 2019-02-19 microbiome_online.Rmd
updated

@@
 
-ps_merge <- merge_phyloseq(pseq, atlas1006)
+ps_merge <- merge_phyloseq(esophagus, atlas1006)
 
@@
 
+#
+data(esophagus)
+tree <- phy_tree(esophagus)
+otu  <- otu_table(esophagus)
+otutree0 <- phyloseq(otu, tree)
+ plot_tree(otutree0)
+otutree1 <- merge_taxa(otutree0, 1:8, 2)
+plot_tree(esophagus, ladderize="left")
+
+stomach <- subset_samples(physeq3, Sample_type=="Stomach")
+data(atlas1006)   
+# Rename the example data (which is a phyloseq object)
+pseq <- atlas1006
+x <- tax_glom(pseq, taxrank="Phylum")
+pseq <- tax_glom(pseq, taxrank="Genus")
+
+rank_names(pseq)
+library(HITChipDB)
+ data.dir <- system.file("HITChip.tab", package = "microbiome")
+  dat <- import_hitchip(data.dir)
+
+  pseq <- HITChipDB::read_hitchip(data.dir, method = "frpa")$pseq
+
+  treefile = read_tree("HITdb.tre", errorIfNULL = T)
+
+    library(microbiome)
+  data(peerj32)
+  otu <- peerj32$microbes
+  meta <- peerj32$meta
+
 Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.
@@
 ```{r}
-	setwd("/omega/bashscripts")
+	setwd("/omega/bashscripts/microbiome")
 
@@
 library(phylogeo)
+library("HITChipDB")
+library(phangorn)
 
@@
 pseq <- atlas1006
+#pseq <- tax_glom(pseq, taxrank="Phylum")
+#pseq <- tax_glom(pseq, taxrank="Genus")
+```
+
+Taxonomic Filtering
+
+In many biological settings, the set of all organisms from all samples are well-represented in the available taxonomic reference database.
+When (and only when) this is the case, it is reasonable or even advisable to filter taxonomic features for which a high-rank taxonomy could not be assigned.
+Such ambiguous features in this setting are almost always sequence artifacts that don’t exist in nature.
+It should be obvious that such a filter is not appropriate for samples from poorly characterized or novel specimens, at least until the possibility of taxonomic novelty can be satisfactorily rejected.
+Phylum is a useful taxonomic rank to consider using for this purpose, but others may work effectively for your data.
+
+```{r}
+
+table(tax_table(pseq)[, "Phylum"], exclude = NULL)
+
+# Compute prevalence of each feature, store as data.frame
+prevdf = apply(X = otu_table(pseq),
+               MARGIN = ifelse(taxa_are_rows(pseq), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+# Add taxonomy and total read counts to this data.frame
+prevdf = data.frame(Prevalence = prevdf,
+                    TotalAbundance = taxa_sums(pseq),
+                    tax_table(pseq))
+
+# Are there phyla that are comprised of mostly low-prevalence features? Compute the total and average prevalences of the features in each phylum.
+tab <- plyr::ddply(prevdf, "Phylum", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
+
+# Plot
+stargazer(tab, summary=FALSE, type = "text")
+
+#IF NEED TO REMOVE
+# Define phyla to filter
+#filterPhyla = c("Fusobacteria", "Deinococcus-Thermus")
+# Filter entries with unidentified Phylum.
+#ps1 = subset_taxa(ps, !Phylum %in% filterPhyla)
+#ps1
+
+```
+
+Prevalence Filtering
+The previous filtering steps are considered supervised, because they relied on prior information that is external to this experiment (a taxonomic reference database).
+This next filtering step is completely unsupervised, relying only on the data in this experiment, and a parameter that we will choose after exploring the data.
+Thus, this filtering step can be applied even in settings where taxonomic annotation is unavailable or unreliable.
+
+First, explore the relationship of prevalence and total read count for each feature.
+Sometimes this reveals outliers that should probably be removed, and also provides insight into the ranges of either feature that might be useful.
+This aspect depends quite a lot on the experimental design and goals of the downstream inference, so keep these in mind.
+It may even be the case that different types of downstream inference require different choices here.
+There is no reason to expect ahead of time that a single filtering workflow is appropriate for all analysis.
+
+Each point in Figure 2 is a different taxa.
+Exploration of the data in this way is often useful for selecting filtering parameters, like the minimum prevalence criteria we will used to filter the data above.
+
+Sometimes a natural separation in the dataset reveals itself, or at least, a conservative choice that is in a stable region for which small changes to the choice would have minor or no effect on the biological interpreation (stability).
+Here no natural separation is immediately evident, but it looks like we might reasonably define a prevalence threshold in a range of zero to ten percent or so.
+Take care that this choice does not introduce bias into a downstream analysis of association of differential abundance.
+
+The following uses five percent of all samples as the prevalence threshold.
+
+```{r}
+# Subset to the remaining phyla
+prevdf1 = subset(prevdf, Phylum %in% get_taxa_unique(pseq))
+ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(pseq),color=Phylum)) +
+  # Include a guess for parameter
+  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
+  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
+  facet_wrap(~Phylum) + theme(legend.position="none")
+```
+
+The following uses five percent of all samples as the prevalence threshold.
+
+```{r}
+# Define prevalence threshold as 5% of total samples
+prevalenceThreshold = 0.05 * nsamples(pseq)
+keepTaxa = rownames(prevdf1)[(prevdf1$Prevalence >= prevalenceThreshold)]
+ps2 = prune_taxa(keepTaxa, pseq)
+```
+
+Agglomerate taxa
+When there is known to be a lot of species or sub-species functional redundancy in a microbial community, it might be useful to agglomerate the data features corresponding to closely related taxa.
+Ideally we would know the functional redundancies perfectly ahead of time, in which case we would agglomerate taxa using those defined relationships and the function in phyloseq.
+That kind of exquisite functional data is usually not available, and different pairs of microbes will have different sets of overlapping functions, complicating the matter of defining appropriate grouping criteria.
+
+While not necessarily the most useful or functionally-accurate criteria for grouping microbial features (sometimes far from accurate), taxonomic agglomeration has the advantage of being much easier to define ahead of time.
+This is because taxonomies are usually defined with a comparatively simple tree-like graph structure that has a fixed number of internal nodes, called “ranks”.
+This structure is simple enough for the phyloseq package to represent taxonomies as table of taxonomy labels.
+Taxonomic agglomeration groups all the “leaves” in the hierarchy that descend from the user-prescribed agglomerating rank, this is sometimes called ‘glomming’.
+
+The following example code shows how one would combine all features that descend from the same genus.
+
+```{r}
+# How many genera would be present after filtering?
+length(get_taxa_unique(ps2, taxonomic.rank = "Genus"))
+## [1] 49
+ps3 = tax_glom(ps2, "Genus", NArm = TRUE)
+ps4 = tax_glom(ps2, "Phylum", NArm = TRUE)
+
+#Here phyloseq’s plot_tree() function compare the original unfiltered data, the tree after taxonoic agglomeration, and the tree after phylogenetic agglomeration.
+These are stored as separate plot objects, then rendered together in one combined graphic using  gridExtra::grid.arrange.
+
+multiPlotTitleTextSize = 15
+p2tree = plot_tree(ps2, method = "treeonly",
+                   ladderize = "left",
+                   title = "Before Agglomeration") +
+  theme(plot.title = element_text(size = multiPlotTitleTextSize))
+p3tree = plot_tree(ps3, method = "treeonly",
+                   ladderize = "left", title = "By Genus") +
+  theme(plot.title = element_text(size = multiPlotTitleTextSize))
+p4tree = plot_tree(ps4, method = "treeonly",
+                   ladderize = "left", title = "By Phylum") +
+  theme(plot.title = element_text(size = multiPlotTitleTextSize))
+
+# group plots together
+grid.arrange(nrow = 1, p2tree, p3tree, p4tree)
+
+```
+
+```{r}
+library(dada2)
+library(DECIPHER)
+ref_fasta <- "data/rdp_train_set_14.fa.gz"
+taxtab <- assignTaxonomy(seqtab, refFasta = ref_fasta)
+colnames(taxtab) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")
+
+taxonomy <- tax_table(pseq)
 
@@
 ```{r}
+
 tab <- global(pseq, index = "all")
@@
 kable(head(tab))
+
+alpha2 <- . %>%
+  prune_taxa(taxa_sums(.) > 0, .) %>%
+  diversities(., index = "all")  %>%
+  kable(head(.))
+
 ```
@@
 plot(tab)
+
+```
+
+```{r}
+# CORE
+
+# Calculate compositional version of the data
+# (relative abundances)
+pseq.rel <- microbiome::transform(pseq, "compositional")
+
+#Relative population frequencies; at 1% compositional abundance threshold:
+
+head(prevalence(pseq.rel, detection = 1, sort = TRUE))
+
+# Absolute population frequencies (sample count):
+
+head(prevalence(pseq.rel, detection = 1, sort = TRUE, count = TRUE))
+
+# This returns the taxa that exceed the given prevalence and detection thresholds
+pseq.core <- core(pseq.rel, detection = 0, prevalence = .5)
+
+# Retrieving the associated taxa names from the phyloseq object:
+core.taxa <- taxa(pseq.core)
+
+# Total core abundance in each sample (sum of abundances of the core members):
+core.abundance <- sample_sums(core(pseq.rel, detection = .01, prevalence = .95))
+
+# With compositional (relative) abundances
+det <- c(0, 0.1, 0.5, 2, 5, 20)/100
+prevalences <- seq(.05, 1, .05)
+plot_core(pseq.rel, prevalences = prevalences, detections = det, plot.type = "lineplot") + xlab("Relative Abundance (%)")
+
+# COREHEAT MAPS
+
+# Core with compositionals:
+prevalences <- seq(.05, 1, .05)
+detections <- 10^seq(log10(1e-3), log10(.2), length = 10)
+
+# Also define gray color palette
+gray <- gray(seq(0,1,length=5))
+p <- plot_core(pseq.rel, plot.type = "heatmap", colours = gray,
+    prevalences = prevalences, detections = detections) +
+    xlab("Detection Threshold (Relative Abundance (%))")
+print(p)
+
+# Same with the viridis color palette
+# color-blind friendly and uniform
+# options: viridis, magma, plasma, inferno
+# https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
+# Also discrete=TRUE versions available
+library(viridis)
+
+## Loading required package: viridisLite
+
+print(p + scale_fill_viridis())
+
+## Scale for 'fill' is already present. Adding another scale for 'fill',
+## which will replace the existing scale.
+
+# Core with absolute counts and horizontal view:
+# and minimum population prevalence (given as percentage)
+detections <- 10^seq(log10(1), log10(max(abundances(pseq))/10), length = 10)
+
+library(RColorBrewer)
+p <- plot_core(pseq, plot.type = "heatmap", 
+             prevalences = prevalences,
+             detections = detections,
+         colours = rev(brewer.pal(5, "Spectral")),
+         min.prevalence = .2, horizontal = TRUE)
+print(p)
 

### 2019-02-19 Installs_RPackages.R

@@ -5,6 +5,7 @@
 
 # alter the directory. Remember microbiome is a GITHUB respository 
 # THe file io defined in Stre_list_RPackages.R 
-load("/Users/rezahaji-dehabadi/scripts/microbiome/Required_coreRpackages.rda")
+load("/omega/bashscripts/microbiome/Required_coreRpackages.rda")
 
 for (count in 1:length(installedpackages)) install.packages(installedpackages[count])
+

### 2019-06-18 GlobalMicrobiome_NS.Rmd
updated

---
title: "MRD_final_microbiome"
author: "Reis-Dehabadi, M"
date: "4/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/omega/1006") 
```

```{r}

##### FUNCTIONS ########

##### Normalization #######

# Better rounding function than R's base round
myround <- function(x) { trunc(x + 0.5) }

# Scales reads by 
# 1) taking proportions
# 2) multiplying by a given library size of n
# 3) rounding 
# Default for n is the minimum sample size in your library
# Default for round is floor
scale_reads <- function(physeq, n = min(sample_sums(physeq)), round = "floor") {

  # transform counts to n
  physeq.scale <- transform_sample_counts(physeq, 
    function(x) {(n * x/sum(x))}
  )

  # Pick the rounding functions
  if (round == "floor"){
    otu_table(physeq.scale) <- floor(otu_table(physeq.scale))
  } else if (round == "round"){
    otu_table(physeq.scale) <- myround(otu_table(physeq.scale))
  }

  # Prune taxa and return new phyloseq object
  physeq.scale <- prune_taxa(taxa_sums(physeq.scale) > 0, physeq.scale)
  return(physeq.scale)
}

##### ADONIS ###########

# Function to run adonis test on a phyloseq object and a variable from metadata
# Make sure OTU data is standardized/normalized before 
phyloseq_to_adonis <- function(physeq, distmat = NULL, dist = "bray", formula) {

  if(!is.null(distmat)) {
    phydist <- distmat
  } else {
    phydist <- phyloseq::distance(physeq, dist)
  }

  metadata <- as(sample_data(physeq), "data.frame")

  # Adonis test
  f <- reformulate(formula, response = "phydist")
  adonis.test <- adonis(f, data = metadata)
  print(adonis.test)

  # Run homogeneity of dispersion test if there is only 1 variable
  if (grepl("\\+", formula)) {
    l <- list(
      dist = phydist, 
      formula = f, 
      adonis = adonis.test
    )
  } else {
    group <- metadata[ ,formula]
    beta <- betadisper(phydist, group)
    disper.test = permutest(beta)
    print(disper.test)

    l <- list(
      dist = phydist, 
      formula = f, 
      adonis = adonis.test, 
      disper = disper.test
    )
  }
  return (l)
}

########## Bar Plots #################

# This function takes a phyloseq object, agglomerates OTUs to the desired taxonomic rank, 
# prunes out OTUs below a certain relative proportion in a sample (ie 1% ) 
# and melts the phyloseq object into long format which is suitable for ggplot stacked barplots.
taxglom_and_melt <- function(physeq, taxrank, prune){

  # Agglomerate all otu's by given taxonomic level
  pglom <- tax_glom(physeq, taxrank = taxrank)

  # Create a new phyloseq object which removes taxa from each sample below the prune parameter
  pglom_prune <- transform_sample_counts(pglom,function(x) {x/sum(x)})
  otu_table(pglom_prune)[otu_table(pglom_prune) < prune] <- 0
  pglom_prune <- prune_taxa(taxa_sums(pglom_prune) > 0, pglom_prune)

  # Melt into long format and sort by taxonomy
  physeq_long <- psmelt(pglom_prune)
  physeq_long <- physeq_long[order(physeq_long[ ,taxrank]), ]

  # Return long data frame
  return(physeq_long)
}

###### Merge functions ############

# Merge samples by averaging OTU counts instead of summing
merge_samples_mean <- function(physeq, group, round){
  # Calculate the number of samples in each group
  group_sums <- as.matrix(table(sample_data(physeq)[ ,group]))[,1]

  # Merge samples by summing
  merged <- merge_samples(physeq, group)

  # Divide summed OTU counts by number of samples in each group to get mean
  # Calculation is done while taxa are columns
  x <- as.matrix(otu_table(merged))
  if(taxa_are_rows(merged)){ x<-t(x) }

  # Pick the rounding functions
  if (round == "floor"){
    out <- floor(t(x/group_sums))
  } else if (round == "round"){
    out <- myround(t(x/group_sums))
  }

  # Return new phyloseq object with taxa as rows
  out <- otu_table(out, taxa_are_rows = TRUE)
  otu_table(merged) <- out
  return(merged)
}

# Merge samples, just including OTUs that were present in all merged samples
# Call this function before running merge_samples()
merge_OTU_intersect <- function(physeq, group){

  # Make sure we're not starting with more taxa than we need 
  physeq <- prune_taxa(taxa_sums(physeq) > 0, physeq)

  s <- data.frame(sample_data(physeq))
  l <- levels(s[,group])
  o <- otu_table(physeq)

  # Loop through each category
  for (cat in 1:length(l)) {

    # Get the index of all samples in that category
    w <- which(s[,group]==l[cat])

    # subset to just those columns of OTU table
    cat.sub<-o[,w]
    print(dim(cat.sub))

    # Find the indices of 0's in the OTU table
    zeros <- apply(cat.sub, 1, function(r) any(r == 0))

    # If an OTU had a 0 in at least one sample, change all samples to 0
    cat.sub[zeros,] <- 0
  }

  o[,w] <- cat.sub
  otu_table(physeq) <- o

  return(physeq)

}

```

Atlas1006
https://rdrr.io/github/microbiome/microbiome/man/atlas1006.html
Using dataset: atlas1006: HITChip Atlas with 1006 Western adults
This data set contains genus-level microbiota profiling with HITChip for 1006 western adults with no reported health complications. 
Format: phyloseq-class format
Usage: data(atlas1006)

## Data Set-up
```{r Data set-up}

#setwd("/Users/maryam/Scripts/Microbiome")

	## Get data
	library(foreign)
	library(stargazer)
	library(psych)
	library(plyr)
	library(xtable)
	library(car)
	#library(dplyr)
	library(scales)
	library(reshape) 
	library(gmodels)
	library(nnet)
	library(tidyr)
	library(sjmisc)
	library(RColorBrewer)
  library(htmltools)
	library(broom)

  #library(hrbrthemes)
library(gcookbook)
library(tidyverse)
 #library(gdtools)

#The phyloseq package seeks to address issues with multiple microbiome analysis packages by providing a set of functions that internally manage the organizing, linking, storing, and analyzing of phylogenetic sequencing data. In general, this package is used for UniFrac analyses.
library(phyloseq)
library(microbiome)
library(knitr)
library(ggpubr)

  #library(phylogeo)
library(dplyr)

#Graphing package used in phyloseq. To edit the default setting of a plot, you need to use functions in this package.
library(ggplot2)

library(viridis) 
library(cowplot)

# install.packages("devtools")
library(devtools)
# install_github("opisthokonta/tsnemicrobiota")
library(tsnemicrobiota)

# Data from http://dx.doi.org/10.1038/ncomms7342
data(atlas1006)   
pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006

```

## MND,PD,AD Variable
```{r mnd variable}

# Create mnd variable
# grabs the nationality from phyloseq
mbnational <- get_variable(pseq, "nationality" )
# check you've installed this library
library(forcats)

# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
mndmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

pdmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

admbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

# reorder
mndmbnational <- factor(mndmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
pdmbnational <- factor(pdmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
admbnational <- factor(admbnational, levels = (c("LOW", "MEDIUM", "HIGH")))

# creates a new variable in the phyloseq called mnd
sample_data(pseq)$mnd = mndmbnational
sample_data(pseq)$pd = pdmbnational
sample_data(pseq)$ad = admbnational

# checks that it has worked.
get_variable(pseq, "mnd")

```

# Clostridum grouping
```{r Clostridium grouping}

clostridium <- subset_taxa(pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XVIa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales")

pseq.new <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacilli"| Phylum =="Proteobacteria"  | Phylum =="Verruccomicrobia"  | Phylum =="Bacteroidetes" | Phylum =="Asteroleplasma"| Phylum =="Spirochaetes" | Phylum =="Fusobacteria" | Phylum =="Cyanobacteria" | Phylum =="Uncultured Mollucutes")

```

## pseq summary
```{r pseq}

pseq

```

```{r List of genera in phylum}

# List of Genera in the Bacteroideted Phylum
taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
print(taxa)

```

```{r MND table summary}
table(sample_data(pseq)$nationality)
table(sample_data(pseq)$mnd)
table(sample_data(pseq)$pd)
table(sample_data(pseq)$ad)
```

## Summary REMOVE?
```{r summary}

# Summarise contents of phyloseq
summarize_phyloseq(pseq)
# Sparsity is how populated is the data with zeros

kable(head(tax_table(pseq)))
tax_table(pseq)

# Check the data
print(pseq)

# Display variables from the mapping file
sample_variables(pseq)

# Total number of taxa in the entire data
ntaxa(pseq)

# Taxonomic ranks
rank_names(pseq)

```

## Check OTUs/Phylums
Prune(remove) taxa because we want to remove OTUs/taxa that aren't in any of the samples
Don't need to prune this taxa because OTUs are found in all samples.
Check this everytime you use the subset_samples function because of the pseq object being filtered. 
```{r check OTUs}
# Check if any OTUs are not present in any samples
any(taxa_sums(pseq) == 0)
# FALSE, therefore, OTUs are found in all samples and we don't need to prune the taxa. If the answer is TRUE, some OTUs are not present in any samples. This is usually the case when data is subset to remove some samples. OTUs unique to those sample are not removed along with the samples. Therefore, it is important to check this everytime the phyloseq object is filtered for samples using `subset_samples` function.

head(meta(pseq))

```

## ALPHA DIVERSITY
Diversity: Shannon, Simpson
Richness: Chao, ACE

## Prepare Data for Alpha Diversity Visualisation
```{r prepare data for alpha diversity visualisation}

# Prepare data for vizualisation

# Alpha diversity again
png("alpha.png")
tab <- diversities(pseq, index = "all")
stargazer(tab, type = "text",summary = NULL)
kable(head(tab))
plot(tab)
dev.off()

# Get metadata
pseq.meta <- meta(pseq)
kable(head(pseq.meta))
stargazer(pseq.meta, type = "text",summary = NULL)

# Add the diversity table to metadata
pseq.meta$Shannon <- tab$shannon 
pseq.meta$InverseSimpson <- tab$inverse_simpson
pseq.meta$fisher <- tab$fisher
pseq.meta$chao1 <- tab$chao1

```

## Nationality Violin Plot
```{r nationality violin plot}

# Alpha diversity is the species richness of a place
# High alpha diversity = high no. of species

pseq.meta$nationality <- factor(pseq.meta$nationality, levels = c("Scandinavia","EasternEurope","CentralEurope", "SouthEurope", "UKIE", "US"))

# Violin plot
# the number in the palettes must equal the number of groups in nationlaity.pairs
p1 <- ggviolin(pseq.meta, x = "nationality", y = "Shannon",
 add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
  print(p1)
ggsave("nation_violin_shannon.png")

p2 <- ggviolin(pseq.meta, x = "nationality", y = "InverseSimpson",
 add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
print(p2)
ggsave("nation_violin_simpson.png")

p3 <- ggviolin(pseq.meta, x = "nationality", y = "fisher",
 add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("") + ylab("Fisher")
print(p3)
ggsave("nation_violin_fisher.png")

plot_grid(p1, p2, p3, align = "h" , ncol = 2, nrow = 2)
ggsave("nation_violin.png")

```

```{r}

p <- plot_richness(pseq, x="nationality", measures=c("Chao1", "Shannon", "Simpson")) 
p + geom_boxplot()
ggsave("alpha_nationality.png")

```

## MND Violin Plot
```{r mnd violin plot}

# Add the diversity table to metadata
pseq.meta$Shannon <- tab$shannon
pseq.meta$InverseSimpson <- tab$inverse_simpson

# Compare differences in Shannon index between mnd group of the study subjects
pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")

pseq.meta$pd <- factor(pseq.meta$pd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")

pseq.meta$ad <- factor(pseq.meta$ad, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")

# Create a list of pairwise comaprisons
mnd <- levels(pseq.meta$mnd) # get the variables
pd <- levels(pseq.meta$pd) # get the variables
ad <- levels(pseq.meta$ad) # get the variables

# Make a pairwise list that we want to compare.
mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))

pd.pairs <- combn(seq_along(pd), 2, simplify = FALSE, FUN = function(i)pd[i])
pd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))

ad.pairs <- combn(seq_along(ad), 2, simplify = FALSE, FUN = function(i)ad[i])
ad.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))

# Shannon
# the number in the palettes must equal the number of groups in bmi.pairs
p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
p1 <- p1 + stat_compare_means(comparisons = mnd.pairs, paired = FALSE) + xlab("MND Incidence") + theme(legend.position="none")
print(p1)
ggsave("mnd_violin__shannon_stats.png")

# Inverse Simpson
p2 <- ggviolin(pseq.meta, x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
p2 <- p2 + stat_compare_means(comparisons = mnd.pairs) 
print(p2)
ggsave("mnd_violin_simpson.png")

```

## BETA DIVERSITY
Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
WE want to: quantify community divergence within a given sample set
Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.

```{r Phylum NMDS}

png("Phylum_NMDS.png")
pseq.ord <- ordinate(pseq, "NMDS", "bray")
p1 = plot_ordination(pseq, pseq.ord, type = "taxa", color = "mnd", title = "Taxa by Phylum")
p1 + facet_wrap(~Phylum, 3)
dev.off()

```

Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance. This was applied in group-level comparisons for instance in Salonen et al. ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
```{r Calculate group divergences}

# Use these to compare microbiota divergence within each group

p.low <- subset_samples(pseq, mnd == "LOW")
p.low.d <- divergence(p.low)

p.med <- subset_samples(pseq, mnd == "MEDIUM")
p.med.d <- divergence(p.med)

p.high <- subset_samples(pseq, mnd == "HIGH")
p.high.d <- divergence(p.high)

p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
plot(p)

```
 The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):

# VISUALISE MICORBIOME VARIATION
```{r convert to compositional data}

# Convert to compositional data
pseq.rel <- microbiome::transform(pseq, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

```

NMDS/Bray
Bray-Curtis dissimilarity does not consider phylogenetic relationships between OTUs. There are several distance methods and a list can be obtained by typying ?distanceMethodList in the console pane.
```{r}

pseq.ord <- ordinate(pseq.rel, "NMDS", "bray")

p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
ggsave("NMDSt.png")

p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
ggsave("NMDS_facet.png")

```

MDS ordination on Bray-Curtis distance
Performs principal coordinate analysis (also called principle coordinate decomposition, multidimensional scaling (MDS), or classical scaling) of a distance matrix (Gower 1966), including two correction methods for negative eigenvalues. See pcoa for further details.
```{r}

pseq.ord <- ordinate(pseq.rel, "MDS", "bray")

p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
ggsave("MDS.png")

p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
ggsave("NMDS_facet.png")

```

PCoA for compositional data with Bray-Curtis distances
```{r}

p <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
p2 <- p + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b")

p1 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
print(p1)

```

```{r PCoA Bray}

# Pick core taxa
pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)

# PCoA for compositional data with Bray-Curtis distances
png("PCoA_bray.png")
p <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
                      method = "PCoA", distance = "bray") +
       labs(title = paste("PCoA / Compositional / Bray-Curtis"))
print(p)
dev.off()

```

t-SNE is a popular new ordination technique
```{r}

tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
  color = 'mnd', title='t-SNE (Bray)')
tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
ggsave("t_sne.png")

```

## PERMANOVA significance test for group-level differences
Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition

```{r Redefine data}

# Convert to compositional data
pseqLvH <- merge_phyloseq(p.low,p.high)
pseqLvM <- merge_phyloseq(p.low,p.med)

pseq.rel <- microbiome::transform(pseqLvH, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

```

```{r PERMANOVA}

# samples x species as input
library(vegan)
permanova <- adonis(t(otu) ~ mnd,
               data = meta, permutations=999, method = "bray")

# P-value
print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
summary(permanova$aov.tab)
```

Check that variance homogeneity assumptions hold (to ensure the reliability of the results):Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
```{r ANOVA}

dist <- vegdist(t(otu))
anov <- anova(betadisper(dist, meta$mnd))
summary(anov)
tidy(anov)

```

Show coefficients for the top taxa separating the groups
```{r top taxa L vs H}

coef <- coefficients(permanova)["mnd1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")

ggsave("LvH.png")

```

```{r L vs M}

pseqLvM <- merge_phyloseq(p.low,p.med)

pseq.rel <- microbiome::transform(pseqLvM, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

library(vegan)
permanova <- adonis(t(otu) ~ mnd,
               data = meta, permutations=999, method = "bray")

# P-value
print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])

dist <- vegdist(t(otu))
anov <- anova(betadisper(dist, meta$mnd))
summary(anov)
tidy(anov)

coef <- coefficients(permanova)["mnd1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))

barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")

ggsave("LvM.png", width = 20, height = 20, units = "cm")

```

## MICROBIOME COMPOSITION

```{r Phylum #}

# Show available ranks in the dataset
rank_names(pseq)

# Create table, number of features for each phyla
table(tax_table(pseq)[, "Phylum"], exclude = NULL)

```
This shows a few phyla for which only one feature was observed. Those may be worth filtering.

```{r Total Phylum abundnace bar}

png("Phylum_abundance.png")
p<- plot_bar(pseq,x="Phylum",fill = "Phylum") +geom_bar(stat="identity")

phylumGlommed = tax_glom(pseq, "Phylum")
plot_bar(phylumGlommed, x="Phylum", fill ="Phylum") +geom_bar(stat="identity")

pd <- p$data %>% 
    as_tibble %>%
    mutate(Genus = as.character(Genus))
genus_abun <- pd %>%
    group_by(Genus) %>%
    summarize(Abundance = sum(Abundance)) %>%
    arrange(Abundance)
genus_levels <- genus_abun$Genus
pd0 <- pd %>%
    mutate(Genus = factor(Genus, genus_levels))
ggplot(pd0, aes(x = Sample, y = Abundance, fill = Genus)) + 
    geom_bar(stat = "identity")
dev.off()

```

```{r Phylum prevalence table}

prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
prev.otu + theme(legend.position="none")
ggsave("phylum_prevalence.png")

tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
tab + theme(legend.position="none")
ggsave("Phylum_prevalence_detection5.png")

```

theme(axis.text.x = element_text(angle = 90)
+ guide_italics
  ggtitle("Relative abundance")  + 
```{r Relative abundance}

pseq.ph <- aggregate_taxa(pseq.new, "Phylum")

pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
  theme(legend.position = "bottom") + theme_bw() + 
  theme(axis.text.x=element_blank()) + 
  theme(legend.title = element_text(size=18)) + scale_fill_brewer(palette = "Paired")
print(plot.comp.rel)
ggsave("phylum_relative_abundance11.png")

plot.comp.rel + theme(legend.position="none")
ggsave("phylum_relative_abundance12.png")

```

guide_italics
ggtitle("Relative abundance") +  
```{r}

clos.ph <- aggregate_taxa(clostridium, "Phylum")

clos.ph.rel <- microbiome::transform(clos.ph, "compositional")

plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
  theme(legend.position = "bottom", ) + theme_bw() + 
  theme(axis.text.x=element_blank()) + 
  theme(legend.title = element_text(size=18))
plot.comp.rel + scale_fill_brewer(palette = "Paired")

plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
 theme(legend.position = "bottom") + theme_bw() +
 theme(axis.text.x=element_blank()) +
 theme(legend.title = element_text(size=18),legend.text = element_text( face="italic") ) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("my awesome title"))

ggsave("phylum_relative_abundance_clos.png")

plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none")
ggsave("phylum_relative_abundance_clos2.png")

```

```{r Phylum Bar Plot}

levels(pseq.meta$mnd)

pseq.ph <- aggregate_taxa(pseq, "Phylum")

pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot")

plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
ggsave("phylum_abundance_bar.png")

```

```{r pseq.new bar plot}

p.new <- aggregate_taxa(pseq.new, "Phylum")

p.new.rel <- microbiome::transform(p.new, "compositional")

plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")

plot.p.new.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
ggsave("pnew_abundance_bar.png")

```

```{r CLostridium Bar Plot}

clos <- aggregate_taxa(clostridium, "Phylum")

pseq.cl.rel <- microbiome::transform(clos, "compositional")

plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")

plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
ggsave("cl_abundance_bar.png")

```

```{r Clostridium genus bar plots}

pseq.cl.rel <- microbiome::transform(clostridium, "compositional")

plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")

plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")

```

```{r Top Phylum Heatmap}

pseq.ph <- aggregate_taxa(pseq, "Phylum")

pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")

plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
ggsave("phylum_histogram.png")

```

```{r p.new heatmap}

pseq.n.ph <- aggregate_taxa(pseq.new, "Phylum")

pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")

plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")

p <- plot.n.comp.rel + coord_flip()
p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
ggsave("pnew_histogram.png")

```

```{r clostridium heatmap}

pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")

pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")

plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")

p <- plot.c.comp.rel + coord_flip()
p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
ggsave("clos_histogram.png")

```

```{r Network}

png("network_bray_phylum.png")
jg = make_network(pseq, "taxa", "bray", 0.3)
plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
dev.off()

```

## Community composition plotting
Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results. These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted. Interpretation of these plots should always be subject to subsequent statistical analysis.
```{r community-composition-plots}
# Create a data table for ggplot
# agglomerate at phylum level
pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")

# Transform to rel. abundance (or use ps0.ra)
pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})

# Melt to long format for easy ggploting
pseq_phylum <- psmelt(pseq_phyl)

     # filter(Abundance > 0.01)                             # Filter out low abundance taxa

# Plot - Phylum
p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(title = "Abundant Phylum (> 1%)")

p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
plot(p)
ggsave("phylum_abun_mnd.png")

```

```{r}

clos <- aggregate_taxa(clostridium, "Phylum")

# agglomerate at phylum level
clos_phy <- tax_glom(clos, taxrank = "Phylum")

# Transform to rel. abundance (or use ps0.ra)
clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})

# Melt to long format for easy ggploting
clos_phylum <- psmelt(clos_phyl)

# Plot - Phylum
p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(title = "Abundant Phylum of Clostridium(> 1%)")

p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
plot(p)
ggsave("phylum_abun_mnd_clos.png")

```

```{r}

get_taxa_unique(pseq, "Phylum")

```

```{r}

get_taxa_unique(pseq, "Genus")

```

```{r}

get_taxa_unique(clostridium, "Genus")

```

```{r}

df <- psmelt(pseq)
kable(head(df))

```

```{r}

head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])

```

### 2019-06-18 microbiome_online.Rmd
updated

@@
 
-ps_merge <- merge_phyloseq(esophagus, atlas1006)
+mergedGP = merge_samples(TwinsUK,atlas1006)
+
+ps_merge <- merge_phyloseq(atlas1006, tax_table(TwinsUK))
+
+ps_merge <- merge_phyloseq(TwinsUK, atlas1006)
 
@@
 
-sample_data(batmicrobiome)
+sample_data(atlas1006)
+
+ sample_data(atlas1006)
+
+ data=subset(pseq.meta, !is.na(mnd))
 
@@
 
+pseq_temp <- baseline(atlas1006)
+
+temp <- complete.cases(sample_data(atlas1006))
+
 sample_data(atlas1006)$human = factor( get_variable(atlas1006) %in% c("Feces", "Mock", "Skin", "Tongue") )
@@
 sample_data(atlas1006)$human = sample(50:1000)
+
+table(sample_data(atlas1006)$nationality)
 
@@
 
+clostridium <- subset_taxa(pseq, Phylum =="Clostridium cluster I"| "Clostridium cluster XI"|"Clostridium cluster XVI"| "Clostridium cluster III"| "Clostridium cluster XIII"| "Clostridium cluster XVII"| "Clostridium cluster IV"| "Clostridium cluster XVIa"| "Clostridium cluster XVIII" |"Clostridium cluster IX"| "Clostridium cluster XV"| "Uncultured Clostridiales")
+
 stomach <- subset_samples(physeq3, Sample_type=="Stomach")
@@
 
+  # drops it into different countries
+  pseq.subset <- subset_samples(pseq, nationality == "SouthEurope"& gender=="female" )
+
+  clostridium <- subset_samples(pseq, Phylum =="Clostridium")
+
+  library(car)
+recode(sample_data(pseq)$nationality, "c('SouthEurope', 'CentralEurope')='A+B'")
+
+  df %>%
+   mutate(x= recode(sample_data(pseq)$nationality, "c('SouthEurope', 'CentralEurope')='A+B'"))
+
+sample_data(pseq)$bob[subset_samples(pseq, nationality == "SouthEurope"& nationality == "Scandinavia" )] <- "TEST"
+
+# grabs the nationality from phyloseq
+test <- get_variable(pseq,"nationality" ) 
+# check you've installed this library
+library(forcats)
+# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
+test <- fct_collapse(test, AB = c("SouthEurope","CentralEurope"))
+# creates a new variable in the phyloseq called mnd
+sample_data(pseq)$mnd = test
+# checks that it has worked. 
+get_variable(pseq,"mnd" )
+
+data$grip6max[data$grprespd==3 & data$sex==1] <- 29.46
+
+# Gives an overview of the data set
+sample_data(cdi_schubert)
+sample_data(ps_merge)
+sample_data(pseq)$nationality
+
+https://www.nature.com/articles/s41467-017-01973-8.pdf?origin=ppub
+
+dietswap
+
+ps_merge <- merge_phyloseq(cdi_schubert, atlas1006)
+
+)
+
 Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.
@@
 ```{r}
-	setwd("/omega/bashscripts/microbiome")
+	setwd("/omega/bashscripts/microbiomedatarepo/datasets/microbiomeDB")
 
@@
 library(phangorn)
+library(microbiomeutilities)
 
@@
 
+p <- ggplot(tab) + 
 ```

### 2019-06-18 GlobalMicrobiome_NS.Rmd
updated

@@
 plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-ggsave("NMDSt.png")
+ggsave("NMDStMND.png")
 
 p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
-ggsave("NMDS_facet.png")
+ggsave("NMDS_facetMND.png")
+
+p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
+plot(p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
+ggsave("NMDStpd.png")
+
+p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "PD Prevalence")
+ggsave("NMDS_facetpd.png")
+
+p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
+plot(p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
+ggsave("NMDStpd.png")
+
+p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "PD Prevalence")
+ggsave("NMDS_facetpd.png")
 
@@
 ggsave("NMDS_facet.png")
+
+p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
+plot(p2 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
+ggsave("MDSpd.png")
+
+p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, 3) + labs(color = "PD Prevalence")
+ggsave("NMDS_facetPD.png")
+
+p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
+plot(p3 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
+ggsave("MDSad.png")
+
+p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, 3) + labs(color = "AD Prevalence")
+ggsave("NMDS_facetAD.png")
 

### 2019-06-19 GlobalMicrobiome_NS.Rmd
updated

@@
 knitr::opts_chunk$set(echo = TRUE)
-setwd("/omega/1006") 
+knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
+
+#test
 ```
 
-```{r}
+```{r functions}
 
 ##### FUNCTIONS ########
-
 ##### Normalization #######
@@
 
-Atlas1006
-https://rdrr.io/github/microbiome/microbiome/man/atlas1006.html
-Using dataset: atlas1006: HITChip Atlas with 1006 Western adults
-This data set contains genus-level microbiota profiling with HITChip for 1006 western adults with no reported health complications. 
-Format: phyloseq-class format
-Usage: data(atlas1006)
-
 ## Data Set-up
@@
 library(ggpubr)
-
 #library(phylogeo)
 library(dplyr)
-
 #Graphing package used in phyloseq. To edit the default setting of a plot, you need to use functions in this package.
 library(ggplot2)
-
 library(viridis) 
 library(cowplot)
-
 # install.packages("devtools")
@@
 library(tsnemicrobiota)
+# check you've installed this library
+library(forcats)
 
 # Data from http://dx.doi.org/10.1038/ncomms7342
-data(atlas1006)   
+#This data set contains genus-level microbiota profiling with HITChip for 1006 western adults with no reported health complications.
+
 pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
@@
 mbnational <- get_variable(pseq, "nationality" )
-# check you've installed this library
-library(forcats)
 

### 2019-06-19 GlobalMicrobiome_NS.Rmd
updated

@@
 # checks that it has worked.
-get_variable(pseq, "mnd")
+#get_variable(pseq, "mnd")
 
@@
 ```{r pseq}
-
 pseq
-
 ```
@@
 ```{r List of genera in phylum}
-
 # List of Genera in the Bacteroideted Phylum
@@
 print(taxa)
-
 ```
@@
 # Sparsity is how populated is the data with zeros
-
 kable(head(tax_table(pseq)))
@@
 # Prepare data for vizualisation
-
-# Alpha diversity again
+# Alpha diversity for the entire data set
 png("alpha.png")
@@
 
-Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
-This was applied in group-level comparisons for instance in Salonen et al.
-ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
-```{r Calculate group divergences}
-
-# Use these to compare microbiota divergence within each group
-
-p.low <- subset_samples(pseq, mnd == "LOW")
-p.low.d <- divergence(p.low)
-
-p.med <- subset_samples(pseq, mnd == "MEDIUM")
-p.med.d <- divergence(p.med)
-
-p.high <- subset_samples(pseq, mnd == "HIGH")
-p.high.d <- divergence(p.high)
-
-p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
-plot(p)
-
-```
  The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
@@
 
-p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "PD Prevalence")
+p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, nrow = 2) + labs(color = "PD Prevalence")
 ggsave("NMDS_facetpd.png")
 
-p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
-plot(p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
-ggsave("NMDStpd.png")
+p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
+plot(p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
+ggsave("NMDStad.png")
 
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "PD Prevalence")
-ggsave("NMDS_facetpd.png")
+p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, nrow = 2) + labs(color = "AD Prevalence")
+ggsave("NMDS_facetad.png")
 
@@
 
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, 3) + labs(color = "AD Prevalence")
+p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, 3) + labs(color = "AD Prevalence")
 ggsave("NMDS_facetAD.png")
@@
 tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+#MND
 tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
@@
 tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
-ggsave("t_sne.png")
+ggsave("mnd_t_sne.png")
+# PD
+tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
+  color = 'pd', title='t-SNE (Bray)')
+tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence")
+ggsave("pd_t_sne.png")
+
+# AD
+tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
+  color = 'ad', title='t-SNE (Bray)')
+tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence")
+ggsave("ad_t_sne.png")
+
+```
+
+Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
+This was applied in group-level comparisons for instance in Salonen et al.
+ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
+```{r Calculate group divergences}
+
+# Use these to compare microbiota divergence within each group
+
+#MND
+mnd.p.low <- subset_samples(pseq, mnd == "LOW")
+mnd.p.low.d <- divergence(mnd.p.low)
+
+mnd.p.med <- subset_samples(pseq, mnd == "MEDIUM")
+p.med.d <- divergence(mnd.p.med)
+
+mnd.p.high <- subset_samples(pseq, mnd == "HIGH")
+p.high.d <- divergence(mnd.p.high)
+
+mnd.p <- boxplot(list(Low = mnd.p.low.d, Medium = mnd.p.med.d, High = mnd.p.high.d))
+plot(mnd.p)
+
+# PD
+
+pd.p.low <- subset_samples(pseq, pd == "LOW")
+pd.p.low.d <- divergence(pd.p.low)
+
+pd.p.med <- subset_samples(pseq, pd == "MEDIUM")
+p.med.d <- divergence(pd.p.med)
+
+pd.p.high <- subset_samples(pseq, pd == "HIGH")
+p.high.d <- divergence(pd.p.high)
+
+pd.p <- boxplot(list(Low = pd.p.low.d, Medium = pd.p.med.d, High = pd.p.high.d))
+plot(pd.p)
+
+# AD
+
+ad.p.low <- subset_samples(pseq, ad == "LOW")
+ad.p.low.d <- divergence(ad.p.low)
+
+ad.p.med <- subset_samples(pseq, ad == "MEDIUM")
+p.med.d <- divergence(ad.p.med)
+
+ad.p.high <- subset_samples(pseq, ad == "HIGH")
+p.high.d <- divergence(ad.p.high)
+
+ad.p <- boxplot(list(Low = ad.p.low.d, Medium = ad.p.med.d, High = ad.p.high.d))
+plot(ad.p)
 
@@
 # Convert to compositional data
-pseqLvH <- merge_phyloseq(p.low,p.high)
-pseqLvM <- merge_phyloseq(p.low,p.med)
+# MND
+mnd.pseqLvH <- merge_phyloseq(mnd.p.low,mnd.p.high)
+mnd.pseqLvM <- merge_phyloseq(mnd.p.low,mnd.p.med)
 
-pseq.rel <- microbiome::transform(pseqLvH, "compositional")
-otu <- abundances(pseq.rel)
-meta <- meta(pseq.rel)
+mnd.pseq.rel <- microbiome::transform(mnd.pseqLvH, "compositional")
+mnd.otu <- abundances(mnd.pseq.rel)
+mnd.meta <- meta(mnd.pseq.rel)
+
+# PD
+
+pd.pseqLvH <- merge_phyloseq(pd.p.low,pd.p.high)
+pd.pseqLvM <- merge_phyloseq(pd.p.low,pd.p.med)
+
+pd.pseq.rel <- microbiome::transform(pd.pseqLvH, "compositional")
+pd.otu <- abundances(pd.pseq.rel)
+pd.meta <- meta(pd.pseq.rel)
+
+# AD
+
+ad.pseqLvH <- merge_phyloseq(ad.p.low,ad.p.high)
+ad.pseqLvM <- merge_phyloseq(ad.p.low,ad.p.med)
+
+ad.pseq.rel <- microbiome::transform(ad.pseqLvH, "compositional")
+ad.otu <- abundances(ad.pseq.rel)
+ad.meta <- meta(ad.pseq.rel)
 

### 2019-06-19 GlobalMicrobiome_NS.Rmd
updated

@@
 knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
-
+# test
 #test
@@
 library(vegan)
-permanova <- adonis(t(otu) ~ mnd,
+
+# MND
+mnd.permanova <- adonis(t(mnd.otu) ~ mnd,
                data = meta, permutations=999, method = "bray")
-
 # P-value
-print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
-summary(permanova$aov.tab)
+print(as.data.frame(mnd.permanova$aov.tab)["mnd", "Pr(>F)"])
+summary(mnd.permanova$aov.tab)
+
+# PD
+pd.permanova <- adonis(t(pd.otu) ~ pd,
+               data = meta, permutations=999, method = "bray")
+# P-value
+print(as.data.frame(pd.permanova$aov.tab)["pd", "Pr(>F)"])
+summary(pd.permanova$aov.tab)
+
+# AD
+ad.permanova <- adonis(t(ad.otu) ~ ad,
+               data = meta, permutations=999, method = "bray")
+# P-value
+print(as.data.frame(ad.permanova$aov.tab)["ad", "Pr(>F)"])
+summary(ad.permanova$aov.tab)
+
 ```
@@
 ```{r ANOVA}
-
-dist <- vegdist(t(otu))
-anov <- anova(betadisper(dist, meta$mnd))
-summary(anov)
-tidy(anov)
+#MND
+mnd.dist <- vegdist(t(mnd.otu))
+mnd.anov <- anova(betadisper(mnd.dist, meta$mnd))
+summary(mnd.anov)
+tidy(mnd.anov)
+
+#PD
+pd.dist <- vegdist(t(pd.otu))
+pd.anov <- anova(betadisper(pd.dist, meta$pd))
+summary(pd.anov)
+tidy(pd.anov)
+
+#AD
+ad.dist <- vegdist(t(ad.otu))
+ad.anov <- anova(betadisper(ad.dist, meta$ad))
+summary(ad.anov)
+tidy(ad.anov)
 

### 2019-06-19 GlobalMicrobiome_NS.Rmd
updated

@@
 
-coef <- coefficients(permanova)["mnd1",]
-top.coef <- coef[rev(order(abs(coef)))[1:20]]
+# MDN
+mnd.coef <- coefficients(mnd.permanova)["mnd1",]
+mnd.top.coef <- coef[rev(order(abs(mnd.coef)))[1:20]]
 par(mar = c(3, 14, 2, 1))
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
-
-ggsave("LvH.png")
+barplot(sort(mnd.top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus") 
+ggsave("mndLvH.png")
+
+# PD
+pd.coef <- coefficients(pd.permanova)["pd1",]
+pd.top.coef <- coef[rev(order(abs(pd.coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(pd.top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus") 
+ggsave("pdLvH.png")
+
+# PD
+ad.coef <- coefficients(ad.permanova)["ad1",]
+ad.top.coef <- coef[rev(order(abs(ad.coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(ad.top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus") 
+ggsave("adLvH.png")
 
@@
 
-pseqLvM <- merge_phyloseq(p.low,p.med)
-
-pseq.rel <- microbiome::transform(pseqLvM, "compositional")
-otu <- abundances(pseq.rel)
-meta <- meta(pseq.rel)
-
+#MND
+mnd.pseqLvM <- merge_phyloseq(mnd.p.low,mnd.p.med)
+mnd.pseq.rel <- microbiome::transform(mnd.pseqLvM, "compositional")
+mnd.otu <- abundances(mnd.pseq.rel)
+mnd.meta <- meta(mnd.pseq.rel)
+
+#PD
+pd.pseqLvM <- merge_phyloseq(pd.p.low,pd.p.med)
+pd.pseq.rel <- microbiome::transform(pd.pseqLvM, "compositional")
+pd.otu <- abundances(pd.pseq.rel)
+pd.meta <- meta(pd.pseq.rel)
+
+#AD
+ad.pseqLvM <- merge_phyloseq(ad.p.low,ad.p.med)
+ad.pseq.rel <- microbiome::transform(ad.pseqLvM, "compositional")
+ad.otu <- abundances(ad.pseq.rel)
+ad.meta <- meta(ad.pseq.rel)
+
+# MND
 library(vegan)
-permanova <- adonis(t(otu) ~ mnd,
+mnd.permanova <- adonis(t(mnd.otu) ~ mnd,
                data = meta, permutations=999, method = "bray")
-
 # P-value
-print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
-
-dist <- vegdist(t(otu))
-anov <- anova(betadisper(dist, meta$mnd))
-summary(anov)
-tidy(anov)
-
-coef <- coefficients(permanova)["mnd1",]
-top.coef <- coef[rev(order(abs(coef)))[1:20]]
+print(as.data.frame(mnd.permanova$aov.tab)["mnd", "Pr(>F)"])
+mnd.dist <- vegdist(t(mnd.otu))
+mnd.anov <- anova(betadisper(mnd.dist, mnd.meta$mnd))
+summary(mnd.anov)
+tidy(mnd.anov)
+mnd.coef <- coefficients(mnd.permanova)["mnd1",]
+mnd.top.coef <- coef[rev(order(abs(mnd.coef)))[1:20]]
 par(mar = c(3, 14, 2, 1))
-
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
-
-ggsave("LvM.png", width = 20, height = 20, units = "cm")
+barplot(sort(mnd.top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
+ggsave("mndLvM.png", width = 20, height = 20, units = "cm")
+
+# PD
+library(vegan)
+pd.permanova <- adonis(t(pd.otu) ~ pd,
+               data = meta, permutations=999, method = "bray")
+# P-value
+print(as.data.frame(pd.permanova$aov.tab)["pd", "Pr(>F)"])
+pd.dist <- vegdist(t(pd.otu))
+pd.anov <- anova(betadisper(pd.dist, pd.meta$pd))
+summary(pd.anov)
+tidy(pd.anov)
+pd.coef <- coefficients(pd.permanova)["pd1",]
+pd.top.coef <- coef[rev(order(abs(pd.coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(pd.top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
+ggsave("pdLvM.png", width = 20, height = 20, units = "cm")
+
+# AD
+library(vegan)
+ad.permanova <- adonis(t(ad.otu) ~ ad,
+               data = meta, permutations=999, method = "bray")
+# P-value
+print(as.data.frame(ad.permanova$aov.tab)["ad", "Pr(>F)"])
+ad.dist <- vegdist(t(ad.otu))
+ad.anov <- anova(betadisper(ad.dist, ad.meta$ad))
+summary(ad.anov)
+tidy(ad.anov)
+ad.coef <- coefficients(ad.permanova)["ad1",]
+ad.top.coef <- coef[rev(order(abs(ad.coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(ad.top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
+ggsave("adLvM.png", width = 20, height = 20, units = "cm")
 
@@
 pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
 pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
+#MND
 plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot")
-
 plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("phylum_abundance_bar.png")
-
+ggsave("mndphylum_abundance_bar.png")
+#PD
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "pd", x.label = "PD Incidence", plot.type = "barplot")
+plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+ggsave("pdphylum_abundance_bar.png")
+#AD
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "ad", x.label = "AD Incidence", plot.type = "barplot")
+plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+ggsave("adphylum_abundance_bar.png")
 ```
@@
 p.new.rel <- microbiome::transform(p.new, "compositional")
-
+#MND
 plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
 plot.p.new.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("pnew_abundance_bar.png")
+ggsave("mndpnew_abundance_bar.png")
+#PD
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "pd", x.label = "sample", plot.type = "barplot")
+plot.p.new.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+ggsave("pdpnew_abundance_bar.png")
+#AD
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "ad", x.label = "sample", plot.type = "barplot")
+plot.p.new.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+ggsave("adpnew_abundance_bar.png")
 
@@
 
+#MND
 plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
 plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("cl_abundance_bar.png")
+ggsave("mndcl_abundance_bar.png")
+
+#PD
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "pd", x.label = "sample", plot.type = "barplot")
+plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+ggsave("pdcl_abundance_bar.png")
+
+#AD
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "ad", x.label = "sample", plot.type = "barplot")
+plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+ggsave("adcl_abundance_bar.png")
 
@@
 plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
 plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")

### 2019-06-19 GlobalMicrobiome_NS.Rmd
updated

@@
 pseq.cl.rel <- microbiome::transform(clostridium, "compositional")
-
+#MND
 plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
 plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A") 
+ggsave("mndcl_relabundance_bar.png")
+
+#PD
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "pd", x.label = "sample", plot.type = "barplot")
+plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A") 
+ggsave("pdcl_relabundance_bar.png")
+
+#AD
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "ad", x.label = "sample", plot.type = "barplot")
+plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A") 
+ggsave("adcl_relabundance_bar.png")
 
@@
 pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
 pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
 plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
-
 plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
@@
 pseq.n.ph <- aggregate_taxa(pseq.new, "Phylum")
-
 pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
-
 plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
-
 p <- plot.n.comp.rel + coord_flip()
@@
 pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")
-
 pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")
-
 plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")
-
 p <- plot.c.comp.rel + coord_flip()
@@
 
-# Plot - Phylum
+# Plot - Phylum MND
 p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
@@
 
+# Plot - Phylum PD
+p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~pd, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum (> 1%)")
+
+p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+plot(p)
+ggsave("phylum_abun_pd.png")
+
+# Plot - Phylum AD
+p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~ad, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum (> 1%)")
+
+p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+plot(p)
+ggsave("phylum_abun_ad.png")
+
 ```
@@
 clos <- aggregate_taxa(clostridium, "Phylum")
-
 # agglomerate at phylum level
 clos_phy <- tax_glom(clos, taxrank = "Phylum")                  
-
 # Transform to rel. abundance (or use ps0.ra)
 clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)}) 
-
 # Melt to long format for easy ggploting
@@
 
-# Plot - Phylum
+# Plot - Phylum MND
 p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
@@
 
+# Plot - Phylum PD
+p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~pd, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum of Clostridium(> 1%)")
+
+p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+plot(p)
+ggsave("phylum_abun_pd_clos.png")
+
+# Plot - Phylum AD
+p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~ad, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum of Clostridium(> 1%)")
+
+p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+plot(p)
+ggsave("phylum_abun_ad_clos.png")
+
 ```

### 2019-06-25 import_popln_data.Rmd
updated

--- 2019-06-19 GlobalMicrobiome_NS.Rmd
@@
 knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
-# test
-#test
-```
-
-```{r functions}
-
-##### FUNCTIONS ########
-##### Normalization #######
-
-# Better rounding function than R's base round
-myround <- function(x) { trunc(x + 0.5) }
-
-# Scales reads by 
-# 1) taking proportions
-# 2) multiplying by a given library size of n
-# 3) rounding 
-# Default for n is the minimum sample size in your library
-# Default for round is floor
-scale_reads <- function(physeq, n = min(sample_sums(physeq)), round = "floor") {
-
-  # transform counts to n
-  physeq.scale <- transform_sample_counts(physeq, 
-    function(x) {(n * x/sum(x))}
-  )
-
-  # Pick the rounding functions
-  if (round == "floor"){
-    otu_table(physeq.scale) <- floor(otu_table(physeq.scale))
-  } else if (round == "round"){
-    otu_table(physeq.scale) <- myround(otu_table(physeq.scale))
-  }
-
-  # Prune taxa and return new phyloseq object
-  physeq.scale <- prune_taxa(taxa_sums(physeq.scale) > 0, physeq.scale)
-  return(physeq.scale)
-}
-
-##### ADONIS ###########
-
-# Function to run adonis test on a phyloseq object and a variable from metadata
-# Make sure OTU data is standardized/normalized before 
-phyloseq_to_adonis <- function(physeq, distmat = NULL, dist = "bray", formula) {
-
-  if(!is.null(distmat)) {
-    phydist <- distmat
-  } else {
-    phydist <- phyloseq::distance(physeq, dist)
-  }
-
-  metadata <- as(sample_data(physeq), "data.frame")
-
-  # Adonis test
-  f <- reformulate(formula, response = "phydist")
-  adonis.test <- adonis(f, data = metadata)
-  print(adonis.test)
-
-  # Run homogeneity of dispersion test if there is only 1 variable
-  if (grepl("\\+", formula)) {
-    l <- list(
-      dist = phydist, 
-      formula = f, 
-      adonis = adonis.test
-    )
-  } else {
-    group <- metadata[ ,formula]
-    beta <- betadisper(phydist, group)
-    disper.test = permutest(beta)
-    print(disper.test)
-
-    l <- list(
-      dist = phydist, 
-      formula = f, 
-      adonis = adonis.test, 
-      disper = disper.test
-    )
-  }
-  return (l)
-}
-
-########## Bar Plots #################
-
-# This function takes a phyloseq object, agglomerates OTUs to the desired taxonomic rank, 
-# prunes out OTUs below a certain relative proportion in a sample (ie 1% ) 
-# and melts the phyloseq object into long format which is suitable for ggplot stacked barplots.
-taxglom_and_melt <- function(physeq, taxrank, prune){
-
-  # Agglomerate all otu's by given taxonomic level
-  pglom <- tax_glom(physeq, taxrank = taxrank)
-
-  # Create a new phyloseq object which removes taxa from each sample below the prune parameter
-  pglom_prune <- transform_sample_counts(pglom,function(x) {x/sum(x)})
-  otu_table(pglom_prune)[otu_table(pglom_prune) < prune] <- 0
-  pglom_prune <- prune_taxa(taxa_sums(pglom_prune) > 0, pglom_prune)
-
-  # Melt into long format and sort by taxonomy
-  physeq_long <- psmelt(pglom_prune)
-  physeq_long <- physeq_long[order(physeq_long[ ,taxrank]), ]
-
-  # Return long data frame
-  return(physeq_long)
-}
-
-###### Merge functions ############
-
-# Merge samples by averaging OTU counts instead of summing
-merge_samples_mean <- function(physeq, group, round){
-  # Calculate the number of samples in each group
-  group_sums <- as.matrix(table(sample_data(physeq)[ ,group]))[,1]
-
-  # Merge samples by summing
-  merged <- merge_samples(physeq, group)
-
-  # Divide summed OTU counts by number of samples in each group to get mean
-  # Calculation is done while taxa are columns
-  x <- as.matrix(otu_table(merged))
-  if(taxa_are_rows(merged)){ x<-t(x) }
-
-  # Pick the rounding functions
-  if (round == "floor"){
-    out <- floor(t(x/group_sums))
-  } else if (round == "round"){
-    out <- myround(t(x/group_sums))
-  }
-
-  # Return new phyloseq object with taxa as rows
-  out <- otu_table(out, taxa_are_rows = TRUE)
-  otu_table(merged) <- out
-  return(merged)
-}
-
-# Merge samples, just including OTUs that were present in all merged samples
-# Call this function before running merge_samples()
-merge_OTU_intersect <- function(physeq, group){
-
-  # Make sure we're not starting with more taxa than we need 
-  physeq <- prune_taxa(taxa_sums(physeq) > 0, physeq)
-
-  s <- data.frame(sample_data(physeq))
-  l <- levels(s[,group])
-  o <- otu_table(physeq)
-
-  # Loop through each category
-  for (cat in 1:length(l)) {
-
-    # Get the index of all samples in that category
-    w <- which(s[,group]==l[cat])
-
-    # subset to just those columns of OTU table
-    cat.sub<-o[,w]
-    print(dim(cat.sub))
-
-    # Find the indices of 0's in the OTU table
-    zeros <- apply(cat.sub, 1, function(r) any(r == 0))
-
-    # If an OTU had a 0 in at least one sample, change all samples to 0
-    cat.sub[zeros,] <- 0
-  }
-
-  o[,w] <- cat.sub
-  otu_table(physeq) <- o
-
-  return(physeq)
-
-}
 
@@
 
-pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
+setwd("~/globalmicrobiome")
+mnd.incidence <- read_csv("mnd.csv",col_names = TRUE)
+mnd.incidence <- as.data.frame(mnd.incidence)
+
+mb.gather <- function (x) {
+    mb.group <- data.frame("Location" = c("Central Europe","Eastern Europe","Scandinavia", "Southern Europe","UKIE","USA"), "Mean" = c(1:6)) 
+  c.europe <- filter(x,Location=="Belgium"|Location=="Denmark"|Location=="Germany"|Location=="The Netherlands")
+    e.europe <- filter(mnd.incidence,Location=="Poland")
+    scan <- filter(mnd.incidence,Location=="Finland"|Location=="Norway"|Location=="Sweden")
+    s.europe <- filter(mnd.incidence,Location=="France"|Location=="Italy"|Location=="Serbia"|Location=="Spain")
+    ukie <- filter(mnd.incidence,Location=="United Kingdom"|Location=="Ireland")
+    usa <- filter(mnd.incidence,Location=="United States")
+    mb.group[1,2] <- mean(c.europe$Value)
+    mb.group[2,2] <- mean(e.europe$Value)
+    mb.group[3,2] <- mean(scan$Value)
+    mb.group[4,2] <- mean(s.europe$Value)
+    mb.group[5,2] <- mean(ukie$Value)
+    mb.group[6,2] <- mean(usa$Value)
+    return(mb.group)
+    }
+
+temp <- mb.gather(mnd.incidence)
 
 ```
-
-## MND,PD,AD Variable
-```{r mnd variable}
-
-# Create mnd variable
-# grabs the nationality from phyloseq
-mbnational <- get_variable(pseq, "nationality" )
-
-# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-mndmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-pdmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-admbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-# reorder
-mndmbnational <- factor(mndmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-pdmbnational <- factor(pdmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-admbnational <- factor(admbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-
-# creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = mndmbnational
-sample_data(pseq)$pd = pdmbnational
-sample_data(pseq)$ad = admbnational
-
-# checks that it has worked.
-#get_variable(pseq, "mnd")
-
-```
-
-# Clostridum grouping
-```{r Clostridium grouping}
-
-clostridium <- subset_taxa(pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XVIa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales")
-
-pseq.new <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacilli"| Phylum =="Proteobacteria"  | Phylum =="Verruccomicrobia"  | Phylum =="Bacteroidetes" | Phylum =="Asteroleplasma"| Phylum =="Spirochaetes" | Phylum =="Fusobacteria" | Phylum =="Cyanobacteria" | Phylum =="Uncultured Mollucutes")
-
-```
-
-## pseq summary
-```{r pseq}
-pseq
-```
-
-```{r List of genera in phylum}
-# List of Genera in the Bacteroideted Phylum
-taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
-print(taxa)
-```
-
-```{r MND table summary}
-table(sample_data(pseq)$nationality)
-table(sample_data(pseq)$mnd)
-table(sample_data(pseq)$pd)
-table(sample_data(pseq)$ad)
-```
-
-## Summary REMOVE?
-```{r summary}
-
-# Summarise contents of phyloseq
-summarize_phyloseq(pseq)
-# Sparsity is how populated is the data with zeros
-kable(head(tax_table(pseq)))
-tax_table(pseq)
-
-# Check the data
-print(pseq)
-
-# Display variables from the mapping file
-sample_variables(pseq)
-
-# Total number of taxa in the entire data
-ntaxa(pseq)
-
-# Taxonomic ranks
-rank_names(pseq)
-
-```
-
-## Check OTUs/Phylums
-Prune(remove) taxa because we want to remove OTUs/taxa that aren't in any of the samples
-Don't need to prune this taxa because OTUs are found in all samples.
-Check this everytime you use the subset_samples function because of the pseq object being filtered. 
-```{r check OTUs}
-# Check if any OTUs are not present in any samples
-any(taxa_sums(pseq) == 0)
-# FALSE, therefore, OTUs are found in all samples and we don't need to prune the taxa.
-If the answer is TRUE, some OTUs are not present in any samples.
-This is usually the case when data is subset to remove some samples.
-OTUs unique to those sample are not removed along with the samples.
-Therefore, it is important to check this everytime the phyloseq object is filtered for samples using `subset_samples` function.
-
-head(meta(pseq))
-
-```
-
-## ALPHA DIVERSITY
-Diversity: Shannon, Simpson
-Richness: Chao, ACE
-
-## Prepare Data for Alpha Diversity Visualisation
-```{r prepare data for alpha diversity visualisation}
-
-# Prepare data for vizualisation
-# Alpha diversity for the entire data set
-png("alpha.png")
-tab <- diversities(pseq, index = "all")
-stargazer(tab, type = "text",summary = NULL)
-kable(head(tab))
-plot(tab)
-dev.off()
-
-# Get metadata
-pseq.meta <- meta(pseq)
-kable(head(pseq.meta))
-stargazer(pseq.meta, type = "text",summary = NULL)
-
-# Add the diversity table to metadata
-pseq.meta$Shannon <- tab$shannon 
-pseq.meta$InverseSimpson <- tab$inverse_simpson
-pseq.meta$fisher <- tab$fisher
-pseq.meta$chao1 <- tab$chao1
-
-```
-
-## Nationality Violin Plot
-```{r nationality violin plot}
-
-# Alpha diversity is the species richness of a place
-# High alpha diversity = high no. of species
-
-pseq.meta$nationality <- factor(pseq.meta$nationality, levels = c("Scandinavia","EasternEurope","CentralEurope", "SouthEurope", "UKIE", "US"))
-
-# Violin plot
-# the number in the palettes must equal the number of groups in nationlaity.pairs
-p1 <- ggviolin(pseq.meta, x = "nationality", y = "Shannon",
-      add = "boxplot", fill = "nationality", palette = "paired") +   theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-  print(p1)
-ggsave("nation_violin_shannon.png")
-
-p2 <- ggviolin(pseq.meta, x = "nationality", y = "InverseSimpson",
- add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-print(p2)
-ggsave("nation_violin_simpson.png")
-
-p3 <- ggviolin(pseq.meta, x = "nationality", y = "fisher",
- add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("") + ylab("Fisher")
-print(p3)
-ggsave("nation_violin_fisher.png")
-
-plot_grid(p1, p2, p3, align = "h" , ncol = 2, nrow = 2)
-ggsave("nation_violin.png")
-
-```
-
-```{r}
-
-p <- plot_richness(pseq, x="nationality", measures=c("Chao1", "Shannon", "Simpson")) 
-p + geom_boxplot()
-ggsave("alpha_nationality.png")
-
-```
-
-## MND Violin Plot
-```{r mnd violin plot}
-
-# Add the diversity table to metadata
-pseq.meta$Shannon <- tab$shannon
-pseq.meta$InverseSimpson <- tab$inverse_simpson
-
-# Compare differences in Shannon index between mnd group of the study subjects
-pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
-pseq.meta$pd <- factor(pseq.meta$pd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
-pseq.meta$ad <- factor(pseq.meta$ad, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
-# Create a list of pairwise comaprisons
-mnd <- levels(pseq.meta$mnd) # get the variables
-pd <- levels(pseq.meta$pd) # get the variables
-ad <- levels(pseq.meta$ad) # get the variables
-
-# Make a pairwise list that we want to compare.
-mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
-mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-pd.pairs <- combn(seq_along(pd), 2, simplify = FALSE, FUN = function(i)pd[i])
-pd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-ad.pairs <- combn(seq_along(ad), 2, simplify = FALSE, FUN = function(i)ad[i])
-ad.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-# Shannon
-# the number in the palettes must equal the number of groups in bmi.pairs
-p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
-p1 <- p1 + stat_compare_means(comparisons = mnd.pairs, paired = FALSE) + xlab("MND Incidence") + theme(legend.position="none")
-print(p1)
-ggsave("mnd_violin__shannon_stats.png")
-
-# Inverse Simpson
-p2 <- ggviolin(pseq.meta, x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
-p2 <- p2 + stat_compare_means(comparisons = mnd.pairs) 
-print(p2)
-ggsave("mnd_violin_simpson.png")
-
-```
-
-## BETA DIVERSITY
-Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
-WE want to: quantify community divergence within a given sample set
-Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.
-
-```{r Phylum NMDS}
-
-png("Phylum_NMDS.png")
-pseq.ord <- ordinate(pseq, "NMDS", "bray")
-p1 = plot_ordination(pseq, pseq.ord, type = "taxa", color = "mnd", title = "Taxa by Phylum")
-p1 + facet_wrap(~Phylum, 3)
-dev.off()
-
-```
-
- The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
-
-# VISUALISE MICORBIOME VARIATION
-```{r convert to compositional data}
-
-# Convert to compositional data
-pseq.rel <- microbiome::transform(pseq, "compositional")
-otu <- abundances(pseq.rel)
-meta <- meta(pseq.rel)
-
-```
-
-NMDS/Bray
-Bray-Curtis dissimilarity does not consider phylogenetic relationships between OTUs. There are several distance methods and a list can be obtained by typying ?distanceMethodList in the console pane.
-```{r}
-
-pseq.ord <- ordinate(pseq.rel, "NMDS", "bray")
-
-p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
-plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-ggsave("NMDStMND.png")
-
-p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
-ggsave("NMDS_facetMND.png")
-
-p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
-plot(p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
-ggsave("NMDStpd.png")
-
-p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, nrow = 2) + labs(color = "PD Prevalence")
-ggsave("NMDS_facetpd.png")
-
-p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
-plot(p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
-ggsave("NMDStad.png")
-
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, nrow = 2) + labs(color = "AD Prevalence")
-ggsave("NMDS_facetad.png")
-
-```
-
-MDS ordination on Bray-Curtis distance
-Performs principal coordinate analysis (also called principle coordinate decomposition, multidimensional scaling (MDS), or classical scaling) of a distance matrix (Gower 1966), including two correction methods for negative eigenvalues.
-See pcoa for further details.
-```{r}
-
-pseq.ord <- ordinate(pseq.rel, "MDS", "bray")
-
-p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
-plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-ggsave("MDS.png")
-
-p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
-ggsave("NMDS_facet.png")
-
-p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
-plot(p2 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
-ggsave("MDSpd.png")
-
-p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, 3) + labs(color = "PD Prevalence")
-ggsave("NMDS_facetPD.png")
-
-p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
-plot(p3 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
-ggsave("MDSad.png")
-
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, 3) + labs(color = "AD Prevalence")
-ggsave("NMDS_facetAD.png")
-
-```
-
-PCoA for compositional data with Bray-Curtis distances
-```{r}
-
-p <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
-p2 <- p + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b")
-
-p1 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
-print(p1)
-
-```
-
-```{r PCoA Bray}
-
-# Pick core taxa
-pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)
-
-# PCoA for compositional data with Bray-Curtis distances
-png("PCoA_bray.png")
-p <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
-                      method = "PCoA", distance = "bray") +
-       labs(title = paste("PCoA / Compositional / Bray-Curtis"))
-print(p)
-dev.off()
-
-```
-
-t-SNE is a popular new ordination technique
-```{r}
-
-tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
-#MND
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'mnd', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
-ggsave("mnd_t_sne.png")
-# PD
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'pd', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence")
-ggsave("pd_t_sne.png")
-
-# AD
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'ad', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence")
-ggsave("ad_t_sne.png")
-
-```
-
-Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
-This was applied in group-level comparisons for instance in Salonen et al.
-ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
-```{r Calculate group divergences}
-
-# Use these to compare microbiota divergence within each group
-
-#MND
-mnd.p.low <- subset_samples(pseq, mnd == "LOW")
-mnd.p.low.d <- divergence(mnd.p.low)
-
-mnd.p.med <- subset_samples(pseq, mnd == "MEDIUM")
-p.med.d <- divergence(mnd.p.med)
-
-mnd.p.high <- subset_samples(pseq, mnd == "HIGH")
-p.high.d <- divergence(mnd.p.high)
-
-mnd.p <- boxplot(list(Low = mnd.p.low.d, Medium = mnd.p.med.d, High = mnd.p.high.d))
-plot(mnd.p)
-
-# PD
-
-pd.p.low <- subset_samples(pseq, pd == "LOW")
-pd.p.low.d <- divergence(pd.p.low)
-
-pd.p.med <- subset_samples(pseq, pd == "MEDIUM")
-p.med.d <- divergence(pd.p.med)
-
-pd.p.high <- subset_samples(pseq, pd == "HIGH")
-p.high.d <- divergence(pd.p.high)
-
-pd.p <- boxplot(list(Low = pd.p.low.d, Medium = pd.p.med.d, High = pd.p.high.d))
-plot(pd.p)
-
-# AD
-
-ad.p.low <- subset_samples(pseq, ad == "LOW")
-ad.p.low.d <- divergence(ad.p.low)
-
-ad.p.med <- subset_samples(pseq, ad == "MEDIUM")
-p.med.d <- divergence(ad.p.med)
-
-ad.p.high <- subset_samples(pseq, ad == "HIGH")
-p.high.d <- divergence(ad.p.high)
-
-ad.p <- boxplot(list(Low = ad.p.low.d, Medium = ad.p.med.d, High = ad.p.high.d))
-plot(ad.p)
-
-```
-
-## PERMANOVA significance test for group-level differences
-Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
-
-```{r Redefine data}
-
-# Convert to compositional data
-# MND
-mnd.pseqLvH <- merge_phyloseq(mnd.p.low,mnd.p.high)
-mnd.pseqLvM <- merge_phyloseq(mnd.p.low,mnd.p.med)
-
-mnd.pseq.rel <- microbiome::transform(mnd.pseqLvH, "compositional")
-mnd.otu <- abundances(mnd.pseq.rel)
-mnd.meta <- meta(mnd.pseq.rel)
-
-# PD
-
-pd.pseqLvH <- merge_phyloseq(pd.p.low,pd.p.high)
-pd.pseqLvM <- merge_phyloseq(pd.p.low,pd.p.med)
-
-pd.pseq.rel <- microbiome::transform(pd.pseqLvH, "compositional")
-pd.otu <- abundances(pd.pseq.rel)
-pd.meta <- meta(pd.pseq.rel)
-
-# AD
-
-ad.pseqLvH <- merge_phyloseq(ad.p.low,ad.p.high)
-ad.pseqLvM <- merge_phyloseq(ad.p.low,ad.p.med)
-
-ad.pseq.rel <- microbiome::transform(ad.pseqLvH, "compositional")
-ad.otu <- abundances(ad.pseq.rel)
-ad.meta <- meta(ad.pseq.rel)
-
-```
-
-```{r PERMANOVA}
-
-# samples x species as input
-library(vegan)
-
-# MND
-mnd.permanova <- adonis(t(mnd.otu) ~ mnd,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(mnd.permanova$aov.tab)["mnd", "Pr(>F)"])
-summary(mnd.permanova$aov.tab)
-
-# PD
-pd.permanova <- adonis(t(pd.otu) ~ pd,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(pd.permanova$aov.tab)["pd", "Pr(>F)"])
-summary(pd.permanova$aov.tab)
-
-# AD
-ad.permanova <- adonis(t(ad.otu) ~ ad,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(ad.permanova$aov.tab)["ad", "Pr(>F)"])
-summary(ad.permanova$aov.tab)
-
-```
-
-Check that variance homogeneity assumptions hold (to ensure the reliability of the results):Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
-```{r ANOVA}
-#MND
-mnd.dist <- vegdist(t(mnd.otu))
-mnd.anov <- anova(betadisper(mnd.dist, meta$mnd))
-summary(mnd.anov)
-tidy(mnd.anov)
-
-#PD
-pd.dist <- vegdist(t(pd.otu))
-pd.anov <- anova(betadisper(pd.dist, meta$pd))
-summary(pd.anov)
-tidy(pd.anov)
-
-#AD
-ad.dist <- vegdist(t(ad.otu))
-ad.anov <- anova(betadisper(ad.dist, meta$ad))
-summary(ad.anov)
-tidy(ad.anov)
-
-```
-
-Show coefficients for the top taxa separating the groups
-```{r top taxa L vs H}
-
-coef <- coefficients(permanova)["mnd1",]
-top.coef <- coef[rev(order(abs(coef)))[1:20]]
-par(mar = c(3, 14, 2, 1))
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
-
-ggsave("LvH.png")
-
-```
-
-```{r L vs M}
-
-pseqLvM <- merge_phyloseq(p.low,p.med)
-
-pseq.rel <- microbiome::transform(pseqLvM, "compositional")
-otu <- abundances(pseq.rel)
-meta <- meta(pseq.rel)
-
-library(vegan)
-permanova <- adonis(t(otu) ~ mnd,
-               data = meta, permutations=999, method = "bray")
-
-# P-value
-print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
-
-dist <- vegdist(t(otu))
-anov <- anova(betadisper(dist, meta$mnd))
-summary(anov)
-tidy(anov)
-
-coef <- coefficients(permanova)["mnd1",]
-top.coef <- coef[rev(order(abs(coef)))[1:20]]
-par(mar = c(3, 14, 2, 1))
-
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
-
-ggsave("LvM.png", width = 20, height = 20, units = "cm")
-
-```
-
-## MICROBIOME COMPOSITION
-
-```{r Phylum #}
-
-# Show available ranks in the dataset
-rank_names(pseq)
-
-# Create table, number of features for each phyla
-table(tax_table(pseq)[, "Phylum"], exclude = NULL)
-
-```
-This shows a few phyla for which only one feature was observed. Those may be worth filtering.
-
-```{r Total Phylum abundnace bar}
-
-png("Phylum_abundance.png")
-p<- plot_bar(pseq,x="Phylum",fill = "Phylum") +geom_bar(stat="identity")
-
-phylumGlommed = tax_glom(pseq, "Phylum")
-plot_bar(phylumGlommed, x="Phylum", fill ="Phylum") +geom_bar(stat="identity")
-
-pd <- p$data %>% 
-    as_tibble %>%
-    mutate(Genus = as.character(Genus))
-genus_abun <- pd %>%
-    group_by(Genus) %>%
-    summarize(Abundance = sum(Abundance)) %>%
-    arrange(Abundance)
-genus_levels <- genus_abun$Genus
-pd0 <- pd %>%
-    mutate(Genus = factor(Genus, genus_levels))
-ggplot(pd0, aes(x = Sample, y = Abundance, fill = Genus)) + 
-    geom_bar(stat = "identity")
-dev.off()
-
-```
-
-```{r Phylum prevalence table}
-
-prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
-prev.otu + theme(legend.position="none")
-ggsave("phylum_prevalence.png")
-
-tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
-tab + theme(legend.position="none")
-ggsave("Phylum_prevalence_detection5.png")
-
-```
-
-theme(axis.text.x = element_text(angle = 90)
-+ guide_italics
-  ggtitle("Relative abundance")  + 
-```{r Relative abundance}
-
-pseq.ph <- aggregate_taxa(pseq.new, "Phylum")
-
-pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
-  theme(legend.position = "bottom") + theme_bw() + 
-  theme(axis.text.x=element_blank()) + 
-  theme(legend.title = element_text(size=18)) + scale_fill_brewer(palette = "Paired")
-print(plot.comp.rel)
-ggsave("phylum_relative_abundance11.png")
-
-plot.comp.rel + theme(legend.position="none")
-ggsave("phylum_relative_abundance12.png")
-
-```
-
-guide_italics
-ggtitle("Relative abundance") +  
-```{r}
-
-clos.ph <- aggregate_taxa(clostridium, "Phylum")
-
-clos.ph.rel <- microbiome::transform(clos.ph, "compositional")
-
-plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
-  theme(legend.position = "bottom", ) + theme_bw() + 
-  theme(axis.text.x=element_blank()) + 
-  theme(legend.title = element_text(size=18))
-plot.comp.rel + scale_fill_brewer(palette = "Paired")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
- theme(legend.position = "bottom") + theme_bw() +
- theme(axis.text.x=element_blank()) +
- theme(legend.title = element_text(size=18),legend.text = element_text( face="italic") ) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("my awesome title"))
-
-ggsave("phylum_relative_abundance_clos.png")
-
-plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none")
-ggsave("phylum_relative_abundance_clos2.png")
-
-```
-
-```{r Phylum Bar Plot}
-
-levels(pseq.meta$mnd)
-
-pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
-pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot")
-
-plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("phylum_abundance_bar.png")
-
-```
-
-```{r pseq.new bar plot}
-
-p.new <- aggregate_taxa(pseq.new, "Phylum")
-
-p.new.rel <- microbiome::transform(p.new, "compositional")
-
-plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.p.new.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("pnew_abundance_bar.png")
-
-```
-
-```{r CLostridium Bar Plot}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.cl.rel <- microbiome::transform(clos, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("cl_abundance_bar.png")
-
-```
-
-```{r Clostridium genus bar plots}
-
-pseq.cl.rel <- microbiome::transform(clostridium, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")
-
-```
-
-```{r Top Phylum Heatmap}
-
-pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
-pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
-ggsave("phylum_histogram.png")
-
-```
-
-```{r p.new heatmap}
-
-pseq.n.ph <- aggregate_taxa(pseq.new, "Phylum")
-
-pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
-
-plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-p <- plot.n.comp.rel + coord_flip()
-p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-ggsave("pnew_histogram.png")
-
-```
-
-```{r clostridium heatmap}
-
-pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")
-
-plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-p <- plot.c.comp.rel + coord_flip()
-p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-ggsave("clos_histogram.png")
-
-```
-
-```{r Network}
-
-png("network_bray_phylum.png")
-jg = make_network(pseq, "taxa", "bray", 0.3)
-plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
-dev.off()
-
-```
-
-## Community composition plotting
-Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results.
-These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted.
-Interpretation of these plots should always be subject to subsequent statistical analysis.
-```{r community-composition-plots}
-# Create a data table for ggplot
-# agglomerate at phylum level
-pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
-
-# Transform to rel. abundance (or use ps0.ra)
-pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
-
-# Melt to long format for easy ggploting
-pseq_phylum <- psmelt(pseq_phyl)
-
-     # filter(Abundance > 0.01)                             # Filter out low abundance taxa
-
-# Plot - Phylum
-p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-  geom_bar(stat = "identity", width = 1) +
-  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-  theme(axis.text.x = element_blank()) +
-  theme(axis.title.x = element_blank()) +
-  labs(title = "Abundant Phylum (> 1%)")
-
-p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-plot(p)
-ggsave("phylum_abun_mnd.png")
-
-```
-
-```{r}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-# agglomerate at phylum level
-clos_phy <- tax_glom(clos, taxrank = "Phylum")
-
-# Transform to rel. abundance (or use ps0.ra)
-clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})
-
-# Melt to long format for easy ggploting
-clos_phylum <- psmelt(clos_phyl)
-
-# Plot - Phylum
-p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-  geom_bar(stat = "identity", width = 1) +
-  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-  theme(axis.text.x = element_blank()) +
-  theme(axis.title.x = element_blank()) +
-  labs(title = "Abundant Phylum of Clostridium(> 1%)")
-
-p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-plot(p)
-ggsave("phylum_abun_mnd_clos.png")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Phylum")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Genus")
-
-```
-
-```{r}
-
-get_taxa_unique(clostridium, "Genus")
-
-```
-
-```{r}
-
-df <- psmelt(pseq)
-kable(head(df))
-
-```
-
-```{r}
-
-head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])
-
-```

### 2019-06-25 import_popln_data.Rmd
updated

@@
 mb.gather <- function (x) {
+    # Creates an empty data frame
     mb.group <- data.frame("Location" = c("Central Europe","Eastern Europe","Scandinavia", "Southern Europe","UKIE","USA"), "Mean" = c(1:6)) 
+     # groups the countries to match the microbiome classification 
   c.europe <- filter(x,Location=="Belgium"|Location=="Denmark"|Location=="Germany"|Location=="The Netherlands")
@@
     usa <- filter(mnd.incidence,Location=="United States")
+    # gmeans the data and put it into the data frame
     mb.group[1,2] <- mean(c.europe$Value)

### 2019-06-25 import_popln_data.Rmd
updated

@@
 
-setwd("~/globalmicrobiome")
-mnd.incidence <- read_csv("mnd.csv",col_names = TRUE)
-mnd.incidence <- as.data.frame(mnd.incidence)
-
 mb.gather <- function (x) {
@@
 
+setwd("~/globalmicrobiome")
+# files nwed to be in the owrking directory
+mnd.incidence <- read_csv("MND-De-B.csv",col_names = TRUE)
+mnd.incidence <- as.data.frame(mnd.incidence)
 temp <- mb.gather(mnd.incidence)

### 2019-06-25 import_popln_data.Rmd
updated

--- 2019-06-25 import_popln_data.Rmd
@@
 
-setwd("~/globalmicrobiome")
-mnd.incidence <- read_csv("mnd.csv",col_names = TRUE)
-mnd.incidence <- as.data.frame(mnd.incidence)
-
 mb.gather <- function (x) {
+    # Creates an empty data frame
     mb.group <- data.frame("Location" = c("Central Europe","Eastern Europe","Scandinavia", "Southern Europe","UKIE","USA"), "Mean" = c(1:6)) 
+     # groups the countries to match the microbiome classification 
   c.europe <- filter(x,Location=="Belgium"|Location=="Denmark"|Location=="Germany"|Location=="The Netherlands")
-    e.europe <- filter(mnd.incidence,Location=="Poland")
-    scan <- filter(mnd.incidence,Location=="Finland"|Location=="Norway"|Location=="Sweden")
-    s.europe <- filter(mnd.incidence,Location=="France"|Location=="Italy"|Location=="Serbia"|Location=="Spain")
-    ukie <- filter(mnd.incidence,Location=="United Kingdom"|Location=="Ireland")
-    usa <- filter(mnd.incidence,Location=="United States")
+    e.europe <- filter(x,Location=="Poland")
+    scan <- filter(x,Location=="Finland"|Location=="Norway"|Location=="Sweden")
+    s.europe <- filter(x,Location=="France"|Location=="Italy"|Location=="Serbia"|Location=="Spain")
+    ukie <- filter(x,Location=="United Kingdom"|Location=="Ireland")
+    usa <- filter(x,Location=="United States")
+    # gmeans the data and put it into the data frame
     mb.group[1,2] <- mean(c.europe$Value)
@@
 
-temp <- mb.gather(mnd.incidence)
+setwd("~/globalmicrobiome")
+# files nwed to be in the owrking directory
+mnd.incidence <- as.data.frame(read_csv("MND-I-B.csv",col_names = TRUE))
+mnd.prev <- as.data.frame(read_csv("MND-P-B.csv",col_names = TRUE))
+mnd.incidence <- mb.gather(mnd.incidence)
+mnd.prev <- mb.gather(mnd.prev)
 

### 2019-06-26 import_popln_data.Rmd
updated

@@
 # files nwed to be in the owrking directory
+
+# For MND
 mnd.incidence <- as.data.frame(read_csv("MND-I-B.csv",col_names = TRUE))
 mnd.prev <- as.data.frame(read_csv("MND-P-B.csv",col_names = TRUE))
+
+# Run function and change name
 mnd.incidence <- mb.gather(mnd.incidence)
+colnames(mnd.incidence)[2] <- "incidence"
 mnd.prev <- mb.gather(mnd.prev)
+colnames(mnd.prev)[2] <- "prevalance"
+
+mnd.ip <- right_join(mnd.prev, mnd.incidence, by = "Location")
+
+# For IPD
+ipd.incidence <- as.data.frame(read_csv("PD-I-B.csv",col_names = TRUE))
+ipd.prev <- as.data.frame(read_csv("PD-P-B.csv",col_names = TRUE))
+
+# Run function and change name
+ipd.incidence <- mb.gather(ipd.incidence)
+colnames(ipd.incidence)[2] <- "incidence"
+ipd.prev <- mb.gather(ipd.prev)
+colnames(ipd.prev)[2] <- "prevalance"
+
+ipd.ip <- right_join(ipd.prev, ipd.incidence, by = "Location")
+
+# For AD
+ad.incidence <- as.data.frame(read_csv("AD-I-B.csv",col_names = TRUE))
+ad.prev <- as.data.frame(read_csv("AD-P-B.csv",col_names = TRUE))
+
+# Run function and change name
+ad.incidence <- mb.gather(ad.incidence)
+colnames(ad.incidence)[2] <- "incidence"
+ad.prev <- mb.gather(ad.prev)
+colnames(ad.prev)[2] <- "prevalance"
+
+ad.ip <- right_join(ad.prev, ad.incidence, by = "Location")
 

### 2019-06-26 import_popln_data.Rmd
updated

--- 2019-06-25 import_popln_data.Rmd
@@
 # files nwed to be in the owrking directory
-mnd.incidence <- as.data.frame(read_csv("MND-I-B.csv",col_names = TRUE))
-mnd.prev <- as.data.frame(read_csv("MND-P-B.csv",col_names = TRUE))
+
+# For MND
+mnd.incidence <- as.data.frame(read_csv("./rawdata/MND-I-B.csv",col_names = TRUE))
+mnd.prev <- as.data.frame(read_csv("./rawdata/MND-P-B.csv",col_names = TRUE))
+
+mnd.roc.i <- as.data.frame(read_csv("./rawdata/MND-I-B-RoC.csv",col_names = TRUE))
+
+mnd.roc.p <- as.data.frame(read_csv("./rawdata/MND-P-B-RoC.csv",col_names = TRUE))
+
+# Run function and change name
 mnd.incidence <- mb.gather(mnd.incidence)
+colnames(mnd.incidence)[2] <- "incidence"
+
+mnd.roc.i <- mb.gather(mnd.roc.i)
+colnames(mnd.roc.i)[2] <- "ROC incidence"
+
 mnd.prev <- mb.gather(mnd.prev)
+colnames(mnd.prev)[2] <- "prevalance"
+
+mnd.roc.p <- mb.gather(mnd.roc.p)
+colnames(mnd.roc.p)[2] <- "ROC prevalance"
+# Join them together. 
+mnd.ip <- right_join(mnd.prev, mnd.roc.p, by = "Location")
+mnd.ip <- right_join(mnd.ip, mnd.incidence, by = "Location")
+mnd.ip <- right_join(mnd.ip, mnd.roc.i, by = "Location")
+
+# For IPD
+
+ipd.incidence <- as.data.frame(read_csv("./rawdata/PD-I-B.csv",col_names = TRUE))
+ipd.prev <- as.data.frame(read_csv("./rawdata/PD-P-B.csv",col_names = TRUE))
+
+ipd.roc.i <- as.data.frame(read_csv("./rawdata/PD-I-B-RoC.csv",col_names = TRUE))
+
+ipd.roc.p <- as.data.frame(read_csv("./rawdata/PD-P-B-RoC.csv",col_names = TRUE))
+
+# Run function and change name
+ipd.incidence <- mb.gather(ipd.incidence)
+colnames(ipd.incidence)[2] <- "incidence"
+
+ipd.roc.i <- mb.gather(ipd.roc.i)
+colnames(ipd.roc.i)[2] <- "ROC incidence"
+
+ipd.prev <- mb.gather(ipd.prev)
+colnames(ipd.prev)[2] <- "prevalance"
+
+ipd.roc.p <- mb.gather(ipd.roc.p)
+colnames(ipd.roc.p)[2] <- "ROC prevalance"
+# Join them together. 
+ipd.ip <- right_join(ipd.prev, ipd.roc.p, by = "Location")
+ipd.ip <- right_join(ipd.ip, ipd.incidence, by = "Location")
+ipd.ip <- right_join(ipd.ip, ipd.roc.i, by = "Location")
+
+# For AD
+ad.incidence <- as.data.frame(read_csv("./rawdata/AD-I-B.csv",col_names = TRUE))
+ad.prev <- as.data.frame(read_csv("./rawdata/AD-P-B.csv",col_names = TRUE))
+
+ad.roc.i <- as.data.frame(read_csv("./rawdata/AD-I-B-RoC.csv",col_names = TRUE))
+
+ad.roc.p <- as.data.frame(read_csv("./rawdata/AD-P-B-RoC.csv",col_names = TRUE))
+
+# Run function and change name
+ad.incidence <- mb.gather(ad.incidence)
+colnames(ad.incidence)[2] <- "incidence"
+
+ad.roc.i <- mb.gather(ad.roc.i)
+colnames(ad.roc.i)[2] <- "ROC incidence"
+
+ad.prev <- mb.gather(ad.prev)
+colnames(ad.prev)[2] <- "prevalance"
+
+ad.roc.p <- mb.gather(ad.roc.p)
+colnames(ad.roc.p)[2] <- "ROC prevalance"
+# Join them together. 
+ad.ip <- right_join(ad.prev, ad.roc.p, by = "Location")
+ad.ip <- right_join(ad.ip, ad.incidence, by = "Location")
+ad.ip <- right_join(ad.ip, ad.roc.i, by = "Location")
+
+plot(ad.ip$incidence,ad.ip$prevalance)
+plot(ipd.ip$incidence,ipd.ip$prevalance)
+plot(mnd.ip$incidence,mnd.ip$prevalance)
 

### 2019-06-26 import_popln_data.Rmd
updated

--- 2019-06-25 import_popln_data.Rmd
@@
 # files nwed to be in the owrking directory
-mnd.incidence <- as.data.frame(read_csv("MND-I-B.csv",col_names = TRUE))
-mnd.prev <- as.data.frame(read_csv("MND-P-B.csv",col_names = TRUE))
+
+# For MND
+mnd.incidence <- as.data.frame(read_csv("./rawdata/MND-I-B.csv",col_names = TRUE))
+
+mnd.i.12 <- as.data.frame(read_csv("./rawdata/MND-I-B-12.csv",col_names = TRUE))
+mnd.i.13 <- as.data.frame(read_csv("./rawdata/MND-I-B-13.csv",col_names = TRUE))
+mnd.i.14 <- as.data.frame(read_csv("./rawdata/MND-I-B-14.csv",col_names = TRUE))
+mnd.i.15 <- as.data.frame(read_csv("./rawdata/MND-I-B-15.csv",col_names = TRUE))
+mnd.i.16 <- as.data.frame(read_csv("./rawdata/MND-I-B-16.csv",col_names = TRUE))
+
+mnd.prev <- as.data.frame(read_csv("./rawdata/MND-P-B.csv",col_names = TRUE))
+mnd.p.12 <- as.data.frame(read_csv("./rawdata/MND-P-B-12.csv",col_names = TRUE))
+mnd.p.13 <- as.data.frame(read_csv("./rawdata/MND-P-B-13.csv",col_names = TRUE))
+mnd.p.14 <- as.data.frame(read_csv("./rawdata/MND-P-B-14.csv",col_names = TRUE))
+mnd.p.15 <- as.data.frame(read_csv("./rawdata/MND-P-B-15.csv",col_names = TRUE))
+mnd.p.16 <- as.data.frame(read_csv("./rawdata/MND-P-B-16.csv",col_names = TRUE))
+
+mnd.roc.i <- as.data.frame(read_csv("./rawdata/MND-I-B-RoC.csv",col_names = TRUE))
+
+mnd.roc.p <- as.data.frame(read_csv("./rawdata/MND-P-B-RoC.csv",col_names = TRUE))
+
+# Run function and change name
 mnd.incidence <- mb.gather(mnd.incidence)
+colnames(mnd.incidence)[2] <- "incidence 2017"
+
+mnd.i.12 <- mb.gather(mnd.i.12)
+colnames(mnd.i.12)[2] <- "incidence 2012"
+mnd.i.13 <- mb.gather(mnd.i.13)
+colnames(mnd.i.13)[2] <- "incidence 2013"
+mnd.i.14 <- mb.gather(mnd.i.14)
+colnames(mnd.i.14)[2] <- "incidence 2014"
+mnd.i.15 <- mb.gather(mnd.i.15)
+colnames(mnd.i.15)[2] <- "incidence 2015"
+mnd.i.16 <- mb.gather(mnd.i.16)
+colnames(mnd.i.16)[2] <- "incidence 2016"
+mnd.roc.i <- mb.gather(mnd.roc.i)
+colnames(mnd.roc.i)[2] <- "ROC incidence"
+
 mnd.prev <- mb.gather(mnd.prev)
+colnames(mnd.prev)[2] <- "prevalance 2017"
+mnd.p.12 <- mb.gather(mnd.p.12)
+colnames(mnd.p.12)[2] <- "prevalance 2012"
+mnd.p.13 <- mb.gather(mnd.p.13)
+colnames(mnd.p.13)[2] <- "prevalance 2013"
+mnd.p.14 <- mb.gather(mnd.p.14)
+colnames(mnd.p.14)[2] <- "prevalance 2014"
+mnd.p.15 <- mb.gather(mnd.p.15)
+colnames(mnd.p.15)[2] <- "prevalance 2015"
+mnd.p.16 <- mb.gather(mnd.p.16)
+colnames(mnd.p.16)[2] <- "prevalance 2016"
+mnd.roc.p <- mb.gather(mnd.roc.p)
+colnames(mnd.roc.p)[2] <- "ROC prevalance"
+
+# Join them together.
+
+mnd.i <- list( mnd.i.12, mnd.i.13,mnd.i.14,mnd.i.15,mnd.i.16,mnd.incidence,mnd.roc.i) %>% reduce(left_join, by = "Location")
+
+mnd.p <- list( mnd.p.12, mnd.p.13,mnd.p.14,mnd.p.15,mnd.p.16,mnd.incidence,mnd.roc.p) %>% reduce(left_join, by = "Location")
+
+# For IPD
+
+ipd.incidence <- as.data.frame(read_csv("./rawdata/PD-I-B.csv",col_names = TRUE))
+
+ipd.i.12 <- as.data.frame(read_csv("./rawdata/PD-I-B-12.csv",col_names = TRUE))
+ipd.i.13 <- as.data.frame(read_csv("./rawdata/PD-I-B-13.csv",col_names = TRUE))
+ipd.i.14 <- as.data.frame(read_csv("./rawdata/PD-I-B-14.csv",col_names = TRUE))
+ipd.i.15 <- as.data.frame(read_csv("./rawdata/PD-I-B-15.csv",col_names = TRUE))
+ipd.i.16 <- as.data.frame(read_csv("./rawdata/PD-I-B-16.csv",col_names = TRUE))
+
+ipd.prev <- as.data.frame(read_csv("./rawdata/PD-P-B.csv",col_names = TRUE))
+ipd.p.12 <- as.data.frame(read_csv("./rawdata/PD-P-B-12.csv",col_names = TRUE))
+ipd.p.13 <- as.data.frame(read_csv("./rawdata/PD-P-B-13.csv",col_names = TRUE))
+ipd.p.14 <- as.data.frame(read_csv("./rawdata/PD-P-B-14.csv",col_names = TRUE))
+ipd.p.15 <- as.data.frame(read_csv("./rawdata/PD-P-B-15.csv",col_names = TRUE))
+ipd.p.16 <- as.data.frame(read_csv("./rawdata/PD-P-B-16.csv",col_names = TRUE))
+
+ipd.roc.i <- as.data.frame(read_csv("./rawdata/PD-I-B-RoC.csv",col_names = TRUE))
+
+ipd.roc.p <- as.data.frame(read_csv("./rawdata/PD-P-B-RoC.csv",col_names = TRUE))
+
+# Run function and change name
+ipd.incidence <- mb.gather(ipd.incidence)
+colnames(ipd.incidence)[2] <- "incidence 2017"
+
+ipd.i.12 <- mb.gather(ipd.i.12)
+colnames(ipd.i.12)[2] <- "incidence 2012"
+ipd.i.13 <- mb.gather(ipd.i.13)
+colnames(ipd.i.13)[2] <- "incidence 2013"
+ipd.i.14 <- mb.gather(ipd.i.14)
+colnames(ipd.i.14)[2] <- "incidence 2014"
+ipd.i.15 <- mb.gather(ipd.i.15)
+colnames(ipd.i.15)[2] <- "incidence 2015"
+ipd.i.16 <- mb.gather(ipd.i.16)
+colnames(ipd.i.16)[2] <- "incidence 2016"
+ipd.roc.i <- mb.gather(ipd.roc.i)
+colnames(ipd.roc.i)[2] <- "ROC incidence"
+
+ipd.prev <- mb.gather(ipd.prev)
+colnames(ipd.prev)[2] <- "prevalance 2017"
+ipd.p.12 <- mb.gather(ipd.p.12)
+colnames(ipd.p.12)[2] <- "prevalance 2012"
+ipd.p.13 <- mb.gather(ipd.p.13)
+colnames(ipd.p.13)[2] <- "prevalance 2013"
+ipd.p.14 <- mb.gather(ipd.p.14)
+colnames(ipd.p.14)[2] <- "prevalance 2014"
+ipd.p.15 <- mb.gather(ipd.p.15)
+colnames(ipd.p.15)[2] <- "prevalance 2015"
+ipd.p.16 <- mb.gather(ipd.p.16)
+colnames(ipd.p.16)[2] <- "prevalance 2016"
+ipd.roc.p <- mb.gather(ipd.roc.p)
+colnames(ipd.roc.p)[2] <- "ROC prevalance"
+
+# Join them together.
+
+ipd.i <- list( ipd.i.12, ipd.i.13,ipd.i.14,ipd.i.15,ipd.i.16,ipd.incidence,ipd.roc.i) %>% reduce(left_join, by = "Location")
+
+ipd.p <- list( ipd.p.12, ipd.p.13,ipd.p.14,ipd.p.15,ipd.p.16,ipd.incidence,ipd.roc.p) %>% reduce(left_join, by = "Location")
+
+# For AD
+
+ad.incidence <- as.data.frame(read_csv("./rawdata/AD-I-B.csv",col_names = TRUE))
+
+ad.i.12 <- as.data.frame(read_csv("./rawdata/AD-I-B-12.csv",col_names = TRUE))
+ad.i.13 <- as.data.frame(read_csv("./rawdata/AD-I-B-13.csv",col_names = TRUE))
+ad.i.14 <- as.data.frame(read_csv("./rawdata/AD-I-B-14.csv",col_names = TRUE))
+ad.i.15 <- as.data.frame(read_csv("./rawdata/AD-I-B-15.csv",col_names = TRUE))
+ad.i.16 <- as.data.frame(read_csv("./rawdata/AD-I-B-16.csv",col_names = TRUE))
+
+ad.prev <- as.data.frame(read_csv("./rawdata/AD-P-B.csv",col_names = TRUE))
+ad.p.12 <- as.data.frame(read_csv("./rawdata/AD-P-B-12.csv",col_names = TRUE))
+ad.p.13 <- as.data.frame(read_csv("./rawdata/AD-P-B-13.csv",col_names = TRUE))
+ad.p.14 <- as.data.frame(read_csv("./rawdata/AD-P-B-14.csv",col_names = TRUE))
+ad.p.15 <- as.data.frame(read_csv("./rawdata/AD-P-B-15.csv",col_names = TRUE))
+ad.p.16 <- as.data.frame(read_csv("./rawdata/AD-P-B-16.csv",col_names = TRUE))
+
+ad.roc.i <- as.data.frame(read_csv("./rawdata/AD-I-B-RoC.csv",col_names = TRUE))
+
+ad.roc.p <- as.data.frame(read_csv("./rawdata/AD-P-B-RoC.csv",col_names = TRUE))
+
+# Run function and change name
+ad.incidence <- mb.gather(ad.incidence)
+colnames(ad.incidence)[2] <- "incidence 2017"
+
+ad.i.12 <- mb.gather(ad.i.12)
+colnames(ad.i.12)[2] <- "incidence 2012"
+ad.i.13 <- mb.gather(ad.i.13)
+colnames(ad.i.13)[2] <- "incidence 2013"
+ad.i.14 <- mb.gather(ad.i.14)
+colnames(ad.i.14)[2] <- "incidence 2014"
+ad.i.15 <- mb.gather(ad.i.15)
+colnames(ad.i.15)[2] <- "incidence 2015"
+ad.i.16 <- mb.gather(ad.i.16)
+colnames(ad.i.16)[2] <- "incidence 2016"
+ad.roc.i <- mb.gather(ad.roc.i)
+colnames(ad.roc.i)[2] <- "ROC incidence"
+
+ad.prev <- mb.gather(ad.prev)
+colnames(ad.prev)[2] <- "prevalance 2017"
+ad.p.12 <- mb.gather(ad.p.12)
+colnames(ad.p.12)[2] <- "prevalance 2012"
+ad.p.13 <- mb.gather(ad.p.13)
+colnames(ad.p.13)[2] <- "prevalance 2013"
+ad.p.14 <- mb.gather(ad.p.14)
+colnames(ad.p.14)[2] <- "prevalance 2014"
+ad.p.15 <- mb.gather(ad.p.15)
+colnames(ad.p.15)[2] <- "prevalance 2015"
+ad.p.16 <- mb.gather(ad.p.16)
+colnames(ad.p.16)[2] <- "prevalance 2016"
+ad.roc.p <- mb.gather(ad.roc.p)
+colnames(ad.roc.p)[2] <- "ROC prevalance"
+
+# Join them together.
+
+ad.i <- list( ad.i.12, ad.i.13,ad.i.14,ad.i.15,ad.i.16,ad.incidence,ad.roc.i) %>% reduce(left_join, by = "Location")
+
+ad.p <- list( ad.p.12, ad.p.13,ad.p.14,ad.p.15,ad.p.16,ad.incidence,ad.roc.p) %>% reduce(left_join, by = "Location")
 

### 2019-06-28 import_popln_data.Rmd
updated

@@
 mnd.incidence <- mb.gather(mnd.incidence)
-colnames(mnd.incidence)[2] <- "incidence 2017"
+colnames(mnd.incidence)[2] <- 2017
 
 mnd.i.12 <- mb.gather(mnd.i.12)
-colnames(mnd.i.12)[2] <- "incidence 2012"
+colnames(mnd.i.12)[2] <- 2012
 mnd.i.13 <- mb.gather(mnd.i.13)
-colnames(mnd.i.13)[2] <- "incidence 2013"
+colnames(mnd.i.13)[2] <- 2013
 mnd.i.14 <- mb.gather(mnd.i.14)
-colnames(mnd.i.14)[2] <- "incidence 2014"
+colnames(mnd.i.14)[2] <- 2014
 mnd.i.15 <- mb.gather(mnd.i.15)
-colnames(mnd.i.15)[2] <- "incidence 2015"
+colnames(mnd.i.15)[2] <- 2015
 mnd.i.16 <- mb.gather(mnd.i.16)
-colnames(mnd.i.16)[2] <- "incidence 2016"
-mnd.roc.i <- mb.gather(mnd.roc.i)
-colnames(mnd.roc.i)[2] <- "ROC incidence"
+colnames(mnd.i.16)[2] <- 2016
+#mnd.roc.i <- mb.gather(mnd.roc.i)
+#colnames(mnd.roc.i)[2] <- "ROC incidence"
 
 mnd.prev <- mb.gather(mnd.prev)
-colnames(mnd.prev)[2] <- "prevalance 2017"
+colnames(mnd.prev)[2] <- 2017
 mnd.p.12 <- mb.gather(mnd.p.12)
-colnames(mnd.p.12)[2] <- "prevalance 2012"
+colnames(mnd.p.12)[2] <- 2012
 mnd.p.13 <- mb.gather(mnd.p.13)
-colnames(mnd.p.13)[2] <- "prevalance 2013"
+colnames(mnd.p.13)[2] <- 2013
 mnd.p.14 <- mb.gather(mnd.p.14)
-colnames(mnd.p.14)[2] <- "prevalance 2014"
+colnames(mnd.p.14)[2] <- 2014
 mnd.p.15 <- mb.gather(mnd.p.15)
-colnames(mnd.p.15)[2] <- "prevalance 2015"
+colnames(mnd.p.15)[2] <- 2015
 mnd.p.16 <- mb.gather(mnd.p.16)
-colnames(mnd.p.16)[2] <- "prevalance 2016"
-mnd.roc.p <- mb.gather(mnd.roc.p)
-colnames(mnd.roc.p)[2] <- "ROC prevalance"
+colnames(mnd.p.16)[2] <- 2016
+#mnd.roc.p <- mb.gather(mnd.roc.p)
+#colnames(mnd.roc.p)[2] <- "ROC prevalance"
 
@@
 
-mnd.i <- list( mnd.i.12, mnd.i.13,mnd.i.14,mnd.i.15,mnd.i.16,mnd.incidence,mnd.roc.i) %>% reduce(left_join, by = "Location")
+mnd.i <- list( mnd.i.12, mnd.i.13,mnd.i.14,mnd.i.15,mnd.i.16,mnd.incidence) %>% reduce(left_join, by = "Location")
 
-mnd.p <- list( mnd.p.12, mnd.p.13,mnd.p.14,mnd.p.15,mnd.p.16,mnd.incidence,mnd.roc.p) %>% reduce(left_join, by = "Location")
+mnd.p <- list( mnd.p.12, mnd.p.13,mnd.p.14,mnd.p.15,mnd.p.16,mnd.incidence) %>% reduce(left_join, by = "Location")
 
@@
 ipd.incidence <- mb.gather(ipd.incidence)
-colnames(ipd.incidence)[2] <- "incidence 2017"
+colnames(ipd.incidence)[2] <- 2017
 
 ipd.i.12 <- mb.gather(ipd.i.12)
-colnames(ipd.i.12)[2] <- "incidence 2012"
+colnames(ipd.i.12)[2] <- 2012
 ipd.i.13 <- mb.gather(ipd.i.13)
-colnames(ipd.i.13)[2] <- "incidence 2013"
+colnames(ipd.i.13)[2] <- 2013
 ipd.i.14 <- mb.gather(ipd.i.14)
-colnames(ipd.i.14)[2] <- "incidence 2014"
+colnames(ipd.i.14)[2] <- 2014
 ipd.i.15 <- mb.gather(ipd.i.15)
-colnames(ipd.i.15)[2] <- "incidence 2015"
+colnames(ipd.i.15)[2] <- 2015
 ipd.i.16 <- mb.gather(ipd.i.16)
-colnames(ipd.i.16)[2] <- "incidence 2016"
-ipd.roc.i <- mb.gather(ipd.roc.i)
-colnames(ipd.roc.i)[2] <- "ROC incidence"
+colnames(ipd.i.16)[2] <- 2016
+#ipd.roc.i <- mb.gather(ipd.roc.i)
+#colnames(ipd.roc.i)[2] <- "ROC incidence"
 
 ipd.prev <- mb.gather(ipd.prev)
-colnames(ipd.prev)[2] <- "prevalance 2017"
+colnames(ipd.prev)[2] <- 2017
 ipd.p.12 <- mb.gather(ipd.p.12)
-colnames(ipd.p.12)[2] <- "prevalance 2012"
+colnames(ipd.p.12)[2] <- 2012
 ipd.p.13 <- mb.gather(ipd.p.13)
-colnames(ipd.p.13)[2] <- "prevalance 2013"
+colnames(ipd.p.13)[2] <- 2013
 ipd.p.14 <- mb.gather(ipd.p.14)
-colnames(ipd.p.14)[2] <- "prevalance 2014"
+colnames(ipd.p.14)[2] <- 2014
 ipd.p.15 <- mb.gather(ipd.p.15)
-colnames(ipd.p.15)[2] <- "prevalance 2015"
+colnames(ipd.p.15)[2] <- 2015
 ipd.p.16 <- mb.gather(ipd.p.16)
-colnames(ipd.p.16)[2] <- "prevalance 2016"
-ipd.roc.p <- mb.gather(ipd.roc.p)
-colnames(ipd.roc.p)[2] <- "ROC prevalance"
+colnames(ipd.p.16)[2] <- 2016
+#ipd.roc.p <- mb.gather(ipd.roc.p)
+#colnames(ipd.roc.p)[2] <- "ROC prevalance"
 
@@
 
-ipd.i <- list( ipd.i.12, ipd.i.13,ipd.i.14,ipd.i.15,ipd.i.16,ipd.incidence,ipd.roc.i) %>% reduce(left_join, by = "Location")
+ipd.i <- list( ipd.i.12, ipd.i.13,ipd.i.14,ipd.i.15,ipd.i.16,ipd.incidence) %>% reduce(left_join, by = "Location")
 
-ipd.p <- list( ipd.p.12, ipd.p.13,ipd.p.14,ipd.p.15,ipd.p.16,ipd.incidence,ipd.roc.p) %>% reduce(left_join, by = "Location")
+ipd.p <- list( ipd.p.12, ipd.p.13,ipd.p.14,ipd.p.15,ipd.p.16,ipd.incidence) %>% reduce(left_join, by = "Location")
 
@@
 ad.incidence <- mb.gather(ad.incidence)
-colnames(ad.incidence)[2] <- "incidence 2017"
+colnames(ad.incidence)[2] <- 2017
 
 ad.i.12 <- mb.gather(ad.i.12)
-colnames(ad.i.12)[2] <- "incidence 2012"
+colnames(ad.i.12)[2] <- 2012
 ad.i.13 <- mb.gather(ad.i.13)
-colnames(ad.i.13)[2] <- "incidence 2013"
+colnames(ad.i.13)[2] <- 2013
 ad.i.14 <- mb.gather(ad.i.14)
-colnames(ad.i.14)[2] <- "incidence 2014"
+colnames(ad.i.14)[2] <- 2014
 ad.i.15 <- mb.gather(ad.i.15)
-colnames(ad.i.15)[2] <- "incidence 2015"
+colnames(ad.i.15)[2] <- 2015
 ad.i.16 <- mb.gather(ad.i.16)
-colnames(ad.i.16)[2] <- "incidence 2016"
-ad.roc.i <- mb.gather(ad.roc.i)
-colnames(ad.roc.i)[2] <- "ROC incidence"
+colnames(ad.i.16)[2] <- 2016
+#ad.roc.i <- mb.gather(ad.roc.i)
+#colnames(ad.roc.i)[2] <- "ROC incidence"
 
 ad.prev <- mb.gather(ad.prev)
-colnames(ad.prev)[2] <- "prevalance 2017"
+colnames(ad.prev)[2] <- 2017
 ad.p.12 <- mb.gather(ad.p.12)
-colnames(ad.p.12)[2] <- "prevalance 2012"
+colnames(ad.p.12)[2] <- 2012
 ad.p.13 <- mb.gather(ad.p.13)
-colnames(ad.p.13)[2] <- "prevalance 2013"
+colnames(ad.p.13)[2] <- 2013
 ad.p.14 <- mb.gather(ad.p.14)
-colnames(ad.p.14)[2] <- "prevalance 2014"
+colnames(ad.p.14)[2] <- 2014
 ad.p.15 <- mb.gather(ad.p.15)
-colnames(ad.p.15)[2] <- "prevalance 2015"
+colnames(ad.p.15)[2] <- 2015
 ad.p.16 <- mb.gather(ad.p.16)
-colnames(ad.p.16)[2] <- "prevalance 2016"
-ad.roc.p <- mb.gather(ad.roc.p)
-colnames(ad.roc.p)[2] <- "ROC prevalance"
+colnames(ad.p.16)[2] <- 2016
+#ad.roc.p <- mb.gather(ad.roc.p)
+#colnames(ad.roc.p)[2] <- "ROC prevalance"
 
@@
 
-ad.i <- list( ad.i.12, ad.i.13,ad.i.14,ad.i.15,ad.i.16,ad.incidence,ad.roc.i) %>% reduce(left_join, by = "Location")
+ad.i <- list( ad.i.12, ad.i.13,ad.i.14,ad.i.15,ad.i.16,ad.incidence) %>% reduce(left_join, by = "Location")
 
-ad.p <- list( ad.p.12, ad.p.13,ad.p.14,ad.p.15,ad.p.16,ad.incidence,ad.roc.p) %>% reduce(left_join, by = "Location")
+ad.p <- list( ad.p.12, ad.p.13,ad.p.14,ad.p.15,ad.p.16,ad.incidence) %>% reduce(left_join, by = "Location")
+
+library(reshape)
+as.numeric.factor <- function(x) {as.numeric(levels(x))[x]}
+
+mnd.i.melt = melt(mnd.i, id = 'Location')
+mnd.i.melt$variable <- as.numeric.factor(mnd.i.melt$variable)
+ggplot(mnd.i.melt, aes(x = variable, y = value, colour = Location)) + 
+  geom_line() + 
+  ylab(label="mnd Incidence") + 
+  xlab("Year")
+
+mnd.p.melt = melt(mnd.i, id = 'Location')
+mnd.p.melt$variable <- as.numeric.factor(mnd.p.melt$variable)
+ggplot(mnd.p.melt, aes(x = variable, y = value, colour = Location)) + 
+  geom_line() + 
+  ylab(label="mnd Prevalance") + 
+  xlab("Year")
+
+ipd.i.melt = melt(ipd.i, id = 'Location')
+ipd.i.melt$variable <- as.numeric.factor(ipd.i.melt$variable)
+ggplot(ipd.i.melt, aes(x = variable, y = value, colour = Location)) + 
+  geom_line() + 
+  ylab(label="ipd Incidence") + 
+  xlab("Year")
+
+ipd.p.melt = melt(ipd.i, id = 'Location')
+ipd.p.melt$variable <- as.numeric.factor(ipd.p.melt$variable)
+ggplot(ipd.p.melt, aes(x = variable, y = value, colour = Location)) + 
+  geom_line() + 
+  ylab(label="ipd Prevalance") + 
+  xlab("Year")
+
+ad.i.melt = melt(ad.i, id = 'Location')
+ad.i.melt$variable <- as.numeric.factor(ad.i.melt$variable)
+ggplot(ad.i.melt, aes(x = variable, y = value, colour = Location)) + 
+  geom_line() + 
+  ylab(label="AD Incidence") + 
+  xlab("Year")
+
+ad.p.melt = melt(ad.i, id = 'Location')
+ad.p.melt$variable <- as.numeric.factor(ad.p.melt$variable)
+ggplot(ad.p.melt, aes(x = variable, y = value, colour = Location)) + 
+  geom_line() + 
+  ylab(label="AD Prevalance") + 
+  xlab("Year")
 

### 2019-06-28 import_popln_data.Rmd
updated

@@
 
+require(cowplot)
+theme_set(theme_cowplot(font_size=12)) # reduce default font size
+
 mnd.i.melt = melt(mnd.i, id = 'Location')
 mnd.i.melt$variable <- as.numeric.factor(mnd.i.melt$variable)
-ggplot(mnd.i.melt, aes(x = variable, y = value, colour = Location)) + 
+p1 <- ggplot(mnd.i.melt, aes(x = variable, y = value, colour = Location)) + 
   geom_line() + 
   ylab(label="mnd Incidence") + 
-  xlab("Year")
+  xlab("Year") + theme(legend.position="none")
 
@@
 mnd.p.melt$variable <- as.numeric.factor(mnd.p.melt$variable)
-ggplot(mnd.p.melt, aes(x = variable, y = value, colour = Location)) + 
+p2 <-  ggplot(mnd.p.melt, aes(x = variable, y = value, colour = Location)) + 
   geom_line() + 
   ylab(label="mnd Prevalance") + 
-  xlab("Year")
+  xlab("Year") + theme(legend.position="none")
 
@@
 ipd.i.melt$variable <- as.numeric.factor(ipd.i.melt$variable)
-ggplot(ipd.i.melt, aes(x = variable, y = value, colour = Location)) + 
+p3 <- ggplot(ipd.i.melt, aes(x = variable, y = value, colour = Location)) + 
   geom_line() + 
   ylab(label="ipd Incidence") + 
-  xlab("Year")
+  xlab("Year") + theme(legend.position="none")
 
@@
 ipd.p.melt$variable <- as.numeric.factor(ipd.p.melt$variable)
-ggplot(ipd.p.melt, aes(x = variable, y = value, colour = Location)) + 
+p4 <-  ggplot(ipd.p.melt, aes(x = variable, y = value, colour = Location)) + 
   geom_line() + 
   ylab(label="ipd Prevalance") + 
-  xlab("Year")
+  xlab("Year") + theme(legend.position="none")
 
@@
 ad.i.melt$variable <- as.numeric.factor(ad.i.melt$variable)
-ggplot(ad.i.melt, aes(x = variable, y = value, colour = Location)) + 
+p5 <-  ggplot(ad.i.melt, aes(x = variable, y = value, colour = Location)) + 
   geom_line() + 
   ylab(label="AD Incidence") + 
-  xlab("Year")
+  xlab("Year") + theme(legend.position="none")
 
@@
 ad.p.melt$variable <- as.numeric.factor(ad.p.melt$variable)
-ggplot(ad.p.melt, aes(x = variable, y = value, colour = Location)) + 
+p6 <- ggplot(ad.p.melt, aes(x = variable, y = value, colour = Location)) + 
   geom_line() + 
   ylab(label="AD Prevalance") + 
-  xlab("Year")
+  xlab("Year") + theme(legend.position="none")
 
+final <- plot_grid(p1, p2,p3,p4,p5,p6, labels = c('A', 'B','C', 'D','E', 'F'),ncol = 2)
+ggsave("overview.jpg", final) 
 ```

### 2019-07-11 Earth_microbiome_MS.Rmd
updated

--- 2019-06-19 GlobalMicrobiome_NS.Rmd
@@
 ---
-title: "MRD_final_microbiome"
-author: "Reis-Dehabadi, M"
-date: "4/10/2019"
+title: "Earth_microbiome_MS"
+author: "Muriel Sirgi"
+date: "03/07/2019"
 output: html_document
@@
 knitr::opts_chunk$set(echo = TRUE)
+#setting the working directory in globalmicrobiome (GitHub directory)
 knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
-# test
-#test
+
 ```
 
-```{r functions}
+##Data Set-up
+```{r R Library Set-up}
 
+## Get data
+
+#libraries for data processing and formatting
+	library(foreign)
+	library(stargazer)
+	library(psych)
+	library(plyr)
+	library(xtable)
+	library(car) #
+
+	#library(dplyr)
+
+	library(scales)
+	library(reshape)
+	library(gmodels) #
+  library(nnet)
+	library(tidyr) #works well with dplyr data pipelines
+
+	library(sjmisc)
+	library(RColorBrewer)
+  library(htmltools) #
+	library(broom)
+
+  #library(hrbrthemes)
+library(gcookbook)
+library(tidyverse)
+ #library(gdtools)
+
+#The phyloseq package seeks to address issues with multiple microbiome analysis packages by providing a set of functions that internally manage the organizing, linking, storing, and analyzing of phylogenetic sequencing data.
+In general, this package is used for UniFrac analyses.
+library(phyloseq)
+library(microbiome)
+library(knitr)
+library(ggpubr)
+#library(phylogeo)
+library(dplyr)
+#Graphing package used in phyloseq. To edit the default setting of a plot, you need to use functions in this package.
+library(ggplot2)
+library(viridis) 
+library(cowplot)
+# install.packages("devtools")
+library(devtools)
+# install_github("opisthokonta/tsnemicrobiota")
+library(tsnemicrobiota)
+# check you've installed this library
+library(forcats)
+#package required for biomformat
+library(jsonlite)
+library(Matrix)
+library(rhdf5)
+#package used to work with biom files in R 
+library(biomformat)
+
+```
+
+##Microbiome Data Import
+```{r Microbiome Data Import}
+#file.choose() has been chosen here to take away any errors pertaining to the file path while investigating the files from the ftp directory
+githubdata_biom = file.choose()
+
+#This function transforms a biom file into a phyloseq class object
+githubdata_phyloseq = read_biom2phyloseq(githubdata_biom)
+
+```
+
+```{r Import GitHUb file (metada)}
+read.table
+```
+
+```{r Microbiome Data Import Draft}
+#import_biom("~/globalmicrobiome/emp_cr_silva_16S_123qc_filtered.biom") 
+#import_biom(BIOMfilename = '~/globalmicrobiome/emp_cr_gg_13_8.release1.biom')
+#biomfile <- file.choose("~/globalmicrobiome/emp_cr_gg_13_8.release1.biom")
+#read_biom2phyloseq(biom.file = biomfile)
+#import_biom('~/globalmicrobiome/emp_cr_gg_13_8.qc_filtered.biom')
+#import_biom('~/globalmicrobiome/emp_cr_silva_16S_123.qc_filtered.biom'
+
+#silva_closedref_biom <- system.file("extdata", "/Users/Muriel/globalmicrobiome/emp_cr_silva_16S_123qc_filtered.biom", package = "phyloseq")
+#sample_metadata("/Users/Muriel/globalmicrobiome/emp_cr_silva_16S_123qc_filtered.biom")
+
+#silva_closedref_biom = system.file("extdata", "~/globalmicrobiome/emp_cr_silva_16S_123qc_filtered.biom", package = "biomformat")
+
+#silva_data = sample_metadata(silva_closedref_biom)
+
+#biomfile = "/Users/Muriel/globalmicrobiome/emp_cr_silva_16S_123qc_filtered.biom"
+#read_biom(biomfile)
+#x = read_biom(biomfile)
+#header(x)
+#biom_shape(x)
+#nrow(x)
+#ncol(x)
+#rownames(x)
+#colnames(x)
+#matrix_element_type(x)
+#biom_data(x)
+
+```
+
+#Microbiome Functions (Nik)
+```{r functions (Nik)}
 ##### FUNCTIONS ########
@@
 
-## Data Set-up
-```{r Data set-up}
-
-#setwd("/Users/maryam/Scripts/Microbiome")
-
-	## Get data
-	library(foreign)
-	library(stargazer)
-	library(psych)
-	library(plyr)
-	library(xtable)
-	library(car)
-	#library(dplyr)
-	library(scales)
-	library(reshape) 
-	library(gmodels)
-	library(nnet)
-	library(tidyr)
-	library(sjmisc)
-	library(RColorBrewer)
-  library(htmltools)
-	library(broom)
-
-  #library(hrbrthemes)
-library(gcookbook)
-library(tidyverse)
- #library(gdtools)
-
-#The phyloseq package seeks to address issues with multiple microbiome analysis packages by providing a set of functions that internally manage the organizing, linking, storing, and analyzing of phylogenetic sequencing data.
-In general, this package is used for UniFrac analyses.
-library(phyloseq)
-library(microbiome)
-library(knitr)
-library(ggpubr)
-#library(phylogeo)
-library(dplyr)
-#Graphing package used in phyloseq. To edit the default setting of a plot, you need to use functions in this package.
-library(ggplot2)
-library(viridis) 
-library(cowplot)
-# install.packages("devtools")
-library(devtools)
-# install_github("opisthokonta/tsnemicrobiota")
-library(tsnemicrobiota)
-# check you've installed this library
-library(forcats)
-
-# Data from http://dx.doi.org/10.1038/ncomms7342
-#This data set contains genus-level microbiota profiling with HITChip for 1006 western adults with no reported health complications.
-
-pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
+```{r }
 
 ```
-
-## MND,PD,AD Variable
-```{r mnd variable}
-
-# Create mnd variable
-# grabs the nationality from phyloseq
-mbnational <- get_variable(pseq, "nationality" )
-
-# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-mndmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-pdmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-admbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-# reorder
-mndmbnational <- factor(mndmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-pdmbnational <- factor(pdmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-admbnational <- factor(admbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-
-# creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = mndmbnational
-sample_data(pseq)$pd = pdmbnational
-sample_data(pseq)$ad = admbnational
-
-# checks that it has worked.
-#get_variable(pseq, "mnd")
-
-```
-
-# Clostridum grouping
-```{r Clostridium grouping}
-
-clostridium <- subset_taxa(pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XVIa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales")
-
-pseq.new <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacilli"| Phylum =="Proteobacteria"  | Phylum =="Verruccomicrobia"  | Phylum =="Bacteroidetes" | Phylum =="Asteroleplasma"| Phylum =="Spirochaetes" | Phylum =="Fusobacteria" | Phylum =="Cyanobacteria" | Phylum =="Uncultured Mollucutes")
-
-```
-
-## pseq summary
-```{r pseq}
-pseq
-```
-
-```{r List of genera in phylum}
-# List of Genera in the Bacteroideted Phylum
-taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
-print(taxa)
-```
-
-```{r MND table summary}
-table(sample_data(pseq)$nationality)
-table(sample_data(pseq)$mnd)
-table(sample_data(pseq)$pd)
-table(sample_data(pseq)$ad)
-```
-
-## Summary REMOVE?
-```{r summary}
-
-# Summarise contents of phyloseq
-summarize_phyloseq(pseq)
-# Sparsity is how populated is the data with zeros
-kable(head(tax_table(pseq)))
-tax_table(pseq)
-
-# Check the data
-print(pseq)
-
-# Display variables from the mapping file
-sample_variables(pseq)
-
-# Total number of taxa in the entire data
-ntaxa(pseq)
-
-# Taxonomic ranks
-rank_names(pseq)
-
-```
-
-## Check OTUs/Phylums
-Prune(remove) taxa because we want to remove OTUs/taxa that aren't in any of the samples
-Don't need to prune this taxa because OTUs are found in all samples.
-Check this everytime you use the subset_samples function because of the pseq object being filtered. 
-```{r check OTUs}
-# Check if any OTUs are not present in any samples
-any(taxa_sums(pseq) == 0)
-# FALSE, therefore, OTUs are found in all samples and we don't need to prune the taxa.
-If the answer is TRUE, some OTUs are not present in any samples.
-This is usually the case when data is subset to remove some samples.
-OTUs unique to those sample are not removed along with the samples.
-Therefore, it is important to check this everytime the phyloseq object is filtered for samples using `subset_samples` function.
-
-head(meta(pseq))
-
-```
-
-## ALPHA DIVERSITY
-Diversity: Shannon, Simpson
-Richness: Chao, ACE
-
-## Prepare Data for Alpha Diversity Visualisation
-```{r prepare data for alpha diversity visualisation}
-
-# Prepare data for vizualisation
-# Alpha diversity for the entire data set
-png("alpha.png")
-tab <- diversities(pseq, index = "all")
-stargazer(tab, type = "text",summary = NULL)
-kable(head(tab))
-plot(tab)
-dev.off()
-
-# Get metadata
-pseq.meta <- meta(pseq)
-kable(head(pseq.meta))
-stargazer(pseq.meta, type = "text",summary = NULL)
-
-# Add the diversity table to metadata
-pseq.meta$Shannon <- tab$shannon 
-pseq.meta$InverseSimpson <- tab$inverse_simpson
-pseq.meta$fisher <- tab$fisher
-pseq.meta$chao1 <- tab$chao1
-
-```
-
-## Nationality Violin Plot
-```{r nationality violin plot}
-
-# Alpha diversity is the species richness of a place
-# High alpha diversity = high no. of species
-
-pseq.meta$nationality <- factor(pseq.meta$nationality, levels = c("Scandinavia","EasternEurope","CentralEurope", "SouthEurope", "UKIE", "US"))
-
-# Violin plot
-# the number in the palettes must equal the number of groups in nationlaity.pairs
-p1 <- ggviolin(pseq.meta, x = "nationality", y = "Shannon",
-      add = "boxplot", fill = "nationality", palette = "paired") +   theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-  print(p1)
-ggsave("nation_violin_shannon.png")
-
-p2 <- ggviolin(pseq.meta, x = "nationality", y = "InverseSimpson",
- add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-print(p2)
-ggsave("nation_violin_simpson.png")
-
-p3 <- ggviolin(pseq.meta, x = "nationality", y = "fisher",
- add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("") + ylab("Fisher")
-print(p3)
-ggsave("nation_violin_fisher.png")
-
-plot_grid(p1, p2, p3, align = "h" , ncol = 2, nrow = 2)
-ggsave("nation_violin.png")
-
-```
-
-```{r}
-
-p <- plot_richness(pseq, x="nationality", measures=c("Chao1", "Shannon", "Simpson")) 
-p + geom_boxplot()
-ggsave("alpha_nationality.png")
-
-```
-
-## MND Violin Plot
-```{r mnd violin plot}
-
-# Add the diversity table to metadata
-pseq.meta$Shannon <- tab$shannon
-pseq.meta$InverseSimpson <- tab$inverse_simpson
-
-# Compare differences in Shannon index between mnd group of the study subjects
-pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
-pseq.meta$pd <- factor(pseq.meta$pd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
-pseq.meta$ad <- factor(pseq.meta$ad, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
-# Create a list of pairwise comaprisons
-mnd <- levels(pseq.meta$mnd) # get the variables
-pd <- levels(pseq.meta$pd) # get the variables
-ad <- levels(pseq.meta$ad) # get the variables
-
-# Make a pairwise list that we want to compare.
-mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
-mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-pd.pairs <- combn(seq_along(pd), 2, simplify = FALSE, FUN = function(i)pd[i])
-pd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-ad.pairs <- combn(seq_along(ad), 2, simplify = FALSE, FUN = function(i)ad[i])
-ad.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-# Shannon
-# the number in the palettes must equal the number of groups in bmi.pairs
-p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
-p1 <- p1 + stat_compare_means(comparisons = mnd.pairs, paired = FALSE) + xlab("MND Incidence") + theme(legend.position="none")
-print(p1)
-ggsave("mnd_violin__shannon_stats.png")
-
-# Inverse Simpson
-p2 <- ggviolin(pseq.meta, x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
-p2 <- p2 + stat_compare_means(comparisons = mnd.pairs) 
-print(p2)
-ggsave("mnd_violin_simpson.png")
-
-```
-
-## BETA DIVERSITY
-Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
-WE want to: quantify community divergence within a given sample set
-Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.
-
-```{r Phylum NMDS}
-
-png("Phylum_NMDS.png")
-pseq.ord <- ordinate(pseq, "NMDS", "bray")
-p1 = plot_ordination(pseq, pseq.ord, type = "taxa", color = "mnd", title = "Taxa by Phylum")
-p1 + facet_wrap(~Phylum, 3)
-dev.off()
-
-```
-
- The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
-
-# VISUALISE MICORBIOME VARIATION
-```{r convert to compositional data}
-
-# Convert to compositional data
-pseq.rel <- microbiome::transform(pseq, "compositional")
-otu <- abundances(pseq.rel)
-meta <- meta(pseq.rel)
-
-```
-
-NMDS/Bray
-Bray-Curtis dissimilarity does not consider phylogenetic relationships between OTUs. There are several distance methods and a list can be obtained by typying ?distanceMethodList in the console pane.
-```{r}
-
-pseq.ord <- ordinate(pseq.rel, "NMDS", "bray")
-
-p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
-plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-ggsave("NMDStMND.png")
-
-p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
-ggsave("NMDS_facetMND.png")
-
-p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
-plot(p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
-ggsave("NMDStpd.png")
-
-p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, nrow = 2) + labs(color = "PD Prevalence")
-ggsave("NMDS_facetpd.png")
-
-p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
-plot(p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
-ggsave("NMDStad.png")
-
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, nrow = 2) + labs(color = "AD Prevalence")
-ggsave("NMDS_facetad.png")
-
-```
-
-MDS ordination on Bray-Curtis distance
-Performs principal coordinate analysis (also called principle coordinate decomposition, multidimensional scaling (MDS), or classical scaling) of a distance matrix (Gower 1966), including two correction methods for negative eigenvalues.
-See pcoa for further details.
-```{r}
-
-pseq.ord <- ordinate(pseq.rel, "MDS", "bray")
-
-p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
-plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-ggsave("MDS.png")
-
-p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
-ggsave("NMDS_facet.png")
-
-p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
-plot(p2 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
-ggsave("MDSpd.png")
-
-p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, 3) + labs(color = "PD Prevalence")
-ggsave("NMDS_facetPD.png")
-
-p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
-plot(p3 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
-ggsave("MDSad.png")
-
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, 3) + labs(color = "AD Prevalence")
-ggsave("NMDS_facetAD.png")
-
-```
-
-PCoA for compositional data with Bray-Curtis distances
-```{r}
-
-p <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
-p2 <- p + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b")
-
-p1 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
-print(p1)
-
-```
-
-```{r PCoA Bray}
-
-# Pick core taxa
-pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)
-
-# PCoA for compositional data with Bray-Curtis distances
-png("PCoA_bray.png")
-p <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
-                      method = "PCoA", distance = "bray") +
-       labs(title = paste("PCoA / Compositional / Bray-Curtis"))
-print(p)
-dev.off()
-
-```
-
-t-SNE is a popular new ordination technique
-```{r}
-
-tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
-#MND
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'mnd', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
-ggsave("mnd_t_sne.png")
-# PD
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'pd', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence")
-ggsave("pd_t_sne.png")
-
-# AD
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'ad', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence")
-ggsave("ad_t_sne.png")
-
-```
-
-Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
-This was applied in group-level comparisons for instance in Salonen et al.
-ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
-```{r Calculate group divergences}
-
-# Use these to compare microbiota divergence within each group
-
-#MND
-mnd.p.low <- subset_samples(pseq, mnd == "LOW")
-mnd.p.low.d <- divergence(mnd.p.low)
-
-mnd.p.med <- subset_samples(pseq, mnd == "MEDIUM")
-p.med.d <- divergence(mnd.p.med)
-
-mnd.p.high <- subset_samples(pseq, mnd == "HIGH")
-p.high.d <- divergence(mnd.p.high)
-
-mnd.p <- boxplot(list(Low = mnd.p.low.d, Medium = mnd.p.med.d, High = mnd.p.high.d))
-plot(mnd.p)
-
-# PD
-
-pd.p.low <- subset_samples(pseq, pd == "LOW")
-pd.p.low.d <- divergence(pd.p.low)
-
-pd.p.med <- subset_samples(pseq, pd == "MEDIUM")
-p.med.d <- divergence(pd.p.med)
-
-pd.p.high <- subset_samples(pseq, pd == "HIGH")
-p.high.d <- divergence(pd.p.high)
-
-pd.p <- boxplot(list(Low = pd.p.low.d, Medium = pd.p.med.d, High = pd.p.high.d))
-plot(pd.p)
-
-# AD
-
-ad.p.low <- subset_samples(pseq, ad == "LOW")
-ad.p.low.d <- divergence(ad.p.low)
-
-ad.p.med <- subset_samples(pseq, ad == "MEDIUM")
-p.med.d <- divergence(ad.p.med)
-
-ad.p.high <- subset_samples(pseq, ad == "HIGH")
-p.high.d <- divergence(ad.p.high)
-
-ad.p <- boxplot(list(Low = ad.p.low.d, Medium = ad.p.med.d, High = ad.p.high.d))
-plot(ad.p)
-
-```
-
-## PERMANOVA significance test for group-level differences
-Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
-
-```{r Redefine data}
-
-# Convert to compositional data
-# MND
-mnd.pseqLvH <- merge_phyloseq(mnd.p.low,mnd.p.high)
-mnd.pseqLvM <- merge_phyloseq(mnd.p.low,mnd.p.med)
-
-mnd.pseq.rel <- microbiome::transform(mnd.pseqLvH, "compositional")
-mnd.otu <- abundances(mnd.pseq.rel)
-mnd.meta <- meta(mnd.pseq.rel)
-
-# PD
-
-pd.pseqLvH <- merge_phyloseq(pd.p.low,pd.p.high)
-pd.pseqLvM <- merge_phyloseq(pd.p.low,pd.p.med)
-
-pd.pseq.rel <- microbiome::transform(pd.pseqLvH, "compositional")
-pd.otu <- abundances(pd.pseq.rel)
-pd.meta <- meta(pd.pseq.rel)
-
-# AD
-
-ad.pseqLvH <- merge_phyloseq(ad.p.low,ad.p.high)
-ad.pseqLvM <- merge_phyloseq(ad.p.low,ad.p.med)
-
-ad.pseq.rel <- microbiome::transform(ad.pseqLvH, "compositional")
-ad.otu <- abundances(ad.pseq.rel)
-ad.meta <- meta(ad.pseq.rel)
-
-```
-
-```{r PERMANOVA}
-
-# samples x species as input
-library(vegan)
-
-# MND
-mnd.permanova <- adonis(t(mnd.otu) ~ mnd,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(mnd.permanova$aov.tab)["mnd", "Pr(>F)"])
-summary(mnd.permanova$aov.tab)
-
-# PD
-pd.permanova <- adonis(t(pd.otu) ~ pd,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(pd.permanova$aov.tab)["pd", "Pr(>F)"])
-summary(pd.permanova$aov.tab)
-
-# AD
-ad.permanova <- adonis(t(ad.otu) ~ ad,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(ad.permanova$aov.tab)["ad", "Pr(>F)"])
-summary(ad.permanova$aov.tab)
-
-```
-
-Check that variance homogeneity assumptions hold (to ensure the reliability of the results):Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
-```{r ANOVA}
-#MND
-mnd.dist <- vegdist(t(mnd.otu))
-mnd.anov <- anova(betadisper(mnd.dist, meta$mnd))
-summary(mnd.anov)
-tidy(mnd.anov)
-
-#PD
-pd.dist <- vegdist(t(pd.otu))
-pd.anov <- anova(betadisper(pd.dist, meta$pd))
-summary(pd.anov)
-tidy(pd.anov)
-
-#AD
-ad.dist <- vegdist(t(ad.otu))
-ad.anov <- anova(betadisper(ad.dist, meta$ad))
-summary(ad.anov)
-tidy(ad.anov)
-
-```
-
-Show coefficients for the top taxa separating the groups
-```{r top taxa L vs H}
-
-coef <- coefficients(permanova)["mnd1",]
-top.coef <- coef[rev(order(abs(coef)))[1:20]]
-par(mar = c(3, 14, 2, 1))
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
-
-ggsave("LvH.png")
-
-```
-
-```{r L vs M}
-
-pseqLvM <- merge_phyloseq(p.low,p.med)
-
-pseq.rel <- microbiome::transform(pseqLvM, "compositional")
-otu <- abundances(pseq.rel)
-meta <- meta(pseq.rel)
-
-library(vegan)
-permanova <- adonis(t(otu) ~ mnd,
-               data = meta, permutations=999, method = "bray")
-
-# P-value
-print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
-
-dist <- vegdist(t(otu))
-anov <- anova(betadisper(dist, meta$mnd))
-summary(anov)
-tidy(anov)
-
-coef <- coefficients(permanova)["mnd1",]
-top.coef <- coef[rev(order(abs(coef)))[1:20]]
-par(mar = c(3, 14, 2, 1))
-
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
-
-ggsave("LvM.png", width = 20, height = 20, units = "cm")
-
-```
-
-## MICROBIOME COMPOSITION
-
-```{r Phylum #}
-
-# Show available ranks in the dataset
-rank_names(pseq)
-
-# Create table, number of features for each phyla
-table(tax_table(pseq)[, "Phylum"], exclude = NULL)
-
-```
-This shows a few phyla for which only one feature was observed. Those may be worth filtering.
-
-```{r Total Phylum abundnace bar}
-
-png("Phylum_abundance.png")
-p<- plot_bar(pseq,x="Phylum",fill = "Phylum") +geom_bar(stat="identity")
-
-phylumGlommed = tax_glom(pseq, "Phylum")
-plot_bar(phylumGlommed, x="Phylum", fill ="Phylum") +geom_bar(stat="identity")
-
-pd <- p$data %>% 
-    as_tibble %>%
-    mutate(Genus = as.character(Genus))
-genus_abun <- pd %>%
-    group_by(Genus) %>%
-    summarize(Abundance = sum(Abundance)) %>%
-    arrange(Abundance)
-genus_levels <- genus_abun$Genus
-pd0 <- pd %>%
-    mutate(Genus = factor(Genus, genus_levels))
-ggplot(pd0, aes(x = Sample, y = Abundance, fill = Genus)) + 
-    geom_bar(stat = "identity")
-dev.off()
-
-```
-
-```{r Phylum prevalence table}
-
-prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
-prev.otu + theme(legend.position="none")
-ggsave("phylum_prevalence.png")
-
-tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
-tab + theme(legend.position="none")
-ggsave("Phylum_prevalence_detection5.png")
-
-```
-
-theme(axis.text.x = element_text(angle = 90)
-+ guide_italics
-  ggtitle("Relative abundance")  + 
-```{r Relative abundance}
-
-pseq.ph <- aggregate_taxa(pseq.new, "Phylum")
-
-pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
-  theme(legend.position = "bottom") + theme_bw() + 
-  theme(axis.text.x=element_blank()) + 
-  theme(legend.title = element_text(size=18)) + scale_fill_brewer(palette = "Paired")
-print(plot.comp.rel)
-ggsave("phylum_relative_abundance11.png")
-
-plot.comp.rel + theme(legend.position="none")
-ggsave("phylum_relative_abundance12.png")
-
-```
-
-guide_italics
-ggtitle("Relative abundance") +  
-```{r}
-
-clos.ph <- aggregate_taxa(clostridium, "Phylum")
-
-clos.ph.rel <- microbiome::transform(clos.ph, "compositional")
-
-plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
-  theme(legend.position = "bottom", ) + theme_bw() + 
-  theme(axis.text.x=element_blank()) + 
-  theme(legend.title = element_text(size=18))
-plot.comp.rel + scale_fill_brewer(palette = "Paired")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
- theme(legend.position = "bottom") + theme_bw() +
- theme(axis.text.x=element_blank()) +
- theme(legend.title = element_text(size=18),legend.text = element_text( face="italic") ) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("my awesome title"))
-
-ggsave("phylum_relative_abundance_clos.png")
-
-plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none")
-ggsave("phylum_relative_abundance_clos2.png")
-
-```
-
-```{r Phylum Bar Plot}
-
-levels(pseq.meta$mnd)
-
-pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
-pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot")
-
-plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("phylum_abundance_bar.png")
-
-```
-
-```{r pseq.new bar plot}
-
-p.new <- aggregate_taxa(pseq.new, "Phylum")
-
-p.new.rel <- microbiome::transform(p.new, "compositional")
-
-plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.p.new.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("pnew_abundance_bar.png")
-
-```
-
-```{r CLostridium Bar Plot}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.cl.rel <- microbiome::transform(clos, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-ggsave("cl_abundance_bar.png")
-
-```
-
-```{r Clostridium genus bar plots}
-
-pseq.cl.rel <- microbiome::transform(clostridium, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")
-
-```
-
-```{r Top Phylum Heatmap}
-
-pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
-pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
-ggsave("phylum_histogram.png")
-
-```
-
-```{r p.new heatmap}
-
-pseq.n.ph <- aggregate_taxa(pseq.new, "Phylum")
-
-pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
-
-plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-p <- plot.n.comp.rel + coord_flip()
-p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-ggsave("pnew_histogram.png")
-
-```
-
-```{r clostridium heatmap}
-
-pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")
-
-plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-p <- plot.c.comp.rel + coord_flip()
-p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-ggsave("clos_histogram.png")
-
-```
-
-```{r Network}
-
-png("network_bray_phylum.png")
-jg = make_network(pseq, "taxa", "bray", 0.3)
-plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
-dev.off()
-
-```
-
-## Community composition plotting
-Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results.
-These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted.
-Interpretation of these plots should always be subject to subsequent statistical analysis.
-```{r community-composition-plots}
-# Create a data table for ggplot
-# agglomerate at phylum level
-pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
-
-# Transform to rel. abundance (or use ps0.ra)
-pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
-
-# Melt to long format for easy ggploting
-pseq_phylum <- psmelt(pseq_phyl)
-
-     # filter(Abundance > 0.01)                             # Filter out low abundance taxa
-
-# Plot - Phylum
-p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-  geom_bar(stat = "identity", width = 1) +
-  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-  theme(axis.text.x = element_blank()) +
-  theme(axis.title.x = element_blank()) +
-  labs(title = "Abundant Phylum (> 1%)")
-
-p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-plot(p)
-ggsave("phylum_abun_mnd.png")
-
-```
-
-```{r}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-# agglomerate at phylum level
-clos_phy <- tax_glom(clos, taxrank = "Phylum")
-
-# Transform to rel. abundance (or use ps0.ra)
-clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})
-
-# Melt to long format for easy ggploting
-clos_phylum <- psmelt(clos_phyl)
-
-# Plot - Phylum
-p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-  geom_bar(stat = "identity", width = 1) +
-  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-  theme(axis.text.x = element_blank()) +
-  theme(axis.title.x = element_blank()) +
-  labs(title = "Abundant Phylum of Clostridium(> 1%)")
-
-p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-plot(p)
-ggsave("phylum_abun_mnd_clos.png")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Phylum")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Genus")
-
-```
-
-```{r}
-
-get_taxa_unique(clostridium, "Genus")
-
-```
-
-```{r}
-
-df <- psmelt(pseq)
-kable(head(df))
-
-```
-
-```{r}
-
-head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])
-
-```

### 2019-07-15 Earth_microbiome_MS.Rmd
updated

@@
 
-```
-
-```{r Import GitHUb file (metada)}
-read.table
 ```
@@
 
-```{r }
+#Read phyloseq object
+```{r OTU table read}
 
 ```
+
+#READ ME
+
+Files downloaded from the emp ftp directory could not be reecognised a JSON or HDF5 file as required by the biomformat package and therefore could not be read in R.
+This is a significant problem that explains the use of a GitHub file (emp_deblur_150bp.subset_2k.rare_5000.biom").
+This file was used as a test to verify the working order of certain sections of the code.
+
+Table selected to run code ("/Users/Muriel/Downloads/emp_deblur_150bp.subset_2k.rare_5000.biom") does not feature sample data - it may be necessary to look into other files or acquire the sample data from another file using the sample references

### 2019-07-15 Earth_microbiome_MS.Rmd
updated

@@
 #file.choose() has been chosen here to take away any errors pertaining to the file path while investigating the files from the ftp directory
-githubdata_biom = file.choose()
+githubdata_biom <- file.choose()
 
 #This function transforms a biom file into a phyloseq class object
-githubdata_phyloseq = read_biom2phyloseq(githubdata_biom)
+githubdata_phyloseq <- read_biom2phyloseq(githubdata_biom)
 
@@
 #biom_data(x)
+
+#ftpdata_silva_qrbiom = import_qiime("/Users/Muriel/globalmicrobiome/emp_cr_silva_16S_123qc_filtered.biom")
+#ftpdata_silva_phyloseq = read_biom2phyloseq(ftpdata_silva_qrbiom)
 
@@
 #Read phyloseq object
-```{r OTU table read}
+```{r Pseq read}
+#Methods obtained from Callahan et al.
+
+rank_names(githubdata_phyloseq)
 

### 2019-07-15 Earth_microbiome_MS.Rmd
updated

@@
 #Read phyloseq object
-```{r Pseq read}
+```{r Taxonomic filtering}
 #Methods obtained from Callahan et al.
 
+#Show taxonomic ranks in the phyloseq object
 rank_names(githubdata_phyloseq)
+
+#Create table showing the phyla present in the table 
+table(tax_table(githubdata_phyloseq)[, "Phylum"], exclude = NULL)
+#It is noted here that some phyla are named following a different denomination system
+
+#Remove unchar
 
@@
 Table selected to run code ("/Users/Muriel/Downloads/emp_deblur_150bp.subset_2k.rare_5000.biom") does not feature sample data - it may be necessary to look into other files or acquire the sample data from another file using the sample references
+
+Some of the phyla in the table

### 2019-07-15 Earth_microbiome_MS.Rmd
updated

@@
 #Read phyloseq object
+```{r Phyloseq object - first read}
+#Methods from MRD microbiome file
+#Print phyloseq object to check its component parts
+print(githubdata_phyloseq)
+
+#Check sparsity of physeq data
+kable(head(tax_table(githubdata_phyloseq)))
+tax_table(githubdata_phyloseq)
+
+#summarize_phyloseq(githubdata_phyloseq)
+#The above code ran unsuccessfully due to the fact that the data does not include sample data
+```
+
 ```{r Taxonomic filtering}

### 2019-07-15 Earth_microbiome_MS.Rmd
updated

@@
 	library(stargazer)
-	library(psych)
+	library(githubdata_phyloseqych)
 	library(plyr)
@@
 
+#Taxonomic filtering
 ```{r Taxonomic filtering}
@@
 table(tax_table(githubdata_phyloseq)[, "Phylum"], exclude = NULL)
-#It is noted here that some phyla are named following a different denomination system
+#It is noted here that some phyla are named following a different convention
 
-#Remove unchar
+#Remove ambiguous phyla (if any)
+githubdata_phyloseq <- subset_taxa(githubdata_phyloseq, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))
+
+#Compute prevalence of each Phyla, store as data.frame
+prevdf = apply(X = otu_table(githubdata_phyloseq),
+               MARGIN = ifelse(taxa_are_rows(githubdata_phyloseq), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+#Add taxonomy and total read counts to this data.frame
+prevdf = data.frame(Prevalence = prevdf,
+                    TotalAbundance = taxa_sums(githubdata_phyloseq),
+                    tax_table(githubdata_phyloseq))
+
+#Create table dispaying the average prevalence of each phyla per sample and the total reads for all phyla in the data frame
+plyr::ddply(prevdf, "Phylum", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
 
@@
 
-#READ ME
+#READ ME - Problems encountered
 
@@
 
-Some of the phyla in the table
+Some of the phyla in the table are named following a differenct convention.
+If this file is used for the data this will have to be kept in mind.
+It is possible that other files for EMP data also share this problem.

### 2019-07-15 Earth_microbiome_MS.Rmd
updated

@@
 
+#Filter out low-prevalance phyla
+  #1. Define phyla to filter
+filterPhyla = c("p__Caldiserica", "p__Deferribacteres", "p__LD1", "p__MVP-21", "p__OctSpA1-106", "p__TPD-58", "p__SC4", "p__VHS-B3-43", "p__WS1" )
+  #2. Filter entries
+github_filtered = subset_taxa(githubdata_phyloseq, !Phylum %in% filterPhyla)
+github_filtered
+  #3. Create data frame in R Markdown for easier comparison (see prevdf)
+prevdf_filtered = apply(X = otu_table(github_filtered),
+               MARGIN = ifelse(taxa_are_rows(github_filtered), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+plyr::ddply(prevdf_filtered, "Phylum", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
+
 ```

### 2019-07-15 Earth_microbiome_MS.Rmd
updated

--- 2019-07-15 Earth_microbiome_MS.Rmd
@@
 
+#Filter out low-prevalance phyla
+  #1. Define phyla to filter
+filterPhyla = c("p__Caldiserica", "p__Deferribacteres", "p__LD1", "p__MVP-21", "p__OctSpA1-106", "p__TPD-58", "p__SC4", "p__VHS-B3-43", "p__WS1" )
+  #2. Filter entries
+github_filtered = subset_taxa(githubdata_phyloseq, !Phylum %in% filterPhyla)
+github_filtered
+  #3. Create data frame in R Markdown for easier comparison (see prevdf)
+prevdf_filtered = apply(X = otu_table(github_filtered),
+               MARGIN = ifelse(taxa_are_rows(github_filtered), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevdf_filtered = data.frame(Prevalence = prevdf_filtered,
+                    TotalAbundance = taxa_sums(github_filtered),
+                    tax_table(github_filtered))
+
+plyr::ddply(prevdf_filtered, "Phylum", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
+
 ```

### 2019-07-16 Earth_microbiome_MS.Rmd
updated

@@
 #Check sparsity of physeq data
-kable(head(tax_table(githubdata_phyloseq)))
-tax_table(githubdata_phyloseq)
+#kable(head(tax_table(githubdata_phyloseq)))
+#tax_table(githubdata_phyloseq)
 
@@
 github_filtered
-  #3. Create data frame in R Markdown for easier comparison (see prevdf)
+  #3. Create data frame in R Markdown to display filtered prevalence to allow for an easier comparison (see prevdf)
 prevdf_filtered = apply(X = otu_table(github_filtered),
@@
 
-#READ ME - Problems encountered
+#READ ME
+
+####Problems encountered
 
@@
 
-Some of the phyla in the table are named following a differenct convention.
+Some of the phyla in the table are named following what appears to be a different convention.
+However, further research indicates thas these are uncharacterised phyla (likely with no cultured representatives) present in the Silva and Greengenes databases.
 If this file is used for the data this will have to be kept in mind.
-It is possible that other files for EMP data also share this problem.
+
+####Notes
+
+I believe taxonomic filtering should not be arbitrary.
+This can be done by manually choosing phyla and writing them into the function but can likely be done mathematically by choosing a minimum prevalence (total or average)
+Prevalence filtering can be used as an alternative (should be consistent with the gut microbiome filtering?)
+
+AC1 (doi: https://dx.doi.org/10.1073%2Fpnas.1015676108)

### 2019-07-16 Earth_microbiome_MS copy.Rmd
updated

--- 2019-07-16 Earth_microbiome_MS.Rmd
@@
 	library(stargazer)
-	library(githubdata_phyloseqych)
+	library(psych)
 	library(plyr)
@@
 #Check sparsity of physeq data
-#kable(head(tax_table(githubdata_phyloseq)))
-#tax_table(githubdata_phyloseq)
+kable(head(tax_table(githubdata_phyloseq)))
+tax_table(githubdata_phyloseq)
 
@@
 ```
+
+#Check/Prune OTUs
+```{r Check and prune OTUs}
+any(taxa_sums(githubdata_phyloseq) == 0)
+#FALSE, therefore pruning is unecessary.
+```
+
+The following methods are arguably different options that can be used to work with reads of low-prevalence phyla.
+These are taxonomic filtering (removing low-prevalence phyla), prevalence filtering (removing phyla that do not meet a set prevalence), and agglomerating taxa/merging OTUs.
 
@@
 
+```{r Prevalence filtering}
+#This method could be preferred to filter out plow-prevalence phyla
+
+#Define the prevalence threshold
+  #Considering the diversity of this data set I will arbitrarily set the prevalence   threshold at 5% (due to its size, I also considered 10%)
+prevalence_threshold = 0.05 * nsamples(githubdata_phyloseq)
+prevalence_threshold
+
+#
+
+```
+
 #READ ME
@@
 Prevalence filtering can be used as an alternative (should be consistent with the gut microbiome filtering?)
+See functions as an alternative
 
 AC1 (doi: https://dx.doi.org/10.1073%2Fpnas.1015676108)
+https://www.uniprot.org/taxonomy/1052815

### 2019-07-16 Earth_microbiome_MS copy.Rmd
updated

@@
 
+#Prevalence Filtering
 ```{r Prevalence filtering}
 #This method could be preferred to filter out plow-prevalence phyla
+#Additionally, it could be useful to review potential sample data to verify the diversity of environments covered by this data set
 
@@
 
-#
+#Prune taxa using the previously set prevalence threshold
+keepTaxa = rownames(prevdf)[(prevdf$Prevalence >= prevalence_threshold)]
+github_filtered2 = prune_taxa(keepTaxa, githubdata_phyloseq)
+
+#Display information
+print(github_filtered2)
+
+#display information about other tables for a comparison value. 
+print(githubdata_phyloseq)
+print(github_filtered)
+
+```
+Considering the comparison created above, I believe this method is not ideal unless the average prevalence of phyla within the data set is considered.
+
+#Merge OTUs
+```{r Merge OTUs}
 
@@
 
+Methods of filtering?
+
 ####Notes

### 2019-07-16 Earth_microbiome_MS copy.Rmd
updated

@@
 #Create table dispaying the average prevalence of each phyla per sample and the total reads for all phyla in the data frame
-plyr::ddply(prevdf, "Phylum", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
+plyr::ddply(prevdf, "Phylum (unfiltered)", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
 
@@
 
+```{r Prevalence Calculations}
+#Calculate the mean prevalence from the prevdf
+```
+
+```{r Phyloseq filtering}
+#Function is simple and could be an alternative - need to check how it fits into code
+#https://rdrr.io/github/vmikk/metagMisc/src/R/phyloseq_filter.R#sym-phyloseq_filter_prevalence
+
+#github_filtered3 <- phyloseq_filter_prevalence(githubdata_phyloseq, prev.trh = 0.05, abund.trh = NULL,
+  #threshold_condition = "OR", abund.type = "mean")
+
+#print(github_filtered3)
+```
+
 #Merge OTUs

### 2019-08-01 MRD_final_microbiome .Rmd
updated

--- 2019-06-19 GlobalMicrobiome_NS.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
-knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
-# test
-#test
-```
-
-```{r functions}
+```
+
+```{r}
 
 ##### FUNCTIONS ########
+
 ##### Normalization #######
@@
 
+## R Markdown
+
+This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+
+When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
+You can embed an R code chunk like this:
+
+Atlas1006
+https://rdrr.io/github/microbiome/microbiome/man/atlas1006.html
+Using dataset: atlas1006: HITChip Atlas with 1006 Western adults
+This data set contains genus-level microbiota profiling with HITChip for 1006 western adults with no reported health complications. 
+Format: phyloseq-class format
+Usage: data(atlas1006)
+
 ## Data Set-up
@@
 
-#setwd("/Users/maryam/Scripts/Microbiome")
+setwd("/Users/maryam/Scripts/Microbiome")
 
@@
 library(ggpubr)
+
   #library(phylogeo)
 library(dplyr)
+
 #Graphing package used in phyloseq. To edit the default setting of a plot, you need to use functions in this package.
 library(ggplot2)
+
 library(viridis) 
 library(cowplot)
+
 # install.packages("devtools")
@@
 library(tsnemicrobiota)
+
+# Data from http://dx.doi.org/10.1038/ncomms7342
+data(atlas1006)
+
+pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
+
+```
+
+## MND Variable
+```{r mnd variable}
+
+# Create mnd variable
+# grabs the nationality from phyloseq
+test <- get_variable(pseq, "nationality" )
+
 # check you've installed this library
@@
 
-# Data from http://dx.doi.org/10.1038/ncomms7342
-#This data set contains genus-level microbiota profiling with HITChip for 1006 western adults with no reported health complications.
-
-pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
-
-```
-
-## MND,PD,AD Variable
-```{r mnd variable}
-
-# Create mnd variable
-# grabs the nationality from phyloseq
-mbnational <- get_variable(pseq, "nationality" )
-
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-mndmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-pdmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-admbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+test <- fct_collapse(test, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
 
 # reorder
-mndmbnational <- factor(mndmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-pdmbnational <- factor(pdmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
-admbnational <- factor(admbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
+test <- factor(test, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(test)
 
 # creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = mndmbnational
-sample_data(pseq)$pd = pdmbnational
-sample_data(pseq)$ad = admbnational
+sample_data(pseq)$mnd = test
 
 # checks that it has worked.
-#get_variable(pseq, "mnd")
+get_variable(pseq, "mnd")
 
@@
 
-clostridium <- subset_taxa(pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XVIa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales")
-
-pseq.new <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacilli"| Phylum =="Proteobacteria"  | Phylum =="Verruccomicrobia"  | Phylum =="Bacteroidetes" | Phylum =="Asteroleplasma"| Phylum =="Spirochaetes" | Phylum =="Fusobacteria" | Phylum =="Cyanobacteria" | Phylum =="Uncultured Mollucutes")
+clostridium <- subset_taxa(pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XIVa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales")
+
+pseq.new <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacilli"| Phylum =="Proteobacteria"  | Phylum =="Verrucomicrobia"  | Phylum =="Bacteroidetes" | Phylum =="Asteroleplasma"| Phylum =="Spirochaetes" | Phylum =="Fusobacteria" | Phylum =="Cyanobacteria" | Phylum =="Uncultured Mollicutes")
 
@@
 ```{r pseq}
+
 pseq
-```
-
+
+```
+
+#### List of genera in phylums
 ```{r List of genera in phylum}
+
 # List of Genera in the Bacteroideted Phylum
@@
 print(taxa)
+
+map_levels(NULL, "Phylum", "Genus", pseq)
+
 ```
@@
 ```{r MND table summary}
+
+table(sample_data(pseq)$mnd)
+
 table(sample_data(pseq)$nationality)
-table(sample_data(pseq)$mnd)
-table(sample_data(pseq)$pd)
-table(sample_data(pseq)$ad)
-```
-
-## Summary REMOVE?
+
+```
+
+## Summary
 ```{r summary}
@@
 # Sparsity is how populated is the data with zeros
+
 kable(head(tax_table(pseq)))
@@
 rank_names(pseq)
+
+```
+
+```{r average}
+
+library(purrr)
+library(pastecs)
+pseq.meta %>% split(.$mnd) %>% map(stat.desc)
 
@@
 # Prepare data for vizualisation
-# Alpha diversity for the entire data set
+
+# Alpha diversity again
 png("alpha.png")
@@
 
-pseq.meta$pd <- factor(pseq.meta$pd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
-pseq.meta$ad <- factor(pseq.meta$ad, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-
 # Create a list of pairwise comaprisons
 mnd <- levels(pseq.meta$mnd) # get the variables
-pd <- levels(pseq.meta$pd) # get the variables
-ad <- levels(pseq.meta$ad) # get the variables
 
@@
 
-pd.pairs <- combn(seq_along(pd), 2, simplify = FALSE, FUN = function(i)pd[i])
-pd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-ad.pairs <- combn(seq_along(ad), 2, simplify = FALSE, FUN = function(i)ad[i])
-ad.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
 # Shannon
 # the number in the palettes must equal the number of groups in bmi.pairs
-p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
-p1 <- p1 + stat_compare_means(comparisons = mnd.pairs, paired = FALSE) + xlab("MND Incidence") + theme(legend.position="none")
+p1 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+p1 <- p1 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
 print(p1)
@@
 # Inverse Simpson
-p2 <- ggviolin(pseq.meta, x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
+p2 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
 p2 <- p2 + stat_compare_means(comparisons = mnd.pairs) 
@@
 pseq.ord <- ordinate(pseq, "NMDS", "bray")
+
 p1 = plot_ordination(pseq, pseq.ord, type = "taxa", color = "mnd", title = "Taxa by Phylum")
@@
 
+Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
+This was applied in group-level comparisons for instance in Salonen et al.
+ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
+```{r Calculate group divergences}
+
+# Use these to compare microbiota divergence within each group
+
+p.low <- subset_samples(pseq, mnd == "LOW")
+p.low.d <- divergence(p.low)
+
+p.med <- subset_samples(pseq, mnd == "MEDIUM")
+p.med.d <- divergence(p.med)
+
+p.high <- subset_samples(pseq, mnd == "HIGH")
+p.high.d <- divergence(p.high)
+
+p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+plot(p)
+
+```
  The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
@@
 plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-ggsave("NMDStMND.png")
+ggsave("NMDSt.png")
 
 p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
-ggsave("NMDS_facetMND.png")
-
-p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
-plot(p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
-ggsave("NMDStpd.png")
-
-p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, nrow = 2) + labs(color = "PD Prevalence")
-ggsave("NMDS_facetpd.png")
-
-p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
-plot(p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
-ggsave("NMDStad.png")
-
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, nrow = 2) + labs(color = "AD Prevalence")
-ggsave("NMDS_facetad.png")
+ggsave("NMDS_facet.png")
 
@@
 p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
-ggsave("NMDS_facet.png")
-
-p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
-plot(p2 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
-ggsave("MDSpd.png")
-
-p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, 3) + labs(color = "PD Prevalence")
-ggsave("NMDS_facetPD.png")
-
-p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
-plot(p3 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
-ggsave("MDSad.png")
-
-p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, 3) + labs(color = "AD Prevalence")
-ggsave("NMDS_facetAD.png")
+ggsave("MDS_facet.png")
 
@@
 t-SNE is a popular new ordination technique
-```{r}
+, title='t-SNE (Bray)'
+```{r t-SNE}
 
 tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
-#MND
 tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'mnd', title='t-SNE (Bray)')
+  color = 'mnd')
 tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
-ggsave("mnd_t_sne.png")
-# PD
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'pd', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence")
-ggsave("pd_t_sne.png")
-
-# AD
-tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
-  color = 'ad', title='t-SNE (Bray)')
-tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence")
-ggsave("ad_t_sne.png")
-
-```
-
-Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
-This was applied in group-level comparisons for instance in Salonen et al.
-ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
-```{r Calculate group divergences}
-
-# Use these to compare microbiota divergence within each group
-
-#MND
-mnd.p.low <- subset_samples(pseq, mnd == "LOW")
-mnd.p.low.d <- divergence(mnd.p.low)
-
-mnd.p.med <- subset_samples(pseq, mnd == "MEDIUM")
-p.med.d <- divergence(mnd.p.med)
-
-mnd.p.high <- subset_samples(pseq, mnd == "HIGH")
-p.high.d <- divergence(mnd.p.high)
-
-mnd.p <- boxplot(list(Low = mnd.p.low.d, Medium = mnd.p.med.d, High = mnd.p.high.d))
-plot(mnd.p)
-
-# PD
-
-pd.p.low <- subset_samples(pseq, pd == "LOW")
-pd.p.low.d <- divergence(pd.p.low)
-
-pd.p.med <- subset_samples(pseq, pd == "MEDIUM")
-p.med.d <- divergence(pd.p.med)
-
-pd.p.high <- subset_samples(pseq, pd == "HIGH")
-p.high.d <- divergence(pd.p.high)
-
-pd.p <- boxplot(list(Low = pd.p.low.d, Medium = pd.p.med.d, High = pd.p.high.d))
-plot(pd.p)
-
-# AD
-
-ad.p.low <- subset_samples(pseq, ad == "LOW")
-ad.p.low.d <- divergence(ad.p.low)
-
-ad.p.med <- subset_samples(pseq, ad == "MEDIUM")
-p.med.d <- divergence(ad.p.med)
-
-ad.p.high <- subset_samples(pseq, ad == "HIGH")
-p.high.d <- divergence(ad.p.high)
-
-ad.p <- boxplot(list(Low = ad.p.low.d, Medium = ad.p.med.d, High = ad.p.high.d))
-plot(ad.p)
+ggsave("t_sne.png")
 
@@
 # Convert to compositional data
-# MND
-mnd.pseqLvH <- merge_phyloseq(mnd.p.low,mnd.p.high)
-mnd.pseqLvM <- merge_phyloseq(mnd.p.low,mnd.p.med)
-
-mnd.pseq.rel <- microbiome::transform(mnd.pseqLvH, "compositional")
-mnd.otu <- abundances(mnd.pseq.rel)
-mnd.meta <- meta(mnd.pseq.rel)
-
-# PD
-
-pd.pseqLvH <- merge_phyloseq(pd.p.low,pd.p.high)
-pd.pseqLvM <- merge_phyloseq(pd.p.low,pd.p.med)
-
-pd.pseq.rel <- microbiome::transform(pd.pseqLvH, "compositional")
-pd.otu <- abundances(pd.pseq.rel)
-pd.meta <- meta(pd.pseq.rel)
-
-# AD
-
-ad.pseqLvH <- merge_phyloseq(ad.p.low,ad.p.high)
-ad.pseqLvM <- merge_phyloseq(ad.p.low,ad.p.med)
-
-ad.pseq.rel <- microbiome::transform(ad.pseqLvH, "compositional")
-ad.otu <- abundances(ad.pseq.rel)
-ad.meta <- meta(ad.pseq.rel)
-
-```
-
-```{r PERMANOVA}
-
-# samples x species as input
-library(vegan)
-
-# MND
-mnd.permanova <- adonis(t(mnd.otu) ~ mnd,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(mnd.permanova$aov.tab)["mnd", "Pr(>F)"])
-summary(mnd.permanova$aov.tab)
-
-# PD
-pd.permanova <- adonis(t(pd.otu) ~ pd,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(pd.permanova$aov.tab)["pd", "Pr(>F)"])
-summary(pd.permanova$aov.tab)
-
-# AD
-ad.permanova <- adonis(t(ad.otu) ~ ad,
-               data = meta, permutations=999, method = "bray")
-# P-value
-print(as.data.frame(ad.permanova$aov.tab)["ad", "Pr(>F)"])
-summary(ad.permanova$aov.tab)
-
-```
-
-Check that variance homogeneity assumptions hold (to ensure the reliability of the results):Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
-```{r ANOVA}
-#MND
-mnd.dist <- vegdist(t(mnd.otu))
-mnd.anov <- anova(betadisper(mnd.dist, meta$mnd))
-summary(mnd.anov)
-tidy(mnd.anov)
-
-#PD
-pd.dist <- vegdist(t(pd.otu))
-pd.anov <- anova(betadisper(pd.dist, meta$pd))
-summary(pd.anov)
-tidy(pd.anov)
-
-#AD
-ad.dist <- vegdist(t(ad.otu))
-ad.anov <- anova(betadisper(ad.dist, meta$ad))
-summary(ad.anov)
-tidy(ad.anov)
-
-```
-
-Show coefficients for the top taxa separating the groups
-```{r top taxa L vs H}
-
-coef <- coefficients(permanova)["mnd1",]
-top.coef <- coef[rev(order(abs(coef)))[1:20]]
-par(mar = c(3, 14, 2, 1))
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
-
-ggsave("LvH.png")
-
-```
-
-```{r L vs M}
-
-pseqLvM <- merge_phyloseq(p.low,p.med)
-
-pseq.rel <- microbiome::transform(pseqLvM, "compositional")
+pseqLvH <- merge_phyloseq(p.low,p.high)
+
+pseq.rel <- microbiome::transform(pseqLvH, "compositional")
 otu <- abundances(pseq.rel)
@@
 
+```
+
+```{r PERMANOVA}
+
+# samples x species as input
 library(vegan)
@@
 print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+
+summary(permanova$aov.tab)
+
+```
+
+Homogeneity of variance is an assumption underlying both t tests and F tests (analyses of variance, ANOVAs) in which the population variances (i.e., the distribution, or “spread,” of scores around the mean) of two or more samples are considered equal.
+Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
+```{r ANOVA}
 
@@
 tidy(anov)
+summary(anov$aov.tab)
+
+```
+
+Show coefficients for the top taxa separating the groups
+```{r top taxa L vs H}
 
@@
 par(mar = c(3, 14, 2, 1))
-
-barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
-
-ggsave("LvM.png", width = 20, height = 20, units = "cm")
+barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis=3, main = "Top Genus")
+ggsave("LvH.png", width = 20, height = 20, units = "cm", dpi=700)
+
+```
+
+```{r L vs M}
+
+pseqLvM <- merge_phyloseq(p.low,p.med)
+
+pseq.rel <- microbiome::transform(pseqLvM, "compositional")
+otu <- abundances(pseq.rel)
+meta <- meta(pseq.rel)
+
+library(vegan)
+permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+
+# P-value
+print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+summary(permanova$aov.tab)
+
+dist <- vegdist(t(otu))
+anov <- anova(betadisper(dist, meta$mnd))
+summary(anov)
+tidy(anov)
+
+coef <- coefficients(permanova)["mnd1",]
+top.coef <- coef[rev(order(abs(coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+
+png(filename="LvvM.png")
+barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis=3, main = "Top Genus")
+dev.off()
 
@@
 
-png("Phylum_abundance.png")
-p<- plot_bar(pseq,x="Phylum",fill = "Phylum") +geom_bar(stat="identity")
-
-phylumGlommed = tax_glom(pseq, "Phylum")
-plot_bar(phylumGlommed, x="Phylum", fill ="Phylum") +geom_bar(stat="identity")
-
-pd <- p$data %>% 
-    as_tibble %>%
-    mutate(Genus = as.character(Genus))
-genus_abun <- pd %>%
-    group_by(Genus) %>%
-    summarize(Abundance = sum(Abundance)) %>%
-    arrange(Abundance)
-genus_levels <- genus_abun$Genus
-pd0 <- pd %>%
-    mutate(Genus = factor(Genus, genus_levels))
-ggplot(pd0, aes(x = Sample, y = Abundance, fill = Genus)) + 
-    geom_bar(stat = "identity")
-dev.off()
+plot_bar(pseq,x="Phylum",fill = "Phylum") +geom_bar(stat="identity") + theme(legend.position="none") + theme(axis.text.x = element_text(face="italic"))
+ggsave("Phylum_abundance.png", height=5, width=5, units='in', dpi=700)
 
@@
 
-theme(axis.text.x = element_text(angle = 90)
-+ guide_italics
-  ggtitle("Relative abundance")  + 
 ```{r Relative abundance}
@@
 theme(axis.text.x=element_blank()) +
-  theme(legend.title = element_text(size=18)) + scale_fill_brewer(palette = "Paired")
+c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
+
 print(plot.comp.rel)
+
 ggsave("phylum_relative_abundance11.png")
@@
   theme(axis.text.x=element_blank()) + 
-  theme(legend.title = element_text(size=18))
-plot.comp.rel + scale_fill_brewer(palette = "Paired")
-
-plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
- theme(legend.position = "bottom") + theme_bw() +
- theme(axis.text.x=element_blank()) +
- theme(legend.title = element_text(size=18),legend.text = element_text( face="italic") ) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("my awesome title"))
+  theme(legend.title = element_text(size=18), legend.text = element_text( face="italic")) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend(title="Phylum"))
+
+print(plot.comp.rel)
 
@@
 
-plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none")
+plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none") + guide_italics
 ggsave("phylum_relative_abundance_clos2.png")
@@
 
-plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot")
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
 
@@
 
+```{r}
+
+top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
+
+top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
+
+pseq.ph.rel <- microbiome::transform(top5P, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+```
+
+```{r}
+
+top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
+
+top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
+
+pseq.ph.rel <- microbiome::transform(top10P, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+```
+
 ```{r pseq.new bar plot}
@@
 
-plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.p.new.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
 ggsave("pnew_abundance_bar.png")
@@
 
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.cl.rel + scale_fill_brewer(palette = "Paired")
 ggsave("cl_abundance_bar.png")
@@
 ```
+
+# F/B ratio
+```{r}
+
+firmicutes.bact <- subset_taxa(pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XIVa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales" | Phylum =="Bacilli" | Phylum =="Uncultured Mollicutes" | Phylum =="Bacteroidetes")
+
+p.f <- aggregate_taxa(firmicutes.bact, "Phylum")
+
+p.f.rel <- microbiome::transform(p.f, "compositional")
+
+plot.p.f.rel <- plot_composition(p.f.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.p.f.rel + scale_fill_brewer(palette = "Paired")
+
+```
+
+###### INFO FOR PAPER
+```{r Eggerthella lenta}
+
+# Eggerthella is a bacterial genus of Actinobacteria
+
+# List of Genera in the Actinobacteria Phylum
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Actinobacteria
+print(taxa)
+
+# Abundance of Eggerthella
+head(abundances(pseq)["Eggerthella lenta et rel.",])
+
+pseq.bac <- subset_taxa(pseq, Phylum == "Actinobacteria")
+
+pseq.bac.g <- subset_taxa(pseq.bac, Genus == "Eggerthella lenta et rel.")
+
+# Pick the taxa abundances for a given sample
+samplename <- sample_names(pseq)[,"Eggerthella lenta et rel."]
+
+# Pick abundances for a particular taxon
+# eg.abundances <- abundances(pseq, transform='compositional')[, samplename]
+
+```
+
+```{r Coprobacillus cateniformis}
+
+# Coprobacillus cateniformis is a bacterium from the genus of Coprobacillus
+# Coprobacillus is of phylum Firmicutes 
+# Phylum = Firmicutes 
+# Class = Erysipelotrichi
+
+# This class of the Firmicutes Phylum isnt present in this data
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacilli
+print(taxa)
+
+```
+
+```{r Parabacteroides goldsteinii and Parabacteroides distasonis}
+
+# Parabacteroides goldsteinii 
+# Phylum = Bacteriodetes
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
+print(taxa)
+
+# Parabacteroides distasonis et rel. is present
+# RESEARCH DIFFERENCES
+
+```
+
+```{r Lactobacillus murinus and Lactobacillus gasseri}
+
+# Lactobacillus murinus and gasseri
+# Phylum = Firmicutes
+# Class = Bacilli
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacilli
+print(taxa)
+
+# "Lactobacillus gasseri et rel."     
+# "Lactobacillus plantarum et rel."  
+#       -  closely related
+# "Lactobacillus salivarius et rel." 
+#       -  closely related according to https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-018-0440-5
+# "Lactococcus"
+
+```
+
+```{r Prevotella melaninogenica}
+
+# Prevotella melaninogenica
+# Phylum = Bacteroidetes
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
+print(taxa)
+
+```
+
+```{r Eisenbergiella tayi}
+
+# Eisenbergiella tayi
+# Phylum = Firmicutes
+# Class = Clostridia
+# Order = Clostridiales
+# Family = Lachnospiraceae
+# Produce butyric acid
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$
+print(taxa)
+
+# Not found
+
+```
+
+```{r Subdoligranulum variabile}
+
+# Subdoligranulum variabile
+# Phylum = Clostridium cluster IV
+# Closely related to Faecalibacterium prausnitzii
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Clostridium_cluster_IV
+print(taxa)
+
+```
+
+```{r Ruminococcus torques}
+
+# Ruminococcus torques
+# Phylum = Firmicutes
+# Class = Clostridia
+# Order = Clostridiales
+# Family = Ruminococcaceae / Lachnospiraceae
+# Genus = Ruminococcus
+
+# Closely related to "Ruminococcus gnavus et rel."
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$
+print(taxa)
+
+```
+
+```{r Akkermansia muciniphila}
+
+# Akkermansia muciniphila
+# Phylum = Verrucomicrobia
+# Genus = Akkermansia
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Verrucomicrobia
+print(taxa)
+
+```
+
+```{r Paper relative abundance}
+
+paper <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacteroidetes"| Phylum =="Bacilli"  | Phylum =="Clostridium cluster IV"  | Phylum =="Verrucomicrobia")
+
+p.f <- aggregate_taxa(paper, "Phylum")
+
+p.f.rel <- microbiome::transform(p.f, "compositional")
+
+plot.p.f.rel <- plot_composition(p.f.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.p.f.rel + scale_fill_brewer(palette = "Paired")
+
+plot(plot.p.f.rel + scale_fill_brewer(palette = "Paired") )
+
+```
+
+```{r Paper relative abundance}
+
+paper <- subset_taxa(pseq, Genus =="Eggerthella lenta et rel." | Genus =="Parabacteroides distasonis et rel." | Genus =="Lactobacillus plantarum et rel."  | Genus =="Lactobacillus salivarius et rel." | Genus =="Prevotella melaninogenica et rel." | Genus =="Subdoligranulum variable at rel." | Genus =="Ruminococcus gnavus et rel." | Genus =="Akkermansia")
+
+p.g.rel <- microbiome::transform(paper, "compositional")
+
+plot.p.g.rel <- plot_composition(p.g.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.p.g.rel + scale_fill_brewer(palette = "Paired")
+
+plot(plot.p.g.rel + scale_fill_brewer(palette = "Paired") )
+
+```

### 2019-08-04 GlobalMicrobiome_NS.Rmd
updated

@@
 #test
+#test
+#clash 1
+#clash2maryam
+
 ```
@@
 ```
+
+###### INFO FOR PAPER - Specific to MND
+```{r Eggerthella lenta}
+
+# Eggerthella is a bacterial genus of Actinobacteria
+
+# List of Genera in the Actinobacteria Phylum
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Actinobacteria
+print(taxa)
+
+# Abundance of Eggerthella
+head(abundances(pseq)["Eggerthella lenta et rel.",])
+
+pseq.bac <- subset_taxa(pseq, Phylum == "Actinobacteria")
+
+pseq.bac.g <- subset_taxa(pseq.bac, Genus == "Eggerthella lenta et rel.")
+
+# Pick the taxa abundances for a given sample
+# samplename <- sample_names(pseq)[,"Eggerthella lenta et rel."]
+
+# Pick abundances for a particular taxon
+# eg.abundances <- abundances(pseq, transform='compositional')[, samplename]
+
+```
+
+```{r Coprobacillus cateniformis}
+
+# Coprobacillus cateniformis is a bacterium from the genus of Coprobacillus
+# Coprobacillus is of phylum Firmicutes 
+# Phylum = Firmicutes 
+# Class = Erysipelotrichi
+
+# This class of the Firmicutes Phylum isnt present in this data
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacilli
+print(taxa)
+
+```
+
+```{r Parabacteroides goldsteinii and Parabacteroides distasonis}
+
+# Parabacteroides goldsteinii 
+# Phylum = Bacteriodetes
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
+print(taxa)
+
+# Parabacteroides distasonis et rel. is present
+# RESEARCH DIFFERENCES
+
+```
+
+```{r Lactobacillus murinus and Lactobacillus gasseri}
+
+# Lactobacillus murinus and gasseri
+# Phylum = Firmicutes
+# Class = Bacilli
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacilli
+print(taxa)
+
+# "Lactobacillus gasseri et rel."     
+# "Lactobacillus plantarum et rel."  
+#       -  closely related
+# "Lactobacillus salivarius et rel." 
+#       -  closely related according to https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-018-0440-5
+# "Lactococcus"
+
+```
+
+```{r Prevotella melaninogenica}
+
+# Prevotella melaninogenica
+# Phylum = Bacteroidetes
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
+print(taxa)
+
+```
+
+```{r Eisenbergiella tayi}
+
+# Eisenbergiella tayi
+# Phylum = Firmicutes
+# Class = Clostridia
+# Order = Clostridiales
+# Family = Lachnospiraceae
+# Produce butyric acid
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$
+print(taxa)
+
+# Not found
+
+```
+
+```{r Subdoligranulum variabile}
+
+# Subdoligranulum variabile
+# Phylum = Clostridium cluster IV
+# Closely related to Faecalibacterium prausnitzii
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Clostridium_cluster_IV
+print(taxa)
+
+```
+
+```{r Ruminococcus torques}
+
+# Ruminococcus torques
+# Phylum = Firmicutes
+# Class = Clostridia
+# Order = Clostridiales
+# Family = Ruminococcaceae / Lachnospiraceae
+# Genus = Ruminococcus
+
+# Closely related to "Ruminococcus gnavus et rel."
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$
+print(taxa)
+
+```
+
+```{r Akkermansia muciniphila}
+
+# Akkermansia muciniphila
+# Phylum = Verrucomicrobia
+# Genus = Akkermansia
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Verrucomicrobia
+print(taxa)
+
+```
+
+```{r Paper relative abundance}
+
+paper <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacteroidetes"| Phylum =="Bacilli"  | Phylum =="Clostridium cluster IV"  | Phylum =="Verrucomicrobia")
+
+p.f <- aggregate_taxa(paper, "Phylum")
+
+p.f.rel <- microbiome::transform(p.f, "compositional")
+
+plot.p.f.rel <- plot_composition(p.f.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.p.f.rel + scale_fill_brewer(palette = "Paired")
+
+plot(plot.p.f.rel + scale_fill_brewer(palette = "Paired") )
+
+```
+
+```{r Paper relative abundance}
+
+paper <- subset_taxa(pseq, Genus =="Eggerthella lenta et rel." | Genus =="Parabacteroides distasonis et rel." | Genus =="Lactobacillus plantarum et rel."  | Genus =="Lactobacillus salivarius et rel." | Genus =="Prevotella melaninogenica et rel." | Genus =="Subdoligranulum variable at rel." | Genus =="Ruminococcus gnavus et rel." | Genus =="Akkermansia")
+
+p.g.rel <- microbiome::transform(paper, "compositional")
+
+plot.p.g.rel <- plot_composition(p.g.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.p.g.rel + scale_fill_brewer(palette = "Paired")
+
+plot(plot.p.g.rel + scale_fill_brewer(palette = "Paired") )
+
+paper@tax_table
+
+```

### 2019-08-04 MRD_final_microbiome .Rmd
updated

@@
 
-###### INFO FOR PAPER
+###### INFO FOR PAPER - Specific to MND
+
+```{r Taxa ranks in data set}
+
+# Show available ranks in the dataset
+rank_names(ps)
+
+```
+
+```{r Taxa ranks in data set}
+
+# Create table, number of features for each phyla
+table(tax_table(pseq)[, "Phylum"], exclude = NULL)
+
+```
+
+```{r Violin plot}
+
+# How many genera would be present after filtering?
+length(get_taxa_unique(pseq, taxonomic.rank = "Genus"))
+ps3 = tax_glom(pseq, "Genus", NArm = TRUE)
+
+## VIOLIN PLOT EXAMPLE
+
+# Defining a custom plot function, plot_abundance(), that uses phyloseq’s function to define a relative abundance graphic
+# use this to compare more easily differences in scale and distribution of the abundance values in our phyloseq object before and after transformation
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Eggerthella lenta et rel."))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
+
+```
+
+```{r Load packages}
+
+library("knitr")
+library("BiocStyle")
+
+```
+
 ```{r Eggerthella lenta}
@@
 
-# Pick the taxa abundances for a given sample
-samplename <- sample_names(pseq)[,"Eggerthella lenta et rel."]
-
-# Pick abundances for a particular taxon
-# eg.abundances <- abundances(pseq, transform='compositional')[, samplename]
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Eggerthella lenta et rel."))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
 
@@
 # RESEARCH DIFFERENCES
+
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Parabacteroides distasonis et rel."))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
 
@@
 
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Lactobacillus gasseri et rel."))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
+
 ```
@@
 
-# Prevotella melaninogenica
+# Prevotella melaninogenica et rel.
 # Phylum = Bacteroidetes
@@
 print(taxa)
+
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Prevotella melaninogenica et rel."))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
 
@@
 
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Faecalibacterium prausnitzii et rel."))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
+
 ```
@@
 
-taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$`Clostridium cluster XIVa`
 print(taxa)
+
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Ruminococcus gnavus et rel."))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
 
@@
 
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq, Genus %in% c("Akkermansia"))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
+
 ```
@@
 
-```
+paper@tax_table
+
+```
+
+Note: Use of transformed/normalised and/or filtered datasets is optional however, it is recommended.
+IPCO internally check if there are zero abundance features and also check for common features (OTUs/genus/species) between reference taxonomy and queried dataset.
+```{r IPCO}
+
+# 1. Load reference data
+install_github("https://github.com/IPCO-Rlibrary/IPCO")
+library(ade4)
+
+# 2. Normalise/Transform data
+# Hellinger transformation
+transform_data(pseq, relab)
+
+# 3. Inferring functionality
+# IPCO(R, L, Q)
+# R dataset where rows are functions/pathways and columns are samples
+# L dataset where rows are taxa/OTUs and columns are the same samples as dataset R
+# Q dataset where rows are taxa/OTUs common between L and Q and columns are the samples for which functionality will be inferred
+
+# Requires The references functional (R) and its paired taxonomy dataset (L) and the taxonomy dataset (Q) for which functions need to be inferred
+
+IPCO(R, L, Q)
+
+# 4. Checking for significant covariance between reference taxonomic and functional datasets 
+```

### 2019-08-09 Earth_microbiome_MS copy.Rmd
updated

@@
 #Create table dispaying the average prevalence of each phyla per sample and the total reads for all phyla in the data frame
-plyr::ddply(prevdf, "Phylum (unfiltered)", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
+plyr::ddply(prevdf, "Phylum", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
 

### 2019-08-09 IPCO_NS.Rmd
updated

---
title: "Untitled"
author: "NSharma"
date: "5 May 2016"
output: html_document
---

```{r setup, include=FALSE}

	library(foreign)
	library(stargazer)
	library(psych)
	library(plyr)
	library(xtable)
	library(car)
	#library(dplyr)
	library(scales)
	library(reshape) 
	library(gmodels)
	library(nnet)
	library(tidyr)
	library(sjmisc)
	library(RColorBrewer)
  library(htmltools)
	library(broom)

library(gcookbook)
library(tidyverse)
 #library(gdtools)

#The phyloseq package seeks to address issues with multiple microbiome analysis packages by providing a set of functions that internally manage the organizing, linking, storing, and analyzing of phylogenetic sequencing data. In general, this package is used for UniFrac analyses.
library(phyloseq)
library(microbiome)
library(knitr)
library(ggpubr)
#library(phylogeo)
library(dplyr)
#Graphing package used in phyloseq. To edit the default setting of a plot, you need to use functions in this package.
library(ggplot2)
library(viridis) 
library(cowplot)
# install.packages("devtools")
library(devtools)
# install_github("opisthokonta/tsnemicrobiota")
library(tsnemicrobiota)
# check you've installed this library
library(forcats)
library(IPCO)
library(phylosmith)
library(phyloseq)

data(atlas1006)   
pseq <- baseline(atlas1006)

##Step 1: Extract the required matrix from Phyloseq object
##first convert the OTU level phyloseq object to species/genus level phyloseq object. Depending on the phyloseq version the taxrank or taxlevel parameter would be used.
#taxlevel parameter is probably for older version to the best of my knowledge

#phyloseq_genus <- tax_glom(pseq, taxlevel="Genus")

#taxrank parameter is in the latest version
phyloseq_genus <- tax_glom(pseq, taxrank="Genus")

#If the above step is already done, then you can start from here.
#Convert phyloseq genus/species level object to matrix
genus_table <- as(otu_table(phyloseq_genus), "matrix")

##Step 2: Using IPCO
#load refernce data IPCO_Healthy/IPCO_HMP
library(IPCO)
data(IPCO_Healthy)

#Do normalisation/transform as suggested as an where required. If the phyloseq data is already normalised and doesnot wish to renormalise, then the user doesnot have to follow IPCO normalisation.
#We recommend Hellinger transformation. To be used on count/prop/relative abundance data
ref_genus <- transform_data(IPCO_Healthy$Genus,"relab")
ref_func <- transform_data(IPCO_Healthy$MetaCyc,"prop")
##or
ref_func <- transform_data(IPCO_Healthy$KEGG,"prop")

#if query genus table is count data
query_genus <- transform_data(genus_table,"count")

#Run IPCO
#Note: all samples should be in columns and features (func/taxa) in rows. If not, transpose the matrix
inferred_MetaCyc <- IPCO(ref_func,ref_genus,query_genus)

data <- data("IPCO_Healthy")

write_phyloseq(pseq, type = "all", path = getwd())

tdata <- transform_data(data, type = "proportion")
tdata <- transform_data(IPCO_Healthy, proportion)

```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-08-12 IPCO_NS.Rmd
updated

@@
 author: "NSharma"
-date: "5 May 2016"
+date: "2019"
 output: html_document
@@
 ```{r setup, include=FALSE}
-
-	library(foreign)
-	library(stargazer)
-	library(psych)
-	library(plyr)
-	library(xtable)
-	library(car)
-	#library(dplyr)
-	library(scales)
-	library(reshape) 
-	library(gmodels)
-	library(nnet)
-	library(tidyr)
-	library(sjmisc)
-	library(RColorBrewer)
-  library(htmltools)
-	library(broom)
 
@@
 #Do normalisation/transform as suggested as an where required. If the phyloseq data is already normalised and doesnot wish to renormalise, then the user doesnot have to follow IPCO normalisation.
+
 #We recommend Hellinger transformation. To be used on count/prop/relative abundance data
 ref_genus <- transform_data(IPCO_Healthy$Genus,"relab")
-ref_func <- transform_data(IPCO_Healthy$MetaCyc,"prop")
+ref_func <- transform_data(IPCO_Healthy$MetaCyc,"proportion")
+
+#ref_func <- filter_functionality(ref_func, "metacyc",IPCO_Healthy$MetaCyc_coverage,threshold=0.4) # Filtering
+
+ref_func <- filter_functionality(ref_func, "metacyc", IPCO_Healthy$MetaCyc_coverage)
 ##or
-ref_func <- transform_data(IPCO_Healthy$KEGG,"prop")
+#ref_func <- transform_data(IPCO_Healthy$KEGG,"proportion")
 
 #if query genus table is count data
-query_genus <- transform_data(genus_table,"count")
+#query_genus <- transform_data(genus_table,"count")
 
@@
 #Note: all samples should be in columns and features (func/taxa) in rows. If not, transpose the matrix
-inferred_MetaCyc <- IPCO(ref_func,ref_genus,query_genus)
+inferred_MetaCyc <- IPCO(ref_func,ref_genus,genus_table)
 
-data <- data("IPCO_Healthy")
+library(made4)
 
-write_phyloseq(pseq, type = "all", path = getwd())
+overview(inferred_MetaCyc)
 
-tdata <- transform_data(data, type = "proportion")
-tdata <- transform_data(IPCO_Healthy, proportion)
+k.coa<- ord(inferred_MetaCyc, type="coa")
+summary(k.coa$ord)
+
+plotgenes(k.coa)
+
+heatmap(inferred_MetaCyc)
+
+p <- plot_ly(inferred_MetaCyc = volcano, type = "heatmap")
+
+levelplot(inferred_MetaCyc[1:ncol(inferred_MetaCyc),ncol(inferred_MetaCyc):1])
+
+coinertia_resuls <- check_coinertia(ref_func,genus_table)
 
@@
 ## R Markdown
-
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
-```
-
-## Including Plots
-
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-08-23 IPCO_NS.Rmd
updated

@@
 ```{r setup, include=FALSE}
+
+setwd("/Users/Nik/globalmicrobiome")
 
@@
 
+# Create mnd variable
+# grabs the nationality from phyloseq
+con.class <- get_variable(pseq, "nationality" )
+
+# check you've installed this library
+library(forcats)
+
+# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
+mnd.class <- fct_collapse(con.class, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+
+# reorder
+mnd.class <- factor(mnd.class, levels = (c("LOW", "MEDIUM", "HIGH")))
+
 library(made4)
+library(ade4)
 
@@
 
+```
+
+```{r , include=FALSE}
+
+#p <- overview(inferred_MetaCyc,classvec=mnd.class, labels=c("red", "blue", "yellow"))
+
+#Correspondence Analysis
+#The function ord simplifies the running of ordination methods such as principal component, correspondence or non-symmetric correspondence analysis. It provides a wrapper
+#which can call each of these methods in ade4. To run a correspondence analysis [6] on
+#this dataset.
+
 k.coa<- ord(inferred_MetaCyc, type="coa")
 summary(k.coa$ord)
+p <- plot(k.coa)
+
+p <- plot(k.coa, classvec=mnd.class, arraycol=c("red", "blue", "yellow" ), genecol="grey3")
+
+```
+
+```{r , include=FALSE}
+plotarrays(k.coa, arraylabels=mnd.class)
 
 plotgenes(k.coa)
+
+k.coa2<-ord(inferred_MetaCyc, classvec=mnd.class)
+p <- plot(k.coa2)
+
+#Classification and Class Prediction using Between Group
+#Analysis
+#Between Group Analysis (BGA) is a supervised classification method [3]. The basis
+#of BGA is to ordinate the groups rather than the individual samples. In tests on two
+#microarray gene expression datasets, BGA performed comparably to supervised classification methods, including support vector machines and artifical neural networks [2].
+#To train a dataset, use bga, the projection of test data can be assessed using suppl.
+#One leave out cross validation can be perfo
+
+k.bga<-bga(inferred_MetaCyc, type="coa", classvec=mnd.class)
+between.graph(k.bga, ax=1) # Show the separation on the first axes(ax
+
+#Meta-analysis of microarray gene expression
+#Coinertia analysis cia [4] has been successfully applied to the cross-platform comparison
+#(meta-analysis) of microarray gene expression datasets [8]. CIA is a multivariate method
+#that identifies trends or co-relationships in multiple datasets which contain the same
+#samples. That is either the rows or the columns of a matrix must be ”matchable”. CIA
+#can be applied to datasets where the number of variables (genes) far exceeds the number
+#of samples (arrays) such is the case with microarray analyses. cia calls coinertia in
+#the ade4 package. See the CIA vignette for more details on this method.
+
+#By default the variables (genes) are labelled with the rownames of the matrix. Typically these are spot IDs or Affymetrix accession numbers which are not very easy to
+#interpret. But these can be easily labeled by your own labels. For example its often
+#useful to labels using HUGO gene symbols. We find the Bioconductor annotate and
+#annaffy annotation packages are very useful for this. Alternatively we also use biomaRt
+#or Resourcerer or the Stanford Source database.
+#In this example we provide annotation from the Source database in khan$annotation.
+#The gene symbol are in the column khan$annotation$Symbol
+
+#plot(k.bga, genelabels=gene.symbs)    need to define gene;symbs
 

### 2019-08-23 IPCO_NS.Rmd
updated

@@
 ---
+
+Things to complete in this script:
+
+1. group the pathways
+
+2. Will this tidy up the figures?
+
+3. tidy up the figures anyway!
+
+4. stats
 
@@
 ```{r , include=FALSE}
-plotarrays(k.coa, arraylabels=mnd.class)
+#   plotarrays(k.coa, arraylabels=mnd.class)
 
@@
 p <- plot(k.coa2)
+```
+
+```{r , include=FALSE}
 
@@
 #plot(k.bga, genelabels=gene.symbs)    need to define gene;symbs
+```
+
+```{r , include=FALSE}
 
@@
 
-p <- plot_ly(inferred_MetaCyc = volcano, type = "heatmap")
+library(pheatmap)
 
-levelplot(inferred_MetaCyc[1:ncol(inferred_MetaCyc),ncol(inferred_MetaCyc):1])
+pheatmap(inferred_MetaCyc)
+
+pheatmap(inferred_MetaCyc, cluster_cols = mnd.class)
 

### 2019-09-27 curatedMetagenomic_NS.Rmd
updated

---
title: "curatedMetagenomicData"
author: "NSharma"
date: "26/09/2019"
output: html_document
---

```{r setup, include=FALSE}

#BiocManager::install("curatedMetagenomicData")
#BiocManager::install("ExperimentHub")
library(ExperimentHub)
library(curatedMetagenomicData)
suppressPackageStartupMessages(library(ExperimentHub))
library(phyloseq)
library(microbiome)

eh = ExperimentHub()
myquery = query(eh, "curatedMetagenomicData")

?combined_metadata
View(combined_metadata)
table(combined_metadata$disease)

# Get the stool samples
esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)

# merge
eset <- mergeData(esl)

head( pData(eset ) )

pseq = ExpressionSet2phyloseq(eset )

summarize_phyloseq(pseq)

sample_data( pseq )[1:6, 1:5]

pseq.subset <- subset_samples(pseq, country == "SWE")

pseq.subset <- subset_samples(pseq, disease == "healthy" & age_category =="adult")

s <- rownames(subset(meta(pseq), country == "SWE" ))

sample_variables(pseq)
# get the countres
head(get_variable(pseq, sample_variables(pseq)[5]))

temp1 <- get_variable(pseq, sample_variables(pseq)[5])

```

## R Markdown

### 2019-09-30 curatedMetagenomic_NS.Rmd
updated

@@
 
-sample_data( pseq )[1:6, 1:5]
-
 pseq.subset <- subset_samples(pseq, country == "SWE")
-
 pseq.subset <- subset_samples(pseq, disease == "healthy" & age_category =="adult")
@@
 ## R Markdown
+
+```{r}
+# Create mnd variable
+# grabs the nationality from phyloseq
+test <- get_variable(pseq, "country" )
+
+# check you've installed this library
+library(forcats)
+table(test)
+# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
+
+# Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,
+
+test <- fct_collapse(test, LOW = c("NOR","SWE","FIN") , MEDIUM = c("ITA","AUT","DEU","GBR","ESP","FRA","LUX","NLD"), HIGH = c("USA") )
+
+#test <- fct_collapse(test, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+
+# reorder
+test <- factor(test, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(test)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(pseq)$mnd = test
+
+# checks that it has worked.
+get_variable(pseq, "mnd")
+
+```
+
+```{r MND table summary}
+
+table(sample_data(pseq)$mnd)
+
+table(sample_data(pseq)$country)
+
+```
+## Check OTUs/Phylums
+Prune(remove) taxa because we want to remove OTUs/taxa that aren't in any of the samples
+Don't need to prune this taxa because OTUs are found in all samples.
+Check this everytime you use the subset_samples function because of the pseq object being filtered. 
+```{r check OTUs}
+# Check if any OTUs are not present in any samples
+any(taxa_sums(pseq) == 0)
+# FALSE, therefore, OTUs are found in all samples and we don't need to prune the taxa.
+If the answer is TRUE, some OTUs are not present in any samples.
+This is usually the case when data is subset to remove some samples.
+OTUs unique to those sample are not removed along with the samples.
+Therefore, it is important to check this everytime the phyloseq object is filtered for samples using `subset_samples` function.
+
+head(meta(pseq))
+
+```
+
+## Prepare Data for Alpha Diversity Visualisation
+```{r prepare data for alpha diversity visualisation}
+
+# Prepare data for vizualisation
+
+# Alpha diversity again
+png("alpha.png")
+
+pseq2 <- pseq
+
+#this is a work around. Need to check
+otu_table(pseq2) <- otu_table(round(as((otu_table(pseq2)), "matrix")), taxa_are_rows(pseq2))
+estimate_richness(pseq2)
+
+tab <- diversities(pseq2, index = "all")
+stargazer(tab, type = "text",summary = NULL)
+kable(head(tab))
+plot(tab)
+dev.off()
+
+# Get metadata
+pseq.meta <- meta(pseq)
+kable(head(pseq.meta))
+stargazer(pseq.meta, type = "text",summary = NULL)
+
+# Add the diversity table to metadata
+pseq.meta$Shannon <- tab$shannon 
+pseq.meta$InverseSimpson <- tab$inverse_simpson
+pseq.meta$fisher <- tab$fisher
+pseq.meta$chao1 <- tab$chao1
+
+```

### 2019-10-01 curatedMetagenomic_NS.Rmd
updated

@@
 
-head( pData(eset ) )
+# the relab=FALSE is essential as it Absolute Raw Count Data
+pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
 
-pseq = ExpressionSet2phyloseq(eset )
-
-summarize_phyloseq(pseq)
-
-pseq.subset <- subset_samples(pseq, country == "SWE")
-pseq.subset <- subset_samples(pseq, disease == "healthy" & age_category =="adult")
-
-s <- rownames(subset(meta(pseq), country == "SWE" ))
-
-sample_variables(pseq)
-# get the countres
-head(get_variable(pseq, sample_variables(pseq)[5]))
-
-temp1 <- get_variable(pseq, sample_variables(pseq)[5])
+#For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object.
+curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:
 
@@
 # checks that it has worked.
-get_variable(pseq, "mnd")
+get_variable(pseq.subset, "mnd")
 
@@
 
-table(sample_data(pseq)$mnd)
+table(sample_data(pseq.subset)$mnd)
 
-table(sample_data(pseq)$country)
+table(sample_data(pseq.subset)$country)
 
@@
 # Check if any OTUs are not present in any samples
-any(taxa_sums(pseq) == 0)
+any(taxa_sums(pseq.subset) == 0)
 # FALSE, therefore, OTUs are found in all samples and we don't need to prune the taxa.
@@
 
+# https://bioconductor.riken.jp/packages/3.5/data/experiment/vignettes/curatedMetagenomicData/inst/doc/curatedMetagenomicData.html
 # Prepare data for vizualisation
+
+# Remove   everything without a country
+pseq.subset <- subset_samples(pseq, !is.na(mnd))
+
+# Pick the core (>0.1% relative abundance in >50% of the samples)
+pseq.csubset <- core(pseq.subset, detection = 0.1/100, prevalence = 50/100)
+
+any(taxa_sums(pseq.csubset) == 0)
+
+fpseq.core <- subset_taxa( pseq.core, !is.na(Phylum))
+
+keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
+summary(keepotu)
+
+fpseq <- subset_taxa(pseq, keepotu)
+
+loman.filt = subset_taxa(fpseq, keepotu & !is.na(Strain))
+plot_heatmap(fpseq, method="PCoA", distance="bray")
+
+#keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
+#summary(keepotu)
+
+# subset_taxa(loman.pseq, keepotu)
+#lfilt <- subset_taxa(pseq, keepotu & !is.na(Strain))
+#plot_heatmap(lfilt, method="PCoA", distance="bray")
+
+#lfilt <- subset_taxa(pseq, is.na(Class) & !is.na(Phylum))
+
+#any(taxa_sums(lfilt) == 0)
 
@@
 
-pseq2 <- pseq
+tab <- diversities(pseq.csubset)
 
-#this is a work around. Need to check
-otu_table(pseq2) <- otu_table(round(as((otu_table(pseq2)), "matrix")), taxa_are_rows(pseq2))
-estimate_richness(pseq2)
-
-tab <- diversities(pseq2, index = "all")
+tab <- diversities(pseq.csubset, index = "all")
 stargazer(tab, type = "text",summary = NULL)

### 2019-10-04 curatedMetagenomic_NS.Rmd
updated

@@
 
+wt = UniFrac(pseq.csubset, weighted=TRUE, normalized=FALSE, 
+             parallel=FALSE, fast=TRUE)
+plot(hclust(wt), main="Weighted UniFrac distances")
+
 any(taxa_sums(pseq.csubset) == 0)
@@
 
-loman.filt = subset_taxa(fpseq, keepotu & !is.na(Strain))
-plot_heatmap(fpseq, method="PCoA", distance="bray")
+# This doesnt run. 
+# plot_heatmap(fpseq, method="PCoA", distance="bray")
 
@@
 png("alpha.png")
+
+barplot(sort(taxa_sums(pseq.csubset), TRUE)[1:20] / nsamples(pseq.csubset),
+        ylab = "Total counts", las = 2)
 

### 2019-10-04 MRD_final_microbiome .Rmd
updated

@@
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-test <- fct_collapse(test, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+# test <- fct_collapse(test, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+
+test <- fct_collapse(test, LOW = c("EasternEurope", "SouthEurope") , MEDIUM = c("CentralEurope"), HIGH = c("US", "UKIE", "Scandinavia") )
 
@@
 rank_names(ps)
+
+test <- fct_collapse(test, LOW = c("EasternEurope", "SouthEurope") , MEDIUM = c("CentralEurope"), HIGH = c("US", "UKIE", "Scandinavia") )
+
+# reorder
+test <- factor(test, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(test)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(pseq)$mnd = test
+
+# checks that it has worked.
+get_variable(pseq, "mnd")
 
@@
 
+##IPCO
 Note: Use of transformed/normalised and/or filtered datasets is optional however, it is recommended.
@@
 # Hellinger transformation
-transform_data(pseq, relab)
+# transform_data(pseq, relab)
+
+temp <- data("IPCO_Healthy")
 

### 2019-10-04 curatedMetagenomic_NS.Rmd
updated

@@
 pseq.subset <- subset_samples(pseq, !is.na(mnd))
+pseq.subset <- subset_samples(pseq, disease == "healthy")
 
+keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
+f.pseq <- subset_taxa(pseq, keepotu)
+
+any(taxa_sums(pseq.c) == 0)
+
+genus.pseq <- subset_taxa( pseq.subset, !is.na(Genus))
+phylum.pseq <- subset_taxa( pseq, is.na(Class) & !is.na(Phylum))
+
+taxonomy <- tax_table(genus.pseq)
 # Pick the core (>0.1% relative abundance in >50% of the samples)
-pseq.csubset <- core(pseq.subset, detection = 0.1/100, prevalence = 50/100)
+pseq.subsetc <- core(pseq.subset, detection = 0.1/100, prevalence = 50/100)
 
-wt = UniFrac(pseq.csubset, weighted=TRUE, normalized=FALSE, 
-             parallel=FALSE, fast=TRUE)
-plot(hclust(wt), main="Weighted UniFrac distances")
+map_levels(NULL, "Phylum","Genus",pseq.c)
 
-any(taxa_sums(pseq.csubset) == 0)
+diversities(phylum.pseq, index = "all")
 
-fpseq.core <- subset_taxa( pseq.core, !is.na(Phylum))
+png("alpha.png")
+
+tab <- diversities(pseq.c, index = "all")
+stargazer(tab, type = "text",summary = NULL)
+
+tab <- diversities(pseq.subsetc, index = "Shannon")
+tab <- diversities(phylum.pseq, index = "all")
+stargazer(tab, type = "text",summary = NULL)
+kable(head(tab))
+plot(tab)
+dev.off()
+
+# Get metadata
+pseq.meta <- meta(phylum.pseq)
+kable(head(pseq.meta))
+stargazer(pseq.meta, type = "text",summary = NULL)
+
+# Add the diversity table to metadata
+pseq.meta$Shannon <- tab$shannon 
+pseq.meta$InverseSimpson <- tab$inverse_simpson
+pseq.meta$fisher <- tab$fisher
+pseq.meta$chao1 <- tab$chao1
+#wt = UniFrac(pseq.csubset, weighted=TRUE, normalized=FALSE, 
+#             parallel=FALSE, fast=TRUE)
+#plot(hclust(wt), main="Weighted UniFrac distances")
+
+library(ggpubr)
+
+p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon",
+ add = "boxplot", fill = "mnd", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+  print(p1)
+ggsave("nation_violin_shannon.png")
 
@@
 
-tab <- diversities(pseq.csubset, index = "all")
+tab <- diversities(pseq.subset, index = "all")
 stargazer(tab, type = "text",summary = NULL)

### 2019-10-05 curatedMetagenomic_NS.Rmd
test

@@
 library(microbiome)
+library(forcats)
 
@@
 # grabs the nationality from phyloseq
-test <- get_variable(pseq, "country" )
+mnd.country <- get_variable(pseq, "country" )
+table(mnd.country)
 
-# check you've installed this library
-library(forcats)
-table(test)
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-
 # Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,
 
-test <- fct_collapse(test, LOW = c("NOR","SWE","FIN") , MEDIUM = c("ITA","AUT","DEU","GBR","ESP","FRA","LUX","NLD"), HIGH = c("USA") )
+mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN") , MEDIUM = c("ITA","AUT","DEU","GBR","ESP","FRA","LUX","NLD"), HIGH = c("USA") )
 
-#test <- fct_collapse(test, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
 
 # reorder
-test <- factor(test, levels = (c("LOW", "MEDIUM", "HIGH")))
-levels(test)
+mnd.country <- factor(mnd.country, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(mnd.country)
 
 # creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = test
+sample_data(pseq)$mnd = mnd.country
 
 # checks that it has worked.
-get_variable(pseq.subset, "mnd")
+get_variable(pseq, "mnd")
 

### 2019-10-06 curatedMetagenomic_NS.Rmd
updated

@@
 library(forcats)
+library(stargazer)
+library(ggpubr)
 
@@
 myquery = query(eh, "curatedMetagenomicData")
-
-?combined_metadata
-View(combined_metadata)
-table(combined_metadata$disease)
 
@@
 
-table(sample_data(pseq.subset)$mnd)
+table(sample_data(pseq)$mnd)
 
-table(sample_data(pseq.subset)$country)
+table(sample_data(pseq)$country)
 
@@
 # Check if any OTUs are not present in any samples
-any(taxa_sums(pseq.subset) == 0)
+any(taxa_sums(pseq) == 0)
 # FALSE, therefore, OTUs are found in all samples and we don't need to prune the taxa.
@@
 # Remove   everything without a country
-pseq.subset <- subset_samples(pseq, !is.na(mnd))
-pseq.subset <- subset_samples(pseq, disease == "healthy")
 
 keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
-f.pseq <- subset_taxa(pseq, keepotu)
+pseq.f <- subset_taxa(pseq, keepotu)
+
+pseq.c <- core(pseq, detection = 0.1/100, prevalence = 50/100)
+
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn")
+
+# These are useful is filtering. 
+#sample_variables(pseq.subset)
+#get_variable(pseq.subset, sample_variables(pseq.subset)[4])
 
@@
 
-genus.pseq <- subset_taxa( pseq.subset, !is.na(Genus))
-phylum.pseq <- subset_taxa( pseq, is.na(Class) & !is.na(Phylum))
-
-taxonomy <- tax_table(genus.pseq)
-# Pick the core (>0.1% relative abundance in >50% of the samples)
-pseq.subsetc <- core(pseq.subset, detection = 0.1/100, prevalence = 50/100)
-
-map_levels(NULL, "Phylum","Genus",pseq.c)
-
-diversities(phylum.pseq, index = "all")
-
-png("alpha.png")
-
-tab <- diversities(pseq.c, index = "all")
-stargazer(tab, type = "text",summary = NULL)
-
-tab <- diversities(pseq.subsetc, index = "Shannon")
-tab <- diversities(phylum.pseq, index = "all")
-stargazer(tab, type = "text",summary = NULL)
-kable(head(tab))
-plot(tab)
-dev.off()
+# run the diversities. 	keep an eye on this as only one is working
+tab <- diversities(pseq.subset, index = "Shannon")
 
 # Get metadata
-pseq.meta <- meta(phylum.pseq)
+pseq.meta <- meta(pseq.subset)
 kable(head(pseq.meta))
@@
 pseq.meta$Shannon <- tab$shannon 
-pseq.meta$InverseSimpson <- tab$inverse_simpson
-pseq.meta$fisher <- tab$fisher
-pseq.meta$chao1 <- tab$chao1
-#wt = UniFrac(pseq.csubset, weighted=TRUE, normalized=FALSE, 
-#             parallel=FALSE, fast=TRUE)
-#plot(hclust(wt), main="Weighted UniFrac distances")
-
-library(ggpubr)
+#pseq.meta$InverseSimpson <- tab$inverse_simpson
+#pseq.meta$fisher <- tab$fisher
+#pseq.meta$chao1 <- tab$chao1
 
@@
   print(p1)
-ggsave("nation_violin_shannon.png")
+#ggsave("nation_violin_shannon.png")
 
-keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
-summary(keepotu)
-
-fpseq <- subset_taxa(pseq, keepotu)
-
-# This doesnt run. 
-# plot_heatmap(fpseq, method="PCoA", distance="bray")
-
-#keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
-#summary(keepotu)
-
-# subset_taxa(loman.pseq, keepotu)
-#lfilt <- subset_taxa(pseq, keepotu & !is.na(Strain))
-#plot_heatmap(lfilt, method="PCoA", distance="bray")
-
-#lfilt <- subset_taxa(pseq, is.na(Class) & !is.na(Phylum))
-
-#any(taxa_sums(lfilt) == 0)
-
-# Alpha diversity again
-png("alpha.png")
-
-barplot(sort(taxa_sums(pseq.csubset), TRUE)[1:20] / nsamples(pseq.csubset),
-        ylab = "Total counts", las = 2)
-
-tab <- diversities(pseq.csubset)
-
-tab <- diversities(pseq.subset, index = "all")
-stargazer(tab, type = "text",summary = NULL)
-kable(head(tab))
-plot(tab)
-dev.off()
-
-# Get metadata
-pseq.meta <- meta(pseq)
-kable(head(pseq.meta))
-stargazer(pseq.meta, type = "text",summary = NULL)
-
-# Add the diversity table to metadata
-pseq.meta$Shannon <- tab$shannon 
-pseq.meta$InverseSimpson <- tab$inverse_simpson
-pseq.meta$fisher <- tab$fisher
-pseq.meta$chao1 <- tab$chao1
+# THis subsets the  genus and phylum. 
+#genus.pseq <- subset_taxa( pseq.subset, !is.na(Genus))
+#phylum.pseq <- subset_taxa( pseq, is.na(Class) & !is.na(Phylum))
 

### 2019-10-08 curatedMetagenomic_NS.Rmd
updated

@@
 pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
-
+orig_pseq <- pseq
 #For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object.
@@
 #sample_variables(pseq.subset)
-#get_variable(pseq.subset, sample_variables(pseq.subset)[4])
+#get_variable(pseq.subset, sample_variables(pseq.subset)[13])
 
@@
   print(p1)
+
+  ggsave("nation_violin_shannon.png")
 #ggsave("nation_violin_shannon.png")
@@
 ```
+
+## Nationality Violin Plot
+```{r nationality violin plot}
+
+# Alpha diversity is the species richness of a place
+# High alpha diversity = high no. of species
+
+#pseq.meta$nationality <- factor(pseq.meta$country, levels = c("Scandinavia","EasternEurope","CentralEurope", "SouthEurope", "UKIE", "US"))
+
+# Violin plot
+# the number in the palettes must equal the number of groups in nationlaity.pairs
+#p1 <- ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+# add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+#   print(p1)
+# ggsave("nation_violin_shannon.png")
+
+# p2 <- ggviolin(pseq.meta, x = "nationality", y = "InverseSimpson",
+#  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+# print(p2)
+# ggsave("nation_violin_simpson.png")
+
+# p3 <- ggviolin(pseq.meta, x = "nationality", y = "fisher",
+#  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("") + ylab("Fisher")
+# print(p3)
+# ggsave("nation_violin_fisher.png")
+
+# plot_grid(p1, p2, p3, align = "h" , ncol = 2, nrow = 2)
+# ggsave("nation_violin.png")
+
+```
+
+```{r}
+
+# 
+pseq <- pseq.subset
+# This isnt working. 
+p <- plot_richness(pseq, x="country", measures=c( "Shannon"))
+
+ggsave("alpha_nationality.png")
+
+```
+
+## MND Violin Plot
+```{r mnd violin plot}
+
+# Add the diversity table to metadata
+pseq.meta$Shannon <- tab$shannon
+#pseq.meta$InverseSimpson <- tab$inverse_simpson
+
+# Compare differences in Shannon index between mnd group of the study subjects
+pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+
+# Create a list of pairwise comaprisons
+mnd <- levels(pseq.meta$mnd) # get the variables
+
+# Make a pairwise list that we want to compare.
+mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+# Shannon
+# the number in the palettes must equal the number of groups in bmi.pairs
+p1 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+p1 <- p1 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+print(p1)
+ggsave("mnd_violin__shannon_stats.png")
+
+# Inverse Simpson
+#p2 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
+#p2 <- p2 + stat_compare_means(comparisons = mnd.pairs) 
+#print(p2)
+#ggsave("mnd_violin_simpson.png")
+
+```
+
+## BETA DIVERSITY
+Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
+WE want to: quantify community divergence within a given sample set
+Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.
+
+```{r Phylum NMDS}
+
+png("Phylum_NMDS.png")
+pseq.ord <- ordinate(pseq, "NMDS", "bray")
+
+p1 = plot_ordination(pseq, pseq.ord, type = "taxa", color = "mnd", title = "Taxa by Phylum")
+p1 + facet_wrap(~Phylum, 3)
+dev.off()
+
+```
+
+Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
+This was applied in group-level comparisons for instance in Salonen et al.
+ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
+```{r Calculate group divergences}
+
+# Use these to compare microbiota divergence within each group
+
+p.low <- subset_samples(pseq, mnd == "LOW")
+p.low.d <- divergence(p.low)
+
+p.med <- subset_samples(pseq, mnd == "MEDIUM")
+p.med.d <- divergence(p.med)
+
+p.high <- subset_samples(pseq, mnd == "HIGH")
+p.high.d <- divergence(p.high)
+
+p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+plot(p)
+
+```
+ The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
+
+# VISUALISE MICORBIOME VARIATION
+```{r convert to compositional data}
+
+# Convert to compositional data
+pseq.rel <- microbiome::transform(pseq, "compositional")
+otu <- abundances(pseq.rel)
+meta <- meta(pseq.rel)
+
+```
+
+NMDS/Bray
+Bray-Curtis dissimilarity does not consider phylogenetic relationships between OTUs. There are several distance methods and a list can be obtained by typying ?distanceMethodList in the console pane.
+```{r}
+
+pseq.ord <- ordinate(pseq.rel, "NMDS", "bray")
+
+p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
+plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+ggsave("NMDSt.png")
+
+p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+ggsave("NMDS_facet.png")
+
+```
+
+MDS ordination on Bray-Curtis distance
+Performs principal coordinate analysis (also called principle coordinate decomposition, multidimensional scaling (MDS), or classical scaling) of a distance matrix (Gower 1966), including two correction methods for negative eigenvalues.
+See pcoa for further details.
+```{r}
+
+pseq.ord <- ordinate(pseq.rel, "MDS", "bray")
+
+p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
+plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+ggsave("MDS.png")
+
+p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
+ggsave("MDS_facet.png")
+
+```
+
+PCoA for compositional data with Bray-Curtis distances
+```{r}
+
+p <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+p2 <- p + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b")
+
+p1 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+print(p1)
+
+```
+
+```{r PCoA Bray}
+
+# Pick core taxa
+pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)
+
+# PCoA for compositional data with Bray-Curtis distances
+png("PCoA_bray.png")
+p <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+                      method = "PCoA", distance = "bray") +
+       labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+print(p)
+dev.off()
+
+```
+
+t-SNE is a popular new ordination technique
+, title='t-SNE (Bray)'
+```{r t-SNE}
+
+tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
+  color = 'mnd')
+tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
+ggsave("t_sne.png")
+
+```
+
+## PERMANOVA significance test for group-level differences
+Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
+
+```{r Redefine data}
+
+# Convert to compositional data
+pseqLvH <- merge_phyloseq(p.low,p.high)
+
+pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+otu <- abundances(pseq.rel)
+meta <- meta(pseq.rel)
+
+```
+
+```{r PERMANOVA}
+
+# samples x species as input
+library(vegan)
+permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+
+# P-value
+print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+
+```
+
+Check that variance homogeneity assumptions hold (to ensure the reliability of the results):Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
+```{r ANOVA}
+
+dist <- vegdist(t(otu))
+anov <- anova(betadisper(dist, meta$mnd))
+summary(anov)
+tidy(anov)
+
+```
+
+Show coefficients for the top taxa separating the groups
+```{r top taxa L vs H}
+
+coef <- coefficients(permanova)["mnd1",]
+top.coef <- coef[rev(order(abs(coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
+ggsave("LvH.png")
+
+```
+
+```{r HvL}
+pseqHvL <- merge_phyloseq(p.high,p.low)
+
+pseq.rel <- microbiome::transform(pseqHvL, "compositional")
+otu <- abundances(pseq.rel)
+meta <- meta(pseq.rel)
+
+library(vegan)
+permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=99, method = "bray")
+
+# P-value
+print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+
+dist <- vegdist(t(otu))
+anov <- anova(betadisper(dist, meta$mnd))
+summary(anov)
+tidy(anov)
+
+coef <- coefficients(permanova)["mnd1",]
+top.coef <- coef[rev(order(abs(coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
+
+```
+
+```{r L vs M}
+
+pseqLvM <- merge_phyloseq(p.low,p.med)
+
+pseq.rel <- microbiome::transform(pseqLvM, "compositional")
+otu <- abundances(pseq.rel)
+meta <- meta(pseq.rel)
+
+library(vegan)
+permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+
+# P-value
+print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+
+dist <- vegdist(t(otu))
+anov <- anova(betadisper(dist, meta$mnd))
+summary(anov)
+tidy(anov)
+
+coef <- coefficients(permanova)["mnd1",]
+top.coef <- coef[rev(order(abs(coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
+ggsave("LvM.png")
+
+```
+
+## MICROBIOME COMPOSITION
+
+```{r Phylum #}
+
+# Show available ranks in the dataset
+rank_names(pseq)
+
+# Create table, number of features for each phyla
+table(tax_table(pseq)[, "Phylum"], exclude = NULL)
+
+```
+This shows a few phyla for which only one feature was observed. Those may be worth filtering.
+
+```{r Total Phylum abundnace bar}
+
+png("Phylum_abundance.png")
+p <- plot_bar(pseq,"Phylum") + theme(axis.text.x = element_text(face="italic"))
+print(p)
+dev.off()
+
+```
+
+```{r Phylum prevalence table}
+
+prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
+prev.otu + theme(legend.position="none")
+ggsave("phylum_prevalence.png")
+
+tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
+tab + theme(legend.position="none")
+ggsave("Phylum_prevalence_detection5.png")
+
+```
+
+```{r Relative abundance}
+
+pseq.ph <- aggregate_taxa(pseq.new, "Phylum")
+
+pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
+theme(legend.position = "bottom") + theme_bw() +
+theme(axis.text.x=element_blank()) +
+c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
+
+print(plot.comp.rel)
+
+ggsave("phylum_relative_abundance11.png")
+
+plot.comp.rel + theme(legend.position="none")
+ggsave("phylum_relative_abundance12.png")
+
+```
+
+guide_italics
+ggtitle("Relative abundance") +  
+```{r}
+
+clos.ph <- aggregate_taxa(clostridium, "Phylum")
+
+clos.ph.rel <- microbiome::transform(clos.ph, "compositional")
+
+plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
+  theme(legend.position = "bottom", ) + theme_bw() + 
+  theme(axis.text.x=element_blank()) + 
+  theme(legend.title = element_text(size=18), legend.text = element_text( face="italic")) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend(title="Phylum"))
+
+print(plot.comp.rel)
+
+ggsave("phylum_relative_abundance_clos.png")
+
+plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none") + guide_italics
+ggsave("phylum_relative_abundance_clos2.png")
+
+```
+
+```{r Phylum Bar Plot}
+
+levels(pseq.meta$mnd)
+
+pseq.ph <- aggregate_taxa(pseq, "Phylum")
+
+pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+ggsave("phylum_abundance_bar.png")
+
+```
+
+```{r}
+
+top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
+
+top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
+
+pseq.ph.rel <- microbiome::transform(top5P, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+```
+
+```{r}
+
+top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
+
+top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
+
+pseq.ph.rel <- microbiome::transform(top10P, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+```
+
+```{r pseq.new bar plot}
+
+p.new <- aggregate_taxa(pseq.new, "Phylum")
+
+p.new.rel <- microbiome::transform(p.new, "compositional")
+
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
+ggsave("pnew_abundance_bar.png")
+
+```
+
+```{r CLostridium Bar Plot}
+
+clos <- aggregate_taxa(clostridium, "Phylum")
+
+pseq.cl.rel <- microbiome::transform(clos, "compositional")
+
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.cl.rel + scale_fill_brewer(palette = "Paired")
+ggsave("cl_abundance_bar.png")
+
+```
+
+```{r Clostridium genus bar plots}
+
+pseq.cl.rel <- microbiome::transform(clostridium, "compositional")
+
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
+
+plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")
+
+```
+
+```{r Top Phylum Heatmap}
+
+pseq.ph <- aggregate_taxa(pseq, "Phylum")
+
+pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
+ggsave("phylum_histogram.png")
+
+```
+
+```{r p.new heatmap}
+
+pseq.n.ph <- aggregate_taxa(pseq.new, "Phylum")
+
+pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
+
+plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+p <- plot.n.comp.rel + coord_flip()
+p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
+ggsave("pnew_histogram.png")
+
+```
+
+```{r clostridium heatmap}
+
+pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")
+
+pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")
+
+plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+p <- plot.c.comp.rel + coord_flip()
+p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
+ggsave("clos_histogram.png")
+
+```
+
+```{r Network}
+
+png("network_bray_phylum.png")
+jg = make_network(pseq, "taxa", "bray", 0.3)
+plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
+dev.off()
+
+```
+
+## Community composition plotting
+Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results.
+These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted.
+Interpretation of these plots should always be subject to subsequent statistical analysis.
+```{r community-composition-plots}
+# Create a data table for ggplot
+# agglomerate at phylum level
+pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
+
+# Transform to rel. abundance (or use ps0.ra)
+pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
+
+# Melt to long format for easy ggploting
+pseq_phylum <- psmelt(pseq_phyl)
+
+     # filter(Abundance > 0.01)                             # Filter out low abundance taxa
+
+# Plot - Phylum
+p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum (> 1%)")
+
+p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+plot(p)
+ggsave("phylum_abun_mnd.png")
+
+```
+
+```{r}
+
+clos <- aggregate_taxa(clostridium, "Phylum")
+
+# agglomerate at phylum level
+clos_phy <- tax_glom(clos, taxrank = "Phylum")
+
+# Transform to rel. abundance (or use ps0.ra)
+clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})
+
+# Melt to long format for easy ggploting
+clos_phylum <- psmelt(clos_phyl)
+
+# Plot - Phylum
+p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum of Clostridium(> 1%)")
+
+p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+plot(p)
+ggsave("phylum_abun_mnd_clos.png")
+
+```
+
+```{r}
+
+get_taxa_unique(pseq, "Phylum")
+
+```
+
+```{r}
+
+get_taxa_unique(pseq, "Genus")
+
+```
+
+```{r}
+
+get_taxa_unique(clostridium, "Genus")
+
+```
+
+```{r}
+
+df <- psmelt(pseq)
+kable(head(df))
+
+```
+
+```{r}
+
+head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])
+
+```

### 2019-10-08 curatedMetagenomic_NS.Rmd
updated

@@
 library(ggpubr)
+library(tidyverse)
 
@@
 
-any(taxa_sums(pseq.c) == 0)
+any(taxa_sums(pseq.subset) == 0)
 

### 2019-10-15 curatedMetagenomic_NS.Rmd
updated

--- 2019-10-06 curatedMetagenomic_NS.Rmd
@@
 library(ggpubr)
+library(tidyverse)
+library(cowplot)
+library(vegan)
+library(broom)
+library(tsnemicrobiota)
+library(latex2exp)
 
@@
 # Create mnd variable
+pseq <- orig_pseq
 # grabs the nationality from phyloseq
@@
 
-mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN") , MEDIUM = c("ITA","AUT","DEU","GBR","ESP","FRA","LUX","NLD"), HIGH = c("USA") )
+mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DEN","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
 
@@
 
-```{r MND table summary}
-
-table(sample_data(pseq)$mnd)
-
-table(sample_data(pseq)$country)
-
-```
 ## Check OTUs/Phylums
@@
 
-pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn")
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" & antibiotics_current_use != "yes")
+
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
 
@@
 #sample_variables(pseq.subset)
-#get_variable(pseq.subset, sample_variables(pseq.subset)[4])
+#get_variable(pseq.subset, sample_variables(pseq.subset)[13])
 
-any(taxa_sums(pseq.c) == 0)
+any(taxa_sums(pseq.subset) == 0)
 
@@
 pseq.meta <- meta(pseq.subset)
-kable(head(pseq.meta))
+#kable(head(pseq.meta))
 stargazer(pseq.meta, type = "text",summary = NULL)
-
-# Add the diversity table to metadata
-pseq.meta$Shannon <- tab$shannon 
-#pseq.meta$InverseSimpson <- tab$inverse_simpson
-#pseq.meta$fisher <- tab$fisher
-#pseq.meta$chao1 <- tab$chao1
-
-p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon",
- add = "boxplot", fill = "mnd", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-  print(p1)
-#ggsave("nation_violin_shannon.png")
 
@@
 
+# Shannon
+# the number in the palettes must equal the number of groups in bmi.pairs
+p1 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+p1 <- p1 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+print(p1)
+ggsave("mnd_violin__shannon_stats.png")
+
+alpha_d <- function(pseq.subset) {
+   # gets the meta data 
+   pseq.meta <- meta(pseq.subset)
+   # generates the diversites
+   tab <- diversities(pseq.subset, index = "Shannon")
+   # creates Shannon
+   pseq.meta$Shannon <- tab$shannon
+   mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+   # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+  p1 <-  ggviolin(pseq.meta, x = "mnd", y = "Shannon",
+ add = "boxplot", fill = "mnd", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+  p2<-   ggviolin(pseq.meta, x = "country", y = "Shannon",
+ add = "boxplot", fill = "country", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+# the number in the palettes must equal the number of groups in bmi.pairs
+  p3 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+  p3 <- p3 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+  print(p1)
+  print(p2)
+  p4 <-plot_grid(p1, p2,p3, align = "h" , ncol = 3)
+
+}
+
+beta_d <- function(pseq.subset) {
+  p.low <- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <- divergence(p.low)
+  p.med <- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <- divergence(p.med)
+  p.high <- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <- divergence(p.high)
+  p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+  #plot(p)
+  # NMDS on the whole set???
+  pseq.ord <- ordinate(pseq.subset, "NMDS", "bray")
+  # Convert to compositional data
+  pseq.rel <- microbiome::transform(pseq.subset, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  pseq.nmds <- ordinate(pseq.rel, "NMDS", "bray")
+  #NMDS ordination on Bray-Curtis distance
+  p1 = plot_ordination(pseq.subset, pseq.nmds, type = "taxa", color = "mnd", title = "Taxa by Phylum")
+  p1 + facet_wrap(~Phylum, 3)
+  p2 = plot_ordination(pseq.rel, pseq.nmds, type = "sample", color = "mnd")
+  plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+  p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+
+    # MDS
+   pseq.mds <- ordinate(pseq.rel, "MDS", "bray")
+   p3 = plot_ordination(pseq.rel, pseq.mds, type = "sample", color = "mnd")
+ plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+
+   p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
+   #PCoA for compositional data with Bray-Curtis distances
+    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
+      p5 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+  print(p5)
+  # 
+  #   # Pick core taxa
+    pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)
+  # 
+  #   # PCoA for compositional data with Bray-Curtis distances
+   p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+                        method = "PCoA", distance = "bray") +
+         labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+  # 
+    tsne_res <- tsne_phyloseq(pseq.subset, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+    p7 <- plot_tsne_phyloseq(pseq.subset, tsne_res,
+    color = 'mnd')
+    p7 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
+
+  print(p7)
+
+}
+
+mylab <- function(v1) {
+  deparse(substitute(v1))
+}
+
+pre_perma_d <- function(pseq.subset) { 
+  p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  print(bob)
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+}
+
+pre_perma_d(pseq.subset) 
+perma_d(p.low,p.high)
+
+perma_d <- function(x,y) { # x & y must be either p.low,p.med or p.high
+
+  # Convert to compositional data
+  pseqLvH <- merge_phyloseq(x,y)
+  tempx <- mylab(x)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(3, 14, 2, 1))
+  temp <- TeX("Top Genus $tempx")
+  barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+ }
+
+## MICROBIOME COMPOSITION
+paste("Hello", "world", sep=" ")
+
+comp_d <- function(pseq,mylabel) {
+  png(paste(mylabel,"Phylum_abundance.png",sep="_"))
+  p <- plot_bar(pseq,"Phylum") + theme(axis.text.x = element_text(face="italic"))
+  print(p)
+
+  prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
+  prev.otu + theme(legend.position="none")
+  ggsave(paste(mylabel,"phylum_prevalence.png",sep="_"))
+
+  tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
+  tab + theme(legend.position="none")
+  ggsave("Phylum_prevalence_detection5.png")
+
+  pseq.ph <- aggregate_taxa(pseq, "Phylum")
+  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+  plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
+    theme(legend.position = "bottom") + theme_bw() +
+    theme(axis.text.x=element_blank()) +
+    c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
+  print(plot.comp.rel)
+  ggsave("phylum_relative_abundance11.png")
+  plot.comp.rel + theme(legend.position="none")
+  ggsave("phylum_relative_abundance12.png")
+
+  pseq.ph <- aggregate_taxa(pseq, "Phylum")
+  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND   Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+  ggsave("phylum_abundance_bar.png")
+  # Top 5
+  top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
+
+  top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
+  pseq.ph.rel <- microbiome::transform(top5P, "compositional")
+  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+  # Top 10
+  top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
+
+  top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
+  pseq.ph.rel <- microbiome::transform(top10P, "compositional")
+
+  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+  # new
+  p.new <- aggregate_taxa(pseq, "Phylum")
+  p.new.rel <- microbiome::transform(p.new, "compositional")
+
+  plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+  plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
+  ggsave("pnew_abundance_bar.png")
+
+  # Top Phylum Heatmap
+  pseq.ph <- aggregate_taxa(pseq, "Phylum")
+
+  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+  plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+  plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
+  ggsave("phylum_histogram.png")
+
+  #p.new heatmap
+  pseq.n.ph <- aggregate_taxa(pseq, "Phylum")
+  pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
+
+  plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+  p <- plot.n.comp.rel + coord_flip()
+  p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
+  ggsave("pnew_histogram.png")
+
+  # Newwork
+  png("network_bray_phylum.png")
+  jg = make_network(pseq, "taxa", "bray", 0.3)
+  plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
+  dev.off()
+  #Community-composition-plots
+  # Create a data table for ggplot
+# agglomerate at phylum level
+pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
+
+# Transform to rel. abundance (or use ps0.ra)
+pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
+
+# Melt to long format for easy ggploting
+pseq_phylum <- psmelt(pseq_phyl)
+
+     # filter(Abundance > 0.01)                             # Filter out low abundance taxa
+# Plot - Phylum
+  p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum (> 1%)")
+
+  p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+  plot(p)
+  ggsave("phylum_abun_mnd.png")
+}
+
+alpha_d(pseq.subset         )
+beta_d(pseq.subset)
+
 ```
+
+```{r MND table summary}
+
+library(table1)
+n <- length(sample_data(pseq.subset)$mnd)
+
+data <- data.frame(id=1:n)
+data$mnd <- sample_data(pseq.subset)$mnd
+data$bmi <- sample_data(pseq.subset)$BMI
+data$age <- sample_data(pseq.subset)$age
+data$antibiotics_current_use <- sample_data(pseq.subset)$antibiotics_current_use
+data$smoker <- sample_data(pseq.subset)$smoker
+data$pregnant <- sample_data(pseq.subset)$pregnant
+data$alcohol <- sample_data(pseq.subset)$alcohol
+data$age_category <- sample_data(pseq.subset)$age_category
+
+table1(~ age  +  bmi + smoker + alcohol + age_category | mnd, data=data)
+
+```
+
+## BETA DIVERSITY
+Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
+WE want to: quantify community divergence within a given sample set
+Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.
+
+Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
+This was applied in group-level comparisons for instance in Salonen et al.
+ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
+
+ The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
+
+## PERMANOVA significance test for group-level differences
+Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
+
+## MICROBIOME COMPOSITION
+
+```{r Phylum #}
+
+# Show available ranks in the dataset
+rank_names(pseq)
+
+# Create table, number of features for each phyla
+table(tax_table(pseq)[, "Phylum"], exclude = NULL)
+
+```
+This shows a few phyla for which only one feature was observed. Those may be worth filtering.
+
+guide_italics
+ggtitle("Relative abundance") +  
+```{r}
+
+clos.ph <- aggregate_taxa(clostridium, "Phylum")
+
+clos.ph.rel <- microbiome::transform(clos.ph, "compositional")
+
+plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
+  theme(legend.position = "bottom", ) + theme_bw() + 
+  theme(axis.text.x=element_blank()) + 
+  theme(legend.title = element_text(size=18), legend.text = element_text( face="italic")) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend(title="Phylum"))
+
+print(plot.comp.rel)
+
+ggsave("phylum_relative_abundance_clos.png")
+
+plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none") + guide_italics
+ggsave("phylum_relative_abundance_clos2.png")
+
+```
+
+```{r CLostridium Bar Plot}
+
+clos <- aggregate_taxa(clostridium, "Phylum")
+
+pseq.cl.rel <- microbiome::transform(clos, "compositional")
+
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+plot.cl.rel + scale_fill_brewer(palette = "Paired")
+ggsave("cl_abundance_bar.png")
+
+```
+
+```{r Clostridium genus bar plots}
+
+pseq.cl.rel <- microbiome::transform(clostridium, "compositional")
+
+plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
+
+plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")
+
+```
+
+```{r clostridium heatmap}
+
+pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")
+
+pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")
+
+plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+p <- plot.c.comp.rel + coord_flip()
+p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
+ggsave("clos_histogram.png")
+
+```
+
+## Community composition plotting
+Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results.
+These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted.
+Interpretation of these plots should always be subject to subsequent statistical analysis.
+
+```{r}
+
+clos <- aggregate_taxa(clostridium, "Phylum")
+
+# agglomerate at phylum level
+clos_phy <- tax_glom(clos, taxrank = "Phylum")
+
+# Transform to rel. abundance (or use ps0.ra)
+clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})
+
+# Melt to long format for easy ggploting
+clos_phylum <- psmelt(clos_phyl)
+
+# Plot - Phylum
+p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+  geom_bar(stat = "identity", width = 1) +
+  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
+  theme(axis.text.x = element_blank()) +
+  theme(axis.title.x = element_blank()) +
+  labs(title = "Abundant Phylum of Clostridium(> 1%)")
+
+p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+plot(p)
+ggsave("phylum_abun_mnd_clos.png")
+
+```
+
+```{r}
+
+get_taxa_unique(pseq, "Phylum")
+
+```
+
+```{r}
+
+get_taxa_unique(pseq, "Genus")
+
+```
+
+```{r}
+
+get_taxa_unique(clostridium, "Genus")
+
+```
+
+```{r}
+
+df <- psmelt(pseq)
+kable(head(df))
+
+```
+
+```{r}
+
+head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])
+
+```

### 2019-10-20 curatedMetagenomic_NS.Rmd
updated

@@
 library(latex2exp)
+library(wesanderson)
 
@@
 
+Defining the functions 
+```{r define the functions}
+
+# this defines the colours for the graphs. Uses the wes palettes. Look at https://github.com/karthik/wesanderson for the different themes
+glob_theme <- "FantasticFox1"
+
+# THis is required
+mylab <- function(v1) {
+  deparse(substitute(v1))
+}
+# aplha_d runs the diversity for each country or  groupoing and then compares the three groups
+# example is alpha_d(pseq.subset,"TEST_3") this will generate a png file called TEST_3.png in the 
+# currect working directory. This can be checked with getwd()
+alpha_d <- function(pseq.subset,title_text) {
+   # gets the meta data 
+   pseq.meta <- meta(pseq.subset)
+   # generates the diversites
+   tab <- diversities(pseq.subset, index = "Shannon")
+   # creates Shannon
+   pseq.meta$Shannon <- tab$shannon
+   mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+   # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+  p1<-   ggviolin(pseq.meta, x = "country", y = "Shannon",
+ add = "boxplot", fill = "country", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+# the number in the palettes must equal the number of groups in bmi.pairs
+
+   p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = wes_palette(glob_theme))
+
+    p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+    #this just generates the titles for the output
+    title <- ggdraw() + 
+    draw_label(
+    title_text,
+    fontface = 'bold',
+    x = 0,
+    hjust = 0
+    ) +
+    theme(
+    # add margin on the left of the drawing canvas,
+    # so title is aligned with left edge of first plot
+    plot.margin = margin(0, 0, 0, 7)
+    )
+
+    p3 <<-plot_grid(title,p1,p2, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
+    ggsave(paste("Alpha_diversity_",title_text,".png", sep = "" ), device = "png")
+}
+
+# THis includes the analysis as well so can take some time to run. the graphs are generated with the label of beta diversity. It requires an input of a title. 
+beta_d <- function(pseq.subset,title_text) {
+  #NMDS ordination on Bray-Curtis distance
+  p <<-  boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+ # p1 = plot_ordination(pseq.subset, pseq.nmds, type = "taxa", color = "mnd", title = "Taxa by Phylum") # I dont think that we need this any more. 
+  #p1 + facet_wrap(~Phylum, 3)
+  pseq.nmds <- ordinate(pseq.subset, "NMDS", "bray")
+  p2 <- plot_ordination(pseq.subset, pseq.nmds, type = "sample", color = "mnd")
+
+  p3 <- p2 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND Prevalence") + theme(legend.position = "none")
+  #plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+
+  # This add the facets but I do not think we need that any more. 
+  #p3 <- p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+  ggsave(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), device = "png")
+    # MDS
+   pseq.mds <<- ordinate(pseq.subset, "MDS", "bray")
+   p4 <- plot_ordination(pseq.subset, pseq.mds, type = "sample", color = "mnd")
+ #plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence MDS"))
+
+    p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND Prevalence") + theme(legend.position = "none")
+    ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+
+   #PCoA for compositional data with Bray-Curtis distances
+    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
+    #p6 <- plot_landscape(pseq.subset, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+  #   # PCoA for compositional data with Bray-Curtis distances
+   #p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+    #                    method = "PCoA", distance = "bray") +
+    #     labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+  # 
+    tsne_res <- tsne_phyloseq(pseq.subset, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+    p7 <- plot_tsne_phyloseq(pseq.subset, tsne_res,color = 'mnd')
+    p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND Prevalence") + theme(legend.position = "bottom")
+    ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+       #this just generates the titles for the output
+    title <- ggdraw() + 
+    draw_label(
+    title_text,
+    fontface = 'bold',
+    x = 0,
+    hjust = 0
+    ) +
+    theme(
+    # add margin on the left of the drawing canvas,
+    # so title is aligned with left edge of first plot
+    plot.margin = margin(0, 0, 0, 7)
+    )
+
+    p10 <<-plot_grid(title,p3, p5,p8, ncol = 1, labels = c('','A', 'B','C'),rel_heights = c(0.3, 1.5,1.5,2))
+    ggsave(paste("Beta_diversity_ALL_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+}
+
+pre_perma_d <- function(pseq.subset) { 
+  p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+}
+
+# THis calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
+perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
+  # Convert to compositional data
+  pseqLvH <- merge_phyloseq(x,y) # pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p <- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal()  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
+  ggsave(paste("PERMANOVA_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+
+ }
+
+# comp_d <- function(pseq,subset,title_text) {
+#   png(paste(title_text,"Phylum_abundance.png",sep="_"))
+#   p <- plot_bar(pseq.subset,"Phylum") + theme(axis.text.x = element_text(face="italic"))
+#   print(p)
+
+#   prev.otu <- plot_taxa_prevalence(pseq.subset, "Phylum")
+#   prev.otu + theme(legend.position="none")
+#   ggsave(paste(title_text,"phylum_prevalence.png",sep="_"))
+#   tab <- plot_taxa_prevalence(pseq.subset, "Phylum", detection = 5)
+#   tab + theme(legend.position="none")
+#   ggsave(paste(title_text,"Phylum_prevalence_detection.png",sep="_"))
+
+#   pseq.ph <- aggregate_taxa(pseq.subset, "Phylum")
+#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
+#     theme(legend.position = "bottom") + theme_bw() +
+#     theme(axis.text.x=element_blank()) +
+#     c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
+#   print(plot.comp.rel)
+#   ggsave("phylum_relative_abundance11.png")
+#   plot.comp.rel + theme(legend.position="none")
+#   ggsave("phylum_relative_abundance12.png")
+
+#   pseq.ph <- aggregate_taxa(pseq, "Phylum")
+#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND   Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+#   ggsave("phylum_abundance_bar.png")
+#   # Top 5
+#   top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
+
+#   top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
+#   pseq.ph.rel <- microbiome::transform(top5P, "compositional")
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+#   # Top 10
+#   top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
+
+#   top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
+#   pseq.ph.rel <- microbiome::transform(top10P, "compositional")
+
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+#   # new
+#   p.new <- aggregate_taxa(pseq, "Phylum")
+#   p.new.rel <- microbiome::transform(p.new, "compositional")
+
+#   plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+#   plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
+#   ggsave("pnew_abundance_bar.png")
+
+#   # Top Phylum Heatmap
+#   pseq.ph <- aggregate_taxa(pseq, "Phylum")
+
+#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+#   plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
+#   ggsave("phylum_histogram.png")
+
+#   #p.new heatmap
+#   pseq.n.ph <- aggregate_taxa(pseq, "Phylum")
+#   pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
+
+#   plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+#   p <- plot.n.comp.rel + coord_flip()
+#   p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
+#   ggsave("pnew_histogram.png")
+
+#   # Newwork
+#   png("network_bray_phylum.png")
+#   jg = make_network(pseq, "taxa", "bray", 0.3)
+#   plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
+#   dev.off()
+#   #Community-composition-plots
+#   # Create a data table for ggplot
+# # agglomerate at phylum level
+# pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
+
+# # Transform to rel. abundance (or use ps0.ra)
+# pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
+
+# # Melt to long format for easy ggploting
+# pseq_phylum <- psmelt(pseq_phyl)
+
+#      # filter(Abundance > 0.01)                             # Filter out low abundance taxa
+# # Plot - Phylum
+#   p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+#   geom_bar(stat = "identity", width = 1) +
+#   facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
+#   theme(axis.text.x = element_blank()) +
+#   theme(axis.title.x = element_blank()) +
+#   labs(title = "Abundant Phylum (> 1%)")
+
+#   p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+#   plot(p)
+#   ggsave("phylum_abun_mnd.png")
+# }
+
+```
+
 ## Check OTUs/Phylums
@@
 
-any(taxa_sums(pseq.subset) == 0)
-
-# run the diversities. 	keep an eye on this as only one is working
-tab <- diversities(pseq.subset, index = "Shannon")
-
-# Get metadata
-pseq.meta <- meta(pseq.subset)
-#kable(head(pseq.meta))
-stargazer(pseq.meta, type = "text",summary = NULL)
-
-# THis subsets the  genus and phylum. 
-#genus.pseq <- subset_taxa( pseq.subset, !is.na(Genus))
-#phylum.pseq <- subset_taxa( pseq, is.na(Class) & !is.na(Phylum))
-
-# Shannon
-# the number in the palettes must equal the number of groups in bmi.pairs
-p1 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
-p1 <- p1 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
-print(p1)
-ggsave("mnd_violin__shannon_stats.png")
-
-alpha_d <- function(pseq.subset) {
-   # gets the meta data 
-   pseq.meta <- meta(pseq.subset)
-   # generates the diversites
-   tab <- diversities(pseq.subset, index = "Shannon")
-   # creates Shannon
-   pseq.meta$Shannon <- tab$shannon
-   mnd <- levels(pseq.meta$mnd) # get the variables
-  # Make a pairwise list that we want to compare.
-  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
-  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
-
-   # Compare differences in Shannon index between mnd group of the study subjects
-  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-  # on to the graphs!!!
-  p1 <-  ggviolin(pseq.meta, x = "mnd", y = "Shannon",
- add = "boxplot", fill = "mnd", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-  p2<-   ggviolin(pseq.meta, x = "country", y = "Shannon",
- add = "boxplot", fill = "country", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-# the number in the palettes must equal the number of groups in bmi.pairs
-  p3 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
-  p3 <- p3 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
-  print(p1)
-  print(p2)
-  p4 <-plot_grid(p1, p2,p3, align = "h" , ncol = 3)
-
-}
-
-beta_d <- function(pseq.subset) {
-  p.low <- subset_samples(pseq.subset, mnd == "LOW")
-  p.low.d <- divergence(p.low)
-  p.med <- subset_samples(pseq.subset, mnd == "MEDIUM")
-  p.med.d <- divergence(p.med)
-  p.high <- subset_samples(pseq.subset, mnd == "HIGH")
-  p.high.d <- divergence(p.high)
-  p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
-  #plot(p)
-  # NMDS on the whole set???
-  pseq.ord <- ordinate(pseq.subset, "NMDS", "bray")
-  # Convert to compositional data
-  pseq.rel <- microbiome::transform(pseq.subset, "compositional")
-  otu <- abundances(pseq.rel)
-  meta <- meta(pseq.rel)
-  pseq.nmds <- ordinate(pseq.rel, "NMDS", "bray")
-  #NMDS ordination on Bray-Curtis distance
-  p1 = plot_ordination(pseq.subset, pseq.nmds, type = "taxa", color = "mnd", title = "Taxa by Phylum")
-  p1 + facet_wrap(~Phylum, 3)
-  p2 = plot_ordination(pseq.rel, pseq.nmds, type = "sample", color = "mnd")
-  plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-  p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
-
-    # MDS
-   pseq.mds <- ordinate(pseq.rel, "MDS", "bray")
-   p3 = plot_ordination(pseq.rel, pseq.mds, type = "sample", color = "mnd")
- plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
-
-   p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
-   #PCoA for compositional data with Bray-Curtis distances
-    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
-    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
-      p5 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
-  print(p5)
-  # 
-  #   # Pick core taxa
-    pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)
-  # 
-  #   # PCoA for compositional data with Bray-Curtis distances
-   p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
-                        method = "PCoA", distance = "bray") +
-         labs(title = paste("PCoA / Compositional / Bray-Curtis"))
-  # 
-    tsne_res <- tsne_phyloseq(pseq.subset, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
-    p7 <- plot_tsne_phyloseq(pseq.subset, tsne_res,
-    color = 'mnd')
-    p7 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
-
-  print(p7)
-
-}
-
-mylab <- function(v1) {
-  deparse(substitute(v1))
-}
-
-pre_perma_d <- function(pseq.subset) { 
-  p.low <<- subset_samples(pseq.subset, mnd == "LOW")
-  print(bob)
-  p.low.d <<- divergence(p.low)
-  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
-  p.med.d <<- divergence(p.med)
-  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
-  p.high.d <<- divergence(p.high)
-  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
-
-}
-
+alpha_d(pseq.subset,"TEST_3")
+beta_d(pseq.subset,"TEST_2")
 pre_perma_d(pseq.subset) 
-perma_d(p.low,p.high)
-
-perma_d <- function(x,y) { # x & y must be either p.low,p.med or p.high
-
-  # Convert to compositional data
-  pseqLvH <- merge_phyloseq(x,y)
-  tempx <- mylab(x)
-  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
-  otu <- abundances(pseq.rel)
-  meta <- meta(pseq.rel)
-  permanova <- adonis(t(otu) ~ mnd,
-               data = meta, permutations=999, method = "bray")
- # P-value
-  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
-  dist <- vegdist(t(otu))
-  anov <- anova(betadisper(dist, meta$mnd))
-  summary(anov)
-  tidy(anov)
-  coef <- coefficients(permanova)["mnd1",]
-  top.coef <- coef[rev(order(abs(coef)))[1:20]]
-  par(mar = c(3, 14, 2, 1))
-  temp <- TeX("Top Genus $tempx")
-  barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
- }
-
-## MICROBIOME COMPOSITION
-paste("Hello", "world", sep=" ")
-
-comp_d <- function(pseq,mylabel) {
-  png(paste(mylabel,"Phylum_abundance.png",sep="_"))
-  p <- plot_bar(pseq,"Phylum") + theme(axis.text.x = element_text(face="italic"))
-  print(p)
-
-  prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
-  prev.otu + theme(legend.position="none")
-  ggsave(paste(mylabel,"phylum_prevalence.png",sep="_"))
-
-  tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
-  tab + theme(legend.position="none")
-  ggsave("Phylum_prevalence_detection5.png")
-
-  pseq.ph <- aggregate_taxa(pseq, "Phylum")
-  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-  plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
-    theme(legend.position = "bottom") + theme_bw() +
-    theme(axis.text.x=element_blank()) +
-    c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
-  print(plot.comp.rel)
-  ggsave("phylum_relative_abundance11.png")
-  plot.comp.rel + theme(legend.position="none")
-  ggsave("phylum_relative_abundance12.png")
-
-  pseq.ph <- aggregate_taxa(pseq, "Phylum")
-  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND   Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-  ggsave("phylum_abundance_bar.png")
-  # Top 5
-  top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
-
-  top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
-  pseq.ph.rel <- microbiome::transform(top5P, "compositional")
-  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-
-  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
-
-  # Top 10
-  top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
-
-  top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
-  pseq.ph.rel <- microbiome::transform(top10P, "compositional")
-
-  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-
-  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-  # new
-  p.new <- aggregate_taxa(pseq, "Phylum")
-  p.new.rel <- microbiome::transform(p.new, "compositional")
-
-  plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
-
-  plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
-  ggsave("pnew_abundance_bar.png")
-
-  # Top Phylum Heatmap
-  pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
-  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-  plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-  plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
-  ggsave("phylum_histogram.png")
-
-  #p.new heatmap
-  pseq.n.ph <- aggregate_taxa(pseq, "Phylum")
-  pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
-
-  plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-  p <- plot.n.comp.rel + coord_flip()
-  p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-  ggsave("pnew_histogram.png")
-
-  # Newwork
-  png("network_bray_phylum.png")
-  jg = make_network(pseq, "taxa", "bray", 0.3)
-  plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
-  dev.off()
-  #Community-composition-plots
-  # Create a data table for ggplot
-# agglomerate at phylum level
-pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
-
-# Transform to rel. abundance (or use ps0.ra)
-pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
-
-# Melt to long format for easy ggploting
-pseq_phylum <- psmelt(pseq_phyl)
-
-     # filter(Abundance > 0.01)                             # Filter out low abundance taxa
-# Plot - Phylum
-  p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-  geom_bar(stat = "identity", width = 1) +
-  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-  theme(axis.text.x = element_blank()) +
-  theme(axis.title.x = element_blank()) +
-  labs(title = "Abundant Phylum (> 1%)")
-
-  p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-  plot(p)
-  ggsave("phylum_abun_mnd.png")
-}
-
-alpha_d(pseq.subset         )
-beta_d(pseq.subset)
+perma_d(p.low,p.high,"TESt_highVlow")
 
@@
 ```
+
 This shows a few phyla for which only one feature was observed. Those may be worth filtering.

### 2019-11-11 HvL_curatedMetagenomic_NS.Rmd
updated

--- 2019-10-20 curatedMetagenomic_NS.Rmd
@@
 curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:
+data(atlas1006)
+atlas.pseq <- baseline(atlas1006)
 
@@
 
-mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DEN","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
+# Cut off is 1.8  Low
 
-#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+mnd.country <- fct_collapse(mnd.country, LOW = c("DEU","ESP","ITA","AUT","DNK","LUX" ) , HIGH = c("NOR","FIN","FRA","NLD","GBR","USA","CAN","SWE") )
+
+table(mnd.country)
+#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") ) "ITA","AUT","DEU","ESP","FRA","NLD"
 
 # reorder
-mnd.country <- factor(mnd.country, levels = (c("LOW", "MEDIUM", "HIGH")))
+mnd.country <- factor(mnd.country, levels = (c("LOW", "HIGH")))
 levels(mnd.country)
@@
 get_variable(pseq, "mnd")
+
+# THIS IS FOR THE ATLAS 1006 
+# Create mnd variable
+# grabs the nationality from phyloseq
+mnd.nationality <- get_variable(atlas.pseq, "nationality" )
+
+# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
+mnd.nationality <- fct_collapse(mnd.nationality, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+
+# reorder
+mnd.nationality <- factor(mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(mnd.nationality)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(atlas.pseq)$mnd = mnd.nationality
+
+# checks that it has worked.
+get_variable(atlas.pseq, "mnd")
 
@@
 # currect working directory. This can be checked with getwd()
-alpha_d <- function(pseq.subset,title_text) {
+alpha_d_two <- function(pseq.subset,title_text) {
    # gets the meta data 
@@
    # generates the diversites
-   tab <- diversities(pseq.subset, index = "Shannon")
+   tab <- microbiome::alpha(pseq.subset, index = "shannon")
+
    # creates Shannon
-   pseq.meta$Shannon <- tab$shannon
+   pseq.meta$Shannon <- tab$diversity_shannon
    mnd <- levels(pseq.meta$mnd) # get the variables
@@
   mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
-  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+  mnd.pairs <- list( c("HIGH","LOW"))
 
    # Compare differences in Shannon index between mnd group of the study subjects
-  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
-  # on to the graphs!!!
-  p1<-   ggviolin(pseq.meta, x = "country", y = "Shannon",
- add = "boxplot", fill = "country", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","HIGH"),  exclude = "NA")
+  # on to the graphs!!! country 
+  p1<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+ add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
 # the number in the palettes must equal the number of groups in bmi.pairs
@@
 # THis includes the analysis as well so can take some time to run. the graphs are generated with the label of beta diversity. It requires an input of a title. 
-beta_d <- function(pseq.subset,title_text) {
+beta_d_two <- function(pseq.subset,title_text) {
   #NMDS ordination on Bray-Curtis distance
-  p <<-  boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+  p <<-  boxplot(list(Low = p.low.d, High = p.high.d))
  # p1 = plot_ordination(pseq.subset, pseq.nmds, type = "taxa", color = "mnd", title = "Taxa by Phylum") # I dont think that we need this any more. 
@@
 
-pre_perma_d <- function(pseq.subset) { 
+pre_perma_d_two <- function(pseq.subset) { 
   p.low <<- subset_samples(pseq.subset, mnd == "LOW")
   p.low.d <<- divergence(p.low)
-  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
-  p.med.d <<- divergence(p.med)
   p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
   p.high.d <<- divergence(p.high)
-  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+  p <<- boxplot(list(Low = p.low.d, High = p.high.d))
 
@@
 # THis calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
-perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
-  # Convert to compositional data
-  pseqLvH <- merge_phyloseq(x,y) # pseqLvH <- merge_phyloseq(p.low,p.high)
+perma_d_two <- function(title_text) { # x & y must be either p.low,p.med or p.high
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+  pseqLvH <- merge_phyloseq(p.low,p.high)
   pseq.rel <- microbiome::transform(pseqLvH, "compositional")
@@
   #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
-  p <- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal()  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
+  p <- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
   ggsave(paste("PERMANOVA_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+  ggsave("test_permanova.png", device = "png",width = 297, height = 210, units = "mm")
+ }
+
+testpseq <- atlas1006 %>%
+   subset_samples(DNA_extraction_method == "r") %>%
+   aggregate_taxa(level = "Phylum") %>%
+   transform(transform = "compositional")
+p <- plot_composition(testpseq, sample.sort = "Firmicutes",
+         otu.sort = "abundance", verbose = TRUE) +
+     scale_fill_manual(values = default_colors("Phylum")[taxa(testpseq)])
+
+comp_d <- function(pseq,subset,title_text) {
+  png(paste(title_text,"Phylum_abundance.png",sep="_"))
+  p <- plot_bar(pseq.subset,"Phylum") + theme(axis.text.x = element_text(face="italic"))
+  print(p)
+
+  prev.otu <- plot_taxa_prevalence(pseq.subset, "Phylum" , detection = 5)
+  prev.otu + theme(legend.position="none")
+  ggsave(paste(title_text,"phylum_prevalence.png",sep="_"))
+  tab <- plot_taxa_prevalence(pseq.subset, "Phylum", detection = 5)
+  tab + theme(legend.position="none")
+  ggsave(paste(title_text,"Phylum_prevalence_detection.png",sep="_"))
+
+  pseq.ph <- aggregate_taxa(pseq.subset, "Phylum")
+  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+  plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
+    theme(legend.position = "bottom") + theme_bw() +
+    theme(axis.text.x=element_blank()) +
+    c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
+  print(plot.comp.rel)
+  ggsave("phylum_relative_abundance11.png")
+  plot.comp.rel + theme(legend.position="none")
+  ggsave("phylum_relative_abundance12.png")
+
+  pseq.ph <- aggregate_taxa(pseq.subset, "Phylum")
+  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND   Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+  ggsave("phylum_abundance_bar.png")
+  # Top 5
+  top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
+
+  top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
+u  pseq.ph.rel <- microbiome::transform(top5P, "compositional")
+  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+  # Top 10
+  top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
+
+  top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
+  pseq.ph.rel <- microbiome::transform(top10P, "compositional")
+
+  plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+  plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+  # new
+  p.new <- aggregate_taxa(pseq, "Phylum")
+  p.new.rel <- microbiome::transform(p.new, "compositional")
+
+  plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+  plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
+  ggsave("pnew_abundance_bar.png")
+
+  # Top Phylum Heatmap
+  pseq.ph <- aggregate_taxa(pseq, "Phylum")
+
+  pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+  plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+  plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
+  ggsave("phylum_histogram.png")
+
+  #p.new heatmap
+  pseq.n.ph <- aggregate_taxa(pseq, "Phylum")
+  pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
+
+  plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+  p <- plot.n.comp.rel + coord_flip()
+  p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
+  ggsave("pnew_histogram.png")
+
+  # Newwork
+  png("network_bray_phylum.png")
+  jg = make_network(pseq, "taxa", "bray", 0.3)
+  plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
+  dev.off()
+  #Community-composition-plots
+  # Create a data table for ggplot
+# agglomerate at phylum level
+pseq_phy <- tax_glom(pseq.subset, taxrank = "Phylum")
+
+# Transform to rel. abundance (or use ps0.ra)
+pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
+
+# Melt to long format for easy ggploting
+pseq_phylum <- psmelt(pseq_phyl)
+
+     # filter(Abundance > 0.01)                             # Filter out low abundance taxa
+# Plot - Phylum
+    p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+    geom_bar(stat = "identity", width = 1) +
+    facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
+    theme(axis.text.x = element_blank()) +
+    theme(axis.title.x = element_blank()) +
+    labs(title = "Abundant Phylum (> 1%)")
+
+  p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+  plot(p)
+  ggsave("phylum_abun_mnd.png")
+
+ # these are other functions required along the way.
+
+    # transform counts to n
+  physeq.scale <- transform_sample_counts(physeq, 
+    function(x) {(n * x/sum(x))}
+  )
+
+  # Pick the rounding functions
+  if (round == "floor"){
+    otu_table(physeq.scale) <- floor(otu_table(physeq.scale))
+  } else if (round == "round"){
+    otu_table(physeq.scale) <- myround(otu_table(physeq.scale))
+  }
+
+  # Prune taxa and return new phyloseq object
+  physeq.scale <- prune_taxa(taxa_sums(physeq.scale) > 0, physeq.scale)
+  return(physeq.scale)
+}
+
+##### ADONIS ###########
+
+# Function to run adonis test on a phyloseq object and a variable from metadata
+# Make sure OTU data is standardized/normalized before 
+phyloseq_to_adonis <- function(physeq, distmat = NULL, dist = "bray", formula) {
+
+  if(!is.null(distmat)) {
+    phydist <- distmat
+  } else {
+    phydist <- phyloseq::distance(physeq, dist)
+  }
+
+  metadata <- as(sample_data(physeq), "data.frame")
+
+  # Adonis test
+  f <- reformulate(formula, response = "phydist")
+  adonis.test <- adonis(f, data = metadata)
+  print(adonis.test)
+
+  # Run homogeneity of dispersion test if there is only 1 variable
+  if (grepl("\\+", formula)) {
+    l <- list(
+      dist = phydist, 
+      formula = f, 
+      adonis = adonis.test
+    )
+  } else {
+    group <- metadata[ ,formula]
+    beta <- betadisper(phydist, group)
+    disper.test = permutest(beta)
+    print(disper.test)
+
+    l <- list(
+      dist = phydist, 
+      formula = f, 
+      adonis = adonis.test, 
+      disper = disper.test
+    )
+  }
+  return (l)
+}
+
+########## Bar Plots #################
+
+# This function takes a phyloseq object, agglomerates OTUs to the desired taxonomic rank, 
+# prunes out OTUs below a certain relative proportion in a sample (ie 1% ) 
+# and melts the phyloseq object into long format which is suitable for ggplot stacked barplots.
+taxglom_and_melt <- function(physeq, taxrank, prune){
+
+  # Agglomerate all otu's by given taxonomic level
+  pglom <- tax_glom(physeq, taxrank = taxrank)
+
+  # Create a new phyloseq object which removes taxa from each sample below the prune parameter
+  pglom_prune <- transform_sample_counts(pglom,function(x) {x/sum(x)})
+  otu_table(pglom_prune)[otu_table(pglom_prune) < prune] <- 0
+  pglom_prune <- prune_taxa(taxa_sums(pglom_prune) > 0, pglom_prune)
+
+  # Melt into long format and sort by taxonomy
+  physeq_long <- psmelt(pglom_prune)
+  physeq_long <- physeq_long[order(physeq_long[ ,taxrank]), ]
+
+  # Return long data frame
+  return(physeq_long)
+}
+
+###### Merge functions ############
+
+# Merge samples by averaging OTU counts instead of summing
+merge_samples_mean <- function(physeq, group, round){
+  # Calculate the number of samples in each group
+  group_sums <- as.matrix(table(sample_data(physeq)[ ,group]))[,1]
+
+  # Merge samples by summing
+  merged <- merge_samples(physeq, group)
+
+  # Divide summed OTU counts by number of samples in each group to get mean
+  # Calculation is done while taxa are columns
+  x <- as.matrix(otu_table(merged))
+  if(taxa_are_rows(merged)){ x<-t(x) }
+
+  # Pick the rounding functions
+  if (round == "floor"){
+    out <- floor(t(x/group_sums))
+  } else if (round == "round"){
+    out <- myround(t(x/group_sums))
+  }
+
+  # Return new phyloseq object with taxa as rows
+  out <- otu_table(out, taxa_are_rows = TRUE)
+  otu_table(merged) <- out
+  return(merged)
+}
+
+# Merge samples, just including OTUs that were present in all merged samples
+# Call this function before running merge_samples()
+merge_OTU_intersect <- function(physeq, group){
+
+  # Make sure we're not starting with more taxa than we need 
+  physeq <- prune_taxa(taxa_sums(physeq) > 0, physeq)
+
+  s <- data.frame(sample_data(physeq))
+  l <- levels(s[,group])
+  o <- otu_table(physeq)
+
+  # Loop through each category
+  for (cat in 1:length(l)) {
+
+    # Get the index of all samples in that category
+    w <- which(s[,group]==l[cat])
+
+    # subset to just those columns of OTU table
+    cat.sub<-o[,w]
+    print(dim(cat.sub))
+
+    # Find the indices of 0's in the OTU table
+    zeros <- apply(cat.sub, 1, function(r) any(r == 0))
+
+    # If an OTU had a 0 in at least one sample, change all samples to 0
+    cat.sub[zeros,] <- 0
+  }
+
+  o[,w] <- cat.sub
+  otu_table(physeq) <- o
+
+  return(physeq)
 
@@
 
-# comp_d <- function(pseq,subset,title_text) {
-#   png(paste(title_text,"Phylum_abundance.png",sep="_"))
-#   p <- plot_bar(pseq.subset,"Phylum") + theme(axis.text.x = element_text(face="italic"))
-#   print(p)
-
-#   prev.otu <- plot_taxa_prevalence(pseq.subset, "Phylum")
-#   prev.otu + theme(legend.position="none")
-#   ggsave(paste(title_text,"phylum_prevalence.png",sep="_"))
-#   tab <- plot_taxa_prevalence(pseq.subset, "Phylum", detection = 5)
-#   tab + theme(legend.position="none")
-#   ggsave(paste(title_text,"Phylum_prevalence_detection.png",sep="_"))
-
-#   pseq.ph <- aggregate_taxa(pseq.subset, "Phylum")
-#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-#   plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
-#     theme(legend.position = "bottom") + theme_bw() +
-#     theme(axis.text.x=element_blank()) +
-#     c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
-#   print(plot.comp.rel)
-#   ggsave("phylum_relative_abundance11.png")
-#   plot.comp.rel + theme(legend.position="none")
-#   ggsave("phylum_relative_abundance12.png")
-
-#   pseq.ph <- aggregate_taxa(pseq, "Phylum")
-#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND   Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-#   ggsave("phylum_abundance_bar.png")
-#   # Top 5
-#   top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
-
-#   top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
-#   pseq.ph.rel <- microbiome::transform(top5P, "compositional")
-#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-
-#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
-
-#   # Top 10
-#   top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
-
-#   top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
-#   pseq.ph.rel <- microbiome::transform(top10P, "compositional")
-
-#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-
-#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-#   # new
-#   p.new <- aggregate_taxa(pseq, "Phylum")
-#   p.new.rel <- microbiome::transform(p.new, "compositional")
-
-#   plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
-
-#   plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
-#   ggsave("pnew_abundance_bar.png")
-
-#   # Top Phylum Heatmap
-#   pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
-#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
-#   plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-#   plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
-#   ggsave("phylum_histogram.png")
-
-#   #p.new heatmap
-#   pseq.n.ph <- aggregate_taxa(pseq, "Phylum")
-#   pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
-
-#   plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-#   p <- plot.n.comp.rel + coord_flip()
-#   p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-#   ggsave("pnew_histogram.png")
-
-#   # Newwork
-#   png("network_bray_phylum.png")
-#   jg = make_network(pseq, "taxa", "bray", 0.3)
-#   plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
-#   dev.off()
-#   #Community-composition-plots
-#   # Create a data table for ggplot
-# # agglomerate at phylum level
-# pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
-
-# # Transform to rel. abundance (or use ps0.ra)
-# pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
-
-# # Melt to long format for easy ggploting
-# pseq_phylum <- psmelt(pseq_phyl)
-
-#      # filter(Abundance > 0.01)                             # Filter out low abundance taxa
-# # Plot - Phylum
-#   p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-#   geom_bar(stat = "identity", width = 1) +
-#   facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-#   theme(axis.text.x = element_blank()) +
-#   theme(axis.title.x = element_blank()) +
-#   labs(title = "Abundant Phylum (> 1%)")
-
-#   p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-#   plot(p)
-#   ggsave("phylum_abun_mnd.png")
-# }
+}
 
@@
 
-alpha_d(pseq.subset,"TEST_3")
-beta_d(pseq.subset,"TEST_2")
-pre_perma_d(pseq.subset) 
-perma_d(p.low,p.high,"TESt_highVlow")
+alpha_d_two(pseq.subset,"cHL")
+beta_d_two(pseq.subset,"cHL")
+alpha_d_two(atlas.pseq,"aHL")
+beta_d_two(atlas.pseq,"aHL")
+
+pre_perma_d_two(atlas.pseq) 
+perma_d_two("1303_highVlow")
+pre_perma_d_two(pseq.subset) 
+perma_d_two("1253_highVlow")
 
@@
 Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
+
+NOTHING BELOW THIS WORKS!!!!!!!!
 

### 2019-11-11 curatedMetagenomic_NS_MRD.Rmd
updated

--- 2019-09-30 curatedMetagenomic_NS.Rmd
@@
 library(microbiome)
+library(forcats)
+library(stargazer)
+library(ggpubr)
+library(knitr)
+library(tidyverse)
+library(cowplot)
+library(vegan)
+library(broom)
+library(tsnemicrobiota)
+library(latex2exp)
+library(wesanderson)
 
@@
 
-## R Markdown
+```{r create mnd variable}
 
-```{r}
 # Create mnd variable
 # grabs the nationality from phyloseq
-test <- get_variable(pseq, "country" )
+mnd.country <- get_variable(pseq, "country" )
+table(mnd.country)
 
-# check you've installed this library
-library(forcats)
-table(test)
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-
 # Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,
 
-test <- fct_collapse(test, LOW = c("NOR","SWE","FIN") , MEDIUM = c("ITA","AUT","DEU","GBR","ESP","FRA","LUX","NLD"), HIGH = c("USA") )
+mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","LUX") , MEDIUM = c("DEN","ITA","DEU","ESP","FRA","NLD","GBR"), HIGH = c("USA","CAN") )
 
-#test <- fct_collapse(test, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+# "AUT", "NLD","DEU" was used in medium
+#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
 
 # reorder
-test <- factor(test, levels = (c("LOW", "MEDIUM", "HIGH")))
-levels(test)
+mnd.country <- factor(mnd.country, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(mnd.country)
 
 # creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = test
+sample_data(pseq)$mnd = mnd.country
 
@@
 get_variable(pseq, "mnd")
+
+```
+
+Defining the functions 
+```{r define the functions}
+
+# this defines the colours for the graphs. Uses the wes palettes. Look at https://github.com/karthik/wesanderson for the different themes
+glob_theme <- "FantasticFox1"
+
+# This is required
+mylab <- function(v1) {
+  deparse(substitute(v1))
+}
+
+# ALPHA #####
+# aplha_d runs the diversity for each country or  groupoing and then compares the three groups
+# example is alpha_d(pseq.subset,"TEST_3") this will generate a png file called TEST_3.png in the 
+# currect working directory. This can be checked with getwd()
+alpha_d <- function(pseq.subset,title_text) {
+
+   # gets the meta data 
+   pseq.meta <- meta(pseq.subset)
+
+   # generates the diversites
+   tab <- microbiome::alpha(pseq.subset, index="shannon")
+
+   # creates Shannon
+   pseq.meta$shannon <- tab$diversity_shannon
+   mnd <- levels(pseq.meta$mnd) # get the variables
+
+    # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+   # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+  p1<-   ggviolin(pseq.meta, x = "country", y = "Shannon",
+ add = "boxplot", fill = "country", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+# the number in the palettes must equal the number of groups in bmi.pairs
+
+   p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = wes_palette(glob_theme))
+
+    p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+    #this just generates the titles for the output
+    title <- ggdraw() + 
+    draw_label(
+    title_text,
+    fontface = 'bold',
+    x = 0,
+    hjust = 0
+    ) +
+    theme(
+    # add margin on the left of the drawing canvas,
+    # so title is aligned with left edge of first plot
+    plot.margin = margin(0, 0, 0, 7)
+    )
+
+    p3 <<-plot_grid(title,p1,p2, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
+    ggsave(paste("Alpha_diversity_",title_text,".png", sep = "" ), device = "png")
+}
+
+# BETA #####
+# This includes the analysis as well so can take some time to run. the graphs are generated with the label of beta diversity. It requires an input of a title. 
+beta_d <- function(pseq.subset,title_text) {
+
+   #NMDS ordination on Bray-Curtis distance
+  p <<-  boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+ # p1 = plot_ordination(pseq.subset, pseq.nmds, type = "taxa", color = "mnd", title = "Taxa by Phylum") # I dont think that we need this any more. 
+  #p1 + facet_wrap(~Phylum, 3)
+  pseq.nmds <- ordinate(pseq.subset, "NMDS", "bray")
+  p2 <- plot_ordination(pseq.subset, pseq.nmds, type = "sample", color = "mnd")
+
+  p3 <- p2 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND Prevalence") + theme(legend.position = "none")
+  #plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+
+  # This add the facets but I do not think we need that any more. 
+  #p3 <- p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+  ggsave(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), device = "png")
+    # MDS
+   pseq.mds <<- ordinate(pseq.subset, "MDS", "bray")
+   p4 <- plot_ordination(pseq.subset, pseq.mds, type = "sample", color = "mnd")
+ #plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence MDS"))
+
+    p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND Prevalence") + theme(legend.position = "none")
+    ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+
+   #PCoA for compositional data with Bray-Curtis distances
+    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
+    #p6 <- plot_landscape(pseq.subset, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+  #   # PCoA for compositional data with Bray-Curtis distances
+   #p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+    #                    method = "PCoA", distance = "bray") +
+    #     labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+  # 
+    tsne_res <- tsne_phyloseq(pseq.subset, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+    p7 <- plot_tsne_phyloseq(pseq.subset, tsne_res,color = 'mnd')
+    p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND Prevalence") + theme(legend.position = "bottom")
+    ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+       #this just generates the titles for the output
+    title <- ggdraw() + 
+    draw_label(
+    title_text,
+    fontface = 'bold',
+    x = 0,
+    hjust = 0
+    ) +
+    theme(
+    # add margin on the left of the drawing canvas,
+    # so title is aligned with left edge of first plot
+    plot.margin = margin(0, 0, 0, 7)
+    )
+
+    p10 <<-plot_grid(title,p3, p5,p8, ncol = 1, labels = c('','A', 'B','C'),rel_heights = c(0.3, 1.5,1.5,2))
+    ggsave(paste("Beta_diversity_ALL_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+}
+
+pre_perma_d <- function(pseq.subset) { 
+  p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+}
+
+# This calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
+perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
+  # Convert to compositional data
+  pseqLvH <- merge_phyloseq(x,y) # pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p <- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal()  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
+  ggsave(paste("PERMANOVA_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+
+ }
+
+# comp_d <- function(pseq,subset,title_text) {
+#   png(paste(title_text,"Phylum_abundance.png",sep="_"))
+#   p <- plot_bar(pseq.subset,"Phylum") + theme(axis.text.x = element_text(face="italic"))
+#   print(p)
+
+#   prev.otu <- plot_taxa_prevalence(pseq.subset, "Phylum")
+#   prev.otu + theme(legend.position="none")
+#   ggsave(paste(title_text,"phylum_prevalence.png",sep="_"))
+#   tab <- plot_taxa_prevalence(pseq.subset, "Phylum", detection = 5)
+#   tab + theme(legend.position="none")
+#   ggsave(paste(title_text,"Phylum_prevalence_detection.png",sep="_"))
+
+#   pseq.ph <- aggregate_taxa(pseq.subset, "Phylum")
+#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
+#     theme(legend.position = "bottom") + theme_bw() +
+#     theme(axis.text.x=element_blank()) +
+#     c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))
+#   print(plot.comp.rel)
+#   ggsave("phylum_relative_abundance11.png")
+#   plot.comp.rel + theme(legend.position="none")
+#   ggsave("phylum_relative_abundance12.png")
+
+#   pseq.ph <- aggregate_taxa(pseq, "Phylum")
+#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND   Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+#   ggsave("phylum_abundance_bar.png")
+#   # Top 5
+#   top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
+
+#   top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
+#   pseq.ph.rel <- microbiome::transform(top5P, "compositional")
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")
+
+#   # Top 10
+#   top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
+
+#   top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
+#   pseq.ph.rel <- microbiome::transform(top10P, "compositional")
+
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
+
+#   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
+#   # new
+#   p.new <- aggregate_taxa(pseq, "Phylum")
+#   p.new.rel <- microbiome::transform(p.new, "compositional")
+
+#   plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
+
+#   plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
+#   ggsave("pnew_abundance_bar.png")
+
+#   # Top Phylum Heatmap
+#   pseq.ph <- aggregate_taxa(pseq, "Phylum")
+
+#   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+#   plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+#   plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
+#   ggsave("phylum_histogram.png")
+
+#   #p.new heatmap
+#   pseq.n.ph <- aggregate_taxa(pseq, "Phylum")
+#   pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
+
+#   plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
+
+#   p <- plot.n.comp.rel + coord_flip()
+#   p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
+#   ggsave("pnew_histogram.png")
+
+#   # Newwork
+#   png("network_bray_phylum.png")
+#   jg = make_network(pseq, "taxa", "bray", 0.3)
+#   plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
+#   dev.off()
+#   #Community-composition-plots
+#   # Create a data table for ggplot
+# # agglomerate at phylum level
+# pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
+
+# # Transform to rel. abundance (or use ps0.ra)
+# pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})
+
+# # Melt to long format for easy ggploting
+# pseq_phylum <- psmelt(pseq_phyl)
+
+#      # filter(Abundance > 0.01)                             # Filter out low abundance taxa
+# # Plot - Phylum
+#   p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
+#   geom_bar(stat = "identity", width = 1) +
+#   facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
+#   theme(axis.text.x = element_blank()) +
+#   theme(axis.title.x = element_blank()) +
+#   labs(title = "Abundant Phylum (> 1%)")
+
+#   p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
+#   plot(p)
+#   ggsave("phylum_abun_mnd.png")
+# }
 
@@
 ```
+
+## Prune data
+```{r prune data}
+
+# https://bioconductor.riken.jp/packages/3.5/data/experiment/vignettes/curatedMetagenomicData/inst/doc/curatedMetagenomicData.html
+# Prepare data for vizualisation
+
+# Remove   everything without a country
+
+keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
+pseq.f <- subset_taxa(pseq, keepotu)
+
+pseq.c <- core(pseq, detection = 0.1/100, prevalence = 50/100)
+
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" & antibiotics_current_use != "yes")
+
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
+
+# These are useful if filtering. 
+#sample_variables(pseq.subset)
+#get_variable(pseq.subset, sample_variables(pseq.subset)[13])
+
+#alpha_d(pseq.subset,"TEST_3")
+#beta_d(pseq.subset,"TEST_2")
+#pre_perma_d(pseq.subset) 
+#perma_d(p.low,p.high,"TESt_highVlow")
+
+# check again if any OTUs are not present in any samples
+any(taxa_sums(pseq.c) == 0)
+
+```
+
+```{r MND table summary}
+
+library(table1)
+n <- length(sample_data(pseq.subset)$mnd)
+
+data <- data.frame(id=1:n)
+data$mnd <- sample_data(pseq.subset)$mnd
+data$bmi <- sample_data(pseq.subset)$BMI
+data$age <- sample_data(pseq.subset)$age
+data$antibiotics_current_use <- sample_data(pseq.subset)$antibiotics_current_use
+data$smoker <- sample_data(pseq.subset)$smoker
+data$pregnant <- sample_data(pseq.subset)$pregnant
+data$alcohol <- sample_data(pseq.subset)$alcohol
+data$age_category <- sample_data(pseq.subset)$age_category
+
+table1(~ age  +  bmi + smoker + alcohol + age_category | mnd, data=data)
+
+```
+
+```{r taxa overview}
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq.subset)
+print(taxa)
+
+table(tax_table(pseq.subset)[, "Phylum"], exclude = NULL)
+
+```
+
+```{r phylogenetic tree}
+
+```
+
+```{r COMMANDS}
+
+# alpha_d(pseq.subset,“TEST_3”)
+# beta_d(pseq.subset,“TEST_2")
+# pre_perma_d(pseq.subset)
+# perma_d(p.low,p.high,“TESt_highVlow”)
+
+```
+
 ## Check OTUs/Phylums
@@
 
-head(meta(pseq))
+# remove OTUs with zero occurances
+ps1 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
+
+# check again if any OTUs are not present in any samples
+any(taxa_sums(ps1) == 0)
+
+ntaxa(pseq) - ntaxa(ps1)
+
+# OR can prune taxa through establishing a CORE
+ntaxa(pseq) - ntaxa(pseq.c)
 
@@
 
-## Prepare Data for Alpha Diversity Visualisation
-```{r prepare data for alpha diversity visualisation}
+```{r check distribution or how many reads/samples}
 
+SeqDepth = colSums(otu_table(pseq.subset))
+sample_data(pseq.subset)$SeqDepth = SeqDepth
+qplot(log10(SeqDepth), geom = "histogram") + theme_bw()
+
+ggplot(meta(pseq.subset)) + geom_histogram(aes(x = log10(SeqDepth)), alpha= 0.6) + facet_wrap(~mnd) + theme_bw()
+
+```
+```{r OTU prevalence}
+
+prev.otu <- plot_taxa_prevalence(pseq.subset, "Phylum")
+
+print(prev.otu)
+
+### Should we remove low abundance OTUs????
+
+```
+
+## Alpha Diversity
+```{r alpha diversity }
+
+# https://bioconductor.riken.jp/packages/3.5/data/experiment/vignettes/curatedMetagenomicData/inst/doc/curatedMetagenomicData.html
 # Prepare data for vizualisation
 
-# Alpha diversity again
-png("alpha.png")
+keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
+pseq.f <- subset_taxa(pseq, keepotu)
 
-pseq2 <- pseq
+# Pick the core (>0.1% relative abundance in >50% of the samples)
+pseq.c <- core(pseq, detection = 0.1/100, prevalence = 50/100)
 
-#this is a work around. Need to check
-otu_table(pseq2) <- otu_table(round(as((otu_table(pseq2)), "matrix")), taxa_are_rows(pseq2))
-estimate_richness(pseq2)
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn")
 
-tab <- diversities(pseq2, index = "all")
-stargazer(tab, type = "text",summary = NULL)
-kable(head(tab))
-plot(tab)
-dev.off()
+# These are useful is filtering. 
+#sample_variables(pseq.subset)
+#get_variable(pseq.subset, sample_variables(pseq.subset)[4])
+
+any(taxa_sums(pseq.c) == 0)
+
+# run the diversities. 	keep an eye on this as only one is working
+# tab <- diversities(pseq.subset, index = "shannon")
+tab <- microbiome::alpha(pseq.subset, index="shannon")
 
 # Get metadata
-pseq.meta <- meta(pseq)
-kable(head(pseq.meta))
-stargazer(pseq.meta, type = "text",summary = NULL)
+pseq.meta <- meta(pseq.subset)
+
+# Issues with kable ????????????????
+#kable(head(pseq.meta))
+#stargazer(pseq.meta, type = "text",summary = NULL)
 
 # Add the diversity table to metadata
-pseq.meta$Shannon <- tab$shannon 
-pseq.meta$InverseSimpson <- tab$inverse_simpson
-pseq.meta$fisher <- tab$fisher
-pseq.meta$chao1 <- tab$chao1
+pseq.meta$shannon <- tab$diversity_shannon 
+#pseq.meta$InverseSimpson <- tab$inverse_simpson
+#pseq.meta$fisher <- tab$fisher
+#pseq.meta$chao1 <- tab$chao1
+
+# Compare differences in Shannon index between mnd group of the study subjects
+pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+
+# Create a list of pairwise comaprisons
+mnd <- levels(pseq.meta$mnd) # get the variables
+
+# Make a pairwise list that we want to compare.
+mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+# Shannon violin plot
+# the number in the palettes must equal the number of groups in bmi.pairs
+p1 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+p1 <- p1 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+print(p1)
+#ggsave("mnd_violin__shannon_stats.png")
+
+# This subsets the  genus and phylum. 
+# genus.pseq <- subset_taxa( pseq.subset, !is.na(Genus))
+# phylum.pseq <- subset_taxa( pseq, is.na(Class) & !is.na(Phylum))
 
 ```
+
+# Beta diversity
+```{r beta diveristy }
+
+ #NMDS ordination on Bray-Curtis distance
+  p <<-  boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+  #p1 + facet_wrap(~Phylum, 3)
+  pseq.nmds <- ordinate(pseq, "NMDS", "bray")
+  p2 <- plot_ordination(pseq, pseq.nmds, type = "sample", color = "mnd")
+
+  p3 <- p2 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND Prevalence") + theme(legend.position = "none")
+  #plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+
+  # This add the facets but I do not think we need that any more. 
+  #p3 <- p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+  #### ggsave(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), device = "png")
+    # MDS
+   pseq.mds <<- ordinate(pseq, "MDS", "bray")
+   p4 <- plot_ordination(pseq, pseq.mds, type = "sample", color = "mnd")
+ #plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence MDS"))
+
+    p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND Prevalence") + theme(legend.position = "none")
+   #### ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+
+   #PCoA for compositional data with Bray-Curtis distances
+    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
+    #p6 <- plot_landscape(pseq.subset, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+  #   # PCoA for compositional data with Bray-Curtis distances
+   #p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+    #                    method = "PCoA", distance = "bray") +
+    #     labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+  # 
+    tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+    p7 <- plot_tsne_phyloseq(pseq, tsne_res,color = 'mnd')
+    p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND Prevalence") + theme(legend.position = "bottom")
+   #### ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+       #this just generates the titles for the output
+    title <- ggdraw() + 
+    draw_label(
+    title_text,
+    fontface = 'bold',
+    x = 0,
+    hjust = 0
+    ) +
+    theme(
+    # add margin on the left of the drawing canvas,
+    # so title is aligned with left edge of first plot
+    plot.margin = margin(0, 0, 0, 7)
+    )
+
+    p10 <<-plot_grid(title,p3, p5,p8, ncol = 1, labels = c('','A', 'B','C'),rel_heights = c(0.3, 1.5,1.5,2))
+   ##### ggsave(paste("Beta_diversity_ALL_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+
+```
+
+```{r beta pt.2}
+
+pseq.rel <- microbiome::transform(pseq.subset, "compositional")
+
+bx.ord_pcoa_bray <- ordinate(pseq.rel, "PCoA", "bray")
+
+beta.pseq <- plot_ordination(pseq.rel, 
+                            bx.ord_pcoa_bray, 
+                            color="mnd") + 
+  geom_point(size= 4) + 
+  theme(plot.title = element_text(hjust = 0, size = 12))
+
+beta.ps3 <- beta.pseq + theme_bw(base_size = 14) + 
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
+beta.ps3 + scale_color_brewer(palette = "Dark2") + stat_ellipse()
+
+```
+
+```{r}
+
+bx.ord_pcoa_bray <- ordinate(pseq.rel, "NMDS", "bray")
+
+beta.pseq <- plot_ordination(pseq.rel, 
+                            bx.ord_pcoa_bray, 
+                            color="mnd") + 
+  geom_point(size= 4) + 
+  theme(plot.title = element_text(hjust = 0, size = 12))
+
+beta.ps3 <- beta.pseq + theme_bw(base_size = 14) + 
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
+beta.ps3 + scale_color_brewer(palette = "Dark2") + stat_ellipse()
+
+```
+
+# Prepare for PERMANOVA test
+```{r Calculate group divergences}
+
+# Use these to compare microbiota divergence within each group
+
+p.low <- subset_samples(pseq, mnd == "LOW")
+p.low.d <- divergence(p.low)
+
+p.med <- subset_samples(pseq, mnd == "MEDIUM")
+p.med.d <- divergence(p.med)
+
+p.high <- subset_samples(pseq, mnd == "HIGH")
+p.high.d <- divergence(p.high)
+
+p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+plot(p)
+#  The group tends with smaller values, indicates that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
+
+```
+
+## PERMANOVA significance test for group-level differences
+Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
+
+```{r Redefine data}
+
+# Convert to compositional data
+pseqLvH <- merge_phyloseq(p.low,p.high)
+
+pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+otu <- abundances(pseq.rel)
+meta <- meta(pseq.rel)
+
+```
+
+```{r PERMANOVA}
+
+# samples x species as input
+library(vegan)
+permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+
+# P-value
+print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+
+summary(permanova$aov.tab)
+
+```
+
+Homogeneity of variance is an assumption underlying both t tests and F tests (analyses of variance, ANOVAs) in which the population variances (i.e., the distribution, or “spread,” of scores around the mean) of two or more samples are considered equal.
+Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
+```{r ANOVA}
+
+dist <- vegdist(t(otu))
+anov <- anova(betadisper(dist, meta$mnd))
+summary(anov)
+tidy(anov)
+summary(anov$aov.tab)
+
+```
+
+Show coefficients for the top taxa separating the groups
+```{r top taxa L vs H}
+
+coef <- coefficients(permanova)["mnd1",]
+top.coef <- coef[rev(order(abs(coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis=3, main = "Top Genus")
+#ggsave("LvH.png", width = 20, height = 20, units = "cm", dpi=700)
+
+```
+
+# Taxonomy histogram
+```{r taxonomy histogram}
+
+pseq.sp = subset_taxa(pseq, !is.na(Species) & is.na(Strain))
+par(mar = c(20, 4, 0, 0) + 0.15) #increase margin size on the bottom
+barplot(sort(taxa_sums(pseq.sp), TRUE)[1:20] / nsamples(pseq.sp),
+        ylab = "Total counts", las = 2)
+```
+
+# Phylogenetic tree
+```{r Phylogenetic tree}
+
+plot_tree(pseq.subset, color = "mnd", label.tips = "Species", size = "abundance", plot.margin = 0.5, ladderize = TRUE)
+
+```
+
+```{r phylum composition}
+
+plot_bar(pseq.subset, x="mnd", fill="Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")
+
+```
+
+```{r}
+
+# Create table, number of features for each phyla
+table(tax_table(pseq)[, "Phylum"], exclude = NULL)
+
+table(tax_table(pseq.subset)[, "Phylum"], exclude = NULL)
+
+```
+
+```{r akkermansia}
+
+# use this to compare more easily differences in scale and distribution of the abundance values in our phyloseq object before and after transformation
+
+# How many genera would be present after filtering?
+length(get_taxa_unique(pseq.subset, taxonomic.rank = "Genus"))
+ps3 = tax_glom(pseq.subset, "Genus", NArm = TRUE)
+
+# Akkermansia muciniphila
+# Phylum = Verrucomicrobia
+# Genus = Akkermansia
+
+taxa <- map_levels(NULL, "Phylum", "Genus", pseq.subset)$Verrucomicrobia
+print(taxa)
+
+plot_abundance = function(pseq,title = "",
+                          Facet = "Genus", Color = "Phylum"){
+  # Arbitrary subset, based on Phylum, for plotting
+  p1f = subset_taxa(pseq.subset, Genus %in% c("Akkermansia"))
+  mphyseq = psmelt(p1f)
+  mphyseq <- subset(mphyseq, Abundance > 0)
+  ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance",
+                              color = Color, fill = Color)) +
+    geom_violin(fill = NA) +
+    geom_point(size = 1, alpha = 0.3,
+               position = position_jitter(width = 0.3)) +
+    facet_wrap(facets = Facet) + scale_y_log10()+
+    theme(legend.position="none")
+}
+
+# Transform to relative abundance. Save as new object.
+ps3ra = transform_sample_counts(ps3, function(x){x / sum(x)})
+
+plotBefore = plot_abundance(ps3,"")
+plotAfter = plot_abundance(ps3ra,"")
+# Combine each plot into one graphic.
+grid.arrange(nrow = 2,  plotBefore, plotAfter)
+
+```
+
+```{r subset permanova}
+
+p.low <- subset_samples(pseq.subset, mnd == "LOW")
+p.low.d <- divergence(p.low)
+
+p.med <- subset_samples(pseq.subset, mnd == "MEDIUM")
+p.med.d <- divergence(p.med)
+
+p.high <- subset_samples(pseq.subset, mnd == "HIGH")
+p.high.d <- divergence(p.high)
+
+# Convert to compositional data
+pseqLvH <- merge_phyloseq(p.low,p.high)
+
+pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+otu <- abundances(pseq.rel)
+meta <- meta(pseq.rel)
+
+# samples x species as input
+library(vegan)
+permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+
+# P-value
+print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])
+
+summary(permanova$aov.tab)
+
+coef <- coefficients(permanova)["mnd1",]
+top.coef <- coef[rev(order(abs(coef)))[1:20]]
+par(mar = c(3, 14, 2, 1))
+barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, font.axis=3, main = "Top Genus")
+
+```

### 2019-12-03 combodata_NS.Rmd
updated

--- 2019-10-20 curatedMetagenomic_NS.Rmd
@@
 
+## Setup
 ```{r setup, include=FALSE}
@@
 library(wesanderson)
+library(table1)
 
@@
 orig_pseq <- pseq
+
 #For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object.
 curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:
+data(atlas1006)
+atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
 
@@
 
-## R Markdown
+## MND Countries
+```{r tables}
 
-```{r}
+# THIS FOR THE CURATED DATA SET
+
+# creates a grouping varaible for BMI
+n <- length(sample_data(pseq)$BMI)
+data <- data.frame(id=1:n)
+data$bmi <- sample_data(pseq)$BMI # creates a new variable with BMI in it
+# Classification based upon the ATlas data
+data$bmi[data$bmi <18.5] <- "underweight"
+data$bmi[data$bmi >=45] <- "super obese"
+data$bmi[data$bmi >= 18.5 & data$bmi <25] <- "lean"
+data$bmi[data$bmi >= 25 & data$bmi <30] <- "overweight"
+data$bmi[data$bmi >= 30 & data$bmi <35] <- "obese"
+data$bmi[data$bmi >= 35 & data$bmi <40] <- "severe obese"
+data$bmi[data$bmi >= 40 & data$bmi <45] <- "morbid obese"
+data$bmi <- factor(data$bmi)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(pseq)$bmi_group = data$bmi
+# Check that it works
+get_variable(pseq, "bmi_group")
+
 # Create mnd variable
-pseq <- orig_pseq
+#pseq <- orig_pseq
 # grabs the nationality from phyloseq
-mnd.country <- get_variable(pseq, "country" )
-table(mnd.country)
+c.mnd.country <- get_variable(pseq, "country" )
+table(c.mnd.country)
 
@@
 
-mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DEN","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
+#Adjsuted to match exactly the ATLAS data set
+
+#central Europe = Belgium, Demark, Netherlands, Germany
+#southern Europe = France, Italy, Serbia, Spain
+#UKIE
+#Scandinavia = Norway, Sweden Finland
+#Eastern Europe = Poland
+
+c.mnd.atlas <- fct_collapse(c.mnd.country, Scandinavia = c("NOR","SWE","FIN") ,CentralEurope = c("DNK","DEU","NLD"), SouthEurope = c("ITA","ESP","FRA"), US = "USA", UKIE = "GBR" )
+
+#mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DNK","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
 
@@
 # reorder
-mnd.country <- factor(mnd.country, levels = (c("LOW", "MEDIUM", "HIGH")))
-levels(mnd.country)
+# NOTE THERE IS NO EASTERN EUROPE
+c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
+
+c.mnd.nationality <- factor(c.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(c.mnd.nationality)
 
 # creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = mnd.country
+sample_data(pseq)$mnd = c.mnd.nationality
 
@@
 get_variable(pseq, "mnd")
+table(get_variable(pseq, "mnd"))
+
+###############################################
+
+# THIS IS FOR THE ATLAS 1006 
+# Create mnd variable
+# grabs the nationality from phyloseq
+a.mnd.nationality <- get_variable(atlas.pseq, "nationality" )
+
+# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
+
+# reorder
+a.mnd.nationality <- factor(a.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(a.mnd.nationality)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(atlas.pseq)$mnd = a.mnd.nationality
+#sample_data(atlas.pseq)$country = sample_data(atlas.pseq)$nationality
+# checks that it has worked.
+table(get_variable(atlas.pseq, "mnd"))
 
@@
 
-Defining the functions 
-```{r define the functions}
+##Functions 
+```{r function}
 
@@
 glob_theme <- "FantasticFox1"
+
+my_cur_table <- function(pseq.subset){
+    n <- length(sample_data(pseq.subset)$mnd)
+    data <- data.frame(id=1:n)
+    data$mnd <- sample_data(pseq.subset)$mnd
+    data$bmi <- sample_data(pseq.subset)$BMI
+    data$bmi_group <- sample_data(pseq.subset)$bmi_group
+    data$age <- sample_data(pseq.subset)$age
+    data$antibiotics_current_use <- sample_data(pseq.subset)$antibiotics_current_use
+    data$smoker <- sample_data(pseq.subset)$smoker
+    data$pregnant <- sample_data(pseq.subset)$pregnant
+    data$alcohol <- sample_data(pseq.subset)$alcohol
+    data$age_category <- sample_data(pseq.subset)$age_category
+    table1(~ age  +  bmi + bmi_group + smoker + alcohol + age_category | mnd, data=data)
+}
+
+my_atlas_table <- function(atlas.pseq){
+    n <- length(sample_data(atlas.pseq)$mnd)
+    data <- data.frame(id=1:n)
+    data$mnd <- sample_data(atlas.pseq)$mnd
+    data$bmi <- sample_data(atlas.pseq)$bmi_group
+    data$age <- sample_data(atlas.pseq)$age
+    table1(~ age  +  bmi | mnd, data=data)
+}
 
@@
   # generates the diversites
-   tab <- diversities(pseq.subset, index = "Shannon")
+  tab <- microbiome::alpha(pseq.subset, index = "shannon")
+
   # creates Shannon
-   pseq.meta$Shannon <- tab$shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
   mnd <- levels(pseq.meta$mnd) # get the variables
@@
   # on to the graphs!!!
+  print("The basic aloha diversity analysis has been completed")
+  # This country versus nationality
   p1<-   ggviolin(pseq.meta, x = "country", y = "Shannon",
                   add = "boxplot", fill = "country", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+  print("First graph has completed")
   # the number in the palettes must equal the number of groups in bmi.pairs
@@
   p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+
+  print("Second graph has completed")
   #this just generates the titles for the output
@@
   p3 <<-plot_grid(title,p1,p2, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
-    ggsave(paste("Alpha_diversity_",title_text,".png", sep = "" ), device = "png")
+
+  print("Third graph has completed")
+  ggsave(paste("/Users/Nik/Alpha_diversity_",title_text,".png", sep = "" ), device = "png")
+  #ggsave("/Users/Nik/Alpha_diversity_curate.png")
 }
@@
 perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
-  # Convert to compositional data
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
   pseqLvH <- merge_phyloseq(x,y) # pseqLvH <- merge_phyloseq(p.low,p.high)
@@
   #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
-  p <- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal()  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
   ggsave(paste("PERMANOVA_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
-
+  #ggsave("test_permanova.png", device = "png",width = 297, height = 210, units = "mm")
  }
 
+# testpseq <- atlas1006 %>%
+#    subset_samples(DNA_extraction_method == "r") %>%
+#    aggregate_taxa(level = "Phylum") %>%
+#    transform(transform = "compositional")
+# p <- plot_composition(testpseq, sample.sort = "Firmicutes",
+#          otu.sort = "abundance", verbose = TRUE) +
+#      scale_fill_manual(values = default_colors("Phylum")[taxa(testpseq)]) 
+# 
 # comp_d <- function(pseq,subset,title_text) {
@@
 #   print(p)
-
-#   prev.otu <- plot_taxa_prevalence(pseq.subset, "Phylum")
+# 
+#   prev.otu <- plot_taxa_prevalence(pseq.subset, "Phylum" , detection = 5)
 #   prev.otu + theme(legend.position="none")
@@
 #   ggsave(paste(title_text,"Phylum_prevalence_detection.png",sep="_"))
-
+#   
 #   pseq.ph <- aggregate_taxa(pseq.subset, "Phylum")
@@
 #   ggsave("phylum_relative_abundance12.png") 
-
-#   pseq.ph <- aggregate_taxa(pseq, "Phylum")
+#   
+#   pseq.ph <- aggregate_taxa(pseq.subset, "Phylum")
 #   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
@@
 #   top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]
-
+# 
 #   top5P = subset_taxa(pseq, Genus %in% names(top5P.names))
-#   pseq.ph.rel <- microbiome::transform(top5P, "compositional")
+# u  pseq.ph.rel <- microbiome::transform(top5P, "compositional")
 #   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-
+# 
 #   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
-
+#   
 #   # Top 10
 #   top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]
-
+# 
 #   top10P = subset_taxa(pseq, Genus %in% names(top10P.names))
 #   pseq.ph.rel <- microbiome::transform(top10P, "compositional")
-
+# 
 #   plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))
-
+# 
 #   plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
@@
 #   p.new.rel <- microbiome::transform(p.new, "compositional")
-
+# 
 #   plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
-
+# 
 #   plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
 #   ggsave("pnew_abundance_bar.png")
-
+#   
 #   # Top Phylum Heatmap
 #   pseq.ph <- aggregate_taxa(pseq, "Phylum")
-
+# 
 #   pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
-
+# 
 #   plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")
-
+# 
 #   plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
 #   ggsave("phylum_histogram.png")
-
+#   
 #   #p.new heatmap
@@
 #   pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")
-
+# 
 #   plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")
-
+# 
 #   p <- plot.n.comp.rel + coord_flip()
@@
 #   ggsave("pnew_histogram.png")
-
+#   
 #   # Newwork
@@
 # # agglomerate at phylum level
-# pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")
-
+# pseq_phy <- tax_glom(pseq.subset, taxrank = "Phylum")                  
+# 
 # # Transform to rel. abundance (or use ps0.ra)
 # pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)}) 
-
+# 
 # # Melt to long format for easy ggploting
 # pseq_phylum <- psmelt(pseq_phyl) 
-
+#  
 #      # filter(Abundance > 0.01)                             # Filter out low abundance taxa
@@
 #     labs(title = "Abundant Phylum (> 1%)")
-
+# 
 #   p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
@@
 #   ggsave("phylum_abun_mnd.png")
+# 
+#  # these are other functions required along the way. 
+# 
+#     # transform counts to n
+#   physeq.scale <- transform_sample_counts(physeq, 
+#     function(x) {(n * x/sum(x))}
+#   )
+#   
+#   # Pick the rounding functions
+#   if (round == "floor"){
+#     otu_table(physeq.scale) <- floor(otu_table(physeq.scale))
+#   } else if (round == "round"){
+#     otu_table(physeq.scale) <- myround(otu_table(physeq.scale))
 #   }
+#   
+#   # Prune taxa and return new phyloseq object
+#   physeq.scale <- prune_taxa(taxa_sums(physeq.scale) > 0, physeq.scale)
+#   return(physeq.scale)
+# }
+#
+
+##### ADONIS ###########
+
+# Function to run adonis test on a phyloseq object and a variable from metadata
+# Make sure OTU data is standardized/normalized before 
+phyloseq_to_adonis <- function(physeq, distmat = NULL, dist = "bray", formula) {
+
+  if(!is.null(distmat)) {
+    phydist <- distmat
+  } else {
+    phydist <- phyloseq::distance(physeq, dist)
+  }
+
+  metadata <- as(sample_data(physeq), "data.frame")
+
+  # Adonis test
+  f <- reformulate(formula, response = "phydist")
+  adonis.test <- adonis(f, data = metadata)
+  print(adonis.test)
+
+  # Run homogeneity of dispersion test if there is only 1 variable
+  if (grepl("\\+", formula)) {
+    l <- list(
+      dist = phydist, 
+      formula = f, 
+      adonis = adonis.test
+    )
+  } else {
+    group <- metadata[ ,formula]
+    beta <- betadisper(phydist, group)
+    disper.test = permutest(beta)
+    print(disper.test)
+
+    l <- list(
+      dist = phydist, 
+      formula = f, 
+      adonis = adonis.test, 
+      disper = disper.test
+    )
+  }
+  return (l)
+}
+
+########## Bar Plots #################
+
+# This function takes a phyloseq object, agglomerates OTUs to the desired taxonomic rank, 
+# prunes out OTUs below a certain relative proportion in a sample (ie 1% ) 
+# and melts the phyloseq object into long format which is suitable for ggplot stacked barplots.
+taxglom_and_melt <- function(physeq, taxrank, prune){
+
+  # Agglomerate all otu's by given taxonomic level
+  pglom <- tax_glom(physeq, taxrank = taxrank)
+
+  # Create a new phyloseq object which removes taxa from each sample below the prune parameter
+  pglom_prune <- transform_sample_counts(pglom,function(x) {x/sum(x)})
+  otu_table(pglom_prune)[otu_table(pglom_prune) < prune] <- 0
+  pglom_prune <- prune_taxa(taxa_sums(pglom_prune) > 0, pglom_prune)
+
+  # Melt into long format and sort by taxonomy
+  physeq_long <- psmelt(pglom_prune)
+  physeq_long <- physeq_long[order(physeq_long[ ,taxrank]), ]
+
+  # Return long data frame
+  return(physeq_long)
+}
+
+###### Merge functions ############
+
+# Merge samples by averaging OTU counts instead of summing
+merge_samples_mean <- function(physeq, group, round){
+  # Calculate the number of samples in each group
+  group_sums <- as.matrix(table(sample_data(physeq)[ ,group]))[,1]
+
+  # Merge samples by summing
+  merged <- merge_samples(physeq, group)
+
+  # Divide summed OTU counts by number of samples in each group to get mean
+  # Calculation is done while taxa are columns
+  x <- as.matrix(otu_table(merged))
+  if(taxa_are_rows(merged)){ x<-t(x) }
+
+  # Pick the rounding functions
+  if (round == "floor"){
+    out <- floor(t(x/group_sums))
+  } else if (round == "round"){
+    out <- myround(t(x/group_sums))
+  }
+
+  # Return new phyloseq object with taxa as rows
+  out <- otu_table(out, taxa_are_rows = TRUE)
+  otu_table(merged) <- out
+  return(merged)
+}
+
+# Merge samples, just including OTUs that were present in all merged samples
+# Call this function before running merge_samples()
+merge_OTU_intersect <- function(physeq, group){
+
+  # Make sure we're not starting with more taxa than we need 
+  physeq <- prune_taxa(taxa_sums(physeq) > 0, physeq)
+
+  s <- data.frame(sample_data(physeq))
+  l <- levels(s[,group])
+  o <- otu_table(physeq)
+
+  # Loop through each category
+  for (cat in 1:length(l)) {
+
+    # Get the index of all samples in that category
+    w <- which(s[,group]==l[cat])
+
+    # subset to just those columns of OTU table
+    cat.sub<-o[,w]
+    print(dim(cat.sub))
+
+    # Find the indices of 0's in the OTU table
+    zeros <- apply(cat.sub, 1, function(r) any(r == 0))
+
+    # If an OTU had a 0 in at least one sample, change all samples to 0
+    cat.sub[zeros,] <- 0
+  }
+
+  o[,w] <- cat.sub
+  otu_table(physeq) <- o
+
+  return(physeq)
+
+}
 
@@
 
-## Check OTUs/Phylums
-Prune(remove) taxa because we want to remove OTUs/taxa that aren't in any of the samples
-Don't need to prune this taxa because OTUs are found in all samples.
-Check this everytime you use the subset_samples function because of the pseq object being filtered. 
 ```{r check OTUs}
@@
 
-## Prepare Data for Alpha Diversity Visualisation
-```{r prepare data for alpha diversity visualisation}
+## Filtering
+```{r  data filtering}
 
@@
 
-# Remove   everything without a country
+# Rmove   everything without a country
 
@@
 
-pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" & antibiotics_current_use != "yes")
+#pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" & antibiotics_current_use != "yes")
 
@@
 #get_variable(pseq.subset, sample_variables(pseq.subset)[13])
+my_cur_table(pseq.subset)
+my_cur_table(pseq.c)
 
-alpha_d(pseq.subset,"TEST_3")
-beta_d(pseq.subset,"TEST_2")
+my_atlas_table(atlas.pseq)
+alpha_d(pseq,"final_RLWT_1")
+
+alpha_d(atlas.pseq,"final_RLWT_2")
+
+beta_d(pseq.subset,"final_2")
+
 pre_perma_d(pseq.subset) 
-perma_d(p.low,p.high,"TESt_highVlow")
+perma_d(p.low,p.high,"RLWT_Curatefinal_highVlow")
 
+pre_perma_d(pseq.subset) 
+perma_d(p.med,p.high,"RLWT_Curatefinal_highVmed")
+p
+
+pre_perma_d(atlas.pseq) 
+perma_d(p.low,p.high,"RLWT_Atlasfinal_highVlow")
+
+pre_perma_d(atlas.pseq) 
+perma_d(p.med,p.high,"RLWT_Atlasfinal_highVmed")
+p
+
+plot_abundance.ak = function(pseq.subset,title = "",
+                         Facet = "Genus", Color = "Phylum"){
+ # Arbitrary subset, based on Phylum, for plotting
+ p1f = subset_taxa(pseq.subset, Genus %in% c("Akkermansia"))
+ mphyseq = psmelt(p1f)
+ mphyseq <- subset(mphyseq, Abundance > 0)
+ p <- ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance")) +
+   geom_violin(fill = NA) +
+   geom_point(size = 1, alpha = 0.3,
+              position = position_jitter(width = 0.3)) + scale_y_log10()+
+   theme(legend.position="none") + scale_colour_manual(values = wes_palette(glob_theme)) 
+ library(rms)
+fita <- lrm(mnd ~ gender + log(Abundance) + bmi_group + age_category, data =mphyseq)
+fita <- lrm(mnd ~ gender + log(Abundance) , data =mphyseq)
+fita <- lrm(mnd ~ gender + log(Abundance) + bmi_group, data =mphyseq)
+
+}
+
+newData<-data.frame(JVeg5=factor(Jdata[,"JVeg5"]),scale(Jdata[,c("Elevation","Lat_Y_pos","Coast_dist","Stream_dist")]))
+
+Global <- polr(JVeg5 ~  Elevation +  Lat_Y_pos + Coast_dist + Stream_dist,
+                data=newData, na.action = na.omit, Hess = TRUE)
+
+library(MASS)
+model_fit <- polr(mnd~Abundance, data = mphyseq)
+summary(model_fit)
+
+data(atlas1006)
+plot_regression(diversity ~ mnd, meta(atlas.pseq))
+
+otu <- atlas.pseq$microbes 
+lipids <- peerj32$lipids
+
+# Define data sets to cross-correlate
+x <- log10(otu) # OTU Log10 (44 samples x 130 genera)
+y <- as.matrix(lipids) # Lipids (44 samples x 389 lipids)
+
+# Cross correlate data sets
+correlations <- associate(x, y, method = "spearman", mode = "matrix", p.adj.threshold = 0.05, n.signif = 1)
+
+# Or, alternatively, the same output is also available in a handy table format
+correlation.table <- associate(x, y, method = "spearman", mode = "table", p.adj.threshold = 0.05, n.signif = 1)
+
+kable(head(correlation.table))
 ```
 
-```{r MND table summary}
+```{r MNDtable summary}
 
-library(table1)
-n <- length(sample_data(pseq.subset)$mnd)
-
-data <- data.frame(id=1:n)
-data$mnd <- sample_data(pseq.subset)$mnd
-data$bmi <- sample_data(pseq.subset)$BMI
-data$age <- sample_data(pseq.subset)$age
-data$antibiotics_current_use <- sample_data(pseq.subset)$antibiotics_current_use
-data$smoker <- sample_data(pseq.subset)$smoker
-data$pregnant <- sample_data(pseq.subset)$pregnant
-data$alcohol <- sample_data(pseq.subset)$alcohol
-data$age_category <- sample_data(pseq.subset)$age_category
-
-table1(~ age  +  bmi + smoker + alcohol + age_category | mnd, data=data)
+atlas.pseq
+meta <- meta(atlas.pseq)
 

### 2019-12-03 import_popln_age_data.Rmd
updated

--- 2019-06-28 import_popln_data.Rmd
@@
 
+p7 <- ggplot(ad.p.melt, aes(x = variable, y = value, colour = Location)) + 
+  geom_line() + 
+  ylab(label="AD Prevalance") + 
+  xlab("Year") 
+legend <- get_legend(p7)
+
 final <- plot_grid(p1, p2,p3,p4,p5,p6, labels = c('A', 'B','C', 'D','E', 'F'),ncol = 2)
-ggsave("overview.jpg", final) 
+ggsave("overview.jpg", final,width = 20, height = 20, units = "cm") 
 ```

### 2019-12-03 combodata_NS.Rmd
updated

@@
 ```{r setup, include=FALSE}
+
+setwd("~/globalmicrobiome")
 
@@
 
+#drops unused levels
+c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI") )
+
+sample_data(pseq)$nationality = c.mnd.atlas
+
 #mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DNK","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
@@
 glob_theme <- "FantasticFox1"
-
+setwd("~/globalmicrobiome") 
 my_cur_table <- function(pseq.subset){
@@
   # on to the graphs!!!
-  print("The basic aloha diversity analysis has been completed")
+  print("The basic alpha diversity analysis has been completed")
   # This country versus nationality
-  p1<-   ggviolin(pseq.meta, x = "country", y = "Shannon",
-                  add = "boxplot", fill = "country", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+  p1<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+                  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
   print("First graph has completed")
@@
   print("Third graph has completed")
-  ggsave(paste("/Users/Nik/Alpha_diversity_",title_text,".png", sep = "" ), device = "png")
+  save_plot(paste(title_text,".png", sep = ""),p3,base_height = 11.69 )
   #ggsave("/Users/Nik/Alpha_diversity_curate.png")
@@
   #p3 <- p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
-  ggsave(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), device = "png")
+  save_plot(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), p3)
     # MDS
@@
     p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND Prevalence") + theme(legend.position = "none")
-    ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+    save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p5)
 
@@
     p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND Prevalence") + theme(legend.position = "bottom")
-    ggsave(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), device = "png")
+    save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p8)
        #this just generates the titles for the output
@@
     p10 <<-plot_grid(title,p3, p5,p8, ncol = 1, labels = c('','A', 'B','C'),rel_heights = c(0.3, 1.5,1.5,2))
-    ggsave(paste("Beta_diversity_ALL_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+    save_plot(paste("Beta_diversity_ALL_",title_text,".png", sep = "" ), p10)
 }
@@
   p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
-  ggsave(paste("PERMANOVA_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+  save_plot(paste("PERMANOVA_",title_text,".png", sep = "" ), p)
   #ggsave("test_permanova.png", device = "png",width = 297, height = 210, units = "mm")
@@
 
+setwd("~/globalmicrobiome")
+
 # These are useful is filtering. 
@@
 perma_d(p.med,p.high,"RLWT_Curatefinal_highVmed")
-p
 
@@
 perma_d(p.med,p.high,"RLWT_Atlasfinal_highVmed")
-p
+
+#   not working from here !!!!!!
 

### 2020-01-21 combodata_HL_NS.Rmd
updated

--- 2019-12-03 combodata_NS.Rmd
@@
 date: "26/09/2019"
-output: html_document
+output:
+  word_document: default
+  html_document: default
 ---
 
-## Setup
-```{r setup, include=FALSE}
-
-setwd("~/globalmicrobiome")
-
-#BiocManager::install("curatedMetagenomicData")
-#BiocManager::install("ExperimentHub")
-library(ExperimentHub)
-library(curatedMetagenomicData)
-suppressPackageStartupMessages(library(ExperimentHub))
-library(phyloseq)
-library(microbiome)
-library(forcats)
-library(stargazer)
-library(ggpubr)
-library(tidyverse)
-library(cowplot)
-library(vegan)
-library(broom)
-library(tsnemicrobiota)
-library(latex2exp)
-library(wesanderson)
-library(table1)
-
-eh = ExperimentHub()
-myquery = query(eh, "curatedMetagenomicData")
-
-# Get the stool samples
-esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)
-
-# merge
-eset <- mergeData(esl)
-
-# the relab=FALSE is essential as it Absolute Raw Count Data
-pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
-orig_pseq <- pseq
-
-#For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object.
-curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:
-data(atlas1006)
-atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
-
-```
-
-## MND Countries
-```{r tables}
-
-# THIS FOR THE CURATED DATA SET
-
-# creates a grouping varaible for BMI
-n <- length(sample_data(pseq)$BMI)
-data <- data.frame(id=1:n)
-data$bmi <- sample_data(pseq)$BMI # creates a new variable with BMI in it
-# Classification based upon the ATlas data
-data$bmi[data$bmi <18.5] <- "underweight"
-data$bmi[data$bmi >=45] <- "super obese"
-data$bmi[data$bmi >= 18.5 & data$bmi <25] <- "lean"
-data$bmi[data$bmi >= 25 & data$bmi <30] <- "overweight"
-data$bmi[data$bmi >= 30 & data$bmi <35] <- "obese"
-data$bmi[data$bmi >= 35 & data$bmi <40] <- "severe obese"
-data$bmi[data$bmi >= 40 & data$bmi <45] <- "morbid obese"
-data$bmi <- factor(data$bmi)
-
-# creates a new variable in the phyloseq called mnd
-sample_data(pseq)$bmi_group = data$bmi
-# Check that it works
-get_variable(pseq, "bmi_group")
-
-# Create mnd variable
-#pseq <- orig_pseq
-# grabs the nationality from phyloseq
-c.mnd.country <- get_variable(pseq, "country" )
-table(c.mnd.country)
-
-# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-# Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,
-
-#Adjsuted to match exactly the ATLAS data set
-
-#central Europe = Belgium, Demark, Netherlands, Germany
-#southern Europe = France, Italy, Serbia, Spain
-#UKIE
-#Scandinavia = Norway, Sweden Finland
-#Eastern Europe = Poland
-
-c.mnd.atlas <- fct_collapse(c.mnd.country, Scandinavia = c("NOR","SWE","FIN") ,CentralEurope = c("DNK","DEU","NLD"), SouthEurope = c("ITA","ESP","FRA"), US = "USA", UKIE = "GBR" )
-
-#drops unused levels
-c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI") )
-
-sample_data(pseq)$nationality = c.mnd.atlas
-
-#mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DNK","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
-
-#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-# reorder
-# NOTE THERE IS NO EASTERN EUROPE
-c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
-
-c.mnd.nationality <- factor(c.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
-levels(c.mnd.nationality)
-
-# creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = c.mnd.nationality
-
-# checks that it has worked.
-get_variable(pseq, "mnd")
-table(get_variable(pseq, "mnd"))
-
-###############################################
-
-# THIS IS FOR THE ATLAS 1006 
-# Create mnd variable
-# grabs the nationality from phyloseq
-a.mnd.nationality <- get_variable(atlas.pseq, "nationality" )
-
-# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
-
-# reorder
-a.mnd.nationality <- factor(a.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
-levels(a.mnd.nationality)
-
-# creates a new variable in the phyloseq called mnd
-sample_data(atlas.pseq)$mnd = a.mnd.nationality
-#sample_data(atlas.pseq)$country = sample_data(atlas.pseq)$nationality
-# checks that it has worked.
-table(get_variable(atlas.pseq, "mnd"))
+```{r , include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
 
@@
 glob_theme <- "FantasticFox1"
-setwd("~/globalmicrobiome") 
+
+pre_perma_d <- function(pseq.subset) { 
+  p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+}
+
+pre_perma_d_two <- function(pseq.subset) { 
+  p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, High = p.high.d))
+
+}
+
+perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+  pseqLvH <- merge_phyloseq(x,y) # pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
+  save_plot(paste("PERMANOVA_",title_text,".png", sep = "" ), p)
+  #ggsave("test_permanova.png", device = "png",width = 297, height = 210, units = "mm")
+ }
+
+# THis calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
+perma_d_two <- function(title_text) { # x & y must be either p.low,p.med or p.high
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+  pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p <- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
+  ggsave(paste("PERMANOVA_",title_text,".png", sep = "" ), device = "png",width = 210, height = 297, units = "mm")
+  ggsave("test_permanova.png", device = "png",width = 297, height = 210, units = "mm")
+ }
+
+alpha_d_two <- function(pseq.subset,title_text) {
+   # gets the meta data 
+   pseq.meta <- meta(pseq.subset)
+   # generates the diversites
+   tab <- microbiome::alpha(pseq.subset, index = "shannon")
+
+   # creates Shannon
+   pseq.meta$Shannon <- tab$diversity_shannon
+   mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("HIGH","LOW"))
+
+   # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","HIGH"),  exclude = "NA")
+  # on to the graphs!!! country 
+  p1<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+ add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+# the number in the palettes must equal the number of groups in bmi.pairs
+
+   p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = wes_palette(glob_theme))
+
+    p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+    #this just generates the titles for the output
+    title <- ggdraw() + 
+    draw_label(
+    title_text,
+    fontface = 'bold',
+    x = 0,
+    hjust = 0
+    ) +
+    theme(
+    # add margin on the left of the drawing canvas,
+    # so title is aligned with left edge of first plot
+    plot.margin = margin(0, 0, 0, 7)
+    )
+
+    p3 <<-plot_grid(title,p1,p2, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
+    ggsave(paste("Alpha_diversity_",title_text,".png", sep = "" ), device = "png")
+}
+
+mb.gather <- function (x) {
+    # Creates an empty data frame
+    mb.group <- data.frame("Location" = c("Central Europe","Eastern Europe","Scandinavia", "Southern Europe","UKIE","USA"), "Mean" = c(1:6)) 
+     # groups the countries to match the microbiome classification 
+  c.europe <<- filter(x,Location=="Belgium"|Location=="Denmark"|Location=="Germany"|Location=="The Netherlands")
+    e.europe <- filter(x,Location=="Poland")
+    scan <- filter(x,Location=="Finland"|Location=="Norway"|Location=="Sweden")
+    s.europe <- filter(x,Location=="France"|Location=="Italy"|Location=="Serbia"|Location=="Spain")
+    ukie <- filter(x,Location=="United Kingdom"|Location=="Ireland")
+    usa <- filter(x,Location=="United States")
+    # gmeans the data and put it into the data frame
+
+    mb.group[1,2] <- round(mean(c.europe$Value),digits = 2)
+    mb.group[2,2] <- round(mean(e.europe$Value),digits = 2)
+    mb.group[3,2] <- round(mean(scan$Value),digits = 2)
+    mb.group[4,2] <- round(mean(s.europe$Value),digits = 2)
+    mb.group[5,2] <- round(mean(ukie$Value),digits = 2)
+    mb.group[6,2] <- round(mean(usa$Value),digits = 2)
+
+    mb.group[1,3] <- paste(round(mean(c.europe$Upper.bound),digits = 2), round(mean(c.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[2,3] <- paste(round(mean(e.europe$Upper.bound),digits = 2), round(mean(e.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[3,3] <- paste(round(mean(scan$Upper.bound),digits = 2), round(mean(scan$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[4,3] <- paste(round(mean(s.europe$Upper.bound),digits = 2), round(mean(s.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[5,3] <- paste(round(mean(ukie$Upper.bound),digits = 2), round(mean(ukie$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[6,3] <- paste(round(mean(usa$Upper.bound),digits = 2), round(mean(usa$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    # mb.group[2,3] <- paste(e.europe$Upper.bound,"-", e.europe$Lower.bound)
+    # mb.group[3,3] <- paste(scan$Upper.bound,"-", scan$Lower.bound)
+    # mb.group[4,3] <- paste(s.europe$Upper.bound,"-",s.europe$Lower.bound)
+    # mb.group[5,3] <- paste(ukie$Upper.bound,"-", ukie$Lower.bound )
+    # mb.group[6,3] <- paste(usa$Upper.bound,"-", usa$Lower.bound)
+    return(mb.group)
+    }
+
+mb.cgather <- function (x) {
+    # Creates an empty data frame
+    mb.group <- data.frame("Location" = c("Norway","Sweden","Finland","Denmark","Luxembourg", "Italy","Austria","Germany","Spain","France","Netherlands","United Kingdom","United States","Canada"), "Mean" = c(1:14))
+     # groups the countries to match the microbiome classification 
+  mb.group <- filter(x,Location=="Norway"|Location=="Sweden"|Location=="Finland"|Location=="Denmark"|Location=="Luxembourg"|Location=="Italy"|Location=="Austria"|Location=="Germany"|Location=="Spain"|Location=="France"|Location=="Netherlands"|Location=="United Kingdom"|Location=="United States"|Location=="Canada")
+
+    return(mb.group)
+}
+
 my_cur_table <- function(pseq.subset){
@@
 
-pre_perma_d <- function(pseq.subset) { 
-  p.low <<- subset_samples(pseq.subset, mnd == "LOW")
-  p.low.d <<- divergence(p.low)
-  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
-  p.med.d <<- divergence(p.med)
-  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
-  p.high.d <<- divergence(p.high)
-  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
-
-}
-
 # THis calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
-perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
-  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
-  pseqLvH <- merge_phyloseq(x,y) # pseqLvH <- merge_phyloseq(p.low,p.high)
-  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
-  otu <- abundances(pseq.rel)
-  meta <- meta(pseq.rel)
-  permanova <- adonis(t(otu) ~ mnd,
-               data = meta, permutations=999, method = "bray")
- # P-value
-  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
-  dist <- vegdist(t(otu))
-  anov <- anova(betadisper(dist, meta$mnd))
-  summary(anov)
-  tidy(anov)
-  coef <- coefficients(permanova)["mnd1",]
-  top.coef <- coef[rev(order(abs(coef)))[1:20]]
-  par(mar = c(12, 30, 2, 1))
-  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
-  #temp <- TeX("Top Genus")
-  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
-  #res       = 1500)
-  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
-  #dev.off() 
-  temp <- as.data.frame(top.coef)
-  temp$label <- rownames(temp) 
-  temp$label <- sub("s_*", "", temp[,2])
-  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
-  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
-  save_plot(paste("PERMANOVA_",title_text,".png", sep = "" ), p)
-  #ggsave("test_permanova.png", device = "png",width = 297, height = 210, units = "mm")
- }
 
@@
 
+## Setup
+```{r setup, include=FALSE}
+
+#BiocManager::install("curatedMetagenomicData")
+#BiocManager::install("ExperimentHub")
+library(ExperimentHub)
+library(curatedMetagenomicData)
+suppressPackageStartupMessages(library(ExperimentHub))
+library(phyloseq)
+library(microbiome)
+library(forcats)
+library(stargazer)
+library(ggpubr)
+library(tidyverse)
+library(cowplot)
+library(vegan)
+library(broom)
+library(tsnemicrobiota)
+library(latex2exp)
+library(wesanderson)
+library(table1)
+library(knitr)
+library(kableExtra)
+
+eh = ExperimentHub()
+myquery = query(eh, "curatedMetagenomicData")
+
+# Get the stool samples
+esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)
+
+# merge
+eset <- mergeData(esl)
+
+# the relab=FALSE is essential as it Absolute Raw Count Data
+pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
+orig_pseq <- pseq
+
+#For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object.
+curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:
+data(atlas1006)
+atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
+
+```
+
+## MND Countries
+```{r tables}
+
+# THIS FOR THE CURATED DATA SET
+
+# creates a grouping varaible for BMI
+n <- length(sample_data(pseq)$BMI)
+data <- data.frame(id=1:n)
+data$bmi <- sample_data(pseq)$BMI # creates a new variable with BMI in it
+# Classification based upon the ATlas data
+data$bmi[data$bmi <18.5] <- "underweight"
+data$bmi[data$bmi >=45] <- "super obese"
+data$bmi[data$bmi >= 18.5 & data$bmi <25] <- "lean"
+data$bmi[data$bmi >= 25 & data$bmi <30] <- "overweight"
+data$bmi[data$bmi >= 30 & data$bmi <35] <- "obese"
+data$bmi[data$bmi >= 35 & data$bmi <40] <- "severe obese"
+data$bmi[data$bmi >= 40 & data$bmi <45] <- "morbid obese"
+data$bmi <- factor(data$bmi)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(pseq)$bmi_group = data$bmi
+# Check that it works
+get_variable(pseq, "bmi_group")
+
+# Create mnd variable
+#pseq <- orig_pseq
+# grabs the nationality from phyloseq
+c.mnd.country <- get_variable(pseq, "country" )
+table(c.mnd.country)
+
+# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
+# Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,
+
+#Adjsuted to match exactly the ATLAS data set
+
+#central Europe = Belgium, Demark, Netherlands, Germany
+#southern Europe = France, Italy, Serbia, Spain
+#UKIE
+#Scandinavia = Norway, Sweden Finland
+#Eastern Europe = Poland
+
+c.mnd.atlas <- fct_collapse(c.mnd.country, Scandinavia = c("NOR","SWE","FIN") ,CentralEurope = c("DNK","DEU","NLD"), SouthEurope = c("ITA","ESP","FRA"), US = "USA", UKIE = "GBR" )
+
+#drops unused levels
+c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI","CHN","MDG","MNG","PER","RUS","TZA") )
+
+sample_data(pseq)$nationality = c.mnd.atlas
+
+#mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DNK","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
+
+#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
+
+# reorder
+# NOTE THERE IS NO EASTERN EUROPE
+c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope"), HIGH = c("CentralEurope","US","UKIE") )
+
+c.mnd.nationality <- factor(c.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
+levels(c.mnd.nationality)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(pseq)$mnd = c.mnd.nationality
+
+# checks that it has worked.
+get_variable(pseq, "mnd")
+table(get_variable(pseq, "mnd"))
+
+###############################################
+
+# THIS IS FOR THE ATLAS 1006 
+# Create mnd variable
+# grabs the nationality from phyloseq
+a.mnd.nationality <- get_variable(atlas.pseq, "nationality" )
+
+# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
+#a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia") )
+
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope","SouthEurope") ,  HIGH = c("CentralEurope","US","UKIE","Scandinavia") )
+
+# reorder
+a.mnd.nationality <- factor(a.mnd.nationality, levels = (c("LOW", "HIGH")))
+levels(a.mnd.nationality)
+
+# creates a new variable in the phyloseq called mnd
+sample_data(atlas.pseq)$mnd = a.mnd.nationality
+#sample_data(atlas.pseq)$country = sample_data(atlas.pseq)$nationality
+# checks that it has worked.
+table(get_variable(atlas.pseq, "mnd"))
+
+```
+
+## Incidence of MND
+```{r}
+
+# This block uses the raw data to examin  the incidence of MND to ensure the classificaiton is correct. It assumes that the data is in the correct place so the setwd at the top meeds to be accurate. 
+setwd('~/globalmicrobiome') 
+# For MND  nrows = 196 as thwew are other things after 196
+mnd.i.15_49 <- as.data.frame(read.csv("./rawdata/MND_I_15_49.csv", header=TRUE, nrows = 196))
+
+mnd.i.50_69  <- as.data.frame(read.csv("./rawdata/MND_I_50_69.csv", header=TRUE, nrows = 196))
+mnd.i.70 <- as.data.frame(read.csv("./rawdata/MND_I_70plus.csv", header=TRUE, nrows = 196))
+
+mnd.i.70.test <- mnd.i.70
+# Run function and change name
+mnd.i.15_49 <- mb.gather(mnd.i.15_49)
+colnames(mnd.i.15_49)[2] <- "15-49"
+
+# Run function and change name
+mnd.i.50_69 <- mb.gather(mnd.i.50_69)
+colnames(mnd.i.50_69)[2] <- "50-69"
+
+# Run function and change name
+mnd.i.70 <- mb.gather(mnd.i.70)
+colnames(mnd.i.70)[2] <- "70"
+
+# Join them together.   
+mnd.i <- list( mnd.i.15_49,mnd.i.50_69,mnd.i.70 ) %>% reduce(left_join, by = "Location")
+
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope","SouthEurope") , HIGH = c("CentralEurope","US","UKIE","Scandinavia" ))
+
+mnd.i <- arrange(mnd.i,mnd.i$'70')
+
+mnd.i %>%
+  kable(col.names = c("","Incidence","CI","Incidence","CI","Incidence","CI")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "center") %>%
+add_header_above(c("Location" = 1, "15-49" = 2, "50-69" = 2, "70+" = 2)) %>%
+  add_header_above(c(" ", "Age (years)" = 6)) %>%
+  pack_rows("Low MND group",1,2)%>%
+  pack_rows("High MND group", 3,6)
+
+# Now do the same for the curated data set
+mnd.ic.15_49 <- mb.cgather(mnd.i.15_49)
+colnames(mnd.ic.15_49)[2] <- "15-49"
+
+# Run function and change name
+mnd.ic.50_69 <- mb.cgather(mnd.i.50_69)
+colnames(mnd.ic.50_69)[2] <- "50-69"
+
+# Run function and change name
+mnd.ic.70 <- mb.cgather(mnd.i.70)
+colnames(mnd.ic.70)[2] <- "70"
+
+# Join them together.
+
+mnd.ic <- list( mnd.ic.15_49,mnd.ic.50_69,mnd.ic.70 ) %>% reduce(left_join, by = "Location")
+
+mnd.ic <- dplyr::select(mnd.ic, "Location",  "Value.x", "Lower.bound.x", "Upper.bound.x", "Value.y", "Lower.bound.y", "Upper.bound.y", "Value", "Lower.bound", "Upper.bound" )
+
+# Prep the data for the table
+
+mnd.ic %>%
+  kable(col.names = c("","Incidence","Upper","Lower","Incidence","Upper","Lower","Incidence","Upper","Lower"),digits=3) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "left") %>%
+add_header_above(c("Location" = 1, "15-49" = 3, "50-69" = 3, "70+" = 3)) %>%
+  add_header_above(c(" ", "Age (years)" = 9)) %>%
+  pack_rows("Group 1", 4, 7) %>%
+  pack_rows("Group 1", 4, 7) %>%
+  pack_rows("Group 2", 8, 10)
+
+```
+
 ## Filtering
@@
 alpha_d(pseq,"final_RLWT_1")
+alpha_d_two(pseq,"final_RLWT_1")
+
+pre_perma_d_two(atlas.pseq) 
+perma_d_two("RLWT_highVlow")
 

### 2020-01-21 manuscript_NS.Rmd
updated

--- 2019-12-03 combodata_NS.Rmd
@@
 date: "26/09/2019"
-output: html_document
+output:
+  word_document: default
+  html_document: default
 ---
+
+```{r , include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
+
+```
 
@@
 ```{r setup, include=FALSE}
-
-setwd("~/globalmicrobiome")
 
@@
 library(table1)
+library(knitr)
+library(kableExtra)
 
@@
 atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
+
+```
+
+## Incidence of MND
+```{r}
+
+# This block uses the raw data to examin  the incidence of MND to ensure the classificaiton is correct. It assumes that the data is in the correct place so the setwd at the top meeds to be accurate. 
+setwd('~/globalmicrobiome') 
+# For MND  nrows = 196 as thwew are other things after 196
+mnd.i.15_49 <- as.data.frame(read.csv("./rawdata/MND_I_15_49.csv", header=TRUE, nrows = 196))
+
+mnd.i.50_69  <- as.data.frame(read.csv("./rawdata/MND_I_50_69.csv", header=TRUE, nrows = 196))
+mnd.i.70 <- as.data.frame(read.csv("./rawdata/MND_I_70plus.csv", header=TRUE, nrows = 196))
+
+mnd.i.70.test <- mnd.i.70
+# Run function and change name
+mnd.i.15_49 <- mb.gather(mnd.i.15_49)
+colnames(mnd.i.15_49)[2] <- "15-49"
+
+# Run function and change name
+mnd.i.50_69 <- mb.gather(mnd.i.50_69)
+colnames(mnd.i.50_69)[2] <- "50-69"
+
+# Run function and change name
+mnd.i.70 <- mb.gather(mnd.i.70)
+colnames(mnd.i.70)[2] <- "70"
+
+# Join them together.   
+mnd.i <- list( mnd.i.15_49,mnd.i.50_69,mnd.i.70 ) %>% reduce(left_join, by = "Location")
+
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia" ))
+
+mnd.i %>%
+  kable(col.names = c("","Incidence","CI","Incidence","CI","Incidence","CI")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "center") %>%
+add_header_above(c("Location" = 1, "15-49" = 2, "50-69" = 2, "70+" = 2)) %>%
+  add_header_above(c(" ", "Age (years)" = 6)) %>%
+  pack_rows("High MND group", 5,6)%>%
+  pack_rows("Low MND group", 2, 2) %>%
+  pack_rows("Medium MND group",1,3,4)
+
+# Now do the same for the curated data set
+mnd.ic.15_49 <- mb.cgather(mnd.i.15_49)
+colnames(mnd.ic.15_49)[2] <- "15-49"
+
+# Run function and change name
+mnd.ic.50_69 <- mb.cgather(mnd.i.50_69)
+colnames(mnd.ic.50_69)[2] <- "50-69"
+
+# Run function and change name
+mnd.ic.70 <- mb.cgather(mnd.i.70)
+colnames(mnd.ic.70)[2] <- "70"
+
+# Join them together.
+
+mnd.ic <- list( mnd.ic.15_49,mnd.ic.50_69,mnd.ic.70 ) %>% reduce(left_join, by = "Location")
+
+mnd.ic <- dplyr::select(mnd.ic, "Location",  "Value.x", "Lower.bound.x", "Upper.bound.x", "Value.y", "Lower.bound.y", "Upper.bound.y", "Value", "Lower.bound", "Upper.bound" )
+
+# Prep the data for the table
+
+mnd.ic %>%
+  kable(col.names = c("","Incidence","Upper","Lower","Incidence","Upper","Lower","Incidence","Upper","Lower"),digits=3) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "left") %>%
+add_header_above(c("Location" = 1, "15-49" = 3, "50-69" = 3, "70+" = 3)) %>%
+  add_header_above(c(" ", "Age (years)" = 9)) %>%
+  pack_rows("Group 1", 4, 7) %>%
+  pack_rows("Group 1", 4, 7) %>%
+  pack_rows("Group 2", 8, 10)
 
@@
 #drops unused levels
-c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI") )
+c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI","CHN","MDG","MNG","PER","RUS","TZA") )
 
@@
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia") )
 
@@
 glob_theme <- "FantasticFox1"
-setwd("~/globalmicrobiome") 
+
+mb.gather <- function (x) {
+    # Creates an empty data frame
+    mb.group <- data.frame("Location" = c("Central Europe","Eastern Europe","Scandinavia", "Southern Europe","UKIE","USA"), "Mean" = c(1:6)) 
+     # groups the countries to match the microbiome classification 
+  c.europe <<- filter(x,Location=="Belgium"|Location=="Denmark"|Location=="Germany"|Location=="The Netherlands")
+    e.europe <- filter(x,Location=="Poland")
+    scan <- filter(x,Location=="Finland"|Location=="Norway"|Location=="Sweden")
+    s.europe <- filter(x,Location=="France"|Location=="Italy"|Location=="Serbia"|Location=="Spain")
+    ukie <- filter(x,Location=="United Kingdom"|Location=="Ireland")
+    usa <- filter(x,Location=="United States")
+    # gmeans the data and put it into the data frame
+
+    mb.group[1,2] <- round(mean(c.europe$Value),digits = 2)
+    mb.group[2,2] <- round(mean(e.europe$Value),digits = 2)
+    mb.group[3,2] <- round(mean(scan$Value),digits = 2)
+    mb.group[4,2] <- round(mean(s.europe$Value),digits = 2)
+    mb.group[5,2] <- round(mean(ukie$Value),digits = 2)
+    mb.group[6,2] <- round(mean(usa$Value),digits = 2)
+
+    mb.group[1,3] <- paste(round(mean(c.europe$Upper.bound),digits = 2), round(mean(c.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[2,3] <- paste(round(mean(e.europe$Upper.bound),digits = 2), round(mean(e.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[3,3] <- paste(round(mean(scan$Upper.bound),digits = 2), round(mean(scan$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[4,3] <- paste(round(mean(s.europe$Upper.bound),digits = 2), round(mean(s.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[5,3] <- paste(round(mean(ukie$Upper.bound),digits = 2), round(mean(ukie$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    mb.group[6,3] <- paste(round(mean(usa$Upper.bound),digits = 2), round(mean(usa$Lower.bound),digits = 2), sep = "-",collapse = "-")
+    # mb.group[2,3] <- paste(e.europe$Upper.bound,"-", e.europe$Lower.bound)
+    # mb.group[3,3] <- paste(scan$Upper.bound,"-", scan$Lower.bound)
+    # mb.group[4,3] <- paste(s.europe$Upper.bound,"-",s.europe$Lower.bound)
+    # mb.group[5,3] <- paste(ukie$Upper.bound,"-", ukie$Lower.bound )
+    # mb.group[6,3] <- paste(usa$Upper.bound,"-", usa$Lower.bound)
+    return(mb.group)
+    }
+
+mb.cgather <- function (x) {
+    # Creates an empty data frame
+    mb.group <- data.frame("Location" = c("Norway","Sweden","Finland","Denmark","Luxembourg", "Italy","Austria","Germany","Spain","France","Netherlands","United Kingdom","United States","Canada"), "Mean" = c(1:14))
+     # groups the countries to match the microbiome classification 
+  mb.group <- filter(x,Location=="Norway"|Location=="Sweden"|Location=="Finland"|Location=="Denmark"|Location=="Luxembourg"|Location=="Italy"|Location=="Austria"|Location=="Germany"|Location=="Spain"|Location=="France"|Location=="Netherlands"|Location=="United Kingdom"|Location=="United States"|Location=="Canada")
+
+    return(mb.group)
+}
+
 my_cur_table <- function(pseq.subset){

### 2020-01-21 manuscript_NS.Rmd
updated

@@
 
+#THIS IS NO LONGER NEEDED.
+
 # This block uses the raw data to examin  the incidence of MND to ensure the classificaiton is correct. It assumes that the data is in the correct place so the setwd at the top meeds to be accurate. 
-setwd('~/globalmicrobiome') 
-# For MND  nrows = 196 as thwew are other things after 196
-mnd.i.15_49 <- as.data.frame(read.csv("./rawdata/MND_I_15_49.csv", header=TRUE, nrows = 196))
+# setwd('~/globalmicrobiome') 
+# # For MND  nrows = 196 as thwew are other things after 196
+# mnd.i.15_49 <- as.data.frame(read.csv("./rawdata/MND_I_15_49.csv", header=TRUE, nrows = 196))
 
-mnd.i.50_69  <- as.data.frame(read.csv("./rawdata/MND_I_50_69.csv", header=TRUE, nrows = 196))
-mnd.i.70 <- as.data.frame(read.csv("./rawdata/MND_I_70plus.csv", header=TRUE, nrows = 196))
+# mnd.i.50_69  <- as.data.frame(read.csv("./rawdata/MND_I_50_69.csv", header=TRUE, nrows = 196))
+# mnd.i.70 <- as.data.frame(read.csv("./rawdata/MND_I_70plus.csv", header=TRUE, nrows = 196))
 
-mnd.i.70.test <- mnd.i.70
-# Run function and change name
-mnd.i.15_49 <- mb.gather(mnd.i.15_49)
-colnames(mnd.i.15_49)[2] <- "15-49"
+# mnd.i.70.test <- mnd.i.70
+# # Run function and change name
+# mnd.i.15_49 <- mb.gather(mnd.i.15_49)
+# colnames(mnd.i.15_49)[2] <- "15-49"
 
-# Run function and change name
-mnd.i.50_69 <- mb.gather(mnd.i.50_69)
-colnames(mnd.i.50_69)[2] <- "50-69"
+# # Run function and change name
+# mnd.i.50_69 <- mb.gather(mnd.i.50_69)
+# colnames(mnd.i.50_69)[2] <- "50-69"
 
-# Run function and change name
-mnd.i.70 <- mb.gather(mnd.i.70)
-colnames(mnd.i.70)[2] <- "70"
+# # Run function and change name
+# mnd.i.70 <- mb.gather(mnd.i.70)
+# colnames(mnd.i.70)[2] <- "70"
 
-# Join them together.   
-mnd.i <- list( mnd.i.15_49,mnd.i.50_69,mnd.i.70 ) %>% reduce(left_join, by = "Location")
+# # Join them together.   
+# mnd.i <- list( mnd.i.15_49,mnd.i.50_69,mnd.i.70 ) %>% reduce(left_join, by = "Location")
 
-a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia" ))
+# a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia" ))
 
-mnd.i %>%
-  kable(col.names = c("","Incidence","CI","Incidence","CI","Incidence","CI")) %>%
-  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "center") %>%
-add_header_above(c("Location" = 1, "15-49" = 2, "50-69" = 2, "70+" = 2)) %>%
-  add_header_above(c(" ", "Age (years)" = 6)) %>%
-  pack_rows("High MND group", 5,6)%>%
-  pack_rows("Low MND group", 2, 2) %>%
-  pack_rows("Medium MND group",1,3,4)
+# mnd.i %>%
+#   kable(col.names = c("","Incidence","CI","Incidence","CI","Incidence","CI")) %>%
+#   kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "center") %>%
+# add_header_above(c("Location" = 1, "15-49" = 2, "50-69" = 2, "70+" = 2)) %>%
+#   add_header_above(c(" ", "Age (years)" = 6)) %>%
+#   pack_rows("High MND group", 5,6)%>%
+#   pack_rows("Low MND group", 2, 2) %>%
+#   pack_rows("Medium MND group",1,3,4)
 
-# Now do the same for the curated data set
-mnd.ic.15_49 <- mb.cgather(mnd.i.15_49)
-colnames(mnd.ic.15_49)[2] <- "15-49"
+# # Now do the same for the curated data set
+# mnd.ic.15_49 <- mb.cgather(mnd.i.15_49)
+# colnames(mnd.ic.15_49)[2] <- "15-49"
 
-# Run function and change name
-mnd.ic.50_69 <- mb.cgather(mnd.i.50_69)
-colnames(mnd.ic.50_69)[2] <- "50-69"
+# # Run function and change name
+# mnd.ic.50_69 <- mb.cgather(mnd.i.50_69)
+# colnames(mnd.ic.50_69)[2] <- "50-69"
 
-# Run function and change name
-mnd.ic.70 <- mb.cgather(mnd.i.70)
-colnames(mnd.ic.70)[2] <- "70"
+# # Run function and change name
+# mnd.ic.70 <- mb.cgather(mnd.i.70)
+# colnames(mnd.ic.70)[2] <- "70"
 
-# Join them together.
+# # Join them together.
 
-mnd.ic <- list( mnd.ic.15_49,mnd.ic.50_69,mnd.ic.70 ) %>% reduce(left_join, by = "Location")
+# mnd.ic <- list( mnd.ic.15_49,mnd.ic.50_69,mnd.ic.70 ) %>% reduce(left_join, by = "Location")
 
-mnd.ic <- dplyr::select(mnd.ic, "Location",  "Value.x", "Lower.bound.x", "Upper.bound.x", "Value.y", "Lower.bound.y", "Upper.bound.y", "Value", "Lower.bound", "Upper.bound" )
+# mnd.ic <- dplyr::select(mnd.ic, "Location",  "Value.x", "Lower.bound.x", "Upper.bound.x", "Value.y", "Lower.bound.y", "Upper.bound.y", "Value", "Lower.bound", "Upper.bound" )
 
-# Prep the data for the table
+# # Prep the data for the table
 
-mnd.ic %>%
-  kable(col.names = c("","Incidence","Upper","Lower","Incidence","Upper","Lower","Incidence","Upper","Lower"),digits=3) %>%
-  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "left") %>%
-add_header_above(c("Location" = 1, "15-49" = 3, "50-69" = 3, "70+" = 3)) %>%
-  add_header_above(c(" ", "Age (years)" = 9)) %>%
-  pack_rows("Group 1", 4, 7) %>%
-  pack_rows("Group 1", 4, 7) %>%
-  pack_rows("Group 2", 8, 10)
+# mnd.ic %>%
+#   kable(col.names = c("","Incidence","Upper","Lower","Incidence","Upper","Lower","Incidence","Upper","Lower"),digits=3) %>%
+#   kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "left") %>%
+# add_header_above(c("Location" = 1, "15-49" = 3, "50-69" = 3, "70+" = 3)) %>%
+#   add_header_above(c(" ", "Age (years)" = 9)) %>%
+#   pack_rows("Group 1", 4, 7) %>%
+#   pack_rows("Group 1", 4, 7) %>%
+#   pack_rows("Group 2", 8, 10)
 
@@
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia") )
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
 

### 2020-01-22 manuscript_NS_MRD.Rmd
updated

--- 2020-01-21 manuscript_NS.Rmd
@@
 atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
+atlas.pseq <- subset_samples(atlas.pseq, !is.na(mnd))
 
@@
 # NOTE THERE IS NO EASTERN EUROPE
-c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
+c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE" ), HIGH = c("US") )
 
@@
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ,"UKIE"), HIGH = c("US") )
 
@@
 
-# THis calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
+# This calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
 perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
@@
   #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
-  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa")
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme(axis.text.y = element_text(face="italic"))
   save_plot(paste("PERMANOVA_",title_text,".png", sep = "" ), p)
@@
 
+save_plot(paste("akkermansia",title_text,".png", sep = "" ), p)
 }
+
+plotAtlas = plot_abundance.ak(atlas.pseq, "atlas")
+plotCur = plot_abundance.ak(pseq.subset, "curated")
+grid.arrange(nrow = 2, plotAtlas, plotCur)
 
@@
  The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
-
-## PERMANOVA significance test for group-level differences
-Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
 

### 2020-01-22 manuscript_NS_MRD_atlas.Rmd
updated

--- 2020-01-21 manuscript_NS.Rmd
@@
 ---
-title: "curatedMetagenomicData"
-author: "NSharma"
-date: "26/09/2019"
-output:
-  word_document: default
-  html_document: default
+title: "manuscript_NS_MRD_atlas"
+output: html_document
 ---
 
-```{r , include=FALSE}
+```{r setup, include=FALSE}
+
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-eh = ExperimentHub()
-myquery = query(eh, "curatedMetagenomicData")
-
-# Get the stool samples
-esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)
-
-# merge
-eset <- mergeData(esl)
-
-# the relab=FALSE is essential as it Absolute Raw Count Data
-pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
-orig_pseq <- pseq
-
 #For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object.
@@
 
-## Incidence of MND
-```{r}
-
-# This block uses the raw data to examin  the incidence of MND to ensure the classificaiton is correct. It assumes that the data is in the correct place so the setwd at the top meeds to be accurate. 
-setwd('~/globalmicrobiome') 
-# For MND  nrows = 196 as thwew are other things after 196
-mnd.i.15_49 <- as.data.frame(read.csv("./rawdata/MND_I_15_49.csv", header=TRUE, nrows = 196))
-
-mnd.i.50_69  <- as.data.frame(read.csv("./rawdata/MND_I_50_69.csv", header=TRUE, nrows = 196))
-mnd.i.70 <- as.data.frame(read.csv("./rawdata/MND_I_70plus.csv", header=TRUE, nrows = 196))
-
-mnd.i.70.test <- mnd.i.70
-# Run function and change name
-mnd.i.15_49 <- mb.gather(mnd.i.15_49)
-colnames(mnd.i.15_49)[2] <- "15-49"
-
-# Run function and change name
-mnd.i.50_69 <- mb.gather(mnd.i.50_69)
-colnames(mnd.i.50_69)[2] <- "50-69"
-
-# Run function and change name
-mnd.i.70 <- mb.gather(mnd.i.70)
-colnames(mnd.i.70)[2] <- "70"
-
-# Join them together.   
-mnd.i <- list( mnd.i.15_49,mnd.i.50_69,mnd.i.70 ) %>% reduce(left_join, by = "Location")
-
-a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia" ))
-
-mnd.i %>%
-  kable(col.names = c("","Incidence","CI","Incidence","CI","Incidence","CI")) %>%
-  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "center") %>%
-add_header_above(c("Location" = 1, "15-49" = 2, "50-69" = 2, "70+" = 2)) %>%
-  add_header_above(c(" ", "Age (years)" = 6)) %>%
-  pack_rows("High MND group", 5,6)%>%
-  pack_rows("Low MND group", 2, 2) %>%
-  pack_rows("Medium MND group",1,3,4)
-
-# Now do the same for the curated data set
-mnd.ic.15_49 <- mb.cgather(mnd.i.15_49)
-colnames(mnd.ic.15_49)[2] <- "15-49"
-
-# Run function and change name
-mnd.ic.50_69 <- mb.cgather(mnd.i.50_69)
-colnames(mnd.ic.50_69)[2] <- "50-69"
-
-# Run function and change name
-mnd.ic.70 <- mb.cgather(mnd.i.70)
-colnames(mnd.ic.70)[2] <- "70"
-
-# Join them together.
-
-mnd.ic <- list( mnd.ic.15_49,mnd.ic.50_69,mnd.ic.70 ) %>% reduce(left_join, by = "Location")
-
-mnd.ic <- dplyr::select(mnd.ic, "Location",  "Value.x", "Lower.bound.x", "Upper.bound.x", "Value.y", "Lower.bound.y", "Upper.bound.y", "Value", "Lower.bound", "Upper.bound" )
-
-# Prep the data for the table
-
-mnd.ic %>%
-  kable(col.names = c("","Incidence","Upper","Lower","Incidence","Upper","Lower","Incidence","Upper","Lower"),digits=3) %>%
-  kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "left") %>%
-add_header_above(c("Location" = 1, "15-49" = 3, "50-69" = 3, "70+" = 3)) %>%
-  add_header_above(c(" ", "Age (years)" = 9)) %>%
-  pack_rows("Group 1", 4, 7) %>%
-  pack_rows("Group 1", 4, 7) %>%
-  pack_rows("Group 2", 8, 10)
-
-```
-
 ## MND Countries
@@
 
-# THIS FOR THE CURATED DATA SET
-
-# creates a grouping varaible for BMI
-n <- length(sample_data(pseq)$BMI)
-data <- data.frame(id=1:n)
-data$bmi <- sample_data(pseq)$BMI # creates a new variable with BMI in it
-# Classification based upon the ATlas data
-data$bmi[data$bmi <18.5] <- "underweight"
-data$bmi[data$bmi >=45] <- "super obese"
-data$bmi[data$bmi >= 18.5 & data$bmi <25] <- "lean"
-data$bmi[data$bmi >= 25 & data$bmi <30] <- "overweight"
-data$bmi[data$bmi >= 30 & data$bmi <35] <- "obese"
-data$bmi[data$bmi >= 35 & data$bmi <40] <- "severe obese"
-data$bmi[data$bmi >= 40 & data$bmi <45] <- "morbid obese"
-data$bmi <- factor(data$bmi)
-
-# creates a new variable in the phyloseq called mnd
-sample_data(pseq)$bmi_group = data$bmi
-# Check that it works
-get_variable(pseq, "bmi_group")
-
-# Create mnd variable
-#pseq <- orig_pseq
-# grabs the nationality from phyloseq
-c.mnd.country <- get_variable(pseq, "country" )
-table(c.mnd.country)
-
-# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-# Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,
-
-#Adjsuted to match exactly the ATLAS data set
-
-#central Europe = Belgium, Demark, Netherlands, Germany
-#southern Europe = France, Italy, Serbia, Spain
-#UKIE
-#Scandinavia = Norway, Sweden Finland
-#Eastern Europe = Poland
-
-c.mnd.atlas <- fct_collapse(c.mnd.country, Scandinavia = c("NOR","SWE","FIN") ,CentralEurope = c("DNK","DEU","NLD"), SouthEurope = c("ITA","ESP","FRA"), US = "USA", UKIE = "GBR" )
-
-#drops unused levels
-c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI","CHN","MDG","MNG","PER","RUS","TZA") )
-
-sample_data(pseq)$nationality = c.mnd.atlas
-
-#mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DNK","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )
-
-#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )
-
-# reorder
-# NOTE THERE IS NO EASTERN EUROPE
-c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE") )
-
-c.mnd.nationality <- factor(c.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
-levels(c.mnd.nationality)
-
-# creates a new variable in the phyloseq called mnd
-sample_data(pseq)$mnd = c.mnd.nationality
-
-# checks that it has worked.
-get_variable(pseq, "mnd")
-table(get_variable(pseq, "mnd"))
-
-###############################################
-
-# THIS IS FOR THE ATLAS 1006 
 # Create mnd variable
@@
 # this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
-a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia") )
+a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope","Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE" ), HIGH = c("US") )
 
@@
 glob_theme <- "FantasticFox1"
-
-mb.gather <- function (x) {
-    # Creates an empty data frame
-    mb.group <- data.frame("Location" = c("Central Europe","Eastern Europe","Scandinavia", "Southern Europe","UKIE","USA"), "Mean" = c(1:6)) 
-     # groups the countries to match the microbiome classification 
-  c.europe <<- filter(x,Location=="Belgium"|Location=="Denmark"|Location=="Germany"|Location=="The Netherlands")
-    e.europe <- filter(x,Location=="Poland")
-    scan <- filter(x,Location=="Finland"|Location=="Norway"|Location=="Sweden")
-    s.europe <- filter(x,Location=="France"|Location=="Italy"|Location=="Serbia"|Location=="Spain")
-    ukie <- filter(x,Location=="United Kingdom"|Location=="Ireland")
-    usa <- filter(x,Location=="United States")
-    # gmeans the data and put it into the data frame
-
-    mb.group[1,2] <- round(mean(c.europe$Value),digits = 2)
-    mb.group[2,2] <- round(mean(e.europe$Value),digits = 2)
-    mb.group[3,2] <- round(mean(scan$Value),digits = 2)
-    mb.group[4,2] <- round(mean(s.europe$Value),digits = 2)
-    mb.group[5,2] <- round(mean(ukie$Value),digits = 2)
-    mb.group[6,2] <- round(mean(usa$Value),digits = 2)
-
-    mb.group[1,3] <- paste(round(mean(c.europe$Upper.bound),digits = 2), round(mean(c.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
-    mb.group[2,3] <- paste(round(mean(e.europe$Upper.bound),digits = 2), round(mean(e.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
-    mb.group[3,3] <- paste(round(mean(scan$Upper.bound),digits = 2), round(mean(scan$Lower.bound),digits = 2), sep = "-",collapse = "-")
-    mb.group[4,3] <- paste(round(mean(s.europe$Upper.bound),digits = 2), round(mean(s.europe$Lower.bound),digits = 2), sep = "-",collapse = "-")
-    mb.group[5,3] <- paste(round(mean(ukie$Upper.bound),digits = 2), round(mean(ukie$Lower.bound),digits = 2), sep = "-",collapse = "-")
-    mb.group[6,3] <- paste(round(mean(usa$Upper.bound),digits = 2), round(mean(usa$Lower.bound),digits = 2), sep = "-",collapse = "-")
-    # mb.group[2,3] <- paste(e.europe$Upper.bound,"-", e.europe$Lower.bound)
-    # mb.group[3,3] <- paste(scan$Upper.bound,"-", scan$Lower.bound)
-    # mb.group[4,3] <- paste(s.europe$Upper.bound,"-",s.europe$Lower.bound)
-    # mb.group[5,3] <- paste(ukie$Upper.bound,"-", ukie$Lower.bound )
-    # mb.group[6,3] <- paste(usa$Upper.bound,"-", usa$Lower.bound)
-    return(mb.group)
-    }
-
-mb.cgather <- function (x) {
-    # Creates an empty data frame
-    mb.group <- data.frame("Location" = c("Norway","Sweden","Finland","Denmark","Luxembourg", "Italy","Austria","Germany","Spain","France","Netherlands","United Kingdom","United States","Canada"), "Mean" = c(1:14))
-     # groups the countries to match the microbiome classification 
-  mb.group <- filter(x,Location=="Norway"|Location=="Sweden"|Location=="Finland"|Location=="Denmark"|Location=="Luxembourg"|Location=="Italy"|Location=="Austria"|Location=="Germany"|Location=="Spain"|Location=="France"|Location=="Netherlands"|Location=="United Kingdom"|Location=="United States"|Location=="Canada")
-
-    return(mb.group)
-}
-
-my_cur_table <- function(pseq.subset){
-    n <- length(sample_data(pseq.subset)$mnd)
-    data <- data.frame(id=1:n)
-    data$mnd <- sample_data(pseq.subset)$mnd
-    data$bmi <- sample_data(pseq.subset)$BMI
-    data$bmi_group <- sample_data(pseq.subset)$bmi_group
-    data$age <- sample_data(pseq.subset)$age
-    data$antibiotics_current_use <- sample_data(pseq.subset)$antibiotics_current_use
-    data$smoker <- sample_data(pseq.subset)$smoker
-    data$pregnant <- sample_data(pseq.subset)$pregnant
-    data$alcohol <- sample_data(pseq.subset)$alcohol
-    data$age_category <- sample_data(pseq.subset)$age_category
-    table1(~ age  +  bmi + bmi_group + smoker + alcohol + age_category | mnd, data=data)
-}
 
@@
 
-keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
-pseq.f <- subset_taxa(pseq, keepotu)
+#keepotu = genefilter_sample(pseq, filterfun_sample(topp(0.05)), A=5)
+#pseq.f <- subset_taxa(pseq, keepotu)
 
-pseq.c <- core(pseq, detection = 0.1/100, prevalence = 50/100)
+#pseq.c <- core(pseq, detection = 0.1/100, prevalence = 50/100)
 
@@
 
-pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
+#pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
 
-setwd("~/globalmicrobiome")
+#setwd("~/globalmicrobiome")
 
@@
 #get_variable(pseq.subset, sample_variables(pseq.subset)[13])
-my_cur_table(pseq.subset)
-my_cur_table(pseq.c)
 
@@
 alpha_d(pseq,"final_RLWT_1")
-
-alpha_d(atlas.pseq,"final_RLWT_2")
-
-beta_d(pseq.subset,"final_2")
-
-pre_perma_d(pseq.subset) 
-perma_d(p.low,p.high,"RLWT_Curatefinal_highVlow")
-
-pre_perma_d(pseq.subset) 
-perma_d(p.med,p.high,"RLWT_Curatefinal_highVmed")
 
@@
 
-## BETA DIVERSITY
-Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
-WE want to: quantify community divergence within a given sample set
-Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.
+```{r function list}
 
-Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance.
-This was applied in group-level comparisons for instance in Salonen et al.
-ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
+my_atlas_table(atlas.pseq)
 
- The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
+alpha_d(atlas.pseq, "title")
 
-## PERMANOVA significance test for group-level differences
-Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition
+alpha_d <- function(pseq.subset,title_text) {
+  # gets the meta data 
+  pseq.meta <- meta(pseq.subset)
+  # generates the diversites
+  tab <- microbiome::alpha(pseq.subset, index = "shannon")
 
-NOTHING BELOW THIS WORKS!!!!!!!!
+  # creates Shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
+  mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
 
-## MICROBIOME COMPOSITION
+  # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+  print("The basic alpha diversity analysis has been completed")
+  # This country versus nationality
+  p1<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+                  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+  print("First graph has completed")
+  # the number in the palettes must equal the number of groups in bmi.pairs
 
-```{r Phylum #}
+  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = wes_palette(glob_theme))
 
-# Show available ranks in the dataset
-rank_names(pseq)
+  p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
 
-# Create table, number of features for each phyla
-table(tax_table(pseq)[, "Phylum"], exclude = NULL)
+  print("Second graph has completed")
+  #this just generates the titles for the output
+  title <- ggdraw() + 
+    draw_label(
+      title_text,
+      fontface = 'bold',
+      x = 0,
+      hjust = 0
+    ) +
+    theme(
+      # add margin on the left of the drawing canvas,
+      # so title is aligned with left edge of first plot
+      plot.margin = margin(0, 0, 0, 7)
+    )
+
+  p3 <<-plot_grid(title,p1,p2, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
+
+  print("Third graph has completed")
+  save_plot(paste(title_text,".png", sep = ""),p3,base_height = 11.69 )
+  ggsave("/Users/maryam/Alpha_diversity_curate.png")
+}
 
@@
 
-This shows a few phyla for which only one feature was observed. Those may be worth filtering.
+```{r permanova test}
 
-guide_italics
-ggtitle("Relative abundance") +  
-```{r}
+#pre_perma_d <- function(pseq.subset) { 
+  p.low <<- subset_samples(atlas.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(atlas.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(atlas.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
 
-clos.ph <- aggregate_taxa(clostridium, "Phylum")
+#}
 
-clos.ph.rel <- microbiome::transform(clos.ph, "compositional")
-
-plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
-  theme(legend.position = "bottom", ) + theme_bw() + 
-  theme(axis.text.x=element_blank()) + 
-  theme(legend.title = element_text(size=18), legend.text = element_text( face="italic")) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend(title="Phylum"))
-
-print(plot.comp.rel)
-
-ggsave("phylum_relative_abundance_clos.png")
-
-plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none") + guide_italics
-ggsave("phylum_relative_abundance_clos2.png")
+# THis calcualtes the group PERMANOVA perma_d(p.low,p.high,"TESTHigh_vesus_Low")
+#perma_d <- function(x,y,title_text) { # x & y must be either p.low,p.med or p.high
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+  pseqLvH <- merge_phyloseq(p.low, p.high) # pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 20)  + annotate("text", x = 2.5, y = 0.02, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + guide_italics
+   save_plot(paste("PERMANOVA_",title_text,".png", sep = "" ), p)
+   plot(p)
+  #ggsave("test_permanova.png", device = "png",width = 297, height = 210, units = "mm")
+# }
 
 ```
-
-```{r CLostridium Bar Plot}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.cl.rel <- microbiome::transform(clos, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
-
-plot.cl.rel + scale_fill_brewer(palette = "Paired")
-ggsave("cl_abundance_bar.png")
-
-```
-
-```{r Clostridium genus bar plots}
-
-pseq.cl.rel <- microbiome::transform(clostridium, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")
-
-```
-
-```{r clostridium heatmap}
-
-pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")
-
-plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-p <- plot.c.comp.rel + coord_flip()
-p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-ggsave("clos_histogram.png")
-
-```
-
-## Community composition plotting
-Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results.
-These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted.
-Interpretation of these plots should always be subject to subsequent statistical analysis.
-
-```{r}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-# agglomerate at phylum level
-clos_phy <- tax_glom(clos, taxrank = "Phylum")
-
-# Transform to rel. abundance (or use ps0.ra)
-clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})
-
-# Melt to long format for easy ggploting
-clos_phylum <- psmelt(clos_phyl)
-
-# Plot - Phylum
-p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-  geom_bar(stat = "identity", width = 1) +
-  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-  theme(axis.text.x = element_blank()) +
-  theme(axis.title.x = element_blank()) +
-  labs(title = "Abundant Phylum of Clostridium(> 1%)")
-
-p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-plot(p)
-ggsave("phylum_abun_mnd_clos.png")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Phylum")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Genus")
-
-```
-
-```{r}
-
-get_taxa_unique(clostridium, "Genus")
-
-```
-
-```{r}
-
-df <- psmelt(pseq)
-kable(head(df))
-
-```
-
-```{r}
-
-head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])
-
-```

### 2020-01-22 manuscript_NS_MRD.Rmd
updated

@@
 #drops unused levels
-c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI","CHN","MDG","MNG","PER","RUS","TZA") )
+c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI","CHN","MDG","MNG","PER","RUS","TZA", "KAZ"))
+
+c.mnd.atlas <- subset_samples(c.mnd.atlas, !is.na(nationality) )
 
@@
 
-  p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
+  p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
 
@@
 
-  p3 <- p2 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND Prevalence") + theme(legend.position = "none")
+  p3 <- p2 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND DALYs") + theme(legend.position = "none")
   #plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
@@
 
-    p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND Prevalence") + theme(legend.position = "none")
+    p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND DALYs") + theme(legend.position = "none")
     save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p5)
@@
     p7 <- plot_tsne_phyloseq(pseq.subset, tsne_res,color = 'mnd')
-    p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND Prevalence") + theme(legend.position = "bottom")
+    p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND DALYs") + theme(legend.position = "bottom")
     save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p8)
@@
   #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
-  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme(axis.text.y = element_text(face="italic"))
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme(axis.text.y = element_text(face="italic")) + theme_void()
   save_plot(paste("PERMANOVA_",title_text,".png", sep = "" ), p)
@@
 
-pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & is.na(disease) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
 
@@
 beta_d(pseq.subset,"final_2")
+beta_d(atlas.pseq,"final_3")
 
@@
    theme(legend.position="none") + scale_colour_manual(values = wes_palette(glob_theme)) 
- library(rms)
-fita <- lrm(mnd ~ gender + log(Abundance) + bmi_group + age_category, data =mphyseq)
-fita <- lrm(mnd ~ gender + log(Abundance) , data =mphyseq)
-fita <- lrm(mnd ~ gender + log(Abundance) + bmi_group, data =mphyseq)
+ #library(rms)
+#fita <- lrm(mnd ~ gender + log(Abundance) + bmi_group + age_category, data =mphyseq)
+#fita <- lrm(mnd ~ gender + log(Abundance) , data =mphyseq)
+#fita <- lrm(mnd ~ gender + log(Abundance) + bmi_group, data =mphyseq)
 
-save_plot(paste("akkermansia",title_text,".png", sep = "" ), p)
+save_plot(paste("akkermansia_",title_text,".png", sep = "" ), p)
 }

### 2020-01-27 manuscript_NS_MRD.Rmd
updated

@@
   #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
-  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme(axis.text.y = element_text(face="italic")) + theme_void()
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme_classic() + theme(axis.text.y = element_text(face="italic"))
   save_plot(paste("PERMANOVA_",title_text,".png", sep = "" ), p)
@@
 
-pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & is.na(disease) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
+pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & is.na(disease) | disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
+
+pseq.subset <- subset_samples(pseq.subset, !is.na(nationality) )
+atlas.pseq <- subset_samples(atlas.pseq, !is.na(nationality) )
 
@@
 my_atlas_table(atlas.pseq)
-alpha_d(pseq,"final_RLWT_1")
+alpha_d(pseq.subset,"final_RLWT_1")
 

### 2020-01-27 manuscript_NS_MRD.Rmd
updated

@@
 pseq.subset <- subset_samples(pseq.subset, !is.na(nationality) )
-atlas.pseq <- subset_samples(atlas.pseq, !is.na(nationality) )
+atlas.pseq <- subset_samples(atlas.pseq, !is.na(nationality), !is.na(mnd))
 
@@
               position = position_jitter(width = 0.3)) + scale_y_log10()+
-   theme(legend.position="none") + scale_colour_manual(values = wes_palette(glob_theme)) 
+   theme(legend.position="none") + scale_colour_manual(values = wes_palette(glob_theme)) + theme_classic() + xlab('MND DALYs')
  #library(rms)
@@
 
-save_plot(paste("akkermansia_",title_text,".png", sep = "" ), p)
+#save_plot(paste("akkermansia_",title_text,".png", sep = "" ), p)
 }
 
-plotAtlas = plot_abundance.ak(atlas.pseq, "atlas")
-plotCur = plot_abundance.ak(pseq.subset, "curated")
-grid.arrange(nrow = 2, plotAtlas, plotCur)
+plot_abundance.ak(atlas.pseq, "atlas")
+plot_abundance.ak(pseq.subset, "curated")
+
+#plotAtlas <- plot_abundance.ak(atlas.pseq, "atlas")
+#plotCur <- plot_abundance.ak(pseq.subset, "curated")
+#plot(grid.arrange(nrow = 2, plotAtlas, plotCur))
 

### 2020-01-28 manuscript_NS_MRD.Rmd
updated

@@
 
-pseq.subset <- subset_samples(pseq.subset, !is.na(nationality) )
+pseq.subset <- subset_samples(pseq.subset, !is.na(nationality) , !is.na(mnd) )
 atlas.pseq <- subset_samples(atlas.pseq, !is.na(nationality), !is.na(mnd))
@@
 
-plot_abundance.ak = function(pseq.subset,title = "",
+plot_abundance.ak = function(pseq.subset, title_text,
                          Facet = "Genus", Color = "Phylum"){
@@
 
-#save_plot(paste("akkermansia_",title_text,".png", sep = "" ), p)
+save_plot(paste("akkermansia_",title_text,".png", sep = "" ), p)
 }

### 2020-01-28 manuscript_NS_MRD.Rmd
updated

@@
  mphyseq <- subset(mphyseq, Abundance > 0)
+
+  mnd <- levels(mphyseq$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+  # Compare differences in Shannon index between mnd group of the study subjects
+  mphyseq$mnd <- factor(mphyseq$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+  print("The basic alpha diversity analysis has been completed")
+
  p <- ggplot(data = mphyseq, mapping = aes_string(x = "mnd",y = "Abundance")) +
@@
               position = position_jitter(width = 0.3)) + scale_y_log10()+
-   theme(legend.position="none") + scale_colour_manual(values = wes_palette(glob_theme)) + theme_classic() + xlab('MND DALYs')
+   theme(legend.position="none") + scale_colour_manual(values = wes_palette(glob_theme)) + stat_compare_means(comparisons = mnd.pairs) + theme_classic() + xlab('MND DALYs')
  #library(rms)

### 2020-01-28 manuscript_NS_MRD.Rmd
updated

@@
 
-pseq.subset <- subset_samples(pseq.subset, !is.na(nationality) , !is.na(mnd) )
+pseq.subset <- subset_samples(pseq.subset, !is.na(nationality))
+pseq.subset <- subset_samples(pseq.subset, !is.na(mnd))
+
+#na.exclude(pseq.subset@sam_data[["mnd"]])
+
+# Check mnd variable
+# pseq.subset@sam_data[["mnd"]]
+
 atlas.pseq <- subset_samples(atlas.pseq, !is.na(nationality), !is.na(mnd))

### 2020-02-03 manuscript_NS_MRD.Rmd
updated

@@
 
+```{r Abundance}
+
+abundance <- function(atlas.pseq, title_text){
+  pseq.ph <- aggregate_taxa(atlas.pseq, "Phylum")
+
+pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
+
+plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
+theme(legend.position = "bottom") + theme_bw() +
+theme(axis.text.x=element_blank()) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum")) + theme(legend.text = element_text( face="italic"))
+
+p <- plot.comp.rel
+ save_plot(paste(title_text,".png", sep = ""), p )
+
+}
+
+abundance.bar <- function(atlas.pseq, title_text){
+  p.new <- aggregate_taxa(atlas.pseq, "Phylum")
+
+p.new.rel <- microbiome::transform(p.new, "compositional")
+
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+
+p <- plot.p.new.rel + scale_fill_brewer(palette = "Paired")
+
+ save_plot(paste(title_text,".png", sep = ""), p )
+}
+
+abundance(atlas.pseq, "atlas_abundance")
+abundance(pseq.subset, "pseq_abundance")
+
+abundance.bar(atlas.pseq, "atlas_bar")
+abundance.bar(pseq.subset, "pseq_bar")
+
+```
+
 ```{r MNDtable summary}

### 2020-03-20 siamcat_MND_MRD.Rmd
updated

---
title: "siamcat_MND_MRD"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### ATTEMPT WITH MND ######

```{r}

library(ExperimentHub)
library(curatedMetagenomicData)
suppressPackageStartupMessages(library(ExperimentHub))
library(phyloseq)
library(microbiome)
library(tidyverse) 
library(SIAMCAT) 
library(ggpubr)

eh = ExperimentHub()
myquery = query(eh, "curatedMetagenomicData")

# Get the stool samples
esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)

# merge
eset <- mergeData(esl)

# the relab=FALSE is essential as it Absolute Raw Count Data
pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
orig_pseq <- pseq

# Create mnd variable
#pseq <- orig_pseq
# grabs the nationality from phyloseq
c.mnd.country <- get_variable(pseq, "country" )
table(c.mnd.country)

# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
# Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,

#Adjsuted to match exactly the ATLAS data set

#central Europe = Belgium, Demark, Netherlands, Germany
#southern Europe = France, Italy, Serbia, Spain
#UKIE
#Scandinavia = Norway, Sweden Finland
#Eastern Europe = Poland

c.mnd.atlas <- fct_collapse(c.mnd.country, Scandinavia = c("NOR","SWE","FIN") ,CentralEurope = c("DNK","DEU","NLD"), SouthEurope = c("ITA","ESP","FRA"), US = "USA", UKIE = "GBR" )

#drops unused levels
c.mnd.atlas <- droplevels(c.mnd.atlas,c("ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "BGD","FJI","CHN","MDG","MNG","PER","RUS","TZA", "KAZ"))

#c.mnd.atlas <- subset_samples(c.mnd.atlas, !is.na("country") )
#Error in access(object, "sam_data", errorIfNULL) : sam_data slot is empty.

sample_data(pseq)$nationality = c.mnd.atlas

#mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DNK","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )

#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

# reorder
# NOTE THERE IS NO EASTERN EUROPE
c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE" ), HIGH = c("US") )

c.mnd.nationality <- factor(c.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
levels(c.mnd.nationality)

# creates a new variable in the phyloseq called mnd
sample_data(pseq)$mnd = c.mnd.nationality

# checks that it has worked.
get_variable(pseq, "mnd")
table(get_variable(pseq, "mnd"))

```

The taxonomic abundance profiles and the sample meatadata for the example dataset are available through the curatedMetagenomicsData R package.

clean the data a bit

```{r}

table(pseq@sam_data[["country"]])

table(pseq@sam_data[["mnd"]])

```

The input data should be organized in the same way for every version of SIAMCAT. All files are in tab-separated column format

Label data: First row is expected to be #BINARY:1=[label for cases];-1=[label for controls]
Second row should contain the sample identifiers as tab-separated list (consistent with feature and metadata).
Third row is expected to contain the actual class labels (tab-separated), e.g. 1 for each case and -1 for each
control.
Note: Labels can take other numeric values (but not characters or strings); importantly, the label for cases has to
be greater than the one for controls.

Feature matrix: features (in rows) x samples (in columns)
First row should contain sample labels (consistent with label data), while the first column should contain feature
labels (e.g. taxonomic identifiers). The remaining entries are expected to be real values >= 0 that quantify the
abundance of each feature in each sample.

Metadata (optional): samples (in rows) x metadata (in columns)
Metadata needs to be converted to numerical values by the user (This is necessary for heatmap displays)!

# Create Label
```{r}

# The siamcat object extends on the phyloseq object. Therefore, creating a siamcat object from a phyloseq object is really straightforward. This can be done with the siamcat constructor function. First, however, we need to create a label object:
label <- create.label(meta=sample_data(pseq),
    label = "mnd",
    case = c("MEDIUM", "HIGH"), control = "LOW")

# Assigned LOW as control group

siamcat <- siamcat(phyloseq=pseq, label=label)

```

```{r}

#Accessing and assigning slots
#Each slot in siamcat can be accessed by typing
  #slot_name(siamcat)

#e.g. for the eval_data slot you can types
  #eval_data(siamcat)
#There is one notable exception: the phyloseq slot has to be accessed with physeq(siamcat) due to technical reasons.
physeq(siamcat)

#Slots will be filled during the SIAMCAT workflow by the package’s functions. However, if for any reason a slot needs to be assigned outside of the workflow, the following formula can be used:
    #slot_name(siamcat) <- object_to_assign

#e.g. to assign a new_label object to the label slot:
    #label(siamcat) <- new_label
#Please note that this may lead to unforeseen consequences…

```

```{r}

#There are two slots that have slots inside of them. First, the model_list slot has a models slot that contains the actual list of mlr models -can be accessed via models(siamcat)- and model.type which is a character with the name of the method used to train the model: model_type(siamcat).

#The phyloseq slot has a complex structure. However, unless the phyloseq object is created outside of the SIAMCAT workflow, only two slots of phyloseq slot will be occupied: the otu_table slot containing the features table and the sam_data slot containing metadata information. Both can be accessed by typing either features(siamcat) or meta(siamcat).

#Additional slots inside the phyloseq slots do not have dedicated accessors, but can easily be reached once the phyloseq object is exported from the siamcat object:

phyloseq <- physeq(siamcat)
tax_tab <- tax_table(phyloseq)
head(tax_tab)

```

## Feature Filtering
Now, we can filter feature with low overall abundance and prevalence.
```{r}

siamcat <- filter.features(siamcat, cutoff=1e-04, filter.method = 'abundance')

siamcat <- filter.features(siamcat, cutoff=0.05,
                          filter.method='prevalence',
                          feature.type = 'filtered')

```

A few information about the siamcat object can be accessed with the show function from phyloseq (SIAMCAT builds on the phyloseq data structure):

```{r}

show(siamcat)

```

Since we have quite a lot of microbial species in the dataset at the moment, we can perform unsupervised feature selection using the function filter.features.
```{r}

siamcat <- filter.features(siamcat,
    filter.method = 'abundance',
    cutoff = 0.001)

```

#### Association Testing ####

Associations between microbial species and the label can be tested with the check.associations function. The function computes for each species the significance using a non-parametric Wilcoxon test and different effect sizes for the association (e.g. AUC or fold change).

The function produces a pdf file as output, since the plot is optimized for a landscape DIN-A4 layout, but can also used to plot on an active graphic device, e.g. in RStudio. The resulting plot then looks like that:
```{r}

siamcat <- check.associations(
    siamcat,
    sort.by = 'fc',
    fn.plot = 'assocMND.pdf',
    alpha = 0.05,
    mult.corr = "fdr",
    detect.lim = 10 ^-6,
    plot.type = "quantile.box",
    panels = c("fc", "prevalence", "auroc"))

```

#### Confounder Analysis

We can also check the supplied meta-variables for potential confounding.

```{r}

check.confounders(siamcat, fn.plot = './confoundersMND.pdf')

```

we have more samples than individual subjects:
```{r}

print(length(unique(pseq$subjectID))) 
print(nrow(pseq))

print(length(unique(pseq@sam_data[["subjectID"]])))
#6784 total rows

```

Some subjects (but not all) had been sampled multiple times. Therefore, we want to remove repeated samplings for the same subject, since the samples would otherwise not be indepdenent from another.

The visit number is encoded in the sampleID. Therefore, we can use this information to extract when the samples have been taken and use only the first visit for each subject.

```{r}

pseq.s<- pseq %>%
  select(sampleID, subjectID, study_condition, disease_subtype, 
disease, age, country, number_reads, median_read_length, BMI, mnd) %>%
  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
  mutate(visit=str_remove(visit, '_')) %>% 
  mutate(visit=as.numeric(visit)) %>%
  mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
  group_by(subjectID) %>% 
  filter(visit==min(visit)) %>%
  ungroup() %>%
  mutate(Sample_ID=sampleID) %>%
  mutate(Group=case_when(mnd=='LOW'~'CTR'))

# mutate(Group=case_when(disease=='healthy'~'CTR')
```

First, SIAMCAT needs a feature matrix (can be either a matrix, a data.frame, or a phyloseq-otu_table), which contains values of different features (in rows) for different samples (in columns). For example, the feature matrix included here contains relative abundances for bacterial species calculated with the mOTU profiler for 141 samples:

```{r}

feat.crc.zeller[1:3, 1:3]

# Dimensions of object
dim(feat.crc.zeller)

```

#### Taxonomin Profiles

Load the taxonomic profiles generated with MetaPhlAn2 via the curatedMetagenomicsData R package.

```{r}

# Get the stool samples
esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)

# merge
eset <- mergeData(esl)

# the relab=FALSE is essential as it Absolute Raw Count Data
pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
orig_pseq <- pseq

y <- "*metaphlan_bugs_list.stool*"

feat <- curatedMetagenomicData(x=orig_pseq, dryrun=FALSE)

feat <- feat[[x]]@assayData$exprs

feat <- feat[[y]]@assayData$exprs

x <- 'NielsenHB_2014.metaphlan_bugs_list.stool'
feat <- curatedMetagenomicData(x=x, dryrun=FALSE)

feat <- feat[[x]]@assayData$exprs

```

The MetaPhlAn2 profiles contain information on different taxonomic levels. Therefore, we want to restrict them to species-level profiles. In a second step, we convert them into relative abundances (summing up to 1) instead of using the percentages (summing up to 100) that MetaPhlAn2 outputs.
```{r}

feat <- feat[grep(x=rownames(feat), pattern='s__'),]
feat <- feat[grep(x=rownames(feat),pattern='t__', invert = TRUE),]
feat <- t(t(feat)/100)

```

The feature names are very long and may be a bit un-wieldy for plotting later on, so we shorten them to only the species name:
```{r}

rownames(feat) <- str_extract(rownames(feat), 's__.*$')

```

#### Create the SIAMCAT Object

Now, we have everything ready to create a SIAMCAT object which stores the feature matrix, the meta-variables, and the label. Here, the label is created using the information in the metadata.
```{r}

sc.obj <- siamcat(feat=feat, meta=meta.healthy, label='Group', case='mnd')

sc.obj <- siamcat(meta.healthy, label='Group', case='MEDIUM,HIGH')

```

## Feature Filtering
Now, we can filter feature with low overall abundance and prevalence.
```{r}

sc.obj <- filter.features(sc.obj, cutoff=1e-04, filter.method = 'abundance')

sc.obj <- filter.features(sc.obj, cutoff=0.05,
                          filter.method='prevalence',
                          feature.type = 'filtered')

```

### Association Plot

The check.assocation function calculates the significance of enrichment and metrics of association (such as generalized fold change and single-feautre AUROC).
```{r}

sc.obj <- check.associations(sc.obj, detect.lim = 1e-06, 
                             alpha=0.1, 
                             max.show = 20,
                             plot.type = 'quantile.rect',
                             fn.plot = './association_plot.pdf')

```

### 2020-03-20 siamcat_MRD.Rmd
updated

---
title: "siamcat_MRD"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

SIAMCAT - Statistical Inference of Associations between Microbial Communities And host phenoTypes
Goal: determine changes in community composition that are associated with environmental factors

SIAMCAT provides a full pipeline supporting data preprocessing, statistical association testing, statistical modeling (LASSO logistic regression) including tools for evaluation and interpretation of these models (such as cross validation, parameter selection, ROC analysis and diagnostic model plots)

```{r install}

# can install it from Bioconductor
# if (!requireNamespace("BiocManager", quietly = TRUE))
   # install.packages("BiocManager")
# BiocManager::install("SIAMCAT", version = "3.10")

# alternatively ca install the current development version via devtools
require("devtools")
devtools::install_github(repo = 'zellerlab/siamcat')

setwd("/Users/maryam/Scripts/Microbiome")

# link on how to input own data into siamcat : https://bioconductor.org/packages/release/bioc/vignettes/SIAMCAT/inst/doc/SIAMCAT_read-in.html

```

########### Following https://konradzych.github.io/SIAMCAT/articles/SIAMCAT_vignette.html Vignette

####Step 1: Read/Validate/Filter Data

1. Load the SIAMCAT package and use the files included in SIAMCAT
  The data are the same as used in the publication of Zeller et al., which demonstrated the potential of microbial markers in feacal samples to distinguish patients with colorectal cancer (CRC) from healthy controls.

```{r}

library(SIAMCAT)
library(RColorBrewer)
library(pROC)
library(tidyverse)
library(curatedMetagenomicData)
library(ggpubr)

# Load taxonomic levels onto feat
fn.in.feat  <- system.file(
    "extdata",
    "feat_crc_study-pop-I_N141_tax_profile_mocat_bn_specI_clusters.tsv",
    package = "SIAMCAT"
)

# Label groups into label
fn.in.label <- system.file(
    "extdata",
    "label_crc_study-pop-I_N141_tax_profile_mocat_bn_specI_clusters.tsv",
    package = "SIAMCAT"
)

# Load sample metadata
fn.in.meta  <- system.file(
    "extdata",
    "num_metadata_crc_study-pop-I_N141_tax_profile_mocat_bn_specI_clusters.tsv",
    package = "SIAMCAT"
)

```

2. Access the files with the dedicated SIAMCAT functions and directly construct a SIAMCAT object containing the microbial features, the patient’s labels, and metadata for the patients
```{r}

# CAN'T FIND FUNCTIONS read.features
# feat  <- read.features(fn.in.feat)

# label <- read.labels(fn.in.label)
# meta  <- read.meta(fn.in.meta)
# siamcat <- siamcat(feat, label, meta)

show(siamcat)

# Show phyloseq object
phyloseq <- physeq(siamcat)
show(phyloseq)

```

```{r}

# validate.data function ensures that we have labels for all the samples in features and vice versa.
# ONLY call validate.data before starting the pipeline!
#    siamcat <- validate.data(siamcat, verbose=1)

# Should read:
## Data succesfully validated

```

Data can also be sub-selected based on the available meta-data. For example, if we want to exclude patients that are too young or too old for the question of interest, we can do so easily with:
```{r}

#siamcat <- select.samples(
#    siamcat,
#    filter = 'age',
#    allowed.set = NULL,
#    allowed.range = c(20, 90),
#    verbose = 2
#)

```

####Step 2: Association/Confounder Testing

Confounders are checked with the function check.confounders, which produces a plot for each possible confounder in the metadata and diverts the output into a pdf-file.

```{r}

check.confounders(siamcat,
    fn.plot = 'conf_check.pdf')

```

Associations between microbial markers and the label can be tested with the check.associations function. The function computes a generalized fold change for the marker, the prevalence shift, a single feature AUC, and the significance of the associations. The significance is tested with a Wilcoxon test. The function again produces a pdf-file as output
```{r}

check.associations(
    siamcat,
    sort.by = 'fc',
    fn.plot = 'assoc.pdf',
    alpha = 0.05,
    mult.corr = "fdr",
    detect.lim = 10 ^-6,
    plot.type = "quantile.box",
    panels = c("fc", "prevalence", "auroc"))

```

#### Step 3: Model Building

Construction of machine learning models on the basis of microbial markers. 
SIAMCAT contains functions for data normalization, splitting the data into cross-validation folds, training the model, and making predictions based on cross-validation instances and the trained models.

### Data Normalisation

Data normalization is performed with the normalize.features function. Several control options are available, i.e. the normalization method (log.unit, log.std, rank.unit, rank.std, log.clr) or additional parameters. Here, we use the log.unit method:
```{r}

siamcat <- normalize.features(
    siamcat,
    norm.method = "log.unit",
    norm.param = list(
        log.n0 = 1e-06,
        n.p = 2,
        norm.margin = 1
    ),
    verbose = 2
)

```

### Prepare Cross-Validation

Preparation of the cross-validation fold is a crucial step in machine learning. 
SIAMCAT greatly simplifies the set-up of cross-validation schemes, including stratification of samples or keeping samples inseperable based on metadata. For this small example, we choose a twice-repeated 5-fold cross-validation scheme. The data-split will be saved in the data_split slot of the siamcat object.

```{r}

siamcat <-  create.data.split(
    siamcat,
    num.folds = 5,
    num.resample = 2,
    stratify = TRUE,
    inseparable = NULL,
    verbose = 2
)

```

### Model Training

The actual model training is performed using the function train.model. 
Again, multiple options for customization are available, ranging from the machine learning method to the measure for model selection or customizable parameter set for hyperparameter tuning. Here, we train a Lasso model and enforce at least 5 non-zero coefficients.

```{r}
siamcat <- train.model(
    siamcat,
    method = "lasso",
    stratify = TRUE,
    modsel.crit = list("pr"),
    min.nonzero.coeff = 5,
    param.set = NULL,
    verbose = 3
)

```

The models are saved in the model_list slot of the siamcat object. This slot stores objects of model_list-class. To get the complete list out of the SIAMCAT object:
```{r}

model_list <- model_list(siamcat)

```

This slot also stores information on which method was used to construct the model:
```{r}

model_type(siamcat)

```

Models can also be easily accessed:
```{r}

models <- models(siamcat)
models[[1]]

```

### Make Predictions

Using the data-split and the models trained in previous step, we can use the function make.predictions in order to apply the models on the test instances in the data-split. The predictions will be saved in the pred_matrix slot of the siamcat object.

```{r}

siamcat <- make.predictions(siamcat, verbose=0)
pred_matrix <- pred_matrix(siamcat)
head(pred_matrix)

```

#### Step 4: Model Evaluation and Interpretation

In the final part, we want to find out how well the model performed and which microbial markers had been selected in the model. To do so, we first calculate how well the predictions fit the real data using the function evaluate.predictions. This function calculates the Area Under the Receiver Operating Characteristic (ROC) Curve (AU-ROC) and the Precision Recall (PR) Curve for each resampled cross-validation run. The results of the evaluation will be stored in the eval_data slot of the siamcat object.

```{r}

siamcat <-  evaluate.predictions(siamcat, verbose=2)

```

### Evaluation Plot

To plot the results of the evaluation, we can use the function model.evaluation.plot, which produces a pdf-file showing the ROC and PR Curves for the different resamples runs as well as the mean ROC and PR Curve.

```{r}

model.evaluation.plot(siamcat,'eval_plot.pdf',verbose = 2)

```

Instead of the pdf-output, we can also access the evaluation data in the siamcat object directly and plot e.g. the ROC-Curves:
```{r}

# plot ROC Curves
plot(
    NULL,
    xlim = c(0, 1),
    ylim = c(0, 1),
    xlab = 'False positive rate',
    ylab = 'True positive rate',
    type = 'n'
)
title('ROC curve for the model')
abline(a = 0, b = 1, lty = 3)
# for each resampled CV run
eval_data <- eval_data(siamcat)
for (r in 1:length(eval_data$roc.all)) {
    roc.c = eval_data$roc.all[[r]]
    lines(1 - roc.c$specificities, roc.c$sensitivities, 
        col = gray(runif(1, 0.2, 0.8)))
}
# mean ROC curve
roc.summ = eval_data$roc.average[[1]]
lines(1 - roc.summ$specificities,
    roc.summ$sensitivities,
    col = 'black',
    lwd = 2)
# plot CI
x = as.numeric(rownames(roc.summ$ci))
yl = roc.summ$ci[, 1]
yu = roc.summ$ci[, 3]
polygon(1 - c(x, rev(x)), c(yl, rev(yu)), col = '#88888844' , border = NA)

```

### Interpretation Plot

The final plot produced by SIAMCAT is the model interpretation plot, created by the model.interpretation.plot function. The plot shows for the top selected features the:

- model weights (and how robust they are) as a barplot,
- a heatmap with the z-scores or fold changes for the top selected features, and
- a boxplot showing the proportions of weight per model which is captured by the top selected features.

The function again produces a pdf-file.
```{r}

model.interpretation.plot(
    siamcat,
    fn.plot = 'interpretation.pdf',
    consens.thres = 0.5,
    #norm.models = TRUE,
    limits = c(-3, 3),
    heatmap.type = 'zscore',
    verbose = 2
)

# show the z-score heatmap for the top selected feature
# + model.interpretation.select.features

```

###### END OF THIS VIGNETTE ##########################

##### WORKFLOW FROM https://siamcat.embl.de/articles/SIAMCAT_vignette.html

As example dataset we use the data from the publication of Zeller et al, which demonstrated the potential of microbial species in fecal samples to distinguish patients with colorectal cancer (CRC) from healthy controls.
```{r}

library(SIAMCAT)

data("feat_crc_zeller", package="SIAMCAT")
data("meta_crc_zeller", package="SIAMCAT")

```

First, SIAMCAT needs a feature matrix (can be either a matrix, a data.frame, or a phyloseq-otu_table), which contains values of different features (in rows) for different samples (in columns). For example, the feature matrix included here contains relative abundances for bacterial species calculated with the mOTU profiler for 141 samples:

```{r}

feat.crc.zeller[1:3, 1:3]

# Dimensions of object
dim(feat.crc.zeller)

```

Metadata about the samples in another data.frame:
```{r}

head(meta.crc.zeller)

```

In order to tell SIAMCAT, which samples are cancer cases and which are healthy controls, we can construct a label object from the Group column in the metadata.
```{r}

label.crc.zeller <- create.label(meta=meta.crc.zeller, label='Group', case='CRC')

```

Now we have all the ingredients to create a SIAMCAT object

```{r}

siamcat <- siamcat(feat=feat.crc.zeller,
    label=label.crc.zeller,
    meta=meta.crc.zeller)

```

A few information about the siamcat object can be accessed with the show function from phyloseq (SIAMCAT builds on the phyloseq data structure):

```{r}

show(siamcat)

```

Since we have quite a lot of microbial species in the dataset at the moment, we can perform unsupervised feature selection using the function filter.features.
```{r}

siamcat <- filter.features(siamcat,
    filter.method = 'abundance',
    cutoff = 0.001)

```

#### Association Testing ####

Associations between microbial species and the label can be tested with the check.associations function. The function computes for each species the significance using a non-parametric Wilcoxon test and different effect sizes for the association (e.g. AUC or fold change).

The function produces a pdf file as output, since the plot is optimized for a landscape DIN-A4 layout, but can also used to plot on an active graphic device, e.g. in RStudio. The resulting plot then looks like that:
```{r}

siamcat <- check.associations(
    siamcat,
    sort.by = 'fc',
    fn.plot = 'assoc2.pdf',
    alpha = 0.05,
    mult.corr = "fdr",
    detect.lim = 10 ^-6,
    plot.type = "quantile.box",
    panels = c("fc", "prevalence", "auroc"))

```

##### Model Building ####

One strength of SIAMCAT is the versatile but easy-to-use interface for the construction of machine learning models on the basis of microbial species. SIAMCAT contains functions for data normalization, splitting the data into cross-validation folds, training the model, and making predictions based on cross-validation instances and the trained models.

#### Data Normalization

Data normalization is performed with the normalize.features function. Here, we use the log.unit method, but several other methods and customization options are available (please check the documentation).

```{r}

siamcat <- normalize.features(
    siamcat,
    norm.method = "log.unit",
    norm.param = list(
        log.n0 = 1e-06,
        n.p = 2,
        norm.margin = 1
    )
)

```

#### Perpare Cross-Validation

https://www.youtube.com/watch?v=fSytzGwwBVw

Cross-Validation is basically how we decide which machine learning method would be best for our data set. CV allows us to compare different machine learning methods and get a sense of how well they will work in practice.

Examples of machine learning methods:
1. Logistic Regression
2. K-nearest neighbours
3. Support Vector Machines (SVM) ...etc.

Need to do 2 things with collected data:
1. Estimate parameters for machine learning methods. (training algorithm)
2. Evaluate how well the machine learning method works. (testing algorithm)

Terrible approach to use ALL the data to train algorithm bc wouldn't have any data left to test algorithm. - Data separated into blocks, different block combinations used to train model.

Preparation of the cross-validation fold is a crucial step in machine learning. SIAMCAT greatly simplifies the set-up of cross-validation schemes, including stratification of samples or keeping samples inseperable based on metadata. For this small example, we choose a twice-repeated 5-fold (no. of blocks) cross-validation scheme. The data-split will be saved in the data_split slot of the siamcat object.
```{r}

siamcat <-  create.data.split(
    siamcat,
    num.folds = 5,
    num.resample = 2
)

```

#### Model Training

The actual model training is performed using the function train.model. Again, multiple options for customization are available, ranging from the machine learning method to the measure for model selection or customizable parameter set for hyperparameter tuning.
```{r}

siamcat <- train.model(
    siamcat,
    method = "lasso"
)

siamcat <- train.model(
    siamcat,
    method = "enet"
)

siamcat <- train.model(
    siamcat,
    method = "ridge"
)

siamcat <- train.model(
    siamcat,
    method = "lasso_ll"
)

siamcat <- train.model(
    siamcat,
    method = "ridge_ll"
)

siamcat <- train.model(
    siamcat,
    method = "randomForest"
)

```

The models are saved in the model_list slot of the siamcat object. The model building is performed using the mlr R package. All models can easily be accessed.
```{r}

# get information about the model type
model_type(siamcat)

# access the models
models <- models(siamcat)
models[[1]]

```

#### Make Predicitions

Using the data-split and the models trained in previous step, we can use the function make.predictions in order to apply the models on the test instances in the data-split. The predictions will be saved in the pred_matrix slot of the siamcat object.

```{r}

siamcat <- make.predictions(siamcat)
pred_matrix <- pred_matrix(siamcat)

head(pred_matrix)

```

#### Model Evaluation and Interpretation ####

In the final part, we want to find out how well the model performed and which microbial species had been selected in the model. In order to do so, we first calculate how well the predictions fit the real data using the function evaluate.predictions. This function calculates the Area Under the Receiver Operating Characteristic (ROC) Curve (AU-ROC) and the Precision Recall (PR) Curve for each resampled cross-validation run.

```{r}

siamcat <-  evaluate.predictions(siamcat)

```

#### Evaluation Plot

To plot the results of the evaluation, we can use the function model.evaluation.plot, which produces a pdf-file showing the ROC and PR Curves for the different resamples runs as well as the mean ROC and PR Curve.

```{r}

model.evaluation.plot(siamcat)

```

#### Interpretation Plot

The final plot produced by SIAMCAT is the model interpretation plot, created by the model.interpretation.plot function. The plot shows for the top selected features the:

  - model weights (and how robust they are, i.e. in what proportion of models have they been incorporated) as a barplot

  - a heatmap with the z-scores or fold changes for the top selected features

  - a boxplot showing the proportions of weight per model which is captured by the top selected features.

Additionally, the distribution of metadata is shown in a heatmap below.

The function again produces a pdf-file optimized for a landscape DIN-A4 plotting region.

```{r}

model.interpretation.plot(
    siamcat, 
    fn.plot = 'interpretation2.pdf',
    color.scheme = "BrBG",
    consens.thres = 0.5,
    heatmap.type = "zscore",
    limits = c(-3, 3), detect.lim = 1e-06,
    max.show = 50, prompt=TRUE, verbose = 1)

```

##### END OF THIS VIGNETTE ####

##### STANDARD WORKFLOW FROM SUPPLEMENTARY NOTE 1 ####

Standard workflow of the SIAMCAT package using as an example the dataset from Nielsen et al. Nat Biotechnol 2014. This dataset contains samples from patients with inflammatory bowel disease and from controls.
Running the standard SIAMCAT pipeline, we produce the primary outputs of the package, namely the
• Association plot
• Confounder plots
• Model evaluation plot
• Model interpretation plot

### Loading the Data

The taxonomic abundance profiles and the sample meatadata for the example dataset are available through the curatedMetagenomicsData R package.
```{r}

library("tidyverse") 
library("SIAMCAT") 
library("curatedMetagenomicData") 
library("ggpubr")

meta.nielsen.full <- combined_metadata %>% filter(dataset_name=='NielsenHB_2014')

```

However, we have to clean the data a bit, since the data set contains samples from two different countries and two different disease subtypes.
```{r}

table(meta.nielsen.full$country)

table(meta.nielsen.full$disease, meta.nielsen.full$disease_subtype)

```

Also, we have an additional problem, because we have more samples than individual subjects:
```{r}

print(length(unique(meta.nielsen.full$subjectID)))

print(nrow(meta.nielsen.full))

```

Some subjects (but not all) had been sampled multiple times. Therefore, we want to remove repeated samplings for the same subject, since the samples would otherwise not be indepdenent from another (see also Figure 4 in the main manuscript).

The visit number is encoded in the sampleID. Therefore, we can use this information to extract when the samples have been taken and use only the first visit for each subject.
```{r}

meta.nielsen <- meta.nielsen.full %>%
  select(sampleID, subjectID, study_condition, disease_subtype, 
disease, age, country, number_reads, median_read_length, BMI) %>%
  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
  mutate(visit=str_remove(visit, '_')) %>% 
  mutate(visit=as.numeric(visit)) %>%
  mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
  group_by(subjectID) %>% 
  filter(visit==min(visit)) %>%
  ungroup() %>%
  mutate(Sample_ID=sampleID) %>%
  mutate(Group=case_when(disease=='healthy'~'CTR', TRUE~disease_subtype))

```

Also, we want to restrict the dataset to Spanish samples only and to control samples and ulcerative colitis samples only. Additionally, we transform the metadata (which are stored as a tibble for easier handling) into a dataframe to hand it over to SIAMCAT.
```{r}

meta.nielsen.uc <- meta.nielsen %>% 
  filter(Group %in% c('CTR', 'UC')) %>%
  filter(country== 'ESP') %>%
  as.data.frame()
rownames(meta.nielsen.uc) <- meta.nielsen.uc$sampleID
meta.nielsen.uc$sampleID <- NULL

```

#### Taxonomin Profiles

Load the taxonomic profiles generated with MetaPhlAn2 via the curatedMetagenomicsData R package.

```{r}

x <- 'NielsenHB_2014.metaphlan_bugs_list.stool'
feat <- curatedMetagenomicData(x=x, dryrun=FALSE)

feat <- feat[[x]]@assayData$exprs

```

The MetaPhlAn2 profiles contain information on different taxonomic levels. Therefore, we want to restrict them to species-level profiles. In a second step, we convert them into relative abundances (summing up to 1) instead of using the percentages (summing up to 100) that MetaPhlAn2 outputs.
```{r}

feat <- feat[grep(x=rownames(feat), pattern='s__'),]
feat <- feat[grep(x=rownames(feat),pattern='t__', invert = TRUE),]
feat <- t(t(feat)/100)

```

The feature names are very long and may be a bit un-wieldy for plotting later on, so we shorten them to only the species name:
```{r}

rownames(feat) <- str_extract(rownames(feat), 's__.*$')

```

#### Create the SIAMCAT Object

Now, we have everything ready to create a SIAMCAT object which stores the feature matrix, the meta-variables, and the label. Here, the label is created using the information in the metadata.
```{r}

sc.obj <- siamcat(feat=feat, meta=meta.nielsen.uc, label='Group', case='UC')

```

## Feature Filtering
Now, we can filter feature with low overall abundance and prevalence.
```{r}

sc.obj <- filter.features(sc.obj, cutoff=1e-04, filter.method = 'abundance')

sc.obj <- filter.features(sc.obj, cutoff=0.05,
                          filter.method='prevalence',
                          feature.type = 'filtered')

```

### Association Plot

The check.assocation function calculates the significance of enrichment and metrics of association (such as generalized fold change and single-feautre AUROC).
```{r}

sc.obj <- check.associations(sc.obj, detect.lim = 1e-06, 
                             alpha=0.1, 
                             max.show = 20,
                             plot.type = 'quantile.rect',
                             fn.plot = './association_plot.pdf')

```

#### Confounder Analysis

We can also check the supplied meta-variables for potential confounding.

```{r}

check.confounders(sc.obj, fn.plot = './confounders.pdf')

```

### Machine Learning Workflow

Steps:
• Feature normalization
• Data splitting for cross-validation
• Model training
• Making model predictions (on left-out data)
• Evaluating model predictions (using AUROC and AUPRC)

```{r}

sc.obj <- normalize.features(sc.obj, norm.method = 'log.std', norm.param = list(log.n0=1e-06, sd.min.q=0)) 
## Features normalized successfully.

sc.obj <- create.data.split(sc.obj, num.folds = 10, num.resample = 10) 
## Features splitted for cross-validation successfully.

sc.obj <- train.model(sc.obj, method='lasso')
## Trained lasso models successfully.

sc.obj <- make.predictions(sc.obj)
## Made predictions successfully.

sc.obj <- evaluate.predictions(sc.obj)
## Evaluated predictions successfully.

```

### Model Evaluation Plot
The model evaluation plot will produce one plot with the ROC curve and another one with the precision-recall curve (not shown here).
```{r}

model.evaluation.plot(sc.obj, fn.plot = './eval_plot.pdf')

```

### Model Interpretation Plot
The model interpretation plot can give you additional information about the trained machine learning model. It will show you:
  • the feature importance as barplot,
  • the feature robustness (in how many of the models in the repeated cross-validation this feature has
been selected into the model),
  • the normalized feature abundances across samples as heatmap,
  • the optional metadata as heatmap below, and
  • a boxplot showing the proportion of the model weight that is explained by
the selected features.
```{r}

model.interpretation.plot(sc.obj, consens.thres = 0.8, fn.plot = './interpretation.pdf')

```

### Confounder Analysis
As already mentioned above, the Nielsen dataset contains samples from both Spain and Denmark. Let us see what the result would be if we had not removed them.

First, we create a SIAMCAT object again, this time including the Danish controls:
```{r}

meta.nielsen.uc.dnk <- meta.nielsen %>%
  filter(Group %in% c('CTR', 'UC')) %>%
  # filter(country == ESP) %>% # this time we do not remove Danish samples
  as.data.frame()

rownames(meta.nielsen.uc.dnk) <- meta.nielsen.uc.dnk$sampleID
meta.nielsen.uc.dnk$sampleID <- NULL

sc.obj.dnk <- siamcat(feat=feat, meta=meta.nielsen.uc.dnk, label='Group', case='UC')

sc.obj.dnk <- filter.features(sc.obj.dnk, cutoff = 1e-04,
                              filter.method = 'abundance')

sc.obj.dnk <- filter.features(sc.obj.dnk, cutoff = 0.05,
                              filter.method = 'prevalence',
                              feature.type = 'filtered')

```

The confounder plot would show us that the meta-variable “country” might be problematic:
```{r}

check.confounders(sc.obj.dnk, fn.plot = './confounders_dnk.pdf')

```

### Association Testing

```{r}

sc.obj.dnk <- check.associations(sc.obj.dnk, detect.lim = 1e-06,
                                 alpha = 0.1, max.show = 20,
                                 plot.type = 'quantile.rect',
                                 fn.plot = './associations_plot_dnk.pdf')

```

Confounders can lead to biases in association testing. After using SIAMCAT to test for associations in both datasets (one time including the Danish samples, the other time restricted to samples from Spain only), we can extract the association metrics from both SIAMCAT objects and compare them in a scatter plot.

```{r}

assoc.sp <- associations(sc.obj) 
assoc.sp$species <- rownames(assoc.sp) 
assoc.sp_dnk <- associations(sc.obj.dnk) 
assoc.sp_dnk$species <- rownames(assoc.sp_dnk)

df.plot <- full_join(assoc.sp, assoc.sp_dnk, by='species')
df.plot %>%
  mutate(highlight=str_detect(species, 'Dorea_formicigenerans')) %>%
  ggplot(aes(x=-log10(p.adj.x), y=-log10(p.adj.y), col=highlight)) +
  geom_point(alpha=0.3) + 
  xlab('Spanish samples only\n-log10(q)') +
  ylab('Spanish and Danish samples only\n-log10(q)') +
  theme_classic() +
  theme(panel.grid.major = element_line(colour = 'lightgrey'), aspect.ratio = 1.3) +
  scale_colour_manual(values=c('darkgrey', '#D41645'), guide = FALSE) +
  annotate('text', x=0.7, y=8, label='Dorea formicigenerans')

```
This result shows that several species are only signficant if the Danish control samples are included, but not when considering only the Spanish samples.
As an example, we highlighted the species “Dorea formicigenerans” in the plot above. The test is not significant in the Spanish cohort, but is highly significant when the Danish samples are included.

```{r}

# Extract information out of the siamcat object
feat.dnk <- get.filt_feat.matrix(sc.obj.dnk)
label.dnk <- label(sc.obj.dnk)$label
country <- meta(sc.obj.dnk)$country
names(country) <- rownames(meta(sc.obj.dnk))

df.plot <- tibble(dorea=log10(feat.dnk['s__Dorea_formicigenerans',
                                       names(label.dnk)] + 1e-05),
                  label=label.dnk, country=country) %>%
  mutate(label=case_when(label=='-1'~'CTR', TRUE~"UC")) %>%
  mutate(x_value=paste0(country, '_', label))

df.plot %>%
  ggplot(aes(x=x_value, y=dorea)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.08, stroke=0, alpha=0.2) +
  theme_classic() +
  xlab('') +
  ylab("log10(Dorea_formicigenerans") +
  stat_compare_means(comparisons = list(c('DNK_CTR', 'ESP_CTR'),
                                        c('DNK_CTR', 'ESP_UC'),
                                        c('ESP_CTR', 'ESP_UC'))) +
  theme(aspect.ratio = 1.3)

```

### Machine Learning
The results from the machine learning workflows can also be biased by the differences between countries, leading to exaggerated performance estimates.

```{r}

sc.obj.dnk <- normalize.features(sc.obj.dnk, norm.method = 'log.std', 
                                 norm.param = list(log.n0=1e-06, sd.min.q=0))

sc.obj.dnk <- create.data.split(sc.obj.dnk, num.folds = 10, num.resample = 10)

sc.obj.dnk <- train.model(sc.obj.dnk, method = 'lasso')

sc.obj.dnk <- make.predictions(sc.obj.dnk)

sc.obj.dnk <- evaluate.predictions(sc.obj.dnk)

```

When we compare the performance of the two different models, the model with the Danish and Spanish samples included seems to perform better (higher AUROC value). However, the previous analysis suggests that this performance estimate is biased and exaggerated because differences between Spanish and Danish samples can be very large.

```{r}

model.evaluation.plot("Spanish samples only"=sc.obj,
                      "Danish and Spanish samples"=sc.obj.dnk,
                      fn.plot = './eval_plot_dnk.pdf')

```

To demonstrate how machine learning models can exloit this confounding factor, we can train a model to distinguish between Spanish and Danish control samples. As you can see, the model can distinguish between the two countries with almost perfect accuracy.

```{r}

meta.nielsen.country <- meta.nielsen %>%
  filter(Group %in% c('CTR')) %>% # only control samples
  # filter(country == 'ESP') %>% # this time we do not remove danish samples
  as.data.frame()
rownames(meta.nielsen.country) <- meta.nielsen.country$sampleID
meta.nielsen.country$sampleID <- NULL

sc.obj.country <- siamcat(feat=feat, meta=meta.nielsen.country,
                          label='country', case='ESP')

sc.obj.country <- filter.features(sc.obj.country, cutoff = 1e-04,
                                  filter.method = 'abundance')

sc.obj.country <- filter.features(sc.obj.country, cutoff = 0.05,
                                  filter.method = 'prevalence',
                                  feature.type = 'filtered')

sc.obj.country <- normalize.features(sc.obj.country, norm.method = 'log.std',
                                     norm.param = list(log.n0=1e-06,
                                                       sd.min.q=0))

sc.obj.country <- create.data.split(sc.obj.country,
                                    num.folds = 10, num.resample = 10)

sc.obj.country <- train.model(sc.obj.country, method = 'lasso')

sc.obj.country <- make.predictions(sc.obj.country)

sc.obj.country <- evaluate.predictions(sc.obj.country)

model.evaluation.plot(sc.obj.country, fn.plot = './eval_plot_country.pdf')

```

#### END OF VIGNETTE ####

#### CD META-ANALYSIS from supplementary note II ####

We want to demonstrate how SIAMCAT can facilitate metagenomic meta-analyses. Here, we use the example of Crohn’s disease (CD), since we have taxonomic profiles from five different metagenomic datasets available.

First, we load the data for all studies:

Standard workflow of the SIAMCAT package using as an example the dataset from Nielsen et al. Nat Biotechnol 2014. This dataset contains samples from patients with inflammatory bowel disease and from controls.
Running the standard SIAMCAT pipeline, we produce the primary outputs of the package, namely the
• Association plot
• Confounder plots
• Model evaluation plot
• Model interpretation plot

### Loading the Data

The taxonomic abundance profiles and the sample meatadata for the example dataset are available through the curatedMetagenomicsData R package.
```{r}

meta.nielsen.full <- combined_metadata %>% filter(dataset_name=='NielsenHB_2014')

```

However, we have to clean the data a bit, since the data set contains samples from two different countries and two different disease subtypes.
```{r}

table(meta.nielsen.full$country)

table(meta.nielsen.full$disease, meta.nielsen.full$disease_subtype)

```

Also, we have an additional problem, because we have more samples than individual subjects:
```{r}

print(length(unique(meta.nielsen.full$subjectID)))

print(nrow(meta.nielsen.full))

```

Some subjects (but not all) had been sampled multiple times. Therefore, we want to remove repeated samplings for the same subject, since the samples would otherwise not be indepdenent from another (see also Figure 4 in the main manuscript).

The visit number is encoded in the sampleID. Therefore, we can use this information to extract when the samples have been taken and use only the first visit for each subject.
```{r}

meta.nielsen <- meta.nielsen.full %>%
  select(sampleID, subjectID, study_condition, disease_subtype, 
disease, age, country, number_reads, median_read_length, BMI) %>%
  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
  mutate(visit=str_remove(visit, '_')) %>% 
  mutate(visit=as.numeric(visit)) %>%
  mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
  group_by(subjectID) %>% 
  filter(visit==min(visit)) %>%
  ungroup() %>%
  mutate(Sample_ID=sampleID) %>%
  mutate(Group=case_when(disease=='healthy'~'CD', TRUE~disease_subtype))

```

Also, we want to restrict the dataset to Spanish samples only and to control samples and ulcerative colitis samples only. Additionally, we transform the metadata (which are stored as a tibble for easier handling) into a dataframe to hand it over to SIAMCAT.
```{r}

meta.nielsen.uc <- meta.nielsen %>% 
  filter(Group %in% c('CTR', 'CD')) %>%
  filter(country== 'ESP') %>%
  as.data.frame()
rownames(meta.nielsen.uc) <- meta.nielsen.uc$sampleID
meta.nielsen.uc$sampleID <- NULL

```

#### Taxonomin Profiles

Load the taxonomic profiles generated with MetaPhlAn2 via the curatedMetagenomicsData R package.

```{r}

x <- 'NielsenHB_2014.metaphlan_bugs_list.stool'
feat <- curatedMetagenomicData(x=x, dryrun=FALSE)

feat <- feat[[x]]@assayData$exprs

```

##### END OF THIS VIGNETTE ####

Potential of microbial species in fecal samples to distinguish patients with colorectal cancer (CRC) from healthy controls
```{r load files}

# load files
library(SIAMCAT)

data("feat_crc_zeller", package="SIAMCAT")
data("meta_crc_zeller", package="SIAMCAT")

# show features
feat.crc.zeller[1:3, 1:3]
dim(feat.crc.zeller)

# show meta
head(meta.crc.zeller)

# create label
label.crc.zeller <- create.label(meta=meta.crc.zeller,
    label='Group', case='CRC')

# start
siamcat <- siamcat(feat=feat.crc.zeller,
    label=label.crc.zeller,
    meta=meta.crc.zeller)

# show siamcat
show(siamcat)

# filter feautures
siamcat <- filter.features(siamcat,
    filter.method = 'abundance',
    cutoff = 0.001)

```
```{r code box code from paper}

# conducts entire analysis
# feat = relaive abundance matrix
# meta = meta-variables about samples as a dataframe, containing column called 'disease' which encodes the label

sc.obj <- siamcat(feat=feat, meta=meta, label='disease')
sc.obj <- filter.features(sc.obj, filter.method = 'abundance')

# produces Fig. 1b
sc.obj <- check.associations(sc.obj, 
                             fn.plot = 'associations_plot.pdf')

# produces Fig. 1c
check.confounders(sc.obj,
                  fn.plot = 'confounder_plot.pdf') 
sc.obj <- normalize.features(sc.obj, norm.method = 'log.std') sc.obj <- create.data.split(sc.obj)
sc.obj <- train.model(sc.obj, method='lasso')
sc.obj <- make.predictions(sc.obj)
sc.obj <- evaluate.predictions(sc.obj)

# produces Fig. 1d 
model.evaluation.plot(sc.obj,
                      fn.plot = 'evaluation.pdf')

 # produces Fig. 1e
model.interpretation.plot(sc.obj,consens.thres = 0.8,
                          fn.plot = 'interpretation.pdf')

```

```{r check associations}
# The result of the check.associations function is an association plot. For significantly associated microbial features, the plot shows:
  # - the abundances of the features across the two different classes (CRC vs. controls)
  # - the significance of the enrichment calculated by a Wilcoxon test (after multiple hypothesis testing correction)
  # - the generalized fold change of each feature
  # - the prevalence shift between the two classes, and
  # - the Area Under the Receiver Operating Characteristics Curve (AU-ROC) as non-parametric effect size measure.

# check assosications 
siamcat <- check.associations(
      siamcat,
      sort.by = 'fc',
      alpha = 0.05,
      mult.corr = "fdr",
      detect.lim = 10 ^-6,
      plot.type = "quantile.box",
      panels = c("fc", "prevalence", "auroc"))
```

```{r check associations}
# normalise features
siamcat <- normalize.features(
    siamcat,
    norm.method = "log.unit",
    norm.param = list(
        log.n0 = 1e-06,
        n.p = 2,
        norm.margin = 1
    )
)

# data split
siamcat <-  create.data.split(
    siamcat,
    num.folds = 5,
    num.resample = 2
)

# train model
siamcat <- train.model(
    siamcat,
    method = "lasso"
)

# show models
# get information about the model type
model_type(siamcat)

# access the models
models <- models(siamcat)
models[[1]]

# make predictions ,  message=FALSE, results='hide'
siamcat <- make.predictions(siamcat)
pred_matrix <- pred_matrix(siamcat)

# pred_matrix_head
head(pred_matrix)

# evaluate predictions
siamcat <-  evaluate.predictions(siamcat)

# evaluate plot , fig.height=6, fig.width=6, fig.align='left', message=FALSE
model.evaluation.plot(siamcat)

# model interpretation plot

# After statistical models have been trained to distinguish cancer cases from controls, the models can be investigated by the function model.interpretation.plot. The plots shows:

  # - the median relative feature weight for selected features (barplot on the left)
  # - the robustness of features (i.e. in how many of the models the specific feature has been selected)
  # - the distribution of selected features across samples (central heatmap)
  # - which proportion of the weight of all different models are shown in the plot (boxplot on the right), and
  # - distribution of metadata across samples (heatmap below).

model.interpretation.plot(
      siamcat,
      fn.plot = 'interpretation.pdf',
      consens.thres = 0.5,
      norm.models = TRUE,
      limits = c(-3, 3),
      heatmap.type = 'zscore',
  )

```

```{r}

#### CHECK ASSOCIATIONS

#'@details For each feature, this function calculates different measures of
#'     association between the feature and the label. In detail, these
#'     associations are: \itemize{
#'     \item Significance as computed by a Wilcoxon test followed by multiple
#'     hypothesis testing correction.
#'     \item AUROC (Area Under the Receiver Operating Characteristics Curve)
#'     as a non-parameteric measure of enrichment (corresponds to the effect
#'     size of the Wilcoxon test).
#'     \item The generalized Fold Change (gFC) is a pseudo fold change
#'     which is calculated as geometric mean of the differences between the
#'     quantiles for the different classes found in the label.
#'     \item The prevalence shift between the two different classes found in
#'     the label.
#'     }
#'
#'     Finally, the function produces a plot of the top \code{max.show}
#'     associated features at a user-specified significance level \code{alpha},
#'     showing the distribution of the log10-transformed abundances for both
#'     classes, and user-selected panels for the effect (AU-ROC, Prevalence
#'     Shift, and Fold Change).

check.associations <- function(siamcat, fn.plot=NULL, color.scheme = "RdYlBu",
    alpha = 0.05, mult.corr = "fdr", sort.by = "fc", detect.lim = 1e-06,
    pr.cutoff = 1e-6, max.show = 50, plot.type = "quantile.box",
    panels = c("fc", "auroc"), prompt=TRUE, feature.type='filtered',
    verbose = 1) {

        if (verbose > 1)
            message("+ starting check.associations")
        s.time <- proc.time()[3]

        # check panel and plot.type parameter
        if (!all(panels %in% c("fc", "auroc", "prevalence"))) {
            stop("Unknown panel-type selected!")
        }
        panels <- unique(panels)
        if (length(panels) > 3) {
            warning(
                "Plot layout is not suited for more than 3 panels.
                Continuing with first three panels."
            )
            panels <- panels[seq_len(3)]
        }
        if ((!plot.type %in%
                c("bean", "box", "quantile.box", "quantile.rect")) ||
                length(plot.type) != 1) {
            warning("Plot type has not been specified properly! Continue with",
                    " quantile.box.")
            plot.type <- "quantile.box"
        }
        if (!feature.type %in% c('original', 'filtered', 'normalized')){
            stop("Unrecognised feature type, exiting...\n")
        }
        # get features
        if (feature.type == 'original'){
            feat <- get.orig_feat.matrix(siamcat)
        } else if (feature.type == 'filtered'){
            if (is.null(filt_feat(siamcat, verbose=0))){
                stop('Features have not yet been filtered, exiting...\n')
            }
            feat <- get.filt_feat.matrix(siamcat)
        } else if (feature.type == 'normalized'){
            if (is.null(norm_feat(siamcat, verbose=0))){
                stop('Features have not yet been normalized, exiting...\n')
            }
            feat <- get.norm_feat.matrix(siamcat)
        }

        if (any(is.na(feat))){
            stop('Features contain NAs. Exiting...')
        }
        if ((any(colSums(feat) > 1.01) | any(feat < -0.01)) &
            feature.type != 'normalized'){
            stop('This function expects compositional data. Exiting...')
        }
        # check label
        label <- label(siamcat)
        if (label$type == 'TEST'){
            stop('Can not check assocations for a',
            ' SIAMCAT object with TEST label! Exiting...')
        }
        # check fn.plot
        if (is.null(fn.plot)) {
            message(paste0('### WARNING: Not plotting to a pdf-file.\n',
                '### The plot is optimized for landscape DIN-A4 (or similar) ',
                'layout.\n### Please make sure that your plotting region is',
                ' large enough!!!\n### Use at your own risk...'))
            if (prompt == TRUE){
                continue <- askYesNo('Are you sure that you want to continue?',
                    default = TRUE,
                    prompts = getOption("askYesNo",
                        gettext(c("Yes", "No", "Cancel"))))
            } else {
                continue <- TRUE
            }
            if (!continue || is.na(continue)){
                opt <- options(show.error.messages = FALSE)
                on.exit(options(opt))
                stop('Exiting...')
            }
            par.old <- par(no.readonly=TRUE)
        }
        # either give n_classes colors or color palette
        col <- check.color.scheme(color.scheme, label)

        ### Calculate different effect sizes
        if (verbose > 2)
            message("+++ analysing features\n")
        probs.fc <- seq(.1, .9, .05)
        if (is.null(associations(siamcat, verbose=0))){
            result.list <- analyse.binary.marker(
                feat = feat,
                label = label,
                detect.lim = detect.lim,
                colors = col,
                pr.cutoff = pr.cutoff,
                mult.corr = mult.corr,
                alpha = alpha,
                probs.fc = probs.fc,
                take.log=ifelse(feature.type == 'normalized', FALSE, TRUE),
                verbose = verbose
            )
            # update siamcat
            associations(siamcat) <- list(
                assoc.results=result.list$effect.size,
                assoc.param=list(detect.lim=result.list$detect.lim,
                    pr.cutoff=pr.cutoff, probs.fc=probs.fc,
                    mult.corr=mult.corr, alpha=alpha,
                    feature.type=feature.type))
        } else {
            # if already existing, check parameters
            old.params <- assoc_param(siamcat)
            new.params <- list(detect.lim=detect.lim,
                pr.cutoff=pr.cutoff, probs.fc=probs.fc,
                mult.corr=mult.corr, alpha=alpha,
                feature.type=feature.type)
            # if the same, don't compute again but rather use the old resutls
            if (any(all.equal(new.params, old.params) == TRUE)) {
                result.list <- list()
                result.list$effect.size <- associations(siamcat)
                result.list$detect.lim <- assoc_param(siamcat)$detect.lim
            } else {
                result.list <- analyse.binary.marker(
                    feat = feat,
                    label = label,
                    detect.lim = detect.lim,
                    colors = col,
                    pr.cutoff = pr.cutoff,
                    mult.corr = mult.corr,
                    alpha = alpha,
                    probs.fc = probs.fc,
                    take.log=ifelse(feature.type == 'normalized', FALSE, TRUE),
                    verbose = verbose
                )
                # update siamcat
                associations(siamcat) <- list(
                    assoc.results=result.list$effect.size,
                    assoc.param=list(detect.lim=result.list$detect.lim,
                        pr.cutoff=pr.cutoff, probs.fc=probs.fc,
                        mult.corr=mult.corr, alpha=alpha,
                        feature.type=feature.type))
            }
        }

        ########################################################################
        # extract relevant info for plotting
        temp <- get.plotting.idx(result.list$effect.size, alpha=alpha,
            sort.by=sort.by, max.show=max.show, verbose=verbose)
        if (is.null(temp)){
            return(siamcat)
        }
        effect.size <- result.list$effect.size[temp$idx, , drop=FALSE]
        truncated <- temp$truncated
        detect.lim <- result.list$detect.lim
        feat.red    <- feat[temp$idx, , drop=FALSE]

        if (feature.type == 'normalized'){
            feat.plot <- feat.red
        } else {
            feat.red.log <- log10(feat.red + detect.lim)
            feat.plot <- feat.red.log
        }
        ########################################################################
        ### generate plots with significant associations between
        ##      features and labels

        # make plot matrix dependent on panels parameters
        if (verbose > 2)
            message("+++ preparing plotting layout")
        if (length(panels) == 3) {
            layout.mat <- cbind(2, 1, t(seq(3, length.out = length(panels))))
            widths <- c(0.5, 0.1, rep(0.4 / 3, length(panels)))
        } else {
            layout.mat <- cbind(2, 1, t(seq(3, length.out = length(panels))))
            widths <- c(0.5, 0.1, rep(0.2, length(panels)))
        }
        if (!is.null(fn.plot)) {
            pdf(fn.plot,
                paper = 'special',
                height = 8.27,
                width = 11.69) # format:A4 landscape
            }

        layout(mat = layout.mat, widths = widths)

        ########################################################################
        # PANEL 2: P-VALUES
        # print p-values in second panel of the plot
        associations.pvals.plot(p.vals = effect.size$p.adj,
            alpha = alpha, mult.corr = mult.corr,
            verbose = verbose)

        ########################################################################
        # PANEL 1: DATA
        # prepare margins
        associations.margins.plot(species_names = row.names(feat.red),
            verbose = verbose)

        if (verbose > 2)
            message("+++ plotting results")
        if (plot.type == "bean") {
            associations.bean.plot(feat.plot,
                label,
                col = col,
                take.log=ifelse(feature.type == 'normalized', FALSE, TRUE),
                verbose = verbose)
        } else if (plot.type == "box") {
            associations.box.plot(feat.plot,
                label,
                take.log=ifelse(feature.type == 'normalized', FALSE, TRUE),
                col = col,
                verbose = verbose)
        } else if (plot.type == "quantile.box") {
            associations.quantile.box.plot(feat.plot,
                label,
                col = col,
                take.log=ifelse(feature.type == 'normalized', FALSE, TRUE),
                verbose = verbose)
        } else if (plot.type == "quantile.rect") {
            associations.quantile.rect.plot(feat.plot,
                label,
                col = col,
                take.log=ifelse(feature.type == 'normalized', FALSE, TRUE),
                verbose = verbose)
        }

        # plot title
        xlab <- ifelse(feature.type=='normalized',
                    'Normalized abundance', 'Abundance (log10-scale)')
        if (!truncated) {
            title(main = 'Differentially abundant features', xlab = xlab)
        } else {
            title(
                main = paste(
                    'Differentially abundant features\nshowing top',
                    max.show,
                    'features'
                ), xlab = xlab)
        }

        ########################################################################
        # OTHER PANELS
        for (p in panels) {
            if (p == "fc") {
                associations.fcs.plot(
                    fc.all = effect.size$fc,
                    binary.cols = effect.size$bcol,
                    verbose = verbose
                )
            } else if (p == "prevalence") {
                associations.pr.shift.plot(
                    pr.shifts = effect.size[,c('pr.n', 'pr.p')],
                    col = col,
                    verbose = verbose
                )
            } else if (p == "auroc") {
                associations.aucs.plot(
                    aucs = effect.size[, c('auc', 'auc.ci.l', 'auc.ci.h')],
                    binary.cols = effect.size$bcol,
                    verbose = verbose
                )
            }
        }

        # close pdf device
        if (!is.null(fn.plot)) {
            tmp <- dev.off()
        } else {
            par(par.old)
        }
        e.time <- proc.time()[3]
        if (verbose > 1)
            message(paste(
                "+ finished check.associations in",
                formatC(e.time - s.time, digits = 3),
                "s"
            ))
        if (verbose == 1 & !is.null(fn.plot))
            message(paste(
                "Plotted associations between features and label",
                "successfully to:", fn.plot
            ))
        return(siamcat)
    }

# ##############################################################################
### AUC
#' @keywords internal
associations.aucs.plot <- function(aucs, binary.cols, verbose = 1) {
    if (verbose > 2)
        message("+ starting associations.aucs.plot")
    # set margins
    par(mar = c(5.1, 0, 4.1, 1.6))
    # plot background
    plot(
        NULL,
        xlab = '',
        ylab = '',
        xaxs = 'i',
        yaxs = 'i',
        axes = FALSE,
        xlim = c(0, 1),
        ylim = c(0.5, nrow(aucs) + 0.5),
        type = 'n'
    )
    ticks <- seq(0, 1.0, length.out = 5)
    tick.labels <- formatC(ticks, digits = 2)
    # plot gridlines
    for (v in ticks) {
        abline(v = v,
            lty = 3,
            col = 'lightgrey')
    }
    # make thicker line at .5
    abline(v = .5, lty = 1, col = 'lightgrey')
    # plot single feature aucs
    for (i in seq_len(nrow(aucs))) {
        segments(
            x0 = aucs[i, 2],
            x1 = aucs[i, 3],
            y0 = i,
            col = 'lightgrey',
            lwd = 1.5
        )
        points(aucs[i, 1], i, pch = 18, col = binary.cols[i])
        points(aucs[i, 1],
            i,
            pch = 5,
            col = 'black',
            cex = 0.9)
    }

    # Title and axis label
    axis(
        side = 1,
        at = ticks,
        labels = tick.labels,
        cex.axis = 0.7
    )
    title(main = 'Feature AUCs', xlab = 'AU-ROC')
    if (verbose > 2)
        message("+ finished associations.aucs.plot")
}

# ##############################################################################
### FC
#' @keywords internal
associations.fcs.plot <-
    function(fc.all, binary.cols,    verbose = 1) {
        if (verbose > 2)
            message("+ starting associations.fcs.plot")
        # margins
        par(mar = c(5.1, 0, 4.1, 1.6))
        # get minimum and maximum fcs
        mx <- max(ceiling(abs(
            range(fc.all, na.rm = TRUE, finite = TRUE)
        )))
        mn <- -mx
        # plot background
        plot(
            NULL,
            xlab = '',
            ylab = '',
            xaxs = 'i',
            yaxs = 'i',
            axes = FALSE,
            xlim = c(mn, mx),
            ylim = c(0.2, length(fc.all) + 0.2),
            type = 'n'
        )
        grid(NULL, NA, lty = 3, col = 'lightgrey')
        # plot bars
        barplot(
            fc.all,
            horiz = TRUE,
            width = 0.6,
            space = 2 / 3,
            col = binary.cols,
            axes = FALSE,
            add = TRUE,
            names.arg = FALSE
        )
        # gridlines and axes labels
        ticks <- seq(from = mn,
            to = mx,
            length.out = 5)
        tick.labels <- formatC(ticks, digits = 2)
        axis(
            side = 1,
            at = ticks,
            labels = tick.labels,
            cex.axis = 0.7
        )
        title(main = 'Fold change', xlab = 'Generalized fold change')
        if (verbose > 2)
            message("+ finished associations.fcs.plot")
    }

# ##############################################################################
### PREVALENCE
#' @keywords internal
associations.pr.shift.plot <-
    function(pr.shifts, col, verbose = 1) {
        if (verbose > 2)
            message("+ starting associations.pr.shift.plot")
        # margins
        par(mar = c(5.1, 0, 4.1, 1.6))

        # plot background
        plot(
            NULL,
            xlab = '',
            ylab = '',
            xaxs = 'i',
            yaxs = 'i',
            axes = FALSE,
            xlim = c(0, 1),
            ylim = c(0.2, nrow(pr.shifts) + 0.2),
            type = 'n'
        )
        # gridlines and axes labels
        ticks <- seq(from = 0,
            to = 1,
            length.out = 5)
        for (v in ticks) {
            abline(v = v,
                lty = 3,
                col = 'lightgrey')
        }
        tick.labels <- formatC(ticks * 100, digits = 3)
        axis(
            side = 1,
            at = ticks,
            labels = tick.labels,
            cex.axis = 0.7
        )
        # plot bars
        row.names(pr.shifts) <- NULL
        barplot(
            t(pr.shifts),
            horiz = TRUE,
            axes = FALSE,
            add = TRUE,
            space = c(0, 4 / 3),
            beside = TRUE,
            width = .3,
            col = c(col[1], col[2])
        )
        title(main = 'Prevalence shift', xlab = 'Prevalence [%]')
        if (verbose > 2)
            message("+ finished associations.pr.shift.plot")
    }

# ##############################################################################
# P-VALUES
#' @keywords internal
associations.pvals.plot <- function(p.vals, alpha,  mult.corr,  verbose = 1) {
    if (verbose > 2)
        message("+ starting associations.pvals.plot")
    # margins
    par(mar = c(5.1, .0, 4.1, 1.6))
    p.vals.log <- -log10(p.vals)
    # get minimum and maximum
    mx <-
        max(ceiling(abs(
            range(p.vals.log, na.rm = TRUE, finite = TRUE)
        )))
    mn <- 0
    p.vals.log[is.infinite(p.vals.log)] <- mx
    # plot background
    plot(
        NULL,
        xlab = '',
        ylab = '',
        xaxs = 'i',
        yaxs = 'i',
        axes = FALSE,
        xlim = c(mn, mx),
        ylim = c(0.2, length(p.vals) + 0.2),
        type = 'n'
    )
    grid(NULL, NA, lty = 3, col = 'lightgrey')
    # plot bars
    barplot(
        p.vals.log,
        horiz = TRUE,
        width = 0.6,
        space = 2 / 3,
        col = 'lightgrey',
        axes = FALSE,
        add = TRUE,
        names.arg = FALSE
    )
    # gridlines and axes labels
    ticks <- seq(from = mn, to = mx)
    abline(v = -log10(alpha),
        lty = 1,
        col = 'red')
    tick.labels <- formatC(ticks, digits = 2)
    axis(
        side = 1,
        at = ticks,
        labels = tick.labels,
        cex.axis = 0.7
    )
    if (mult.corr != 'none'){
        title(main = 'Significance', xlab = '-log10(adj. p value)')
    } else {
        title(main='Significance', xlab='-log10(p value)')
    }

    if (verbose > 2)
        message("+ finished associations.pvals.plot")
}

# ##############################################################################
# COLOR
# check if a string is a valid r color reprensentation
# from stackoverflow: Check if character string is a valid color representation
# https://stackoverflow.com/questions/13289009
#' @keywords internal
is.color <- function(x) {
    vapply(
        x,
        FUN = function(z) {
            tryCatch(
                is.matrix(col2rgb(z)),
                error = function(e)
                    FALSE
            )
        },
        FUN.VALUE = logical(1)
    )
}

### check the user-supplied color scheme for validity
### color scheme may either be a single RColorBrewer palette or a vector of
### the same length as the number of classes containing interpretable colors
### as strings
#' @keywords internal
check.color.scheme <- function(color.scheme, label, verbose = 1) {
    if (verbose > 2)
        message("+ starting check.color.scheme")
    n.classes = ifelse(label$type == 'BINARY', 2,
        length(unique(label$label)))

    if (length(color.scheme) == 1 &&
            is.character(color.scheme)) {
        if (n.classes == 2) {
    # if color scheme and binary label, make colors as before
            if (!color.scheme %in% row.names(brewer.pal.info)) {
                warning(
                    "Not a valid RColorBrewer palette name, defaulting to
                    RdBu.\n See brewer.pal.info for more information about
                    RColorBrewer palettes."
                )
                color.scheme <- 'RdYlBu'
            }
            colors <-
                rev(colorRampPalette(brewer.pal(
                    brewer.pal.info[color.scheme,
                        'maxcolors'], color.scheme
                ))(2))
        } else {
    # if color scheme and multiclass label, make colors either directly out
    # of the palette (if n.classes smaller than maxcolors) or like before
            if (!color.scheme %in% row.names(brewer.pal.info)) {
                warning(
                    "Not a valid RColorBrewer palette name, defaulting to
                    Set3.\n See brewer.pal.info for more information about
                    RColorBrewer palettes."
                )
                color.scheme <- 'Set3'
            }
    # if color scheme and multiclass label, check that the palette is not
    # divergent or sequential, but qualitative. Only issue warning.
            if (brewer.pal.info[color.scheme, 'category'] != 'qual')
                warning("Using a divergent or sequential color palette for
                        multiclass data.")
            if (n.classes <= brewer.pal.info[color.scheme, 'maxcolors']) {
                colors <- brewer.pal(n.classes, color.scheme)
            } else {
                warning("The data contains more classes than the color.palette
                    provides.")
                colors <-
                    rev(colorRampPalette(brewer.pal(
                        brewer.pal.info[color.scheme,
                            'maxcolors'], color.scheme
                    ))(n.classes))
            }
        }
        } else if (length(color.scheme == n.classes) &&
                all(is.color(color.scheme))) {
    # if colors, check that all strings are real colors and check that
    # the same length as n classes
    # convert color names to hex representation
            colors <-
                vapply(
                    color.scheme,
                    FUN = function(x) {
                        rgb(t(col2rgb(x)),
                            maxColorValue = 255)
                    },
                    FUN.VALUE = character(1),
                    USE.NAMES = FALSE
                )
        } else {
            stop("Supplied colors do not match the number of classes or are no
                valid colors")
        }
    # add transparency
    colors <- vapply(
        colors,
        FUN = function(x) {
            paste0(x, '85')
        },
        FUN.VALUE = character(1),
        USE.NAMES = FALSE
    )
    if (verbose > 2)
        message("+ finished check.color.scheme")
    return(colors)
        }

#' @keywords internal
create.tints <- function(colour, vec) {
    new.cols <-
        vapply(
            vec,
            FUN = function(x) {
                rgb(matrix(col2rgb(colour) / 255 +
                        (1 - col2rgb(colour) / 255) * x, ncol = 3))
            },
            FUN.VALUE = character(1)
        )
    return(new.cols)
}

#' @keywords internal
change.transparency <- function(col.name) {
    if (nchar(col.name) > 7) {
        # adjust alpha channel by reducing transparency
        a = substr(col.name, nchar(col.name) - 1, nchar(col.name))
        a = 1 - (1 - as.numeric(paste('0x', a, sep = '')) / 255) / 2
        new.col = gsub('..$', toupper(as.hexmode(round(a * 255))), col.name)
    } else {
        new.col <- col.name
    }
    return(new.col)
}

# ##############################################################################
# UTILITY FUNCTIONS
### Prepare margins for the first plots make left margin as big as the
### longest label or maximally 20.1 lines
#' @keywords internal
associations.margins.plot <-
    function(species_names, p.label, verbose = 1) {
        if (verbose > 2)
            message("+ starting associations.margins.plot")
        cex.org <- par()$cex
        par(mar = c(5.1, 18, 4.1, 1.1), cex = 1)
        temp = par()$mai
        cex.labels <- min(.7, (((
            par()$pin[2] / length(species_names)
        ) * .6) /
                max(
                    strheight(species_names, units = 'inches')
                )))
        max_name <- max(strwidth(species_names, units = 'inches',
            cex = cex.labels)) + temp[4]
        temp[2] <- min(temp[2], max_name)
        par(mai = temp, cex = cex.org)
        if (verbose > 2)
            message("+ finished associations.margins.plot")
    }

#' @keywords internal
associations.labels.plot <-
    function(labels, plot.type,    verbose = 1) {
        if (verbose > 2)
            message("+ starting associations.labels.plot")
        adj <- rep(0, length(labels))
        if (plot.type == 'quantile.rect')
            adj <- rep(-0.5, length(labels))
        if (plot.type == 'box')
            adj <- -0.5 + seq_along(labels)
        cex.org <- par()$cex
        par(cex = 1)
        cex.labels <- min(.7, (((
            par()$pin[2] / length(labels)
        ) * .6) /
                max(strheight(labels, units = 'inches'))))
        for (i in seq_along(labels)) {
            mtext(
                labels[i],
                2,
                line = 0,
                at = i + adj[i],
                las = 1,
                cex = cex.labels
            )
        }
        par(cex = cex.org)
        if (verbose > 2)
            message("+ finished associations.labels.plot")
    }

#' @keywords internal
associations.quantiles.plot <- function(quantiles, up = TRUE, col) {
    n.spec <- nrow(quantiles)
    adj.y0 <- ifelse(up, 0, 0.3)
    adj.y1 <- ifelse(up, 0.3, 0)
    #  box
    rect(quantiles[, 2],
        seq_len(n.spec) - adj.y0,
        quantiles[, 4],
        seq_len(n.spec) + adj.y1,
        col = col)
        # 90% interval
    segments(quantiles[, 1], seq_len(n.spec), quantiles[, 5], seq_len(n.spec))
    segments(
        quantiles[, 1],
        y0 = seq_len(n.spec) - adj.y0 / 3 * 2,
        y1 = seq_len(n.spec) + adj.y1 / 3 * 2
    )
    segments(
        quantiles[, 5],
        y0 = seq_len(n.spec) - adj.y0 / 3 * 2,
        y1 = seq_len(n.spec) + adj.y1 / 3 * 2
    )
    # median
    segments(
        quantiles[, 3],
        y0 = seq_len(n.spec) - adj.y0,
        y1 = seq_len(n.spec) + adj.y1,
        lwd = 3
    )
}

# ##############################################################################
# BEAN PLOT
#' @keywords internal
associations.bean.plot <-
    function(data.mat, label, col, take.log=TRUE, verbose = 1) {
        if (verbose > 2)
            message("+ starting associations.bean.plot")

        p.label <- max(label$info)
        n.label <- min(label$info)

        # create data.frame in format for beanplot
        bean.data <- data.frame(data = c(data.mat))
        bean.data$factor <- c(vapply(
            label$label,
            FUN = function(x) {
                paste(rownames(data.mat),
                names(label$info)[match(x, label$info)])
            },
            FUN.VALUE = character(nrow(data.mat)),
            USE.NAMES = FALSE
        ))
        # ensure correct ordering by converting to a factor
        bean.data$factor <- factor(bean.data$factor,
            levels = paste(rep(rownames(data.mat), each = 2),
                            names(label$info[order(label$info)])))

        mn <- floor(c(min(bean.data$data)))
        mx <- ceiling(c(max(bean.data$data)))

        plot(
            NULL,
            xlab = '',
            ylab = '',
            xaxs = 'i',
            yaxs = 'i',
            axes = FALSE,
            xlim = c(mn - 1.5, mx + 1),
            ylim = c(0.45, nrow(data.mat) + 0.6),
            type = 'n'
        )
        ticks <- mn:mx
        for (v in ticks) {
            abline(v = v,
                lty = 3,
                col = 'lightgrey')
        }
        if (take.log){
            tick.labels <- formatC(10 ^ ticks, format = 'E', digits = 0)
            axis(side = 1,at = ticks,labels = tick.labels,cex.axis = 0.7)
        } else {
            axis(1, ticks, cex.axis=0.7)
        }

        beanplot(
            data ~ factor,
            data = bean.data,
            side = "both",
            bw = "nrd0",
            col = list(col[1], col[2]),
            horizontal = TRUE,
            names = c(""),
            show.names = FALSE,
            beanlines = "median",
            maxstripline = 0.2,
            what = c(FALSE, TRUE, TRUE, FALSE),
            axes = FALSE,
            add = TRUE
        )

        legend(
            'topright',
            legend = c(names(which(label$info == p.label)),
                names(which(label$info == n.label))),
            fill = rev(col),
            bty = 'n'
        )
        associations.labels.plot(rownames(data.mat),
            plot.type = 'bean',
            verbose = verbose)
        if (verbose > 2)
            message("+ finished associations.bean.plot")
    }

# ##############################################################################
# BOX PLOT
#' @keywords internal
associations.box.plot <-
    function(data.mat, label, col, take.log=TRUE, verbose = 1) {
        if (verbose > 2)
            message("+ starting associations.box.plot")
        box.colors <- rep(c(col[1], col[2]), nrow(data.mat))

        p.label <- max(label$info)
        n.label <- min(label$info)

        # create data.frame in format for beanplot
        plot.data <- data.frame(data = c(data.mat))
        plot.data$factor <- c(vapply(
            label$label,
            FUN = function(x) {
                paste(rownames(data.mat),
                names(label$info)[match(x, label$info)])
            },
            FUN.VALUE = character(nrow(data.mat)),
            USE.NAMES = FALSE
        ))

        # ensure correct ordering by converting to a factor
        plot.data$factor <- factor(plot.data$factor,
            levels = paste(rep(rownames(data.mat), each = 2),
                            names(label$info[order(label$info)])))

        mn <- floor(c(min(data.mat)))
        mx <- ceiling(c(max(data.mat)))

        plot(
            NULL,
            xlab = '',
            ylab = '',
            xaxs = 'i',
            yaxs = 'i',
            axes = FALSE,
            xlim = c(mn - 0.2, mx + 1),
            ylim = c(+0.5, nrow(data.mat) * 2 + 0.5),
            type = 'n'
        )
        ticks <- mn:mx
        for (v in ticks) {
            abline(v = v,
                lty = 3,
                col = 'lightgrey')
        }
        boxplot(
            data ~ factor,
            data = plot.data,
            horizontal = TRUE,
            names = c(""),
            show.names = FALSE,
            col = box.colors,
            axes = FALSE,
            outcol = c(col[1], col[2]),
            add = TRUE
        )

        if (take.log){
            tick.labels <- formatC(10 ^ ticks, format = 'E', digits = 0)
            axis(side = 1,at = ticks,labels = tick.labels,cex.axis = 0.7)
        } else {
            axis(1, ticks, cex.axis=0.7)
        }
        legend(
            'topright',
            legend = c(names(which(label$info == p.label)),
                names(which(label$info == n.label))),
            fill = rev(col),
            bty = 'n'
        )
        associations.labels.plot(row.names(data.mat),
            plot.type = 'box',
            verbose = verbose)
        if (verbose > 2)
            message("+ finished associations.box.plot")
    }

# ##############################################################################
# QUANTILE BOX PLOT
#' @keywords internal
associations.quantile.box.plot <- function(data.mat, label, take.log=TRUE, col,
    verbose = 1) {
    if (verbose > 2)
        message("+ starting associations.quantile.box.plot")
    pos.col <- col[2]
    neg.col <- col[1]

    p.label <- max(label$info)
    n.label <- min(label$info)
    p.idx <- which(label$label == p.label)
    n.idx <- which(label$label == n.label)
    p.n <- length(which(label$label == p.label))
    n.n <- length(which(label$label == n.label))

    n.spec <- nrow(data.mat)
    if (take.log){
        p.min <- floor(min(data.mat, na.rm = TRUE))
        p.max <- 0
    } else {
        p.min <- floor(min(data.mat, na.rm = TRUE))
        p.max <- ceiling(max(data.mat, na.rm = TRUE))
    }

    plot(
        rep(p.min, n.spec),
        seq_len(n.spec),
        xlab = '',
        ylab = '',
        yaxs = 'i',
        axes = FALSE,
        xlim = c(p.min, p.max),
        ylim = c(0.5, n.spec + 0.5),
        frame.plot = FALSE,
        type = 'n'
    )
    for (v in seq(p.min, p.max, 1)) {
        abline(v = v,
            lty = 3,
            col = 'lightgrey')
    }

    tck = p.min:p.max
    if (take.log){
        axis(1, tck, formatC(10 ^ tck, format='E', digits=0),
            las=1, cex.axis=0.7)
    } else {
        axis(1, tck, las=1, cex.axis=0.7)
    }

    # get quantiles
    quant.probs <- c(0.05, 0.25, 0.5, 0.75, 0.95)
    quantiles.pos = rowQuantiles(data.mat[, p.idx, drop=FALSE],
        probs = quant.probs, na.rm = TRUE, drop=FALSE)
    quantiles.neg = rowQuantiles(data.mat[, n.idx, drop=FALSE],
        probs = quant.probs, na.rm = TRUE, drop=FALSE)

        # inter-quartile range
    associations.quantiles.plot(quantiles.pos, up = TRUE, pos.col)
    associations.quantiles.plot(quantiles.neg, up = FALSE, neg.col)

    # scatter plot on top
    for (i in seq_len(n.spec)) {
        pos.col.t <- change.transparency(pos.col)
        neg.col.t <- change.transparency(neg.col)

        points(
            data.mat[i, p.idx],
            rep(i + 0.15, p.n) + rnorm(p.n, sd = 0.03),
            pch = 16,
            cex = 0.6,
            col = pos.col.t
        )
        points(
            data.mat[i, n.idx],
            rep(i - 0.15, n.n) + rnorm(n.n, sd = 0.03),
            pch = 16,
            cex = 0.6,
            col = neg.col.t
        )
    }
    legend(
        'topright',
        legend = c(names(which(label$info == p.label)),
            names(which(label$info == n.label))),
        fill = rev(col),
        bty = 'n'
    )
    associations.labels.plot(row.names(data.mat),
        plot.type = 'quantile.box',
        verbose = verbose)
    if (verbose > 2)
        message("+ finished associations.quantile.box.plot")
}

# ##############################################################################
# QUANTILE RECT PLOT
#' @keywords internal
associations.quantile.rect.plot <-
    function(data.mat, label, col, take.log=TRUE, verbose = 1) {
        if (verbose > 2)
            message("+ starting associations.quantile.rect.plot")
        n.spec <- nrow(data.mat)
        quant.probs <- seq(from = 0.1, to = 0.9, by = 0.1)

        p.label <- max(label$info)
        n.label <- min(label$info)
        p.idx <- which(label$label == p.label)
        n.idx <- which(label$label == n.label)

        quantiles.pos = rowQuantiles(data.mat[, p.idx, drop=FALSE],
                                        probs = quant.probs,
                                        na.rm = TRUE, drop=FALSE)
        quantiles.neg = rowQuantiles(data.mat[, n.idx, drop=FALSE],
                                        probs = quant.probs,
                                        na.rm = TRUE, drop=FALSE)

        if (take.log){
            p.min <- floor(min(data.mat, na.rm = TRUE))
            p.max <- 0
        } else {
            p.min <- floor(min(data.mat, na.rm = TRUE))
            p.max <- ceiling(max(data.mat, na.rm = TRUE))
        }

        plot(
            rep(p.min, n.spec),
            seq_len(n.spec),
            xlab = '',
            ylab = '',
            yaxs = 'i',
            axes = FALSE,
            xlim = c(p.min, p.max),
            ylim = c(0, n.spec),
            frame.plot = FALSE,
            type = 'n'
        )
        for (v in seq(p.min, p.max, 1)) {
            abline(v = v,
                lty = 3,
                col = 'lightgrey')
        }

        tck = p.min:p.max
        if (take.log){
            axis(1, tck, formatC(10^tck, format='E', digits=0),
                las=1, cex.axis=0.7)
        } else {
            axis(1, tck, las=1, cex.axis=0.7)
        }

        # create different tints of the colours
        colors.p <-
            rev(create.tints(vec = seq(0, 1, length.out = 4),
            colour = col[2]))
        colors.n <-
            rev(create.tints(vec = seq(0, 1, length.out = 4),
            colour = col[1]))

        associations.quantile.rect.sub.plot(quantiles.pos, up = TRUE, colors.p)
        associations.quantile.rect.sub.plot(quantiles.neg, up = FALSE, colors.n)
        associations.quantile.median.sub.plot(quantiles.pos, up = TRUE)
        associations.quantile.median.sub.plot(quantiles.neg, up = FALSE)

        legend(0.3*p.min, n.spec,
            legend = c("Quantiles", "40%-60%", "30%-70%", "20%-80%", "10%-90%",
                "median", "", "", "", "", ""),
            bty = 'n', cex = 1, fill = c('white', rev(colors.p), 'white',
                'white', rev(colors.n), 'white'),
            lwd = 1.3, ncol = 2, border = c("white", "black", "black",
                "black", "black", "white", "white", "black", "black", "black",
                "black", "white"))
        legend(0.3*p.min + abs(0.016*p.min), n.spec,
            legend = c("", "", "", "", "", ""), bty = 'n',
            lty = c(0, 0, 0, 0, 0, 0),
            # cap legend size for diamond (should look
            #   symmetric to other symbols)
            pch = 18, cex = 1,
            pt.cex = c(0, 0, 0, 0, 0, min(35 / n.spec, 2.25)))
        legend('bottomright',
            legend = c(names(which(label$info == max(label$info))),
                names(which(label$info == min(label$info)))),
            fill = rev(col), bty = 'n')
        associations.labels.plot(rownames(data.mat),
            plot.type = 'quantile.rect',
            verbose = verbose)
        if (verbose > 2)
            message("+ finished associations.quantile.rect.plot")
}

#' @keywords internal
associations.quantile.median.sub.plot <-
    function(quantiles, up = TRUE) {
        n.spec <- nrow(quantiles)
        adj.y <- ifelse(up, 0.15,-0.15)
        points(
            quantiles[, ceiling(ncol(quantiles) / 2)],
            y = (0.5:n.spec) + adj.y,
            pch = 18,
            cex = min(35 / n.spec, 4)
        )
    }

#' @keywords internal
associations.quantile.rect.sub.plot <-
    function(quantiles, up = TRUE, colors) {
        n.spec <- nrow(quantiles)
        adj.y0 <- ifelse(up, 0, 0.3)
        adj.y1 <- ifelse(up, 0.3, 0)
        for (i in seq_len(ncol(quantiles) / 2)) {
            rect(
                quantiles[, i],
                (0.5:n.spec) - adj.y0,
                quantiles[, ncol(quantiles) + 1 - i],
                (0.5:n.spec) + adj.y1,
                col = colors[i],
                border = c("black"),
                lwd = 0.9
            )
        }
}

# ##############################################################################
### maker analysis for two-class data
#     calculate p-value with Wilcoxon
#     fold change as normalized absolute difference between quantiles
#     prevalence shift
#     single marker AUC
#' @keywords internal
analyse.binary.marker <- function(feat, label, detect.lim, colors,
    pr.cutoff, mult.corr, alpha, max.show, sort.by, probs.fc = seq(.1, .9, .05),
    take.log=TRUE, verbose = 1) {
    if (verbose > 1)
        message("+ starting analyse.binary.marker")
    s.time <- proc.time()[3]
    ############################################################################
    ### Calculate wilcoxon, pseudo-FC, prevalence shift, and AUC for all feats
    ############################################################################
    if (verbose > 1)
        message('+++ calculating effect size for each feature.')
    if (is.null(detect.lim) & take.log==TRUE) {
        warning(
            "Pseudo-count before log-transformation not supplied! Estimating it
            as 5% percentile.\n"
        )
        detect.lim <- quantile(feat[feat != 0], 0.05)
    }
    if (any(feat[feat != 0] < detect.lim) & take.log==TRUE){
        cnt <- length(which(feat[feat!=0] < detect.lim))
        percentage <- (cnt/length(feat[feat!=0]))*100
        if (percentage >= 5){
            warning(paste0('### Some values (',cnt, ' or ',
            formatC(percentage, digits=2),
            '% of non-zero entries',
            ') are smaller than the given detection limit!'))
        }
    }

    positive.label <- max(label$info)
    negative.label <- min(label$info)

    if (verbose)
        pb <- progress_bar$new(total = nrow(feat))

    effect.size <- data.frame(t(apply(feat, 1, FUN = function(x) {
        # pseudo-fold change as differential quantile area
        if (take.log == TRUE){
            q.p <- quantile(log10(x[which(label$label == positive.label)] +
                detect.lim), probs = probs.fc)
            q.n <- quantile(log10(x[which(label$label == negative.label)] +
                detect.lim), probs = probs.fc)
        } else {
            q.p <- quantile(x[which(label$label == positive.label)],
                probs = probs.fc)
            q.n <- quantile(x[which(label$label == negative.label)],
                probs = probs.fc)
        }
        fc <- sum(q.p - q.n) / length(q.p)

        # wilcoxon
        p.val <- wilcox.test(x[which(label$label == negative.label)],
            x[which(label$label == positive.label)], exact = FALSE)$p.value

        # AU-ROC
        temp <- roc(predictor = x, response = label$label, ci = TRUE,
                    direction = '<', levels = label$info)
        aucs <- c(temp$ci)

        # prevalence shift
        temp.n <- sum(x[which(label$label == negative.label)] >= pr.cutoff) /
            length(which(label$label == negative.label))
        temp.p <- sum(x[which(label$label == positive.label)] >= pr.cutoff) /
            length(which(label$label == positive.label))
        pr.shift <- c(temp.p - temp.n, temp.n, temp.p)
        if (verbose)
            pb$tick()
        return(c('fc' = fc, 'p.val' = p.val, 'auc' = aucs[2],
            'auc.ci.l' = aucs[1], 'auc.ci.h' = aucs[3],
            'pr.shift' = pr.shift[1], 'pr.n' = pr.shift[2],
            'pr.p' = pr.shift[3]))
        }
    )))

    effect.size$bcol <-
        ifelse(effect.size[, 'auc'] >= 0.5, colors[2], colors[1])

    ### Apply multi-hypothesis testing correction
    if (!tolower(mult.corr) %in% c("holm", "hochberg", "hommel", "bonferroni",
        "BH", "BY", "fdr", "none")) {
        stop(
            "! Unknown multiple testing correction method:', mult.corr,'
            Stopping!\n Must of one of c('none','bonferroni', 'holm','fdr',
            'bhy')"
        )
    }
    if (mult.corr == 'none') {
        warning('WARNING: No multiple hypothesis testing performed.')
        effect.size$p.adj <- effect.size$p.val
    } else {
        effect.size$p.adj <-
            p.adjust(effect.size$p.val, method = tolower(mult.corr))
    }

    if (verbose > 1)
        message(
            paste(
                '+++ found',
                sum(effect.size$p.adj < alpha,
                    na.rm = TRUE),
                'significant associations at a significance level <',
                alpha
            )
        )

    e.time <- proc.time()[3]
    if (verbose > 1)
        message(paste(
            "+ finished analyse.binary.marker in",
            formatC(e.time - s.time, digits = 3),
            "s"
        ))
    return(
        list(
            "effect.size" = effect.size,
            "detect.lim" = detect.lim
        )
    )
}

#' @keywords internal
get.plotting.idx <- function(df.results, alpha, sort.by, max.show, verbose){

    idx <- which(df.results$p.adj < alpha)

    if (length(idx) == 0) {
        warning(paste0('No significant associations found.',
        ' No plot will be produced.\n'))
        return(NULL)
    } else if (length(idx) < 5) {
        warning(paste0('Less than 5 associations found. Consider',
        ' changing your alpha value.'))
    }

    idx <- idx[order(df.results$p.adj[idx], decreasing = TRUE)]

    # # truncated the list for the following plots
    truncated = FALSE
    if (length(idx) >= max.show) {
        truncated = TRUE
        idx <- idx[(length(idx) - max.show + 1):length(idx)]
        if (verbose > 1)
            message(
                paste(
                    '+++ truncating the list of significant
                    associations to the top',
                    max.show
                )
            )
    }

    ### Sort features
    if (verbose > 2)
        message('+++ sorting features')
    if (!sort.by %in% c('fc', 'p.val', 'pr.shift', 'auc')) {
            message(paste0(
                '+++ Unknown sorting option: ',
                sort.by,
                '. Instead order by fold change.'
            ))
        sort.by <- 'fc'
    }
    if (sort.by == 'fc') {
        idx <- idx[order(df.results$fc[idx], decreasing = FALSE)]
    } else if (sort.by == 'p.val') {
        idx <- idx[order(df.results$p.adj[idx], decreasing = TRUE)]
    } else if (sort.by == 'pr.shift') {
        idx <- idx[order(df.results$pr.shift[idx], decreasing = FALSE)]
    } else if (sort.by == 'auc'){
        idx <- idx[order(df.results$auc[idx], decreasing = FALSE)]
    }
    return(list('idx'=idx,
                'truncated'=truncated))
}

```

```{r}

check.associations(siamcat)

```
```{r}

```

##### ATTEMPT WITH MND ######

```{r}

library(ExperimentHub)
library(curatedMetagenomicData)
suppressPackageStartupMessages(library(ExperimentHub))
library(phyloseq)
library(microbiome)

eh = ExperimentHub()
myquery = query(eh, "curatedMetagenomicData")

# Get the stool samples
# esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)
meta.healthy <- combined_metadata %>% filter(disease=='healthy')

# merge
# eset <- mergeData(esl)

# the relab=FALSE is essential as it Absolute Raw Count Data
#pseq = ExpressionSet2phyloseq(meta.healthy, relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
#orig_pseq <- pseq

c.mnd.country <- get_variable(meta.healthy, "country" )
table(c.mnd.country)

# Create mnd variable
#pseq <- orig_pseq
# grabs the nationality from phyloseq
#c.mnd.country <- get_variable(pseq, "country" )
#table(c.mnd.country)

# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
# Not used BGD Bangladesh , CAN Canada, CHN, China. DNK denmark, FJI Figi, HUN Hungry, ISL isle of man , MDG, MNG mongolia, PER Iran, RUS RUSSIA,

#Adjsuted to match exactly the ATLAS data set

#central Europe = Belgium, Demark, Netherlands, Germany
#southern Europe = France, Italy, Serbia, Spain
#UKIE
#Scandinavia = Norway, Sweden Finland
#Eastern Europe = Poland

# NOR, GBR not in healthy 
# NOR in Scandanavia

# c.mnd.atlas <- fct_collapse(c.mnd.country, Scandinavia = c("NOR","SWE","FIN") ,CentralEurope = c("DNK","DEU","NLD"), SouthEurope = c("ITA","ESP","FRA"), US = "USA", UKIE = "GBR" )

# c.mnd.atlas <- droplevels(c.mnd.atlas,c("BGD", "ISL","HUN","EST", "SVK", "LUX", "CAN", "AUT", "FJI","CHN","MDG","MNG","PER","RUS","TZA", "KAZ"))

c.mnd.atlas <- fct_collapse(c.mnd.country, Scandinavia = c("SWE","FIN") ,CentralEurope = c("DNK","DEU","NLD"), SouthEurope = c("ITA","ESP","FRA"), US = "USA")

#drops unused levels
c.mnd.atlas <- droplevels(c.mnd.atlas, c("BGD", "CAN", "CHN", "FJI", "KAZ", "MDG", "PER", "RUS", "TZA"))

# c.mnd.atlas <- subset_samples(c.mnd.atlas, !is.na(country) )

sample_data(meta.healthy)$country = c.mnd.atlas

#mnd.country <- fct_collapse(mnd.country, LOW = c("NOR","SWE","FIN","DNK","LUX") , MEDIUM = c("ITA","AUT","DEU","ESP","FRA","NLD"), HIGH = c("GBR","USA","CAN") )

#mnd.country <- fct_collapse(mnd.country, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

# reorder
# NOTE THERE IS NO EASTERN EUROPE
c.mnd.nationality <- fct_collapse(c.mnd.atlas, LOW = c("Scandinavia") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US") )

c.mnd.nationality <- factor(c.mnd.nationality, levels = (c("LOW", "MEDIUM", "HIGH")))
levels(c.mnd.nationality)

# creates a new variable in the phyloseq called mnd
#sample_data(meta.healthy)$mnd = c.mnd.nationality
meta.healthy$mnd = c.mnd.nationality

# only works with phyloseq
# checks that it has worked.
get_variable(meta.healthy, "mnd")
table(get_variable(meta.healthy, "mnd"))

```

However, we have to clean the data a bit, since the data set contains samples from two different countries and two different disease subtypes.
```{r}

table(meta.healthy$country)

table(meta.healthy$mnd)

```

```{r}

# The siamcat object extends on the phyloseq object. Therefore, creating a siamcat object from a phyloseq object is really straightforward. This can be done with the siamcat constructor function. First, however, we need to create a label object:
label <- create.label(meta=sample_data(meta.healthy),
    label = "mnd",
    case = c("MEDIUM", "HIGH"), control = "LOW")

# Assigned LOW as control group

```

we have more samples than individual subjects:
```{r}

print(length(unique(meta.healthy$subjectID))) 
print(nrow(meta.healthy))

```

Some subjects (but not all) had been sampled multiple times. Therefore, we want to remove repeated samplings for the same subject, since the samples would otherwise not be indepdenent from another.

The visit number is encoded in the sampleID. Therefore, we can use this information to extract when the samples have been taken and use only the first visit for each subject.

```{r}

meta.healthy.s<- meta.healthy %>%
  select(sampleID, subjectID, study_condition, disease_subtype, 
disease, age, country, number_reads, median_read_length, BMI, mnd) %>%
  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
  mutate(visit=str_remove(visit, '_')) %>% 
  mutate(visit=as.numeric(visit)) %>%
  mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
  group_by(subjectID) %>% 
  filter(visit==min(visit)) %>%
  ungroup() %>%
  mutate(Sample_ID=sampleID) %>%
  mutate(Group=case_when(mnd=='LOW'~'CTR'))

# mutate(Group=case_when(disease=='healthy'~'CTR')
```

```{r}

print(length(unique(meta.healthy.s$subjectID))) 
print(nrow(meta.healthy.s))

```

#### Taxonomin Profiles

Load the taxonomic profiles generated with MetaPhlAn2 via the curatedMetagenomicsData R package.

```{r}

#x <- '*metaphlan_bugs_list.stool*'
#feat <- curatedMetagenomicData(x=x, dryrun=FALSE)

feat <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)

feat <- feat[[x]]@assayData$exprs

```

The MetaPhlAn2 profiles contain information on different taxonomic levels. Therefore, we want to restrict them to species-level profiles. In a second step, we convert them into relative abundances (summing up to 1) instead of using the percentages (summing up to 100) that MetaPhlAn2 outputs.
```{r}

feat <- feat[grep(x=rownames(feat), pattern='s__'),]
feat <- feat[grep(x=rownames(feat),pattern='t__', invert = TRUE),]
feat <- t(t(feat)/100)

```

The feature names are very long and may be a bit un-wieldy for plotting later on, so we shorten them to only the species name:
```{r}

rownames(feat) <- str_extract(rownames(feat), 's__.*$')

```

#### Create the SIAMCAT Object

Now, we have everything ready to create a SIAMCAT object which stores the feature matrix, the meta-variables, and the label. Here, the label is created using the information in the metadata.
```{r}

sc.obj <- siamcat(feat=feat, meta=meta.healthy, label='Group', case='mnd')

sc.obj <- siamcat(meta.healthy, label='Group', case='MEDIUM,HIGH')

```

## Feature Filtering
Now, we can filter feature with low overall abundance and prevalence.
```{r}

sc.obj <- filter.features(sc.obj, cutoff=1e-04, filter.method = 'abundance')

sc.obj <- filter.features(sc.obj, cutoff=0.05,
                          filter.method='prevalence',
                          feature.type = 'filtered')

```

### Association Plot

The check.assocation function calculates the significance of enrichment and metrics of association (such as generalized fold change and single-feautre AUROC).
```{r}

sc.obj <- check.associations(sc.obj, detect.lim = 1e-06, 
                             alpha=0.1, 
                             max.show = 20,
                             plot.type = 'quantile.rect',
                             fn.plot = './association_plot.pdf')

```

### 2020-03-27 annals_manuscript.Rmd
updated

--- 2020-02-03 manuscript_NS_MRD.Rmd
@@
 ---
-title: "curatedMetagenomicData"
-author: "NSharma"
-date: "26/09/2019"
-output:
-  word_document: default
-  html_document: default
+title: "annals_manuscript"
+output: html_document
 ---
@@
 
-c.mnd.atlas <- subset_samples(c.mnd.atlas, !is.na(nationality) )
+#c.mnd.atlas <- subset_samples(c.mnd.atlas, !is.na(nationality) )
 
@@
 my_atlas_table(atlas.pseq)
-alpha_d(pseq.subset,"final_RLWT_1")
-
-alpha_d(atlas.pseq,"final_RLWT_2")
-
-beta_d(pseq.subset,"final_2")
-beta_d(atlas.pseq,"final_3")
+alpha_d(pseq.subset,"alpha_curated")
+
+alpha_d(atlas.pseq,"alpha_atlas")
+
+beta_d(pseq.subset,"beta_curated")
+beta_d(atlas.pseq,"beta_atlas")
 
 pre_perma_d(pseq.subset) 
-perma_d(p.low,p.high,"RLWT_Curatefinal_highVlow")
+perma_d(p.low,p.high,"permanova_curated_highVlow")
 
 pre_perma_d(pseq.subset) 
-perma_d(p.med,p.high,"RLWT_Curatefinal_highVmed")
+perma_d(p.med,p.high,"permanova_curated_highVmed")
 
 pre_perma_d(atlas.pseq) 
-perma_d(p.low,p.high,"RLWT_Atlasfinal_highVlow")
+perma_d(p.low,p.high,"permanova_atlas_highVlow")
 
 pre_perma_d(atlas.pseq) 
-perma_d(p.med,p.high,"RLWT_Atlasfinal_highVmed")
+perma_d(p.med,p.high,"permanova_atlas_highVmed")
 

### 2020-03-27 siamcat_MND_MRD.Rmd
updated

@@
 
+#install.packages("randomForest")
+
+library(SIAMCAT)
+library(RColorBrewer)
+library(pROC)
+library(tidyverse)
+library(curatedMetagenomicData)
+library(ggpubr)
+
 library(ExperimentHub)
@@
 library(microbiome)
-library(tidyverse) 
-library(SIAMCAT) 
-library(ggpubr)
 
@@
 
+```{r}
+
+#pseq.subset <- subset_samples(pseq, !is.na(mnd) & is.na(disease) | disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
+
+```
+
+# pseq Preprocessing
+Transformed to relative abundance, creating the new pseq.r object, which is then filtered such that only OTUs with a mean greater than 10^-5 are kept.
+
+```{r}
+any(taxa_sums(pseq) == 0)
+# FALSE if OTUs are found in all samples and we don't need to prune the taxa.
+If the answer is TRUE, some OTUs are not present in any samples.
+This is usually the case when data is subset to remove some samples.
+OTUs unique to those sample are not removed along with the samples.
+Therefore, it is important to check this everytime the phyloseq object is filtered for samples using `subset_samples` function.
+head(meta(pseq))
+
+pseq.1 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
+
+# check again if any OTUs are not present in any samples
+any(taxa_sums(pseq.1) == 0)
+
+ntaxa(pseq)
+ntaxa(pseq.1)
+
+ntaxa(pseq) - ntaxa(pseq.1)
+
+head(meta(pseq.1))
+
+```
+
+```{r}
+
+summarize_phyloseq(pseq.1)
+
+```
+
+```{r}
+pseq.1@otu_table [1:3, 1:3]
+
+```
+
+```{r}
+
+#pseq.r  = transform_sample_counts(pseq, function(x) x / sum(x) )
+#pseq.pr = filter_taxa(pseq.r, function(x) var(x) > 1e-5, TRUE)
+
+```
+
+```{r}
+pseq.c <- microbiome::transform(pseq.1, "compositional")
+```
+
 The taxonomic abundance profiles and the sample meatadata for the example dataset are available through the curatedMetagenomicsData R package.
@@
 table(pseq@sam_data[["mnd"]])
+
+```
+
+### Creating Siamcat Object ###
+We have a phyloseq object, therefore:
+Metadata = sam_data
+Features = otu_table
+
+The siamcat object is constructed using the siamcat() function. There are two ways to initialize it:
+
+Features: You can provide a feature matrix, data.frame, or otu_table to the function (together with label and metadata information):
+
+siamcat <- siamcat(feat=feat, label=label, meta=meta)
+phyloseq: The alternative is to create a siamcat object directly out of a phyloseq object:
+
+siamcat <- siamcat(phyloseq=phyloseq, label=label)
+
+```{r}
+
+feat=otu_table(pseq.c)
+meta=sample_data(pseq.c)
+
+#Two ways to create siamcat object:
+
+label <- create.label(meta=sample_data(pseq.c),
+    label = "mnd",
+    case = "HIGH", control = "LOW")
+
+#FIRST WAY:
+# siamcat <- siamcat(feat=feat, label=label, meta=meta)
+# siamcat <- siamcat(feat=otu_table, label=label, meta=sample_data(pseq))
+# Error in validObject(.Object) : invalid class “phyloseq” object: 
+ # An otu_table is required for most analysis / graphics in the phyloseq-package
+
+#SECOND WAY
+# siamcat <- siamcat(phyloseq=phyloseq, label=label)
+siamcat <- siamcat(phyloseq=pseq.c, label=label)
 
@@
 Metadata needs to be converted to numerical values by the user (This is necessary for heatmap displays)!
-
-# Create Label
-```{r}
-
-# The siamcat object extends on the phyloseq object.
-Therefore, creating a siamcat object from a phyloseq object is really straightforward.
-This can be done with the siamcat constructor function.
-First, however, we need to create a label object:
-label <- create.label(meta=sample_data(pseq),
-    label = "mnd",
-    case = c("MEDIUM", "HIGH"), control = "LOW")
-
-# Assigned LOW as control group
-
-siamcat <- siamcat(phyloseq=pseq, label=label)
-
-```
 
@@
 phyloseq <- physeq(siamcat)
-tax_tab <- tax_table(phyloseq)
-head(tax_tab)
+otu_tab <- otu_table(phyloseq)
+head(otu_tab)
+
+```
+
+Data can also be sub-selected based on the available meta-data. For example, if we want to exclude patients that are too young or too old for the question of interest, we can do so easily with:
+```{r}
+
+#siamcat <- select.samples(
+#    siamcat,
+#    filter = 'age',
+#    allowed.set = NULL,
+#    allowed.range = c(20, 90),
+#    verbose = 2
+#)
 
@@
 
+Confounders are checked with the function check.confounders, which produces a plot for each possible confounder in the metadata and diverts the output into a pdf-file.
+
+```{r}
+
+check.confounders(siamcat,
+    fn.plot = 'conf_checkMND.pdf', meta.in = c("mnd", "age", "gender", "BMI", "age_category", "born_method", "feeding_practice"))
+
+```
+
 #### Association Testing ####
@@
 
+library(progress)
+library(pROC)
+library(RColorBrewer)
+
 siamcat <- check.associations(
@@
     panels = c("fc", "prevalence", "auroc"))
+
+```
+
+##### Model Building ####
+
+One strength of SIAMCAT is the versatile but easy-to-use interface for the construction of machine learning models on the basis of microbial species.
+SIAMCAT contains functions for data normalization, splitting the data into cross-validation folds, training the model, and making predictions based on cross-validation instances and the trained models.
+
+#### Data Normalization
+
+Data normalization is performed with the normalize.features function.
+Here, we use the log.unit method, but several other methods and customization options are available (please check the documentation).
+
+```{r}
+
+siamcat <- normalize.features(
+    siamcat,
+    norm.method = "log.unit",
+    norm.param = list(
+        log.n0 = 1e-06,
+        n.p = 2,
+        norm.margin = 1
+    )
+)
+
+```
+
+#### Perpare Cross-Validation
+
+https://www.youtube.com/watch?v=fSytzGwwBVw
+
+Cross-Validation is basically how we decide which machine learning method would be best for our data set.
+CV allows us to compare different machine learning methods and get a sense of how well they will work in practice.
+
+Examples of machine learning methods:
+1. Logistic Regression
+2. K-nearest neighbours
+3. Support Vector Machines (SVM) ...etc.
+
+Need to do 2 things with collected data:
+1. Estimate parameters for machine learning methods. (training algorithm)
+2. Evaluate how well the machine learning method works. (testing algorithm)
+
+Terrible approach to use ALL the data to train algorithm bc wouldn't have any data left to test algorithm. - Data separated into blocks, different block combinations used to train model.
+
+Preparation of the cross-validation fold is a crucial step in machine learning.
+SIAMCAT greatly simplifies the set-up of cross-validation schemes, including stratification of samples or keeping samples inseperable based on metadata.
+For this small example, we choose a twice-repeated 5-fold (no.
+of blocks) cross-validation scheme.
+The data-split will be saved in the data_split slot of the siamcat object.
+```{r}
+
+siamcat <-  create.data.split(
+    siamcat,
+    num.folds = 5,
+    num.resample = 2
+)
+
+```
+
+#### Model Training
+
+The actual model training is performed using the function train.model.
+Again, multiple options for customization are available, ranging from the machine learning method to the measure for model selection or customizable parameter set for hyperparameter tuning.
+```{r}
+
+library(randomForest)
+
+siamcat <- train.model(
+    siamcat,
+    method = "lasso"
+)
+
+#siamcat <- train.model(
+#    siamcat,
+#    method = "enet"
+#)
+
+#siamcat <- train.model(
+#    siamcat,
+#    method = "ridge"
+#)
+
+#siamcat <- train.model(
+#    siamcat,
+#    method = "lasso_ll"
+#)
+
+#siamcat <- train.model(
+#    siamcat,
+#    method = "ridge_ll"
+#)
+
+#siamcat <- train.model(
+#    siamcat,
+#    method = "randomForest"
+#)
+
+```
+
+The models are saved in the model_list slot of the siamcat object. The model building is performed using the mlr R package. All models can easily be accessed.
+```{r}
+
+# get information about the model type
+model_type(siamcat)
+
+# access the models
+models <- models(siamcat)
+models[[1]]
+
+```
+
+#### Make Predicitions
+
+Using the data-split and the models trained in previous step, we can use the function make.predictions in order to apply the models on the test instances in the data-split.
+The predictions will be saved in the pred_matrix slot of the siamcat object.
+
+```{r}
+
+siamcat <- make.predictions(siamcat)
+pred_matrix <- pred_matrix(siamcat)
+
+head(pred_matrix)
+
+```
+
+#### Model Evaluation and Interpretation ####
+
+In the final part, we want to find out how well the model performed and which microbial species had been selected in the model.
+In order to do so, we first calculate how well the predictions fit the real data using the function evaluate.predictions.
+This function calculates the Area Under the Receiver Operating Characteristic (ROC) Curve (AU-ROC) and the Precision Recall (PR) Curve for each resampled cross-validation run.
+
+```{r}
+
+siamcat <-  evaluate.predictions(siamcat)
+
+```
+
+#### Evaluation Plot
+
+To plot the results of the evaluation, we can use the function model.evaluation.plot, which produces a pdf-file showing the ROC and PR Curves for the different resamples runs as well as the mean ROC and PR Curve.
+
+```{r}
+
+model.evaluation.plot(siamcat)
+
+```
+
+#### Interpretation Plot
+
+The final plot produced by SIAMCAT is the model interpretation plot, created by the model.interpretation.plot function. The plot shows for the top selected features the:
+
+  - model weights (and how robust they are, i.e. in what proportion of models have they been incorporated) as a barplot
+
+  - a heatmap with the z-scores or fold changes for the top selected features
+
+  - a boxplot showing the proportions of weight per model which is captured by the top selected features.
+
+Additionally, the distribution of metadata is shown in a heatmap below.
+
+The function again produces a pdf-file optimized for a landscape DIN-A4 plotting region.
+
+```{r}
+
+model.interpretation.plot(
+    siamcat, 
+    fn.plot = 'interpretationMND.pdf',
+    color.scheme = "BrBG",
+    consens.thres = 0.5,
+    heatmap.type = "zscore",
+    limits = c(-3, 3), detect.lim = 1e-06,
+    max.show = 50, prompt=TRUE, verbose = 1)
 
@@
 ```
-
-First, SIAMCAT needs a feature matrix (can be either a matrix, a data.frame, or a phyloseq-otu_table), which contains values of different features (in rows) for different samples (in columns).
-For example, the feature matrix included here contains relative abundances for bacterial species calculated with the mOTU profiler for 141 samples:
-
-```{r}
-
-feat.crc.zeller[1:3, 1:3]
-
-# Dimensions of object
-dim(feat.crc.zeller)
-
-```
-
-#### Taxonomin Profiles
-
-Load the taxonomic profiles generated with MetaPhlAn2 via the curatedMetagenomicsData R package.
-
-```{r}
-
-# Get the stool samples
-esl <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)
-
-# merge
-eset <- mergeData(esl)
-
-# the relab=FALSE is essential as it Absolute Raw Count Data
-pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
-orig_pseq <- pseq
-
-y <- "*metaphlan_bugs_list.stool*"
-
-feat <- curatedMetagenomicData(x=orig_pseq, dryrun=FALSE)
-
-feat <- feat[[x]]@assayData$exprs
-
-feat <- feat[[y]]@assayData$exprs
-
-x <- 'NielsenHB_2014.metaphlan_bugs_list.stool'
-feat <- curatedMetagenomicData(x=x, dryrun=FALSE)
-
-feat <- feat[[x]]@assayData$exprs
-
-```
-
-The MetaPhlAn2 profiles contain information on different taxonomic levels.
-Therefore, we want to restrict them to species-level profiles.
-In a second step, we convert them into relative abundances (summing up to 1) instead of using the percentages (summing up to 100) that MetaPhlAn2 outputs.
-```{r}
-
-feat <- feat[grep(x=rownames(feat), pattern='s__'),]
-feat <- feat[grep(x=rownames(feat),pattern='t__', invert = TRUE),]
-feat <- t(t(feat)/100)
-
-```
-
-The feature names are very long and may be a bit un-wieldy for plotting later on, so we shorten them to only the species name:
-```{r}
-
-rownames(feat) <- str_extract(rownames(feat), 's__.*$')
-
-```
-
-#### Create the SIAMCAT Object
-
-Now, we have everything ready to create a SIAMCAT object which stores the feature matrix, the meta-variables, and the label. Here, the label is created using the information in the metadata.
-```{r}
-
-sc.obj <- siamcat(feat=feat, meta=meta.healthy, label='Group', case='mnd')
-
-sc.obj <- siamcat(meta.healthy, label='Group', case='MEDIUM,HIGH')
-
-```
-
-## Feature Filtering
-Now, we can filter feature with low overall abundance and prevalence.
-```{r}
-
-sc.obj <- filter.features(sc.obj, cutoff=1e-04, filter.method = 'abundance')
-
-sc.obj <- filter.features(sc.obj, cutoff=0.05,
-                          filter.method='prevalence',
-                          feature.type = 'filtered')
-
-```
-
-### Association Plot
-
-The check.assocation function calculates the significance of enrichment and metrics of association (such as generalized fold change and single-feautre AUROC).
-```{r}
-
-sc.obj <- check.associations(sc.obj, detect.lim = 1e-06, 
-                             alpha=0.1, 
-                             max.show = 20,
-                             plot.type = 'quantile.rect',
-                             fn.plot = './association_plot.pdf')
-
-```

### 2020-04-03 annals_manuscript.Rmd
updated

@@
 atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
-atlas.pseq <- subset_samples(atlas.pseq, !is.na(mnd))
+atlas.pseq <- subset_samples(atlas.pseq, !is.na(mnd), !is.na(nationality) )
 
@@
 
-  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = wes_palette(glob_theme))
+  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
 
@@
 
+#Abundance plots
+
 ```{r Abundance}
@@
 
-plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
 
@@
 }
+
+# atlas barplot
+p.new <- aggregate_taxa(atlas.pseq, "Phylum")
+
+p.new.rel <- microbiome::transform(p.new, "compositional")
+
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+
+p <- plot.p.new.rel + scale_fill_brewer(palette = "Paired")
+ggsave("atlas_bar.png")
+
+# pseq barplot
+p.new <- aggregate_taxa(pseq.subset, "Phylum")
+
+p.new.rel <- microbiome::transform(p.new, "compositional")
+
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+
+p1 <- plot.p.new.rel + scale_fill_manual(values=c( "#A5CEE3", "#1F78B4", "#34A02C", "#E3211D", "#FF7E00", "blue", "yellow", "red"))
+ggsave("pseq_bar.png")
+
+p3 <<-plot_grid(title,p,p1, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
+plot(p3)
 
@@
 abundance.bar(atlas.pseq, "atlas_bar")
-abundance.bar(pseq.subset, "pseq_bar")
+#abundance.bar.pseq(pseq.subset, "pseq_bar")
+
+```
+
+# Alpha and beta diversity figure
+```{r}
+
+# gets the meta data 
+  pseq.meta <- meta(atlas.pseq)
+  # generates the diversites
+  tab <- microbiome::alpha(atlas.pseq, index = "shannon")
+
+  # creates Shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
+  mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+  # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+
+  # This country versus nationality
+  p1<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+                  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+
+  # the number in the palettes must equal the number of groups in bmi.pairs
+
+  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+
+  p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
+
+# gets the meta data 
+  pseq.meta <- meta(pseq.subset)
+  # generates the diversites
+  tab <- microbiome::alpha(pseq.subset, index = "shannon")
+
+  # creates Shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
+  mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+  # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+
+  # This country versus nationality
+  p3<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+                  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
+
+  # the number in the palettes must equal the number of groups in bmi.pairs
+
+  p4 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+
+  p4 <<- p4 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
+
+p.1 <<-plot_grid(title, p1, p3, p2, p4, align = "v" , ncol = 2, labels = c('','A', 'B', 'C', 'D'), rel_widths = 1, rel_heights = 2)
+print(p.1)
 
@@
  The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):
-
-NOTHING BELOW THIS WORKS!!!!!!!!
-
-## MICROBIOME COMPOSITION
-
-```{r Phylum #}
-
-# Show available ranks in the dataset
-rank_names(pseq)
-
-# Create table, number of features for each phyla
-table(tax_table(pseq)[, "Phylum"], exclude = NULL)
-
-```
-
-This shows a few phyla for which only one feature was observed. Those may be worth filtering.
-
-guide_italics
-ggtitle("Relative abundance") +  
-```{r}
-
-clos.ph <- aggregate_taxa(clostridium, "Phylum")
-
-clos.ph.rel <- microbiome::transform(clos.ph, "compositional")
-
-plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
-  theme(legend.position = "bottom", ) + theme_bw() + 
-  theme(axis.text.x=element_blank()) + 
-  theme(legend.title = element_text(size=18), legend.text = element_text( face="italic")) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend(title="Phylum"))
-
-print(plot.comp.rel)
-
-ggsave("phylum_relative_abundance_clos.png")
-
-plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none") + guide_italics
-ggsave("phylum_relative_abundance_clos2.png")
-
-```
-
-```{r CLostridium Bar Plot}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.cl.rel <- microbiome::transform(clos, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))
-
-plot.cl.rel + scale_fill_brewer(palette = "Paired")
-ggsave("cl_abundance_bar.png")
-
-```
-
-```{r Clostridium genus bar plots}
-
-pseq.cl.rel <- microbiome::transform(clostridium, "compositional")
-
-plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")
-
-plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")
-
-```
-
-```{r clostridium heatmap}
-
-pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")
-
-pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")
-
-plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")
-
-p <- plot.c.comp.rel + coord_flip()
-p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
-ggsave("clos_histogram.png")
-
-```
-
-## Community composition plotting
-Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results.
-These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted.
-Interpretation of these plots should always be subject to subsequent statistical analysis.
-
-```{r}
-
-clos <- aggregate_taxa(clostridium, "Phylum")
-
-# agglomerate at phylum level
-clos_phy <- tax_glom(clos, taxrank = "Phylum")
-
-# Transform to rel. abundance (or use ps0.ra)
-clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})
-
-# Melt to long format for easy ggploting
-clos_phylum <- psmelt(clos_phyl)
-
-# Plot - Phylum
-p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
-  geom_bar(stat = "identity", width = 1) +
-  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
-  theme(axis.text.x = element_blank()) +
-  theme(axis.title.x = element_blank()) +
-  labs(title = "Abundant Phylum of Clostridium(> 1%)")
-
-p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
-plot(p)
-ggsave("phylum_abun_mnd_clos.png")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Phylum")
-
-```
-
-```{r}
-
-get_taxa_unique(pseq, "Genus")
-
-```
-
-```{r}
-
-get_taxa_unique(clostridium, "Genus")
-
-```
-
-```{r}
-
-df <- psmelt(pseq)
-kable(head(df))
-
-```
-
-```{r}
-
-head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])
-
-```

### 2020-04-03 siamcat_MND_MRD.Rmd
updated

@@
 ```{r}
-pseq.c <- microbiome::transform(pseq.1, "compositional")
+
+pseq.c <- core(pseq, detection = 0.1/100, prevalence = 50/100)
+
+#pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" & antibiotics_current_use != "yes")
+
+pseq.subset <- subset_samples(pseq.c, disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )
+
+pseq.c <- microbiome::transform(pseq.subset, "compositional")
+
 ```
@@
     label = "mnd",
-    case = "HIGH", control = "LOW")
+    case = c("HIGH", "MEDIUM"), control = "LOW")
 
-#FIRST WAY:
+#FRIST WAY
+# siamcat <- siamcat(phyloseq=phyloseq, label=label)
+siamcat <- siamcat(phyloseq=pseq.c, label=label)
+
+#SECOND WAY:
 # siamcat <- siamcat(feat=feat, label=label, meta=meta)
@@
  # An otu_table is required for most analysis / graphics in the phyloseq-package
-
-#SECOND WAY
-# siamcat <- siamcat(phyloseq=phyloseq, label=label)
-siamcat <- siamcat(phyloseq=pseq.c, label=label)
 
@@
 Now, we can filter feature with low overall abundance and prevalence.
+Since we have quite a lot of microbial species in the dataset at the moment, we can perform unsupervised feature selection using the function filter.features.
 ```{r}
@@
 
-Since we have quite a lot of microbial species in the dataset at the moment, we can perform unsupervised feature selection using the function filter.features.
-```{r}
-
-siamcat <- filter.features(siamcat,
-    filter.method = 'abundance',
-    cutoff = 0.001)
-
-```
-
-Confounders are checked with the function check.confounders, which produces a plot for each possible confounder in the metadata and diverts the output into a pdf-file.
-
-```{r}
-
-check.confounders(siamcat,
-    fn.plot = 'conf_checkMND.pdf', meta.in = c("mnd", "age", "gender", "BMI", "age_category", "born_method", "feeding_practice"))
-
-```
-
 #### Association Testing ####
+Checks for associations of single species with disease suing nonparametric Wilcoxon test.
 
@@
 AUC or fold change).
+
+The check.assocation function calculates the significance of enrichment and metrics of association (such as generalized fold change and single-feautre AUROC).
 
@@
     sort.by = 'fc',
-    fn.plot = 'assocMND.pdf',
+    fn.plot = 'assocMND_healthy.pdf',
     alpha = 0.05,
@@
     panels = c("fc", "prevalence", "auroc"))
+
+```
+
+Confounders are checked with the function check.confounders, which produces a plot for each possible confounder in the metadata and diverts the output into a pdf-file.
+
+```{r}
+
+check.confounders(siamcat,
+    fn.plot = 'conf_checkMND_healthy.pdf', meta.in = c("mnd", "gender", "age_category"))
 
@@
 Here, we use the log.unit method, but several other methods and customization options are available (please check the documentation).
+
+Getting all data on the same scale: if the scales for different features are wildly different, this can have a knock-on effect on your ability to learn (depending on what methods you're using to do it).
+Ensuring standardised feature values implicitly weights all features equally in their representation.
 
@@
 
-library(randomForest)
+#library(randomForest)
 
@@
 #### Interpretation Plot
+Displays importance of individual features in the classification model
 
@@
     siamcat, 
-    fn.plot = 'interpretationMND.pdf',
+    fn.plot = 'interpretationMND_healthy.pdf',
     color.scheme = "BrBG",
@@
     max.show = 50, prompt=TRUE, verbose = 1)
+
+model.interpretation.plot(
+    siamcat, 
+    fn.plot = 'interpretationfcMND.pdf',
+   color.scheme = "BrBG",
+    consens.thres = 0.5,
+    heatmap.type = "fc",
+    limits = c(-3, 3), detect.lim = 1e-06,
+    max.show = 50, prompt=TRUE, verbose = 3)
 
@@
 
-check.confounders(siamcat, fn.plot = './confoundersMND.pdf')
+#check.confounders(siamcat, fn.plot = './confoundersMND.pdf')
 
@@
 
-print(length(unique(pseq$subjectID))) 
-print(nrow(pseq))
+#print(length(unique(pseq$subjectID))) 
+#print(nrow(pseq))
 
-print(length(unique(pseq@sam_data[["subjectID"]])))
+#print(length(unique(pseq@sam_data[["subjectID"]])))
 #6784 total rows
@@
 
-pseq.s<- pseq %>%
-  select(sampleID, subjectID, study_condition, disease_subtype, 
-disease, age, country, number_reads, median_read_length, BMI, mnd) %>%
-  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
-  mutate(visit=str_remove(visit, '_')) %>% 
-  mutate(visit=as.numeric(visit)) %>%
-  mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
-  group_by(subjectID) %>% 
-  filter(visit==min(visit)) %>%
-  ungroup() %>%
-  mutate(Sample_ID=sampleID) %>%
-  mutate(Group=case_when(mnd=='LOW'~'CTR'))
+#pseq.s<- pseq %>%
+  #select(sampleID, subjectID, study_condition, disease_subtype, 
+#disease, age, country, number_reads, median_read_length, BMI, mnd) %>%
+#  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
+#  mutate(visit=str_remove(visit, '_')) %>% 
+#  mutate(visit=as.numeric(visit)) %>%
+ # mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
+  #group_by(subjectID) %>% 
+  #filter(visit==min(visit)) %>%
+  #ungroup() %>%
+  #mutate(Sample_ID=sampleID) %>%
+  #mutate(Group=case_when(mnd=='LOW'~'CTR'))
 

### 2020-04-03 annals_manuscript.Rmd
updated

@@
 
-# Alpha and beta diversity figure
+# Alpha figure
 ```{r}
 
-# gets the meta data 
-  pseq.meta <- meta(atlas.pseq)
+pseq.meta <- meta(pseq.subset)
   # generates the diversites
-  tab <- microbiome::alpha(atlas.pseq, index = "shannon")
+  tab <- microbiome::alpha(pseq.subset, index = "shannon")
 
@@
   # on to the graphs!!!
-
+  print("The basic alpha diversity analysis has been completed")
   # This country versus nationality
@@
                   add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
-
+  print("First graph has completed")
   # the number in the palettes must equal the number of groups in bmi.pairs
@@
 
+  print("Second graph has completed")
+
+  p3 <<-plot_grid(p1,p2, align = "v" , ncol = 1, axis = 'tblr', labels = c('C', 'E'),rel_heights = c(5,5,5))
+
+print(p3)
+
 # gets the meta data 
-  pseq.meta <- meta(pseq.subset)
+  pseq.meta <- meta(atlas.pseq)
   # generates the diversites
-  tab <- microbiome::alpha(pseq.subset, index = "shannon")
+  tab <- microbiome::alpha(atlas.pseq, index = "shannon")
 
@@
   # This country versus nationality
-  p3<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+  p1<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
                   add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
@@
 
-  p4 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
-
-  p4 <<- p4 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
-
-p.1 <<-plot_grid(title, p1, p3, p2, p4, align = "v" , ncol = 2, labels = c('','A', 'B', 'C', 'D'), rel_widths = 1, rel_heights = 2)
-print(p.1)
+  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+
+  p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
+
+p4 <<-plot_grid(p1,p2, align = "v" , ncol = 1, axis = 'tblr', labels = c('D', 'F'),rel_heights = c(5,5,5))
+
+print(p4)
+
+p6 <<-plot_grid(p3,p4, align = "v" , ncol = 2, rel_heights = c(10,10,10))
+
+print(p6)
+
+```
+
+# Composition and Alpha 
+```{r}
+
+# pseq barplot
+p.new <- aggregate_taxa(pseq.subset, "Phylum")
+
+p.new.rel <- microbiome::transform(p.new, "compositional")
+
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+
+p <- plot.p.new.rel + scale_fill_manual(values=c( "#A5CEE3", "#1F78B4", "#34A02C", "#E3211D", "#FF7E00", "blue", "yellow", "red"))
+
+# atlas barplot
+p.new <- aggregate_taxa(atlas.pseq, "Phylum")
+
+p.new.rel <- microbiome::transform(p.new, "compositional")
+
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+
+p1 <- plot.p.new.rel + scale_fill_brewer(palette = "Paired")
+
+p7 <<-plot_grid(p,p1, align = "v" , ncol = 2, labels = c('A', 'B'),rel_heights = c(10,10,10))
+plot(p7)
+
+p8 <<-plot_grid(p7, p6, align = "v" , nrow = 2, rel_heights = c(10,30,30))
+
+print(p8)
+
+```
+
+#Beta
+```{r}
+
+ p <<-  boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+ # p1 = plot_ordination(pseq.subset, pseq.nmds, type = "taxa", color = "mnd", title = "Taxa by Phylum") # I dont think that we need this any more. 
+  #p1 + facet_wrap(~Phylum, 3)
+  pseq.nmds <- ordinate(pseq.subset, "NMDS", "bray")
+  p2 <- plot_ordination(pseq.subset, pseq.nmds, type = "sample", color = "mnd")
+
+  p3 <- p2 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND DALYs") + theme(legend.position = "none")
+  #plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+
+  # This add the facets but I do not think we need that any more. 
+  #p3 <- p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+  save_plot(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), p3)
+    # MDS
+   pseq.mds <<- ordinate(pseq.subset, "MDS", "bray")
+   p4 <- plot_ordination(pseq.subset, pseq.mds, type = "sample", color = "mnd")
+ #plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence MDS"))
+
+    p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND DALYs") + theme(legend.position = "none")
+    save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p5)
+
+   #PCoA for compositional data with Bray-Curtis distances
+    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
+    #p6 <- plot_landscape(pseq.subset, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+  #   # PCoA for compositional data with Bray-Curtis distances
+   #p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+    #                    method = "PCoA", distance = "bray") +
+    #     labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+  # 
+    tsne_res <- tsne_phyloseq(pseq.subset, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+    p7 <- plot_tsne_phyloseq(pseq.subset, tsne_res,color = 'mnd')
+    p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND DALYs") + theme(legend.position = "bottom")
+    save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p8)
+
+    p10 <<-plot_grid(title,p3, p5,p8, ncol = 1, labels = c('','A', 'B','C'),rel_heights = c(0.3, 1.5,1.5,2))
+
+    print(p10)
+
+```
+
+# Permanova figure
+```{r}
+
+ p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme_classic() + theme(axis.text.y = element_text(face="italic"))
+
+plot(p, xlim=c(-0.03,0.03))
+
+  # scale_x_continuous(limits=c(-0.03, 0.03), breaks=c(-0.03, -0.02, -0.01, 0, 0.01, 0.02, 0.03))
+
+ p.low <<- subset_samples(atlas.pseq, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(atlas.pseq, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(atlas.pseq, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p2 <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = FALSE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme_classic() + theme(axis.text.y = element_text(face="italic"))
+
+perm <- plot_grid(p,p2, align = "v" , ncol = 1, axis = 'tblr', labels = c('A', 'B'),rel_heights = c(10,10,10))
+
+print(perm)
+ggsave("permanova_annals.png", device = "png",width = 250, height = 280, units = "mm")
 
@@
 
-## BETA DIVERSITY
+BETA DIVERSITY
 Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.

### 2020-04-15 annals_manuscript.Rmd
updated

--- 2020-04-03 annals_manuscript.Rmd
@@
 library(kableExtra)
+library(ggplot2)
+library(grid)
+library(gridExtra)
 
@@
 atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
-atlas.pseq <- subset_samples(atlas.pseq, !is.na(mnd), !is.na(nationality) )
-
-```
-
-## Incidence of MND
-```{r}
-
-#THIS IS NO LONGER NEEDED.
-
-# This block uses the raw data to examin  the incidence of MND to ensure the classificaiton is correct. It assumes that the data is in the correct place so the setwd at the top meeds to be accurate. 
-# setwd('~/globalmicrobiome') 
-# # For MND  nrows = 196 as thwew are other things after 196
-# mnd.i.15_49 <- as.data.frame(read.csv("./rawdata/MND_I_15_49.csv", header=TRUE, nrows = 196))
-
-# mnd.i.50_69  <- as.data.frame(read.csv("./rawdata/MND_I_50_69.csv", header=TRUE, nrows = 196))
-# mnd.i.70 <- as.data.frame(read.csv("./rawdata/MND_I_70plus.csv", header=TRUE, nrows = 196))
-
-# mnd.i.70.test <- mnd.i.70
-# # Run function and change name
-# mnd.i.15_49 <- mb.gather(mnd.i.15_49)
-# colnames(mnd.i.15_49)[2] <- "15-49"
-
-# # Run function and change name
-# mnd.i.50_69 <- mb.gather(mnd.i.50_69)
-# colnames(mnd.i.50_69)[2] <- "50-69"
-
-# # Run function and change name
-# mnd.i.70 <- mb.gather(mnd.i.70)
-# colnames(mnd.i.70)[2] <- "70"
-
-# # Join them together.   
-# mnd.i <- list( mnd.i.15_49,mnd.i.50_69,mnd.i.70 ) %>% reduce(left_join, by = "Location")
-
-# a.mnd.nationality <- fct_collapse(a.mnd.nationality, LOW = c("EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope" ), HIGH = c("US","UKIE","Scandinavia" ))
-
-# mnd.i %>%
-#   kable(col.names = c("","Incidence","CI","Incidence","CI","Incidence","CI")) %>%
-#   kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "center") %>%
-# add_header_above(c("Location" = 1, "15-49" = 2, "50-69" = 2, "70+" = 2)) %>%
-#   add_header_above(c(" ", "Age (years)" = 6)) %>%
-#   pack_rows("High MND group", 5,6)%>%
-#   pack_rows("Low MND group", 2, 2) %>%
-#   pack_rows("Medium MND group",1,3,4)
-
-# # Now do the same for the curated data set
-# mnd.ic.15_49 <- mb.cgather(mnd.i.15_49)
-# colnames(mnd.ic.15_49)[2] <- "15-49"
-
-# # Run function and change name
-# mnd.ic.50_69 <- mb.cgather(mnd.i.50_69)
-# colnames(mnd.ic.50_69)[2] <- "50-69"
-
-# # Run function and change name
-# mnd.ic.70 <- mb.cgather(mnd.i.70)
-# colnames(mnd.ic.70)[2] <- "70"
-
-# # Join them together.
-
-# mnd.ic <- list( mnd.ic.15_49,mnd.ic.50_69,mnd.ic.70 ) %>% reduce(left_join, by = "Location")
-
-# mnd.ic <- dplyr::select(mnd.ic, "Location",  "Value.x", "Lower.bound.x", "Upper.bound.x", "Value.y", "Lower.bound.y", "Upper.bound.y", "Value", "Lower.bound", "Upper.bound" )
-
-# # Prep the data for the table
-
-# mnd.ic %>%
-#   kable(col.names = c("","Incidence","Upper","Lower","Incidence","Upper","Lower","Incidence","Upper","Lower"),digits=3) %>%
-#   kable_styling(bootstrap_options = c("striped", "hover"),full_width = F,position = "left") %>%
-# add_header_above(c("Location" = 1, "15-49" = 3, "50-69" = 3, "70+" = 3)) %>%
-#   add_header_above(c(" ", "Age (years)" = 9)) %>%
-#   pack_rows("Group 1", 4, 7) %>%
-#   pack_rows("Group 1", 4, 7) %>%
-#   pack_rows("Group 2", 8, 10)
 
@@
     data$age <- sample_data(pseq.subset)$age
+     data$country <- sample_data(pseq.subset)$country
     data$antibiotics_current_use <- sample_data(pseq.subset)$antibiotics_current_use
@@
     data$age_category <- sample_data(pseq.subset)$age_category
-    table1(~ age  +  bmi + bmi_group + smoker + alcohol + age_category | mnd, data=data)
+    table1(~ age  +  bmi + bmi_group + smoker + alcohol + age_category + country | mnd, data=data)
 }
@@
     data$bmi <- sample_data(atlas.pseq)$bmi_group
+     data$nationality <- sample_data(atlas.pseq)$nationality
     data$age <- sample_data(atlas.pseq)$age
-    table1(~ age  +  bmi | mnd, data=data)
+    table1(~ age  +  bmi + nationality | mnd, data=data)
 }
@@
 
+#### FIkter curated
 # https://bioconductor.riken.jp/packages/3.5/data/experiment/vignettes/curatedMetagenomicData/inst/doc/curatedMetagenomicData.html
@@
 
-atlas.pseq <- subset_samples(atlas.pseq, !is.na(nationality), !is.na(mnd))
+#atlas.pseq <- subset_samples(atlas.pseq, !is.na(nationality), !is.na(mnd))
+
+#### Filter Atlas
+atlas.pseq <- subset_samples(atlas.pseq, !is.na(mnd), !is.na(nationality))
+
+```
+
+#### List of genera in phylums
+```{r List of genera in phylum}
+
+map_levels(NULL, "Phylum", "Genus", pseq.subset)
+
+map_levels(NULL, "Phylum", "Genus", atlas.pseq)
+
+#taxa <- map_levels(NULL, "Phylum", "Genus", atlas.pseq)$Clostridium_cluster_I
+#taxa <- map_levels(NULL, "Phylum", "Genus", atlas.pseq)$Actinobacteria
+# List of Genera in the Bacteroideted Phylum
+#taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
+#print(taxa)
+
+```
+
+### Clostridium grouping NOT NEEDED ANYMORE since groupoings have been updated
+```{r}
+
+# Clostridum grouping
+
+#clostridium <- subset_taxa(atlas.pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XIVa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales")
+
+#atlas.pseq.new <- subset_taxa(atlas.pseq, Phylum =="Actinobacteria" | Phylum =="Bacilli"| Phylum =="Proteobacteria"  | Phylum =="Verrucomicrobia"  | Phylum =="Bacteroidetes" | Phylum =="Asteroleplasma"| Phylum =="Spirochaetes" | Phylum =="Fusobacteria" | Phylum =="Cyanobacteria" | Phylum =="Uncultured Mollicutes")
+
+```
+
+## Summary
+```{r summary}
+# Summarise contents of phyloseq
+summarize_phyloseq(pseq)
+# Sparsity is how populated is the data with zeros
+kable(head(tax_table(pseq)))
+tax_table(pseq)
+# Check the data
+print(pseq)
+# Display variables from the mapping file
+sample_variables(pseq) 
+# Total number of taxa in the entire data
+ntaxa(pseq) 
+# Taxonomic ranks
+rank_names(pseq) 
+```
+
+```{r}
 
@@
 
-```{r Abundance}
+#Abundance function
+```{r}
 
@@
 
+abundance(atlas.pseq, "atlas_abundance")
+abundance(pseq.subset, "pseq_abundance")
+
+abundance.bar(atlas.pseq, "atlas_bar")
+#abundance.bar.pseq(pseq.subset, "pseq_bar")
+
+```
+
+#Abundance plots figure
+```{r Abundance}
+
 # atlas barplot
@@
 
-plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot", title = "Atlas Abundance Bar Plot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()  + labs(x="MND") + theme(legend.title = element_text(size = 10), legend.text = element_text(size = 10)) + scale_x_discrete(labels=c("LOW" = "Low", "MEDIUM" = "Medium",
+                              "HIGH" = "High"))
+
+###+ theme(legend.title = element_text(size = 7), legend.text = element_text(size = 7))
 
 p <- plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
-ggsave("atlas_bar.png")
+#ggsave("atlas_bar.png")
 
@@
 
-plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()
+plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot", title = "Curated Abundance Bar Plot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text(face="italic")) + theme_classic() + labs(x="MND") + theme(legend.title = element_text(size = 10), legend.text = element_text(size = 10)) + theme(legend.position="none")  + scale_x_discrete(labels=c("LOW" = "Low", "MEDIUM" = "Medium","HIGH" = "High"))
 
 p1 <- plot.p.new.rel + scale_fill_manual(values=c( "#A5CEE3", "#1F78B4", "#34A02C", "#E3211D", "#FF7E00", "blue", "yellow", "red"))
-ggsave("pseq_bar.png")
-
-p3 <<-plot_grid(title,p,p1, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
-plot(p3)
-
-abundance(atlas.pseq, "atlas_abundance")
-abundance(pseq.subset, "pseq_abundance")
-
-abundance.bar(atlas.pseq, "atlas_bar")
-#abundance.bar.pseq(pseq.subset, "pseq_bar")
-
-```
-
-# Alpha and beta diversity figure
+#ggsave("pseq_bar.png") + scale_fill_brewer(palette = "Paired") + labs(title = "B")
+
+# 2. Save the legend
+#+++++++++++++++++++++++
+legend <- get_legend(p)
+# 3. Remove the legend from the box plot
+#+++++++++++++++++++++++
+p<- p + theme(legend.position="none")
+# 4. Arrange ggplot2 graphs with a specific width
+abun_plot <-grid.arrange(p, p1, legend, ncol=3, widths=c(2.5, 2.5, 1))
+
+#abun_plot <- plot_grid(p,p1, align = "h", axis = "bt", ncol = 2, nrow=1, labels = c('A', 'B'), rel_heights = c(10,10)) 
+abun_plot
+plot(abun_plot)
+ggsave(file = "bar_plots.png", dpi = 600, width = 10, height = 5, units = "in")
+
+ggsave("bar_plots.png", width = 10, height = 5)
+
+#,rel_heights = c(30, 30, 30)
+
+#t3 <- grid_arrange_shared_legend(p, p1 , ncol = 2, nrow = 1, position='right')
+#t3
+
+#g_legend<-function(a.gplot){
+#  tmp <- ggplot_gtable(ggplot_build(a.gplot))
+#  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+#  legend <- tmp$grobs[[leg]]
+#  return(legend)}
+
+#library(gridExtra)
+#legend <- g_legend(p + theme(legend.position='right'))
+#f <- grid.arrange(p+theme(legend.position='hidden'), p1+theme(legend.position='hidden'), ncol = 2, nrow=1, legend)
+
+```
+
+# Alpha figure of nationality
 ```{r}
@@
   tab <- microbiome::alpha(atlas.pseq, index = "shannon")
+
+  # creates Shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
+
+  # This country versus nationality
+  p <- ggviolin(pseq.meta, x = "nationality", y = "Shannon", add = "boxplot", fill = "nationality", palette = "paired", title = "Atlas Nationality Violin Plot") + theme(axis.text.x = element_text(angle = 90)) + xlab("") +theme(legend.direction = "vertical") + theme( axis.text.x=element_blank()) + labs(fill = "Nationality") + xlab("Nationality") + theme(title = element_text(size = 8))
+
+pseq.meta <- meta(pseq.subset)
+  # generates the diversites
+  tab <- microbiome::alpha(pseq.subset, index = "shannon")
+
+  # creates Shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
+
+  # This country versus nationality
+  p1 <- ggviolin(pseq.meta, x = "nationality", y = "Shannon",
+                  add = "boxplot", fill = "nationality", palette = "paired", title = "Curated Nationality Violin Plot") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("") + theme( axis.text.x=element_blank()) + labs(fill = "Nationality") + xlab("Nationality") + theme(title = element_text(size = 8))
+
+legend <- get_legend(p) 
+# 3. Remove the legend from the box plot
+#+++++++++++++++++++++++
+p<- p + theme(legend.position="none")
+# 4. Arrange ggplot2 graphs with a specific width
+p2 <- grid.arrange(p, p1, legend, ncol=3, widths=c(10, 10, 3))
+
+#lay <- rbind(c(p, p1),
+            # c(legend))
+
+p2 <- grid.arrange(p, p1, legend, ncol=2, nrow = 2, layout_matrix = rbind(c(1,2), c(3,3)), widths = c(2.7, 2.7), heights = c(2.5, 0.2), legend.direction = "vertical")
+
+ggsave("nationality_violin.png", width = 10, height = 5)
+
+```
+
+# Alpha figure of nationality and MND
+```{r}
+
+pseq.meta <- meta(pseq.subset)
+  # generates the diversites
+  tab <- microbiome::alpha(pseq.subset, index = "shannon")
 
@@
 
+  print("Second graph has completed")
+
+  p3 <<-plot_grid(p1,p2, align = "v" , ncol = 1, axis = 'tblr', labels = c('D', 'F'),rel_heights = c(10,10))
+
+print(p3)
+
 # gets the meta data 
-  pseq.meta <- meta(pseq.subset)
+  pseq.meta <- meta(atlas.pseq)
   # generates the diversites
-  tab <- microbiome::alpha(pseq.subset, index = "shannon")
+  tab <- microbiome::alpha(atlas.pseq, index = "shannon")
 
@@
 
-  p4 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
-
-  p4 <<- p4 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
-
-p.1 <<-plot_grid(title, p1, p3, p2, p4, align = "v" , ncol = 2, labels = c('','A', 'B', 'C', 'D'), rel_widths = 1, rel_heights = 2)
-print(p.1)
+  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
+
+  p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
+
+p4 <<-plot_grid(p1,p2, align = "v" , ncol = 1, axis = 'tblr', labels = c('C', 'E'),rel_heights = c(10,10))
+
+print(p4)
+
+#order is atlas first
+
+p6 <<-plot_grid(p4,p3, align = "v" , ncol = 2, rel_heights = c(20,20))
+
+print(p6)
+
+ggsave("alpha_manuscript.png", width = 10, height = 10)
+
+```
+
+# Composition and Alpha 
+```{r}
+
+p8 <<-plot_grid(abun_plot, p6, align = "v" , nrow=2, rel_heights = c(12,30,30))
+
+print(p8)
+
+ggsave("alpha_comp_manuscript.png", width = 10, height = 15)
+
+```
+
+theme(axis.title.x=element_blank(), axis.text.x=element_blank()
+# Alpha figure of just MND
+```{r}
+
+pseq.meta <- meta(pseq.subset)
+  # generates the diversites
+  tab <- microbiome::alpha(pseq.subset, index = "shannon")
+
+  # creates Shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
+  mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+  # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+  print("The basic alpha diversity analysis has been completed")
+
+  # the number in the palettes must equal the number of groups in bmi.pairs
+
+  p <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red")) + theme( axis.text.x=element_blank()) + scale_fill_discrete(name = "MND DALYs", labels = c("Low", "Medium", "High"))
+
+  p <<- p + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")
+
+  print("Second graph has completed")
+
+# gets the meta data 
+  pseq.meta <- meta(atlas.pseq)
+  # generates the diversites
+  tab <- microbiome::alpha(atlas.pseq, index = "shannon")
+
+  # creates Shannon
+  pseq.meta$Shannon <- tab$diversity_shannon
+  mnd <- levels(pseq.meta$mnd) # get the variables
+  # Make a pairwise list that we want to compare.
+  mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
+  mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))
+
+  # Compare differences in Shannon index between mnd group of the study subjects
+  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
+  # on to the graphs!!!
+
+  # the number in the palettes must equal the number of groups in bmi.pairs
+
+  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red")) + theme( axis.text.x=element_blank()) + labs(fill = "MND") + scale_fill_discrete(name = "MND DALYs", labels = c("Low", "Medium", "High"))
+
+  p2 <<- p2 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="right")
+
+# 2. Save the legend
+#+++++++++++++++++++++++
+legend <- get_legend(p2)
+# 3. Remove the legend from the box plot
+#+++++++++++++++++++++++
+p2<- p2 + theme(legend.position="none")
+# 4. Arrange ggplot2 graphs with a specific width
+p3 <- grid.arrange(p2, p, legend, ncol=3, widths=c(2.3, 2.3, 0.8))
+
+#If want legend bottom centre 
+# grid.arrange(p2, p, legend, ncol=2, nrow = 2, layout_matrix = rbind(c(1,2), c(3,3)), widths = c(2.7, 2.7), heights = c(2.5, 0.2), legend.direction = "horizontal")
+
+  #p3 <<-plot_grid(p2,p, align = "v" , ncol = 2, axis = 'tblr', labels = c('C', 'D'),rel_heights = c(10,10))
+
+print(p3)
+
+ggsave("alpha2_manuscript.png", width = 10, height = 10)
+
+```
+
+# Composition and Alpha  of just MND
+```{r}
+
+p8 <<-plot_grid(abun_plot, p3, align = "v" , nrow=2, rel_heights = c(30,30))
+
+print(p8)
+
+ggsave("alpha_comp_manuscript2.png", width = 10, height = 10)
+
+```
+
+#Beta
+```{r}
+
+ p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+ p <<-  boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+ # p1 = plot_ordination(pseq.subset, pseq.nmds, type = "taxa", color = "mnd", title = "Taxa by Phylum") # I dont think that we need this any more. 
+  #p1 + facet_wrap(~Phylum, 3)
+  pseq.nmds <- ordinate(pseq.subset, "NMDS", "bray")
+  p2 <- plot_ordination(pseq.subset, pseq.nmds, type = "sample", color = "mnd")
+
+  p3 <- p2 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND DALYs") + theme(legend.position = "none")
+  #plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+
+  # This add the facets but I do not think we need that any more. 
+  #p3 <- p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+  ###save_plot(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), p3)
+    # MDS
+   pseq.mds <<- ordinate(pseq.subset, "MDS", "bray")
+   p4 <- plot_ordination(pseq.subset, pseq.mds, type = "sample", color = "mnd")
+ #plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence MDS"))
+
+    p5 <- p4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND DALYs") + theme(legend.position = "none")
+    ###save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p5)
+
+   #PCoA for compositional data with Bray-Curtis distances
+    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
+    #p6 <- plot_landscape(pseq.subset, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+  #   # PCoA for compositional data with Bray-Curtis distances
+   #p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+    #                    method = "PCoA", distance = "bray") +
+    #     labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+  # 
+    tsne_res <- tsne_phyloseq(pseq.subset, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+    p7 <- plot_tsne_phyloseq(pseq.subset, tsne_res,color = 'mnd')
+    p8 <- p7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND DALYs") + theme(legend.position = "bottom")
+   ### save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), p8)
+
+    p10 <<-plot_grid(title,p3, p5,p8, ncol = 1, labels = c('','A', 'B','C'),rel_heights = c(0.3, 1.5,1.5,2))
+
+    print(p10)
+
+# atlas
+ p.low <<- subset_samples(atlas.pseq, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(atlas.pseq, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(atlas.pseq, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+
+ t <<-  boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+  pseq.nmds <- ordinate(atlas.pseq, "NMDS", "bray")
+  t2 <- plot_ordination(atlas.pseq, pseq.nmds, type = "sample", color = "mnd")
+
+  t3 <- pt + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "NMDS",color = "MND DALYs") + theme(legend.position = "none")
+  #plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
+
+  # This add the facets but I do not think we need that any more. 
+  #p3 <- p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
+  save_plot(paste("Beta_diversity_",title_text,"_NMDS.png", sep = "" ), p3)
+    # MDS
+   pseq.mds <<- ordinate(atlas.pseq, "MDS", "bray")
+   t4 <- plot_ordination(atlas.pseq, pseq.mds, type = "sample", color = "mnd")
+ #plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence MDS"))
+
+    t5 <- t4 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "MDS",color = "MND DALYs") + theme(legend.position = "none")
+   ### save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), t5)
+
+   #PCoA for compositional data with Bray-Curtis distances
+    #   p4 <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
+    #   p4 <- p4 + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b") 
+    #p6 <- plot_landscape(pseq.subset, method = "MDS", distance = "bray", col = "mnd", size = 3) 
+  #   # PCoA for compositional data with Bray-Curtis distances
+   #p6 <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
+    #                    method = "PCoA", distance = "bray") +
+    #     labs(title = paste("PCoA / Compositional / Bray-Curtis"))
+  # 
+    tsne_res <- tsne_phyloseq(atlas.pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
+    t7 <- plot_tsne_phyloseq(atlas.pseq, tsne_res,color = 'mnd')
+    t8 <- t7 + scale_colour_manual(values = wes_palette(glob_theme)) + labs(title = "TSNE",color = "MND DALYs") + theme(legend.position = "bottom")
+### save_plot(paste("Beta_diversity_",title_text,"_MDS.png", sep = "" ), t8)
+
+    t10 <<-plot_grid(title,t3, t5, t8, ncol = 1, labels = c('','A', 'B','C'),rel_heights = c(0.3, 1.5,1.5,2))
+
+    print(t10)
+
+beta <<-plot_grid(p10, t10, align = "h" , ncolumn = 2, rel_heights = c(30,30,30))
+
+beta
+
+```
+
+# Permanova figure
+```{r}
+
+ p.low <<- subset_samples(pseq.subset, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(pseq.subset, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(pseq.subset, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = TRUE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Potenital Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme_classic() + theme(axis.text.y = element_text(face="italic")) + ggtitle("Curated PERMANOVA")
+
+plot(p, xlim=c(-0.03,0.03))
+
+  # scale_x_continuous(limits=c(-0.03, 0.03), breaks=c(-0.03, -0.02, -0.01, 0, 0.01, 0.02, 0.03))
+
+ p.low <<- subset_samples(atlas.pseq, mnd == "LOW")
+  p.low.d <<- divergence(p.low)
+  p.med <<- subset_samples(atlas.pseq, mnd == "MEDIUM")
+  p.med.d <<- divergence(p.med)
+  p.high <<- subset_samples(atlas.pseq, mnd == "HIGH")
+  p.high.d <<- divergence(p.high)
+  p <<- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
+
+  # Convert to compositional data perma_d(p.low,p.high,"final_highVlow")
+pseqLvH <- merge_phyloseq(p.low,p.high)
+  pseq.rel <- microbiome::transform(pseqLvH, "compositional")
+  otu <- abundances(pseq.rel)
+  meta <- meta(pseq.rel)
+  permanova <- adonis(t(otu) ~ mnd,
+               data = meta, permutations=999, method = "bray")
+ # P-value
+  print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])  
+  dist <- vegdist(t(otu))
+  anov <- anova(betadisper(dist, meta$mnd))
+  summary(anov)
+  tidy(anov)
+  coef <- coefficients(permanova)["mnd1",]
+  top.coef <- coef[rev(order(abs(coef)))[1:20]]
+  par(mar = c(12, 30, 2, 1))
+  #KEEP THIS  TO ENSURE THERE IS NOT A MISTAKE IN THE REORDERING BELOW
+  #temp <- TeX("Top Genus")
+  #png(paste("PERMOVA_",title_text,".png", sep = "" ), width = 450, height = 210,  units     = "mm",
+  #res       = 1500)
+  #barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = temp)
+  #dev.off() 
+  temp <- as.data.frame(top.coef)
+  temp$label <- rownames(temp) 
+  temp$label <- sub("s_*", "", temp[,2])
+  #temp$colour <- ifelse(temp$top.coef < 0, "firebrick1","steelblue")
+  p2 <<- ggplot(temp,aes(reorder(label, top.coef),top.coef,fill = top.coef > 0)) + geom_col(show.legend = TRUE,) + coord_flip() +  xlab('Taxa') + ylab('p value')  + theme_minimal(base_size = 12)  + annotate("text", x = 2.5, y = 0.01, label = "Harmful Taxa") + annotate("text", x = 19, y = -0.01, label = "Potential Beneficial Taxa") + scale_fill_discrete(name = "Taxa") + theme_classic() + theme(axis.text.y = element_text(face="italic")) + ggtitle("Atlas PERMANOVA")
+
+perm <- plot_grid(p2,p, align = "v" , ncol = 1, axis = 'tblr', labels = c('A', 'B'),rel_heights = c(10,10,10))
+
+print(perm)
+ggsave("permanova_annals.png", device = "png",width = 250, height = 280, units = "mm")
 
